### 2022-09-13 manuscript/manuscript.Rmd
add existing project files

---
title: "S2 pioneer factor MS"
author: "Tyler Gibson"
date: "8/24/2021"
output: html_document
---

# Abstract

# Introduction

# Results

# Discussion

# Materials and Methods

## Reagents

| Reagent                            | Source/vendor            | Catalog/accession \# |
|----------------------------|------------------------|--------------------|
| pMT-puro                           | AddGene                  | 17923                |
| Gibson Cloning                     | New England Biolabs      | E2611                |
| Schneider's Medium                 | Thermo Fisher Scientific | 21720001             |
| Fetal Bovine Serum                 | Omega Scientific         | FB-02                |
| Antibiotic/antimycotic             | Thermo Fisher Scientific | 15240062             |
| Effectene Transfection Reagent     | Qiagen                   | 301427               |
| PVDF Membrane                      |                          |                      |
| **Antibodies**                     |                          |                      |
| Anti-Zld                           | Harrison et al., 2010    |                      |
| Anti-Grh                           | Harrison et al., 2010    |                      |
| Anti-Twi                           | He et al., 2015          |                      |
| Anti-HA-peroxidase clone 3F10      | Roche                    | 12013819001          |
| Anti-Tubulin DM1A                  | Sigma                    | T6199-200UL          |
| goat anti-rabbit IgG-HRP conjugate | Bio-Rad                  | 1706515              |
|                                    |                          |                      |
|                                    |                          |                      |
|                                    |                          |                      |
|                                    |                          |                      |

## Cloning

*zld*, *grh*, or *twi* cDNA (isoforms RB, RH, and RA, respectively) were cloned into pMT-puro plasmid (Addgene) via Gibson cloning (New England Biolabs). For cloning of the Zld and Grh DNA-binding domains, DNA encoding amino acids 1114-1487 (Zld) or amino acids 603-1032 (Grh) was cloned into pMT-puro.

-   **TG2A- Zld RB**

-   **TG3A- Grh PH**

-   **TG4A**

-   **TG6A**

-   **TG8A**

-   **MF8A**

-   GFP-DBD lines

## Cell culture and generation of stable cell lines

Drosophila S2 cells were cultured at 27°C in Schneider's medium (Thermo Fisher Scientific) supplemented with 10% fetal bovine serum (Omega Scientific) and antibiotic/antimycotic (Thermo Fisher Scientific). For the generation of stable cell lines, cells were plated at 5✕10^5^ cells per mL. After 24 hours, cells were transfected with 10 µg plasmid DNA using Effectene transfection reagent (Qiagen). After an additional 24 hours, puromycin was added to a final concentration of 2 µg/mL. Stable cell lines were recovered after 2-3 weeks of selection. Following recovery of stable cell lines, cells were cultured with 1 µg/mL puromycin.

## Induction of transcription factor expression

## Immunoblotting

Proteins were separated on denaturing polyacramide gels before transfer to a 0.45 μm Immobilon-P PVDF membrane (Millipore) in transfer buffer (25 mM Tris, 200 mM Glycine, 20% methanol) for 60 min (75 min for Zld) at 500mA at 4°C. The membranes were blocked with BLOTTO (2.5% non-fat dry milk, 0.5% BSA, 0.5% NP-40, in TBST) for 30 min at room temperature and then incubated with anti-Zld (1:750) (Harrison et al., 2010), anti-Grh (1:1000) (Harrison et al., 2010), anti-Twi (1:1000) (He et al. 2015), anti-HA-peroxidase (1:500) (clone 3F10, Roche), or anti-Tubulin (DM1A, 1:5000) (Sigma), overnight at 4°C. The secondary incubation was performed with goat anti-rabbit IgG-HRP conjugate (1:3000) (Bio-Rad) for 1 hr at room temperature. Blots were treated with SuperSignal West Pico PLUS chemiluminescent substrate (Thermo Fisher Scientific) and visualized using the Azure Biosystems c600 or Kodak/Carestream BioMax Film (VWR).

## Immunostaining

## ChIP-seq

## ChIP-seq analysis

## RNA-seq

## RNA-seq analysis

## ATAC-seq

## ATAC-seq analysis

## CUT&RUN

## CUT&RUN analysis

## Treatment of S2 cells with small molecule inhibitors

## Analysis of previously published data

## Data and code availability

# Figure Legends

**Figure 1: Ectopic pioneer factors open chromatin and activate transcription.**

**Figure 2: Twist binds closed chromatin extensively and drives accessibility at some sites.**

**Figure 3: Some endogenous targets are resistant to binding by ectopic pioneer factors.**

**Figure 4: Binding and opening of closed chromatin is concentration dependent.**

**Figure 5: H3K27me3 and H3K9me3 are barriers to pioneer-factor binding.**

**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function.**

**Figure 7: Pioneer-factor function requires stable chromatin occupancy.**

### 2022-09-13 workflow/scripts/annotate_ChIP_classes.R

@@ -1,0 +1,96 @@
+# setup ------------------------------------------------------------------------
+suppressPackageStartupMessages(library(rtracklayer))
+suppressPackageStartupMessages(library(GenomicRanges))
+suppressPackageStartupMessages(library(tidyverse))
+
+source(snakemake@params[["r_source"]])
+
+# get input file names ---------------------------------------------------------
+ChIP_peaks <- snakemake@input[["ChIP_peaks"]]
+wt_ATAC_peaks <- snakemake@input[["WT_ATAC_peaks"]]
+ATAC_results <- snakemake@input[["ATAC_results"]]
+gtf <- snakemake@input[["gtf_genome_annotation"]]
+RNAseq_results <- snakemake@input[["RNAseq_results"]]
+ChIP_features <- snakemake@input[["ChIP_feature_annotation"]]
+
+# annotate ChIP classes --------------------------------------------------------
+# import ChIP peaks
+peaks_gr <- rtracklayer::import(ChIP_peaks)
+
+peaks_df <- peaks_gr %>% 
+  as.data.frame() %>% 
+  dplyr::rename(peak_chrom = seqnames, peak_start = start, peak_end = end, peak_id = name, MACS2_enrichment = signalValue, MACS2_pvalue = pValue, MACS2_qValue = qValue) %>% 
+  select(-c(width, strand,score,peak))
+
+# read in WT ATAC peaks
+wt_atac <- rtracklayer::import(wt_ATAC_peaks) %>% 
+  resize(width = 201, fix = "center")
+
+# read in diff ATAC results
+diff_atac_df <- read_tsv(ATAC_results)
+diff_atac_gr <- makeGRangesFromDataFrame(diff_atac_df, keep.extra.columns = TRUE)
+
+atac_increased <- diff_atac_df %>% 
+  filter(is_diff, log2FoldChange > 0) %>% 
+  makeGRangesFromDataFrame()
+
+# generate overlap table and annotate ChIP classes
+ChIP_classes <- GRangesList(
+  ChIP_peak = peaks_gr,
+  wt_atac = wt_atac,
+  atac_increased = atac_increased
+) %>% 
+  
+  peak_overlap_table() %>% 
+  filter(ChIP_peak) %>% 
+  mutate(class = case_when(
+    ChIP_peak & wt_atac ~ "i",
+    ChIP_peak & !wt_atac & !atac_increased ~ "ii",
[…]

### 2022-09-13 workflow/scripts/fig1.R

@@ -1,0 +1,1535 @@
+# setup ========================================================================
+library(plotgardener)
+library(tidyverse)
+library(org.Dm.eg.db)
+library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
+library(grImport)
+library(grid)
+library(RColorBrewer)
+
+source("workflow/scripts/plot_heatmap.R")
+
+# function to get range for browser tracks -------------------------------------
+signal_range <- function(x, extra = 0.05) {
+  min <- floor(min(x))
+  max <- ceiling(max(x))
+  range <- max - min
+  margin <- range * extra
+  output <- c(min - margin, max + margin)
+  return(output)
+}
+
+# define input files ===========================================================
+Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
+Zld_WT_ATAC_bw <- snakemake@input[["Zld_WT_ATAC_bw"]]
+Zld_Zld_ATAC_bw <- snakemake@input[["Zld_Zld_ATAC_bw"]]
+Zld_WT_RNAseq_bw <- snakemake@input[["Zld_WT_RNAseq_bw"]]
+Zld_Zld_RNAseq_bw <- snakemake@input[["Zld_Zld_RNAseq_bw"]]
+
+Grh_ChIP_bw <- snakemake@input[["Grh_ChIP_bw"]]
+Grh_WT_ATAC_bw <- snakemake@input[["Grh_WT_ATAC_bw"]]
+Grh_Grh_ATAC_bw <- snakemake@input[["Grh_Grh_ATAC_bw"]]
+Grh_WT_RNAseq_bw <- snakemake@input[["Grh_WT_RNAseq_bw"]]
+Grh_Grh_RNAseq_bw <- snakemake@input[["Grh_Grh_RNAseq_bw"]]
+
+zld_ChIP_classes_fn <- snakemake@input[["zld_ChIP_classes"]]
+grh_ChIP_classes_fn <- snakemake@input[["grh_ChIP_classes"]]
+
+# # create blank layout for plot ===============================================
+pdf(snakemake@output[[1]], useDingbats = FALSE)
+# tiff("results/pub_figs/2022_drafts/fig1.tiff", width = 4.392, height = 3, units = "in", res = 300)
+pageCreate(width = 18.3, height = 12.5, default.units = "cm", showGuides = FALSE)
+
+# general figure settings ======================================================
+# text parameters for Nature Genetics
[…]

### 2022-09-13 workflow/scripts/fig2.R

@@ -1,0 +1,827 @@
+# setup ========================================================================
+library(plotgardener)
+library(tidyverse)
+library(org.Dm.eg.db)
+library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
+library(grImport)
+library(grid)
+
+source("workflow/scripts/plot_heatmap.R")
+
+# function to get range for browser tracks -------------------------------------
+signal_range <- function(x, extra = 0.05) {
+  min <- floor(min(x))
+  max <- ceiling(max(x))
+  range <- max - min
+  margin <- range * extra
+  output <- c(min - margin, max + margin)
+  return(output)
+}
+
+# define input files ===========================================================
+Twi_ChIP_bw <- snakemake@input[["Twi_ChIP_bw"]]
+Twi_WT_ATAC_bw <- snakemake@input[["Twi_WT_ATAC_bw"]]
+Twi_Twi_ATAC_bw <- snakemake@input[["Twi_Twi_ATAC_bw"]]
+
+Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
+Grh_ChIP_bw <- snakemake@input[["Grh_ChIP_bw"]]
+
+twi_ChIP_classes_fn <- snakemake@input[["twi_ChIP_classes"]]
+zld_ChIP_classes_fn <- snakemake@input[["zld_ChIP_classes"]]
+grh_ChIP_classes_fn <- snakemake@input[["grh_ChIP_classes"]]
+
+twi_RNAseq_results_fn <- snakemake@input[["twi_RNAseq_results"]]
+zld_RNAseq_results_fn <- snakemake@input[["zld_RNAseq_results"]]
+grh_RNAseq_results_fn <- snakemake@input[["grh_RNAseq_results"]]
+
+# create blank layout for plot =================================================
+pdf(snakemake@output[[1]], useDingbats = FALSE)
+pageCreate(width = 18.3, height = 12.5, default.units = "cm", showGuides = FALSE)
+
+# general figure settings ======================================================
+# text parameters for Nature Genetics
+panel_label_params <- pgParams(fontsize = 8)
+large_text_params <- pgParams(fontsize = 7)
+small_text_params <- pgParams(fontsize = 5)
+
+# colors
+twi_color <- "#00A08A"
+twi_heatmap_colors <- c("white","#00A08A", "#154734")
+
+zld_color <- "#5BBCD6"
[…]

### 2022-09-13 workflow/scripts/filter_ATAC_results.R

@@ -1,0 +1,37 @@
+# setup -----------------------------------------------------------------------
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(rtracklayer))
+suppressPackageStartupMessages(library(GenomicRanges))
+
+# read DESeq2 results ----------------------------------------------------------
+res_fn <- snakemake@input[[1]]
+results <- read_tsv(res_fn)
+
+
+# read in artifactual regions --------------------------------------------------
+# get name of TF
+factor <- basename(res_fn) %>% 
+  str_split(pattern = "_")
+factor <- factor[[1]][1] %>% 
+  str_split(pattern = "-")
+
+factor <- factor[[1]][2] %>% 
+  str_to_lower()
+
+# import artifacts file
+artifact_fn <- paste0("config/", factor, "_artifacts.bed")
+
+artifacts_gr <- rtracklayer::import(artifact_fn)
+
+# filter results table to exclude artifactual regions --------------------------
+results_gr <- makeGRangesFromDataFrame(results)
+
+overlaps <- findOverlaps(results_gr, artifacts_gr) %>% 
+  queryHits()
+
+filtered_results <- results[-overlaps,] 
+
+# export filtered peaks to file  -----------------------------------------------
+filtered_results %>% 
+  
+  write_tsv(snakemake@output[[1]])

### 2022-09-13 workflow/scripts/filter_ChIP_classes.R

@@ -1,0 +1,30 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(Biostrings))
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(GenomicRanges))
+suppressPackageStartupMessages(library(rtracklayer))
+
+# import input files ===========================================================
+chip_classes <- snakemake@input[[1]] %>% 
+  read_tsv()
+
+nonspecific_sites <- snakemake@input[[2]] %>% 
+  rtracklayer::import()
+
+# filter ChIP classes to exclude nonspecific class I sites =====================
+chip_classes_gr <- chip_classes %>% 
+  makeGRangesFromDataFrame()
+
+overlaps <- findOverlaps(chip_classes_gr, nonspecific_sites)
+
+nonspecific_sites <- chip_classes[overlaps@from,]
+
+chip_classes_filtered <- chip_classes %>% 
+  filter(!((peak_id %in% nonspecific_sites$peak_id) & (class == "i")))
+
+# write output data ============================================================
+chip_classes_filtered %>% 
+  write_tsv(snakemake@output[[1]])
+
+chip_classes_filtered %>% 
+  rtracklayer::export(snakemake@output[[2]])

### 2022-09-13 workflow/scripts/get_motif_classes.R

@@ -1,0 +1,132 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(GenomicRanges))
+suppressPackageStartupMessages(library(rtracklayer))
+suppressPackageStartupMessages(library(BSgenome.Dmelanogaster.UCSC.dm6))
+
+source("../../useful_functions/peak_processing.R")
+
+
+# define input files ===========================================================
+# S2 ChIP peaks and classes
+S2_bound_regions <- c(
+  all_peaks = snakemake@input[["all_peaks"]],
+  class_I_peaks = snakemake@input[["class_I_peaks"]],
+  class_II_peaks = snakemake@input[["class_II_peaks"]],
+  class_III_peaks = snakemake@input[["class_III_peaks"]],
+  motif_instances = snakemake@input[["motifs"]]
+)
+
+# other tissue ChIP peaks
+other_tissue_bound_regions <- c(
+  snakemake@input[["other_tissue_peaks"]]
+)
+names(other_tissue_bound_regions) <- str_replace(basename(other_tissue_bound_regions), ".bed", "")
+
+# Regions with repressive histone modifications
+repressed_regions <- c(
+  H3K27me3_peaks = snakemake@input[["H3K27me3_peaks"]],
+  H3K9me3_peaks = snakemake@input[["H3K9me3_peaks"]]
+)
+
+# nonspecific sites to exclude
+ns_sites <- c(
+  ns_sites = snakemake@input[["nonspecific_sites"]]
+)
+
+# combine all inputs
+input_files <- c(
+  S2_bound_regions,
+  other_tissue_bound_regions,
+  repressed_regions,
+  ns_sites
+)
+
+
+# Define potential binding sites ===============================================
+overlap_table <- input_files %>%
+  map(rtracklayer::import) 
+
+overlap_table$motif_instances <- resize(overlap_table$motif_instances, width = 500, fix = "center")
+
+overlap_table <- overlap_table %>%
+  GRangesList() %>%
+  peak_overlap_table()
+
+# annotate motifs ==============================================================
+other_tissue_names <- names(other_tissue_bound_regions)
+
+overlap_table <- overlap_table %>%
[…]

### 2022-09-13 workflow/scripts/get_nonspecific_peaks.R

@@ -1,0 +1,38 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(Biostrings))
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(GenomicRanges))
+suppressPackageStartupMessages(library(rtracklayer))
+suppressPackageStartupMessages(library(BSgenome.Dmelanogaster.UCSC.dm6))
+suppressPackageStartupMessages(library(Biostrings))
+
+source(snakemake@params[["r_source"]])
+
+
+# define input files ===========================================================
+input_files <- snakemake@input %>% 
+  unlist()
+
+# Define nonspecific binding sites =============================================
+nonspecific_sites <- input_files %>%
+  map(rtracklayer::import) %>%
+  GRangesList() %>%
+  peak_overlap_table() %>% 
+  mutate(n_samples = rowSums(select(.,6:8))) %>%
+  filter(n_samples == 3) %>% 
+  makeGRangesFromDataFrame()
+
+# write nonspecific sites to file ==============================================
+nonspecific_sites$peak_id <- paste0("nonspecific_classI_region_", 1:length(nonspecific_sites))
+seqlevels(nonspecific_sites) <- seqlevels(BSgenome.Dmelanogaster.UCSC.dm6)
+seqinfo(nonspecific_sites) <- seqinfo(BSgenome.Dmelanogaster.UCSC.dm6)
+nonspecific_sites <- trim(nonspecific_sites)
+
+seqs <- getSeq(BSgenome.Dmelanogaster.UCSC.dm6, nonspecific_sites)
+names(seqs) <- nonspecific_sites$peak_id
+
+# write peaks and sequences to files
+writeXStringSet(seqs, snakemake@output[[1]])
+
+nonspecific_sites$name <- nonspecific_sites$peak_id
+rtracklayer::export(nonspecific_sites, snakemake@output[[2]])

### 2022-09-13 workflow/scripts/plot_heatmap.R

@@ -1,0 +1,543 @@
+# functions for plotting deeptools-style heatmaps ================================================
+# function for quickly computing a coverage matrix from a bigwig file. Core code taken from seqPlots package
+coverage_matrix <- function(file,regions, upstream = 1000, downstream = 1000, region_width = 1, bin_size =10) {
+  require(rtracklayer)
+  require(GenomicRanges)
+  
+  # check arguments
+  if (!is.character(file)) stop("file should be a character vector")
+  
+  if (!file.exists(file)) stop("one or more input files not found")
+  
+  # get regions to include in heatmap
+  gr <- GenomicRanges::resize(regions, region_width, fix='center')
+  gr <- suppressWarnings( GenomicRanges::promoters(gr, upstream, downstream ) )
+  gr <- trim(gr)
+  hm_bins  <- seq(-upstream, downstream, by=bin_size )
+  
+  # extract_matrix function taken from seqPlots package
+  extract_matrix <- function(track, gr, size, ignore_strand) {
+    sum <- suppressWarnings(.Call(
+      'BWGFile_summary', path.expand(path(track)),
+      as.character(seqnames(gr)), ranges(gr), 
+      S4Vectors::recycleIntegerArg(size, "size", length(gr)), 
+      "mean", as.numeric(NA_real_), PACKAGE='rtracklayer'
+    ))
+    M <- do.call( rbind, sum )
+    if (!ignore_strand) 
+      M[as.character(strand(gr))=='-',] <- M[
+        as.character(strand(gr))=='-', ncol(M):1]
+    return(M)
+  }
+  
+  # read bigwig file to matrix
+  bwf <- BigWigFile(file)
+  coverage_matrix <- extract_matrix(bwf, gr, length(hm_bins), ignore_strand = TRUE)
+  
+  return(coverage_matrix)
+}
+
+
+plot_heatmap <- function(bw, regions, 
+                         upstream = 1000, downstream = 1000, region_width = 1,
+                         row_split = NULL,
+                         order_by_samples = NULL ,
+                         colors = c("#440154FF", "#21908CFF", "#FDE725FF"),
+                         scale_group = NULL,
+                         individual_scales = FALSE, 
+                         return_heatmap_list = FALSE,
+                         min_percentile = 0.01,
+                         max_percentile = 0.98,
+                         row_order = NULL,
+                         plot_average = TRUE,
[…]

### 2022-09-13 workflow/scripts/process_peaks.R

@@ -1,0 +1,71 @@
+# setup -----------------------------------------------------------------------
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(rtracklayer))
+suppressPackageStartupMessages(library(GenomicRanges))
+
+# define functions -------------------------------------------------------------
+# function to extract peak summits from narrowPeak file
+# peak start and end values will be replaced with the summit location
+extract_summits <- function(gr, extend_width = 1L) {
+  # verify input is a GRanges object
+  if (!inherits(gr, "GRanges") ) {
+    stop("x must be a GRanges object")
+  }
+  # verify that gr object is in narrowPeak format
+  if ( !all(names(mcols(gr))  %in% c("name", "score", "signalValue", "pValue",  "qValue",  "peak"))) {
+    stop(strwrap("GRanges object does not appear to be in narrowPeak format. Object should contain the following metadata columns: name, score, signalValue, pValue, qValue and peak"))
+  }
+  
+  if (all(gr$peak == -1)) {
+    stop("All values for 'peak' column == -1, indicating that summits were not called. Verify that your peak caller called peak summits")
+  }
+  
+  # replace peak start and end values with summit location
+  summit <- start(gr) + gr$peak - 1
+  start(gr) <- summit
+  end(gr) <- summit
+  
+  # resize peaks to desired width
+  gr <- resize(gr, width = extend_width, fix = "center")
+
+  # remove now meaningless peak column
+  gr$peak <- NULL
+  
+  return(gr)
+}
+
+
+# read peaks -------------------------------------------------------------------
+peak_fn <- snakemake@input[[1]]
+peaks <- rtracklayer::import(peak_fn)
+
+
+# extend peak summits ----------------------------------------------------------
+extended_summits <- peaks %>% 
+  extract_summits(extend_width = as.integer(snakemake@params[["extend_width"]]))
+
+# filter out artifactual regions -----------------------------------------------
+# get name of TF
+factor <- basename(peak_fn) %>% 
+  str_split(pattern = "_")
+factor <- factor[[1]][2] %>% 
+  str_remove("a") %>% 
+  str_to_lower()
+
+# import artifacts file
+artifact_fn <- paste0("resources/", factor, "_regions_to_exclude.bed")
+
+artifacts_gr <- rtracklayer::import(artifact_fn)
+
[…]

### 2022-09-13 workflow/scripts/utils.R

@@ -1,0 +1,127 @@
+# function to extract peak summits from narrowPeak file
+# peak start and end values will be replaced with the summit location
+extract_summits <- function(gr, extend_width = 1L) {
+  # verify input is a GRanges object
+  if (!inherits(gr, "GRanges") ) {
+    stop("x must be a GRanges object")
+  }
+  # verify that gr object is in narrowPeak format
+  if ( !all(names(mcols(gr))  %in% c("name", "score", "signalValue", "pValue",  "qValue",  "peak"))) {
+    stop(strwrap("GRanges object does not appear to be in narrowPeak format. Object should contain the following metadata columns: name, score, signalValue, pValue, qValue and peak"))
+  }
+  
+  if (all(gr$peak == -1)) {
+    stop("All values for 'peak' column == -1, indicating that summits were not called. Verify that your peak caller called peak summits")
+  }
+  
+  # replace peak start and end values with summit location
+  summit <- start(gr) + gr$peak - 1
+  start(gr) <- summit
+  end(gr) <- summit
+  
+  # resize peaks to desired width
+  gr <- resize(gr, width = extend_width, fix = "center")
+  
+  # remove now meaningless peak column
+  gr$peak <- NULL
+  
+  return(gr)
+}
+
+# function to merge multiple sets of ChIP peaks into a non-overlapping master set
+# supports multiple methods
+# peaks should be a GRangesList
+# method should be either "overlap" or "merge": 
+# overlap method will determine overlaps among all peak sets and combine non-overlapping peaks
+# merge method will combine overlapping peaks into a single peak
+combine_peaks <- function(peaks, method = "overlap", min_overlap = 1L) {
+  require(GenomicRanges)
+  
+  # check arguments
+  if (length(peaks) < 2) {
+    stop("Only one set of peaks provided. For merging peaks, provide multiple peak sets.")
+  }
+  
+  if (!inherits(peaks, "GRangesList") ) {
+    stop("one or more peak set is not a Granges object. peaks must be a GRangesList")
+  }
+  
+  if (!any(method == c("overlap", "merge"))) {
+    stop("Invalid method provided. method should be 'overlap' or 'merge' ")
+  }
+  
+  if (!is_integer(min_overlap)) {
+    stop("min_overlap must be an integer")
+  }
+  
+  
[…]

### 2022-09-13 manuscript/manuscript.Rmd
fix gene models in browser tracks in fig 1

@@
 # Results
+## Ectopic pioneer factors bind and open closed chromatin
+
+## Chromatin opening correlates with transcriptional activation
+
+## Twist binds closed chromatin extensively and opens chromatin at some sites
+
+## Motif content distinguishes chromatin opening by Twist from Zelda and Grainy head
+
+## Ectopic transcription factors bind nonspecifically to active chromatin
 

### 2022-09-13 workflow/scripts/fig1.R

@@ -1233,9 +1233,9 @@ annoYaxis(
 
 g1 <- plotGenes(
   params = region, assembly = "dm6",
-  x = (ref_x + x_offset_browser), y = (ref_y + 2.6), height = gb_height, width = 2.75, fontsize = small_text_params$fontsize,
+  x = (ref_x + x_offset_browser), y = (ref_y + 2.5), height = 0.6, width = 2.75, fontsize = small_text_params$fontsize,
   fill = c("gray50", "gray50"),
-  just = c("left", "center"),
+  just = c("left", "top"),
   default.units = "cm"
 )
 
@@ -1259,7 +1259,7 @@ annoHighlight(
 # Panel J ======================================================================
 # panel label
 ref_x <- 13
-ref_y <- 3.25 
+ref_y <- 3.6 
 
 
 
@@ -1298,7 +1298,7 @@ j_plot <- zld_chip_classes %>%
 plotGG(
   plot = j_plot,
   x = (ref_x), y = (ref_y),
-  width = 5, height = 3, just = c("left", "top"),
+  width = 5, height = 2.65, just = c("left", "top"),
   default.units = "cm"
 )
 
@@ -1448,9 +1448,9 @@ annoYaxis(
 
 g1 <- plotGenes(
   params = region, assembly = "dm6",
-  x = (ref_x + x_offset_browser), y = (ref_y + 2.6), height = gb_height, width = 2.75, fontsize = small_text_params$fontsize, 
+  x = (ref_x + x_offset_browser), y = (ref_y + 2.5), height = 0.6, width = 2.75, fontsize = small_text_params$fontsize, 
   fill = c("gray50", "gray50"),
-  just = c("left", "center"),
+  just = c("left", "top"),
   default.units = "cm"
 )
 
@@ -1474,7 +1474,7 @@ annoHighlight(
 # Panel L ======================================================================
 # panel label
 ref_x <- 13
-ref_y <- 9.25 
+ref_y <- 9.6
 
 # panel label
 plotText(
@@ -1509,7 +1509,7 @@ l_plot <- grh_chip_classes %>%
 plotGG(
   plot = l_plot,
   x = (ref_x), y = (ref_y),
-  width = 5, height = 3, just = c("left", "top"),
+  width = 5, height = 2.65, just = c("left", "top"),
   default.units = "cm"
 )
 # l_plot <- grh_chip_classes %>% 

### 2022-09-20 manuscript/manuscript.Rmd
add rules for motif enrichment on embryo ChIP-seq data

@@
 # Results
+
 ## Ectopic pioneer factors bind and open closed chromatin
@@
 | goat anti-rabbit IgG-HRP conjugate | Bio-Rad                  | 1706515              |
-|                                    |                          |                      |
+| Anti-HA clone 12CA5                |                          |                      |
 |                                    |                          |                      |

### 2022-09-20 workflow/scripts/get_PWM_instances.R

@@ -1,0 +1,85 @@
+# load packages
+library(tidyverse)
+library(universalmotif)
+library(Biostrings)
+library(BSgenome.Dmelanogaster.UCSC.dm6)
+
+# test reading meme output
+twi_meme <- read_meme("published_ChIPseq/results/motifs/embryo-1-3H_aTwi/meme.txt")
+twi_pwm <- twi_meme[[8]]@motif
+
+grh_meme <- read_meme("published_ChIPseq/results/motifs/embryo-15-16H_aGrh/meme.txt")
+grh_pwm <- grh_meme[[3]]@motif
+
+zld_meme <- read_meme("published_ChIPseq/results/motifs/embryo-nc14_aZld/meme.txt")
+zld_pwm <- zld_meme[[3]]@motif
+
+# score motif occurences in genome
+twi_motif_instances <- matchPWM(twi_pwm, BSgenome.Dmelanogaster.UCSC.dm6, min.score = "80%", with.score = TRUE)
+grh_motif_instances <- matchPWM(grh_pwm, BSgenome.Dmelanogaster.UCSC.dm6, min.score = "80%", with.score = TRUE)
+zld_motif_instances <- matchPWM(zld_pwm, BSgenome.Dmelanogaster.UCSC.dm6, min.score = "80%", with.score = TRUE)
+
+
+
+hist( mcols(twi_motif_instances)$score / maxScore(twi_pwm) )
+hist( mcols(grh_motif_instances)$score / maxScore(grh_pwm) )
+hist( mcols(zld_motif_instances)$score / maxScore(zld_pwm) )
+
+# export motif occurences
+twi_motif_instances %>% 
+  as.data.frame() %>% 
+  write_tsv("results/motif_instances/twi_motifs_80percent.tsv")
+
+zld_motif_instances %>% 
+  as.data.frame() %>% 
+  write_tsv("results/motif_instances/zld_motifs_80percent.tsv")
+
+grh_motif_instances %>% 
+  as.data.frame() %>% 
+  write_tsv("results/motif_instances/grh_motifs_80percent.tsv")
+
+
+twi_motif_instances %>% 
+  plyranges::filter(strand == "+") %>% 
+  plyranges::mutate(score = score / maxScore(twi_pwm) * 100 - 80) %>% 
+  plyranges::reduce_ranges(score = mean(score)) %>% 
+  export("results/motif_instances/twi_motifs_80percent_plus.bw")
+
+twi_motif_instances %>% 
+  plyranges::filter(strand == "-") %>% 
[…]

### 2022-09-20 workflow/scripts/get_embryo_peak_sequences.R

@@ -1,0 +1,30 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(Biostrings))
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(BSgenome.Dmelanogaster.UCSC.dm6))
+suppressPackageStartupMessages(library(rtracklayer))
+
+source(snakemake@params[["r_source"]])
+
+# get peak sequences ===========================================================
+# get input peaks
+peak_file <- snakemake@input[[1]]
+sites_for_seqs <- rtracklayer::import(peak_file)
+
+# extend peak summits ----------------------------------------------------------
+peaks <- sites_for_seqs %>% 
+  extract_summits(extend_width = 201L)
+
+# get underlying peak sequences ------------------------------------------------
+seqlevels(peaks) <- seqlevels(BSgenome.Dmelanogaster.UCSC.dm6)
+seqinfo(peaks) <- seqinfo(BSgenome.Dmelanogaster.UCSC.dm6)
+peaks <- trim(peaks)
+
+peak_seqs <- getSeq(BSgenome.Dmelanogaster.UCSC.dm6, peaks)
+names(peak_seqs) <- peaks$name
+
+
+
+# export to bed file ===========================================================
+writeXStringSet(peak_seqs, snakemake@output[[1]])
+

### 2022-09-20 workflow/scripts/process_peaks.R

@@ -3,37 +3,7 @@ suppressPackageStartupMessages(library(tidyverse))
 suppressPackageStartupMessages(library(rtracklayer))
 suppressPackageStartupMessages(library(GenomicRanges))
 
-# define functions -------------------------------------------------------------
-# function to extract peak summits from narrowPeak file
-# peak start and end values will be replaced with the summit location
-extract_summits <- function(gr, extend_width = 1L) {
-  # verify input is a GRanges object
-  if (!inherits(gr, "GRanges") ) {
-    stop("x must be a GRanges object")
-  }
-  # verify that gr object is in narrowPeak format
-  if ( !all(names(mcols(gr))  %in% c("name", "score", "signalValue", "pValue",  "qValue",  "peak"))) {
-    stop(strwrap("GRanges object does not appear to be in narrowPeak format. Object should contain the following metadata columns: name, score, signalValue, pValue, qValue and peak"))
-  }
-  
-  if (all(gr$peak == -1)) {
-    stop("All values for 'peak' column == -1, indicating that summits were not called. Verify that your peak caller called peak summits")
-  }
-  
-  # replace peak start and end values with summit location
-  summit <- start(gr) + gr$peak - 1
-  start(gr) <- summit
-  end(gr) <- summit
-  
-  # resize peaks to desired width
-  gr <- resize(gr, width = extend_width, fix = "center")
-
-  # remove now meaningless peak column
-  gr$peak <- NULL
-  
-  return(gr)
-}
-
+source(snakemake@params[["r_source"]])
 
 # read peaks -------------------------------------------------------------------
 peak_fn <- snakemake@input[[1]]

### 2022-09-23 manuscript/manuscript.Rmd
update figure titles and add fig 3

@@
 | Reagent                            | Source/vendor            | Catalog/accession \# |
-|----------------------------|------------------------|--------------------|
+|------------------------------------|--------------------------|----------------------|
 | pMT-puro                           | AddGene                  | 17923                |
@@
 
-**Figure 2: Twist binds closed chromatin extensively and drives accessibility at some sites.**
+**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.**
 
-**Figure 3: Some endogenous targets are resistant to binding by ectopic pioneer factors.**
+**Figure 3: Motif content shapes pioneer-factor activity**
 
-**Figure 4: Binding and opening of closed chromatin is concentration dependent.**
+**Figure 4: H3K27me3 and H3K9me3 are barriers to pioneer-factor binding.**
 
-**Figure 5: H3K27me3 and H3K9me3 are barriers to pioneer-factor binding.**
+**Figure 5: Binding and opening of closed chromatin is concentration dependent.**
 

### 2022-09-23 workflow/scripts/fig2.R

@@ -5,6 +5,8 @@ library(org.Dm.eg.db)
 library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
 library(grImport)
 library(grid)
+library(RColorBrewer)
+
 
 source("workflow/scripts/plot_heatmap.R")
 
@@ -46,7 +48,8 @@ small_text_params <- pgParams(fontsize = 5)
 
 # colors
 twi_color <- "#00A08A"
-twi_heatmap_colors <- c("white","#00A08A", "#154734")
+# twi_heatmap_colors <- c("white","#00A08A", "#154734")
+twi_heatmap_colors <- brewer.pal(9, "GnBu")
 
 zld_color <- "#5BBCD6"
 grh_color <- "#F98400"
@@ -498,6 +501,8 @@ plotText(
   x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
 )
 
+
+# generate heatmap
 bw <- c(
   S2_Twi_ChIP =  Twi_ChIP_bw,
   S2_WT_ATAC = Twi_WT_ATAC_bw,
@@ -527,6 +532,7 @@ hm <- plot_heatmap_minimal(
 d_hm <- grid.grabExpr(draw(hm, show_heatmap_legend = FALSE, padding = unit(c(0, 0, 0, 0), "mm")))
 
 
+# place heatmap on page
 plotGG(
   plot = d_hm,
   x = (ref_x + 0.75), y = (ref_y + 0.25),
@@ -567,260 +573,260 @@ plotText(
 
 
 
-# # Panel E ======================================================================
-# panel label
-ref_x <- 0.5
-ref_y <- 6.75
-
-# panel label
-plotText(
-  label = "e", params = panel_label_params, fontface = "bold",
-  x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
-)
-
-# e_plot <- twi_chip_classes %>%
-#   ggplot(aes(x = class, y = atac_FC)) +
-#   geom_violin(fill = twi_color, lwd = 0.1) +
-#   geom_boxplot(width = 0.1, outlier.shape = NA, lwd = 0.1) +
-#   ylab(expression(ATAC~log[2]~fold~change) ) +
-#   theme_classic(base_size = small_text_params$fontsize)
+# # # Panel E ======================================================================
+# # panel label
+# ref_x <- 0.5
+# ref_y <- 6.75
+# 
+# # panel label
+# plotText(
+#   label = "e", params = panel_label_params, fontface = "bold",
[…]

### 2022-09-23 workflow/scripts/get_motif_PWM.R

@@ -1,15 +1,17 @@
-# load packages
-library(tidyverse)
-library(universalmotif)
-library(Biostrings)
+# setup ========================================================================
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(universalmotif))
+suppressPackageStartupMessages(library(Biostrings))
 
-# get input from snakemake
+# get input from snakemake =====================================================
 meme_file <- snakemake@input[["meme_file"]]
-motif_of_interest <- snakemake@input[["meme_motif_number"]]
+motif_of_interest <- snakemake@params[["meme_motif_number"]]
 
-# get motif of interest
+# extract PWM for motif of interest ============================================
 meme <- read_meme(meme_file)
-pwm <- twi_meme[[motif_of_interest]]@motif
+pwm <- meme[[motif_of_interest]]@motif
 
-# write output
-write_tsv(pwm, snakemake@output[[1]])
+# write output =================================================================
+pwm %>% 
+  as.data.frame() %>% 
+  write.table(snakemake@output[[1]], sep = "\t", quote = FALSE)

### 2022-09-23 workflow/scripts/utils.R

@@ -1,4 +1,4 @@
-# function to extract peak summits from narrowPeak file
+# function to extract peak summits from narrowPeak file ==================================================================
 # peak start and end values will be replaced with the summit location
 extract_summits <- function(gr, extend_width = 1L) {
   # verify input is a GRanges object
@@ -28,7 +28,7 @@ extract_summits <- function(gr, extend_width = 1L) {
   return(gr)
 }
 
-# function to merge multiple sets of ChIP peaks into a non-overlapping master set
+# function to merge multiple sets of ChIP peaks into a non-overlapping master set =========================================================
 # supports multiple methods
 # peaks should be a GRangesList
 # method should be either "overlap" or "merge": 
@@ -75,7 +75,7 @@ combine_peaks <- function(peaks, method = "overlap", min_overlap = 1L) {
   
 }
 
-# function to build an overlap table
+# function to build an overlap table =========================================================
 # takes a Granges list object
 # combines all nonoverlapping peaks from all samples, then builds a logical table indicating which peaks from each sample overlap each peak on the master list
 peak_overlap_table <- function(peaks, method = "overlap", min_overlap = 1L, combine_peaks = TRUE) {
@@ -124,4 +124,10 @@ peak_overlap_table <- function(peaks, method = "overlap", min_overlap = 1L, comb
     all_peaks_df[overlaps,sample_name] <- TRUE
   }
   return(all_peaks_df)
+}
+
+# function to check if a sequence is a palindrome ===============================
+is_palindrome <- function(x) {
+  require(Biostrings)
+  return(ifelse(x == reverseComplement(x), TRUE, FALSE) )
 }

### 2022-10-14 manuscript/manuscript.Rmd
update config.yaml for motif analysis and outline supp figs

@@
 | Reagent                            | Source/vendor            | Catalog/accession \# |
-|------------------------------------|--------------------------|----------------------|
+|------------------------------|-----------------------|-------------------|
 | pMT-puro                           | AddGene                  | 17923                |
@@
 
+**Extended data Figure 1: Generation of stable cell lines.**
+
 **Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.**
@@
 **Figure 7: Pioneer-factor function requires stable chromatin occupancy.**
+
+**Information to include in supplemental figures:**
+
+-   Schematic of stable cell line generation.
+
+-   Cell line validation by RNA-seq
+
+-   Verification of protein expression.
+
+-   Determination of physiological protein levels.
+
+-   QC for genomics datasets: e.g. PCA, replicate correlation, example genome browsers of individual replicates.
+
+-   Antibody controls: IgG, IPs in WT cells, etc.
+
+-   CUT&RUN time course for Zld and Grh?
+
+-   De novo motif results
+
+-   Analysis of promoter motifs
+
+-   GO analysis for RNA-seq data
+
+-

### 2022-10-14 workflow/scripts/get_motif_classes.R

@@ -4,7 +4,7 @@ suppressPackageStartupMessages(library(GenomicRanges))
 suppressPackageStartupMessages(library(rtracklayer))
 suppressPackageStartupMessages(library(BSgenome.Dmelanogaster.UCSC.dm6))
 
-source("../../useful_functions/peak_processing.R")
+source("workflow/scripts/utils.R")
 
 
 # define input files ===========================================================

### 2022-11-17 manuscript/manuscript.Rmd
update methods section

@@
 | PVDF Membrane                      |                          |                      |
+| Formaldehyde                       |                          |                      |
 | **Antibodies**                     |                          |                      |
+|                                    |                          |                      |
 | Anti-Zld                           | Harrison et al., 2010    |                      |
@@
 | Anti-HA clone 12CA5                |                          |                      |
-|                                    |                          |                      |
+| Anti-H3K27me3                      |                          |                      |
 |                                    |                          |                      |
@@
 
+Transcription factor expression was induced by adding CuSO~4~ to the cell culture media.
+Expression of protein at physiological levels was achieved with 1000, 100, or 40 µM CuSO~4~ for Zld, Grh or Twi, respectively (Supplemental figure X).
+For experiments with sub- or supraphysiological protein levels (Fig.
+5), the following CuSO~4~ concentrations were used: 500 µM (Zld subphysiological), 1000 µM (Zld physiological) or 1500 µM (Zld supraphysiological); 25 µM (Grh subphysiological), 100 µM (Grh physiological) or 400 µM (Grh supraphysiological); 10 µM (Twi subphysiological), 40 µM (Twi physiological) or 160 µM (Twi supraphysiological).
+For all induction experiments, cells were plated at 1✕10^6^ cells per mL.
+Unless otherwise specified, cells were harvested 48 hours after induction.
+
 ## Immunoblotting
@@
 ## ChIP-seq
+
+For each ChIP experiment, \~50 million cells were crosslinked by adding methanol-free formaldehyde directly to the cell culture medium to a final concentration of 0.8%.
+Cells were rotated on a nutator for 7 minutes at room temperature (RT) before quenching of crosslinking by adding glycine to XX mM.
+Crosslinked cells were pelleted by centrifugation at 600✕g for 3 minutes.
+Cells were washed twice with 1X PBS.
 
@@
 
-## Treatment of S2 cells with small molecule inhibitors
+## Treatment of S2 cells with Tazemetostat
 

### 2022-11-18 manuscript/manuscript.Rmd
update methods section
add results outline to MS

--- 2022-10-14 manuscript/manuscript.Rmd
@@
 
-## Ectopic pioneer factors bind and open closed chromatin
+## Ectopically expressed pioneer factors bind and open closed chromatin
+
+-   Gain of function system is necessary to understand mechanisms of PF activity
+
+-   Establish system for ectopic expression of pioneer factors
+
+-   Identification of physiological expression levels
+
+-   Define binding classes I-III and describe characteristics (distribution, promoters/enhancers, etc.)
 
@@
 
-## Twist binds closed chromatin extensively and opens chromatin at some sites
+-   Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.
 
-## Motif content distinguishes chromatin opening by Twist from Zelda and Grainy head
+-   Comparison of RNA-seq and ATAC data: chromatin opening correlates with transcriptional activation.
 
-## Ectopic transcription factors bind nonspecifically to active chromatin
+-   Class III promoters are most likely to be associated with upregulated genes
+
+## Twist binds closed chromatin extensively and opens chromatin at a subset of sites
+
+-   Introduce Twist and justify the comparison to Zld/Grh.
+
+-   Twi binds closed chromatin and opens at some sites
+
+-   Class I-III sites, distribution, etc.
+
+-   Twi also promotes transcriptional activation.
+
+-   Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.
+
+-   Comparison of RNA-seq and ATAC
+
+## Ectopically expressed transcription factors bind opportunistically to active chromatin
+
+-   Overlap of class I regions suggests nonspecific binding
+
+-   Analysis of published ChIP data shows that class I regions are within active chromatin
+
+-   RNA-seq and ATAC-seq for class I to show that the binding is not functional
+
+## Class II and III regions are mostly in naïve chromatin
+
+-   Analysis of chromatin at class II-III regions: mostly devoid of histone modifications
+
+-   Class III regions have higher H3 and H1 occupancy compared to class II
+
+-   Class II and III have poorly positioned nucleosomes
+
+## Chromatin opening by pioneer factors is driven by motif content
+
+-   De novo motif enrichment for ChIP peaks (supplemental)
+
+-   Motif analysis
+
+## Ectopic Zld, Grh and Twi fail to access many endogenous binding sites
+
+## Repressive chromatin is a barrier to pioneer-factor binding
+
+## Pioneer-factor binding and chromatin opening is concentration dependent
+
+## Regions outside the DNA-binding domain are required for binding and opening of closed chromatin
 

### 2022-11-18 manuscript/manuscript.Rmd

--- 2022-11-17 manuscript/manuscript.Rmd
@@
 
-**Extended data Figure 1: Generation of stable cell lines.**
+**Extended data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.**
 

### 2022-11-18 manuscript/manuscript.qmd
start results for Fig. 1

--- 2022-11-18 manuscript/manuscript.Rmd
@@
 ---
-title: "S2 pioneer factor MS"
-author: "Tyler Gibson"
-date: "8/24/2021"
-output: html_document
+title: "Defining properties regulating pioneer-factor binding and function"
+format: html
+editor: visual
+bibliography: references.bib
 ---
@@
 -   Define binding classes I-III and describe characteristics (distribution, promoters/enhancers, etc.)
+
+-   Time course and dynamics of binding to different classes (supplement)
+
+To investigate the mechanisms of chromatin binding and opening by pioneer factors, we generated stable cell lines inducibly expressing Zelda (Zld) and Grainy head (Grh) in Drosophila S2 cells (Extended Data Fig.
+XX), where neither factor is endogenously expressed.
+We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers (Extended Data Fig.
+1b) [@Cherbas2011].
+We confirmed that Zld and Grh RNA (Extended Data Fig.
+XX) and protein (Extended Data Fig.
+XX) were not present in WT S2 cells.
+The stable cell lines express Zld or Grh under the control of the metallothionein promoter, which allows inducible and tunable expression with the addition of CuSO~4.~ To determine protein levels that would be approximately physiological, we compared Zld or Grh levels in cells with varying concentrations of CuSO~4~ to the levels of endogenous expression found in Drosophila embryos (Extended Data Fig.
+XX).
+Physiological protein levels were achieved at 1000 mM or 100 mM CuSO~4~ for Zld or Grh, respectively.
+
+To determine how exogenous Zld and Grh interact with preexisting chromatin in S2 cells, we expressed each factor at physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
+We identified XXXX peaks for Zld and XXXX for Grh.
 
@@
 
--   Class II and III have poorly positioned nucleosomes
-
 ## Chromatin opening by pioneer factors is driven by motif content
@@
 
--   Motif analysis
+-   Motif analysis: percent with motif, average n motifs, motif quality
+
+-   Motif positioning around nucleosomes? (supplemental)
 
@@
 
+-   Zelda binding is more similar to the brain than the embryo
+
+-   Grh binding is stable across tissues
+
+-   Twi sites in S2 cells have low level binding in embryo, but strongest peaks from embryo are not bound
+
 ## Repressive chromatin is a barrier to pioneer-factor binding
+
+-   Enrichment of repressive marks at sites bound in other tissues
+
+-   Results of Taz experiments (TBD)
+
+-   Comment on sites that are not bound and lack repressive chromatin.
 
@@
 
+-   For all three factors, increased concentration results in increased binding and opening of class II and III sites.
+
+-   For all three factors, increased concentration leads to novel binding sites.
+
+-   For Zld and Twi, increased concentration is insufficient to allow binding to repressed sites.
+
+-   For Grh, increased concentration does allow binding and opening at repressed sites
+
 ## Regions outside the DNA-binding domain are required for binding and opening of closed chromatin
+
+-   Generation of stable cell lines to express DBDs at same concentration as FL proteins
+
+-   Zld DBD is totally unable to bind to closed sites.
+
+-   In supplement, show Zld DBD binding specifically at class II and III sites with CAGGTAG to support that this is not due to a loss of ZnF2
+
+-   Grh DBD has reduced ability to access closed chromatin
+
+-   Neither DBD is sufficient for chromatin opening.
 
@@
 
-| Reagent                            | Source/vendor            | Catalog/accession \# |
-|------------------------------|-----------------------|-------------------|
-| pMT-puro                           | AddGene                  | 17923                |
-| Gibson Cloning                     | New England Biolabs      | E2611                |
-| Schneider's Medium                 | Thermo Fisher Scientific | 21720001             |
-| Fetal Bovine Serum                 | Omega Scientific         | FB-02                |
-| Antibiotic/antimycotic             | Thermo Fisher Scientific | 15240062             |
-| Effectene Transfection Reagent     | Qiagen                   | 301427               |
-| PVDF Membrane                      |                          |                      |
-| Formaldehyde                       |                          |                      |
-| **Antibodies**                     |                          |                      |
-|                                    |                          |                      |
-| Anti-Zld                           | Harrison et al., 2010    |                      |
-| Anti-Grh                           | Harrison et al., 2010    |                      |
-| Anti-Twi                           | He et al., 2015          |                      |
-| Anti-HA-peroxidase clone 3F10      | Roche                    | 12013819001          |
-| Anti-Tubulin DM1A                  | Sigma                    | T6199-200UL          |
-| goat anti-rabbit IgG-HRP conjugate | Bio-Rad                  | 1706515              |
-| Anti-HA clone 12CA5                |                          |                      |
-| Anti-H3K27me3                      |                          |                      |
-|                                    |                          |                      |
-|                                    |                          |                      |
+| Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
+|-------------------|------------------|------------------|------------------|
+| pMT-puro                              | AddGene                   | 17923                |                         |
+| Gibson Cloning                        | New England Biolabs       | E2611                |                         |
+| Schneider's Medium                    | Thermo Fisher Scientific  | 21720001             |                         |
+| Fetal Bovine Serum                    | Omega Scientific          | FB-02                |                         |
+| Antibiotic/antimycotic                | Thermo Fisher Scientific  | 15240062             |                         |
+| Effectene Transfection Reagent        | Qiagen                    | 301427               |                         |
+| PVDF Membrane                         |                           |                      |                         |
+| Formaldehyde                          |                           |                      |                         |
+| Protease inhibitor tablets            |                           |                      |                         |
+| NEBNext Ultra II DNA library prep kit |                           |                      |                         |
+| TruSeq v2 RNA library prep kit        |                           |                      |                         |
+| Trizol                                |                           |                      |                         |
+| **Antibodies**                        |                           |                      |                         |
+| Anti-Zld                              | Harrison et al., 2010     |                      | ChIP: 6 µL , WB: 1:750  |
+| Anti-Grh                              | Harrison et al., 2010     |                      | ChIP: 8 µL, WB: 1:1000  |
+| Anti-Twi                              | He et al., 2015           |                      | ChIP: 10 µg, WB: 1:1000 |
+| Anti-HA-peroxidase clone 3F10         | Roche                     | 12013819001          | WB: 1:500               |
+| Anti-Tubulin DM1A                     | Sigma                     | T6199-200UL          | WB: 1:5000              |
+| goat anti-rabbit IgG-HRP conjugate    | Bio-Rad                   | 1706515              | WB: 1:3000              |
+| Anti-HA clone 12CA5                   | Roche                     |                      | ChIP: 7.5 µL            |
+| Anti-H3K27me3                         | Cell Signaling Technology |                      | WB:, CR:                |
+|                                       |                           |                      |                         |
+|                                       |                           |                      |                         |
 
@@
 5), the following CuSO~4~ concentrations were used: 500 µM (Zld subphysiological), 1000 µM (Zld physiological) or 1500 µM (Zld supraphysiological); 25 µM (Grh subphysiological), 100 µM (Grh physiological) or 400 µM (Grh supraphysiological); 10 µM (Twi subphysiological), 40 µM (Twi physiological) or 160 µM (Twi supraphysiological).
-For all induction experiments, cells were plated at 1✕10^6^ cells per mL.
-Unless otherwise specified, cells were harvested 48 hours after induction.
+For all induction experiments, cells were plated at 1×10^6^ cells per mL.
+Unless otherwise specified, experiments were performed with cells harvested 48 hours after induction.
 
@@
 For each ChIP experiment, \~50 million cells were crosslinked by adding methanol-free formaldehyde directly to the cell culture medium to a final concentration of 0.8%.
-Cells were rotated on a nutator for 7 minutes at room temperature (RT) before quenching of crosslinking by adding glycine to XX mM.
-Crosslinked cells were pelleted by centrifugation at 600✕g for 3 minutes.
+Cells were rotated on a nutator for 7 minutes at room temperature (RT) before quenching of crosslinking by adding glycine to 125 mM.
+Crosslinked cells were pelleted by centrifugation at 600 × *g* for 3 minutes.
 Cells were washed twice with 1X PBS.
+Cells were lysed by resuspending in lysis buffer (50 mM HEPES pH 7.9, 140 mM NaCl, 1 mM EDTA, 10% glycerol, 0.5% NP40, 0.25% Triton-X 100) with protease inhibitors (Pierce) and incubating on ice for 10 minutes.
+The resulting chromatin was pelleted for 5 minutes at 1500 × *g* and resuspended in RIPA buffer.
+Chromatin was sonicated in a Covaris S220 Ultrasonicator (5 cycles of 120 seconds with 60 second delay, 170 peak power, 10% duty factor, 200 cycles/burst).
+Sonicated chromatin was centrifuged for 10 minutes at 10,000 × *g* to pellet insoluble material.
+5% of the supernatant was set aside as input, and the remainder was incubated at 4°C overnight with antibodies (6 µL anti-Zld, 8 µL anti-Grh, 10 µg anti-Twi, 7.5 µL anti-HA).
+20 µL protein A beads (Dynabeads Protein A, ThermoFisher Scientific) blocked with BSA were added and samples were incubated at 4°C for 4 hours.
+Beads were separated on a magnet and washed 3× with low salt wash buffer (10 mM Tris pH 7.6, 1 mM EDTA, 0.1% SDS, 0.1% Na-Deoxycholate, 1% Triton-X 100, 150 mM NaCl), 2× high salt wash buffer (10 mM Tris pH 7.6, 1 mM EDTA, 0.1% SDS, 0.1% sodium deoxycholate, 1% Triton-X 100, 300 mM NaC), 2× LiCl wash buffer (0.25 M LiCl, 0.5% NP40, 0.5% sodium deoxycholate), and 1× with TE buffer with NaCl (10 mM Tris-HCl pH 8.0, 1 mM EDTA, 50 mM NaCl).
+Beads were resuspended in elution buffer (50 mM Tris-HCl pH 8.0, 1% SDS, 10 mM EDTA) and incubated at 65°C for 10 minutes.
+The IP and input samples were treated with 4.5 µL RNAse A for 30 minutes.
+5 µL Proteinase K was added and samples were incubated overnight at 65°C to reverse crosslinking.
+DNA was isolated by phenol:chloroform extraction, precipitated with EtOH, and resuspended in 20 µL H~2~O.Preparation of sequencing libraries was performed using NEBNext Ultra II kit (NEB) with 7 PCR cycles for library amplification.
+Sequencing was performed on the Illumina Hi-Seq4000 using 50bp single-end reads, or on the Illumina NovaSeq 6000 using 150bp paired-end reads.
 
@@
 ## RNA-seq
+
+5×10^6^ cells were harvested and resuspended in 800 µL Trizol (Invitrogen).
+RNA was purified using a chloroform extraction followed by a column-based Quick-RNA MiniPrep extraction (Zymo).
+Poly-A-selected RNA sequencing (RNA-seq) libraries were prepared using the TruSeq RNA sample prep kit v2 (Illumina).
 
@@
 
-**Extended data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.**
-
 **Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.**
@@
 **Figure 7: Pioneer-factor function requires stable chromatin occupancy.**
+
+**Extended data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.**
 

### 2022-11-21 manuscript/manuscript.qmd
update results
update results section

@@
 bibliography: references.bib
+#csl: nature.csl
 ---
@@
 
--   Establish system for ectopic expression of pioneer factors
+-   ~~Establish system for ectopic expression of pioneer factors~~
 
--   Identification of physiological expression levels
+-   ~~Identification of physiological expression levels~~
 
@@
 We identified XXXX peaks for Zld and XXXX for Grh.
+To control for antibody specificity, we performed additional ChIP experiments using IgG, anti-Zld, or anti-Grh (Extended Data Fig.
+XX).
+We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh.
+<!--# Not sure the best way to discuss the control ChIP experiments.
+Should I mention how many peaks we detect in WT cells?
+What supplemental figures should I include?
+Should peak calling be done using IgG or WT cells as control instead of input?
+-->
+
+Having established that Zld and Grh can bind to thousands of sites in S2 cells, we performed ATAC-seq on WT cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
+We performed differential accessibility analysis comparing Zld or Grh-expressing cells to WT cells.
+We detected XXX differentially accessible regions for Zld and XXX for Grh.
 
@@
 
-**Extended data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.**
+**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.**
+
+-   Schematic of cell line generation.
+
+-   RNA-seq data to show expression of S2 and not Kc cell markers.
+
+-   Western blots to determine physiological expression and lack of protein expression in WT cells.
+
+-   RNA-seq data to show leaky expression and strong induction of RNA levels.
+
+**Extended Data Figure 2: Ectopically expressed Zld and Grh bind to thousands of sites in S2 cells.**
+
+-   ChIP-seq replicate correlation
+
+-   Genome browser and heatmaps for all ChIP peaks?
+
+-   IgG and WT controls for antibody specificity
+
+**Extended Data Figure 3: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility.**
+
+-   ATAC PCA plot
+
+-   ATAC replicate correlation
+
+-   volcano and/or MA plots with enrichment of binding
+
+-   Example genome browsers?
+
+-   Analysis of nonspecific changes?
+
+**Extended Data Figure 4: Zld and Grh bind to chromatin rapidly after induction of protein expression**
+
+-   CUT&RUN time course to show rapid binding
+
+-   likely rules out dependence on DNA replication
+
+**Extended Data Figure 5: Zld and Grh binding promotes transcriptional activation at a subset of binding sites**
+
+-   RNA-seq PCA
+
+-   RNA-seq replicate correlation
+
+-   RNA-seq volcano/MA plots
+
+-   RNA-seq GO analysis
 
 **Information to include in supplemental figures:**
-
--   Schematic of stable cell line generation.
-
--   Cell line validation by RNA-seq
-
--   Verification of protein expression.
-
--   Determination of physiological protein levels.
-
--   QC for genomics datasets: e.g. PCA, replicate correlation, example genome browsers of individual replicates.
-
--   Antibody controls: IgG, IPs in WT cells, etc.
-
--   CUT&RUN time course for Zld and Grh?
 
@@
 -   Analysis of promoter motifs
-
--   GO analysis for RNA-seq data
-
--

### 2022-12-06 manuscript/manuscript.qmd
update methods

@@
 
+Raw reads were aligned to the Drosophila melanogaster (dm6) genome using HISAT2 version 2.1.0 with parameters -k 2 --very-sensitive.
+Reads with a mapping quality score \< 30 and reads aligning to the mitochondrial genome or unplaced scaffolds were discarded.
+FeatureCounts (from Subread version 2.0) was used to assign aligned reads to genes (FlyBase gene annotation release r6.45).
+The resultant table of read counts was used to perform differential expression analysis using DESeq2 [@love2014].
+Genes with an adjusted p-value \< 0.05 and a fold change \> 2 were considered differentially expressed.
+
 ## ATAC-seq
@@
 
-## Treatment of S2 cells with Tazemetostat
+## Treatment of S2 cells with tazemetostat
 

### 2022-12-06 manuscript/manuscript.qmd
update MS results

--- 2022-11-21 manuscript/manuscript.qmd
@@
 ---
+
+```{r}
+#| include: false
+library(here)
+library(tidyverse)
+```
 
@@
 
+```{r}
+#| include: false
+# zld_ChIP_classes <- snakemake@input[["zld_ChIP_classes"]]
+# grh_ChIP_classes <- snakemake@input[["grh_ChIP_classes"]]
+
+zld_ChIP_classes <- here("results/ChIP_peak_classes/zld_ChIP_classes.tsv") |> 
+  read_tsv()
+grh_ChIP_classes <- here("results/ChIP_peak_classes/grh_ChIP_classes.tsv") |> 
+  read_tsv()
+
+zld_n_ChIP_peaks <- nrow(zld_ChIP_classes)
+grh_n_ChIP_peaks <- nrow(grh_ChIP_classes)
+```
+
 To determine how exogenous Zld and Grh interact with preexisting chromatin in S2 cells, we expressed each factor at physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
-We identified XXXX peaks for Zld and XXXX for Grh.
-To control for antibody specificity, we performed additional ChIP experiments using IgG, anti-Zld, or anti-Grh (Extended Data Fig.
+We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
+To control for antibody specificity, we performed additional ChIP experiments in WT cells using IgG, anti-Zld, or anti-Grh (Extended Data Fig.
 XX).
@@
 
+```{r}
+#| include: false
+
+zld_ATAC_results <- here("ATACseq/results/DEseq2_results_filtered/S2-Zld_ATACseq_S2-Zld-FL-vs-S2-WT_results.tsv") |> 
+  read_tsv() |> 
+  mutate(direction_of_change = case_when(
+    is_diff & log2FoldChange > 0 ~ "increased",
+    is_diff & log2FoldChange < 0 ~ "decreased",
+    !is_diff ~ "ns"
+  ))
+
+grh_ATAC_results <- here("ATACseq/results/DEseq2_results_filtered/S2-Grh_ATACseq_S2-Grh-FL-vs-S2-WT_results.tsv") |> 
+  read_tsv() |> 
+  mutate(direction_of_change = case_when(
+    is_diff & log2FoldChange > 0 ~ "increased",
+    is_diff & log2FoldChange < 0 ~ "decreased",
+    !is_diff ~ "ns"
+  ))
+
+zld_diff <- zld_ATAC_results |> 
+  pull(is_diff) |> 
+  sum()
+zld_up <- (zld_ATAC_results$direction_of_change == "increased") |> 
+  sum()
+zld_down <- (zld_ATAC_results$direction_of_change == "decreased") |> 
+  sum()
+
+grh_diff <- grh_ATAC_results |> 
+  pull(is_diff) |> 
+  sum()
+grh_up <- (grh_ATAC_results$direction_of_change == "increased") |> 
+  sum()
+grh_down <- (grh_ATAC_results$direction_of_change == "decreased") |> 
+  sum()
+
+```
+
 Having established that Zld and Grh can bind to thousands of sites in S2 cells, we performed ATAC-seq on WT cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
 We performed differential accessibility analysis comparing Zld or Grh-expressing cells to WT cells.
-We detected XXX differentially accessible regions for Zld and XXX for Grh.
+We detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
+XX).
 
@@
 | Anti-HA clone 12CA5                   | Roche                     |                      | ChIP: 7.5 µL            |
-| Anti-H3K27me3                         | Cell Signaling Technology |                      | WB:, CR:                |
+| Anti-H3K27me3                         | Cell Signaling Technology |                      | WB: 1:1000, CR: 0.5 µL  |
 |                                       |                           |                      |                         |
@@
 
+Bowtie2 version 2.4.4 [@Langmead2012] was used to align ChIP-seq reads to the *Drosophila melanogaster* genome (version dm6) with the following non-default parameters: -k 2 \--very-sensitive \--no-mixed \--no-discordant -X 5000.
+Aligned reads were filtered to include only reads with a mapping quality score \> 30.
+Reads mapping to unplaced scaffolds or the mitochondrial genome were discarded.
+
+Alignment
+
+Filtering
+
+peak calling
+
+peak filtering
+
+generation of bigwigs
+
+normalization of bigwigs
+
 ## RNA-seq
@@
 
+-   Western blot time course of protein levels
+
 -   CUT&RUN time course to show rapid binding

### 2022-12-06 manuscript/manuscript.qmd
add CUT&RUN analysis and update MS results

@@
 XX).
+<!--# Add more information here about sites with differential accessibility.
+Include percentage of increased and decreased sites that are bound by Zld or Grh.
+Comment on how much is specific vs.
+secondary effect, and justify defining ChIP peak classes that include only a subset of DA sites.
+-->
+
+To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified 3 classes of binding sites (Fig.
+1a,b).
+Class I sites are accessible in WT cells and remain accessible when bound by ectopically-expressed Zld or Grh.
+Class II sites are inaccessible in WT cells, however binding of Zld or Grh does not lead to chromatin opening.
+Class III sites are inaccessible in WT cells and Zld or Grh binding results in chromatin opening.
+The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig.
+1c,f).
+While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig.
+1d,g).
+Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
+1e,h).
+Together, these data show that when expressed exogenously at physiolgical levels, Zld and Grh can bind and open closed chromatin at a subset of binding sites.
 

### 2022-12-07 manuscript/manuscript.qmd
update MS results

@@
 
--   Define binding classes I-III and describe characteristics (distribution, promoters/enhancers, etc.)
+-   ~~Define binding classes I-III and describe characteristics (distribution, promoters/enhancers, etc.)~~
 
@@
 1d,g).
-Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
+Genome-wide analysis of signa <!--#  -->l intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
 1e,h).
 Together, these data show that when expressed exogenously at physiolgical levels, Zld and Grh can bind and open closed chromatin at a subset of binding sites.
+
+<!--# Could potentially add the cut&run time course here? Think about where to put this, or if it is worth including at all. -->
 
@@
 -   Class III promoters are most likely to be associated with upregulated genes
+
+Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation.
+To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression.
+Differential expression analysis using DESeq2 [@love2014] identified XXX differentially expressed genes for Zld and XXX for Grh (Extended Data Fig.
+XX).
 

### 2022-12-08 manuscript/manuscript.qmd
update MS results

@@
 
-To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified 3 classes of binding sites (Fig.
+To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of binding sites (Fig.
 1a,b).
@@
 XX).
+Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig.
+XX <!--# include statistics  -->), suggesting that binding of these factors mainly leads to transcriptional activation and not repression.
+<!--# Consider mentioning Grh as a repressor here and giving examples of genes where Grh may be repressing.
+Discuss if Grh-mediated transcriptional repression coincides with chromatin opening, chromatin closing, or neither --> Gene ontology (GO) analysis indicated that the genes activated by each factor are consistent with the endogenous functions of Zld and Grh (Extended Data Fig.
+XX).
+<!--# Add more detailed information on GO analysis here -->
+
+To better understand the relationship between chromatin opening and transcriptional activation, we directly compared our RNA-seq and ATAC-seq data.
+Overall, chromatin opening was correlated with transcriptional activation (Fig.
+1i-l).
+sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig.
+XX <!--# add statistics -->).
+When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig.
+1j,l).
+This was particularly evident at class III promoters, which were most likely to correspond to upregulated genes.
+Together, we conclude that while binding and chromatin opening can happen in the absence of transcriptional changes, chromatin opening is correlated with transcriptional activation.
 
@@
 
--   Twi binds closed chromatin and opens at some sites
+-   Twi binds closed chromatin and opens at some sites.
+
+-   Mention analysis of published Abd-A and Ubx cell lines.
 
@@
 
--   Twi also promotes transcriptional activation.
+-   Twi also promotes transcriptional activation
 
@@
 -   Comparison of RNA-seq and ATAC
+
+-   Mention differences between Twi vs. Zld and Grh
+
+Despite the differences between Zld and Grh during Drosophila development, both proteins behaved similarly when expressed in our exogenous S2 cell system.
+To provide additional context to our findings, we performed additional experiments with the transcription factor Twist (Twi).
+Although Twi is also essential for regulating Drosophila embryonic development, Twi has a number if distnict properties from Zld and Grh, making it an interesting point of comparison.
+Twi is a master regulator of mesodermal cell fate (REFS).
+Twi has many binding sites that are dynamic during early embryonic development [@Wilczynski2010a], in contrast to Zld and Grh.
+Although Twi was hypothesized to be a pioneer factor [@Sandmann2007], pioneering activity by Twi had not previously been tested.
+In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012].
+To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig.
+XX).
+We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
+Comparison of Twi ChIP-seq and ATAC-seq data allowed for the identification of the same classes of binding sites that were found for Zld and Grh (Fig 2a,b).
+Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
+Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig.
+2b).
+At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c).
+Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig.
+XX).
+Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig.
+XX).
 

### 2022-12-14 manuscript/manuscript.qmd
update MS results and add Fig 4 draft

@@
 XX), where neither factor is endogenously expressed.
-We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers (Extended Data Fig.
-1b) [@Cherbas2011].
-We confirmed that Zld and Grh RNA (Extended Data Fig.
-XX) and protein (Extended Data Fig.
-XX) were not present in WT S2 cells.
+We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers (Extended Data Fig. 1b) [@Cherbas2011].
+We confirmed that Zld and Grh RNA (Extended Data Fig. XX) and protein (Extended Data Fig. XX) were not present in WT S2 cells.
 The stable cell lines express Zld or Grh under the control of the metallothionein promoter, which allows inducible and tunable expression with the addition of CuSO~4.~ To determine protein levels that would be approximately physiological, we compared Zld or Grh levels in cells with varying concentrations of CuSO~4~ to the levels of endogenous expression found in Drosophila embryos (Extended Data Fig.
@@
 We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
-To control for antibody specificity, we performed additional ChIP experiments in WT cells using IgG, anti-Zld, or anti-Grh (Extended Data Fig.
-XX).
+To control for antibody specificity, we performed additional ChIP experiments in WT cells using IgG, anti-Zld, or anti-Grh (Extended Data Fig. XX).
 We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh.
@@
 
-To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of binding sites (Fig.
-1a,b).
+To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of binding sites (Fig. 1a,b).
 Class I sites are accessible in WT cells and remain accessible when bound by ectopically-expressed Zld or Grh.
@@
 Class III sites are inaccessible in WT cells and Zld or Grh binding results in chromatin opening.
-The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig.
-1c,f).
-While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig.
-1d,g).
-Genome-wide analysis of signa <!--#  -->l intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
+The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig. 1c,f).
+While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
+Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
 1e,h).
@@
 
--   Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.
+-   ~~Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.~~
 
--   Comparison of RNA-seq and ATAC data: chromatin opening correlates with transcriptional activation.
+-   ~~Comparison of RNA-seq and ATAC data: chromatin opening correlates with transcriptional activation.~~
 
--   Class III promoters are most likely to be associated with upregulated genes
+-   ~~Class III promoters are most likely to be associated with upregulated genes~~
 
@@
 To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression.
-Differential expression analysis using DESeq2 [@love2014] identified XXX differentially expressed genes for Zld and XXX for Grh (Extended Data Fig.
-XX).
+Differential expression analysis using DESeq2 [@love2014] identified XXX differentially expressed genes for Zld and XXX for Grh (Extended Data Fig. XX).
 Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig.
@@
 To better understand the relationship between chromatin opening and transcriptional activation, we directly compared our RNA-seq and ATAC-seq data.
-Overall, chromatin opening was correlated with transcriptional activation (Fig.
-1i-l).
-sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig.
-XX <!--# add statistics -->).
-When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig.
-1j,l).
+Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l).
+sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig. XX <!--# add statistics -->).
+When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
 This was particularly evident at class III promoters, which were most likely to correspond to upregulated genes.
@@
 
--   Introduce Twist and justify the comparison to Zld/Grh.
+-   ~~Introduce Twist and justify the comparison to Zld/Grh.~~
 
--   Twi binds closed chromatin and opens at some sites.
+-   ~~Twi binds closed chromatin and opens at some sites.~~
 
@@
 
--   Class I-III sites, distribution, etc.
+-   ~~Class I-III sites, distribution, etc.~~
 
--   Twi also promotes transcriptional activation
+-   ~~Twi also promotes transcriptional activation~~
 
--   Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.
+-   ~~Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.~~
 
--   Comparison of RNA-seq and ATAC
+-   ~~Comparison of RNA-seq and ATAC~~
 
--   Mention differences between Twi vs. Zld and Grh
+-   ~~Mention differences between Twi vs. Zld and Grh~~
 
@@
 In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012].
-To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig.
-XX).
+To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. XX).
 We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
@@
 Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
-Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig.
-2b).
+Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
 At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c).
-Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig.
+Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. XX).
+Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig. XX).
+GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. XX).
+Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
+The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig. XX).
+Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig. XX).
+This was in contrast to Zld and Grh, where induction results in the activation of many genes that are not directly bound by Zld or Grh.
+This suggests that Zld and Grh may activate many transcription factors, which go on to activate their own targets, leading to secondary effects in the RNA-seq data.
+Twi, by contrast, appears to directly activate genes required for mesodermal and muscle cell fate.
+<!--# Is this worth mentioning?
+Might require additional analysis of how many TFs are upregulated in RNA-seq data for each factor.
+Could also require additional motif analysis to identify/support secondary TFs and reason for nonspecific effects.
+--> Similar to Zld and Grh, transcriptional activation was correlated with chromatin opening (Extended Data Fig.
 XX).
-Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig.
-XX).
+Together, our data show that Twi can act as a pioneer factor.
 
@@
 
--   Overlap of class I regions suggests nonspecific binding
+-   ~~Overlap of class I regions suggests nonspecific binding~~
 
--   Analysis of published ChIP data shows that class I regions are within active chromatin
+-   ~~Analysis of published ChIP data shows that class I regions are within active chromatin~~
 
--   RNA-seq and ATAC-seq for class I to show that the binding is not functional
+-   ~~RNA-seq and ATAC-seq for class I to show that the binding is not functional~~
 
-## Class II and III regions are mostly in naïve chromatin
+-   ~~Analysis of chromatin at class II-III regions: mostly devoid of histone modifications~~
 
--   Analysis of chromatin at class II-III regions: mostly devoid of histone modifications
+-   ~~Class III regions have higher H3 and H1 occupancy compared to class II~~
 
--   Class III regions have higher H3 and H1 occupancy compared to class II
+-   Where to put analysis of nucleosome positioning?
+
+While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites are shared across Zld, Grh, and Twi (Fig. 3a).
+This was in contrast to class II and III peaks, which were largely specific to each individual factor (Extended Data Fig. XX).
+Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig. XX).
+We hypothesized that these overlapping class I binding sites may occur at regulatory regions that are active in WT cells.
+To test this, we analyzed published ChIP-seq data from WT S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV.
+Overlapping class I regions had high levels of chromatin marks associated with active regulatory regions (Fig. 3b).
+This enrichment of all three factors at highly active regions suggests that exogenously expressed factors may bind nonspecifically to active regions of the genome.
+This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig. XX).
+
+Considering the enrichment of active chromatin at class I sites, we asked if there were any features of chromatin distinctive to class II or III regions.
+Analysis of published ChIP-seq data demonstrated that class II and III sites have low levels of many histone modifications, chromatin-associated proteins, and transcription factors (Extended Data Fig.
+XX).
+These data are in agreement with previous observations that Oct4, Sox2 and Klf4 access binding sites in naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
+Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. XX).
+Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig. XX).
+Together this suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin.
+Based on these observations, the inability to open class II regions is unlikely to be due to differences in preexisting features of chromatin.
 
@@
 
--   Motif analysis: percent with motif, average n motifs, motif quality
+-   ~~Motif analysis: percent with motif, average n motifs, motif quality~~
 
--   Motif positioning around nucleosomes? (supplemental)
+-   ~~Motif positioning around nucleosomes? (supplemental)~~
+
+To further test the specificity of binding within the three classes of ChIP-seq peaks, we analyzed the enrichment for the canonical Zld, Grh or Twi motifs.
+We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig. 3c).
+When compared to all accessible regions, class I regions had low levels of enrichment for the canonical motifs, supporting that binding to these regions is nonspecific.
+By contrast, class II and III sites were highly enriched for the canonical motifs (Fig. 3c).
+Class III sites had the highest percentage of sites containing a canonical motif.
+We hypothesized that differences in motif content could explain why binding to class II regions does not result in chromatin opening.
+We therefore analyzed the number of motifs per peak and motif strength (when scored against the canonical motif) for class I-III regions.
+For Grh and Twi, but not for Zld, class III sites had more motifs per peak than class II sites (Fig. 3f,h,j).
+For all three factors, class III sites tended to have stronger motifs than class II sites (Fig. 3g,i,k).
+Together, this analysis suggests that a combination of more motifs and stronger motifs at class III sites is what allows Zld, Grh, and Twi to open chromatin at these regions.
+<!--# Add description of de novo motif enrichment, put data in supplement. -->
+
+Previous work has shown that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (refs).
+<!--# Add a few sentences here with more details on past work --> To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
+Analysis of average MNase-seq signal at class I-III peaks indicated that class III sites have high levels of nucleosome occupancy that, on average, occur over the respective motif.
+Hierarchical clustering of MNase-seq data revealed that there was no preference for binding or chromatin opening at sites that had motifs in a particular orientation on nucleosomes (Extended Data Fig.
+XX).
+For class II and III peaks, some motifs were positioned close to the nuclesome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
+This analysis is complicated by the fact that MNase-seq is a bulk assay and may not accurately represent nucleosome positioning in individual cells, especially at regions with poorly positioned nucleosomes.
+To more directly assess the positioning of nucleosomes around motifs, we analyzed a previously published single-molecule dataset using methyltransferase footprinting to determine the locations of nucleosomes on individual chromatin fibers [@Stergachis2020].
+We analyzed the distance from motifs at class II and regions to the nearest nucleosome dyad.
+This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
+Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
+Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
+However, the lack of strong nucleosome positioning at these regions makes it difficult to fully assess the importance of motif positioning.
 
@@
 
--   Results of Taz experiments (TBD)
+-   Results of Taz experiments
 
@@
 
--   Generation of stable cell lines to express DBDs at same concentration as FL proteins
+-   ~~Generation of stable cell lines to express DBDs at same concentration as FL proteins~~
 
--   Zld DBD is totally unable to bind to closed sites.
+-   ~~Zld DBD is totally unable to bind to closed sites.~~
 
--   In supplement, show Zld DBD binding specifically at class II and III sites with CAGGTAG to support that this is not due to a loss of ZnF2
+-   ~~In supplement, show Zld DBD binding specifically at class II and III sites with CAGGTAG to support that this is not due to a loss of ZnF2~~
 
--   Grh DBD has reduced ability to access closed chromatin
+-   ~~Grh DBD has reduced ability to access closed chromatin~~
 
--   Neither DBD is sufficient for chromatin opening.
+-   ~~Neither DBD is sufficient for chromatin opening.~~
+
+Prior work has shown that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs).
+However, little is known about how regions outside the DBD may contribute to pioneering function.
+<!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat. Genet. paper) is outside the DBD -->.
+To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig.
+XX).
+Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was totally unable to bind to class II or III regions (Fig.
+6a,c).
+We had previously shown that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
+Because many Zld binding sites in S2 cells also lack a canonical CAGGTA motif (Fig.
+3, Extended Data Fig XX), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
+To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
+This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required.
+Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
+For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
+Together, these experiments show that regions outside the DBD are essential for binding and opening of closed chromatin.
 
@@
 | Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|-------------------|------------------|------------------|------------------|
+|-----------------------|-----------------|-----------------|-----------------|
 | pMT-puro                              | AddGene                   | 17923                |                         |
@@
 
-**Information to include in supplemental figures:**
+**Extended Data Figure 6: Twist binding leads to chromatin opening and transcriptional activation**
+
+-   Twi ATAC PCA, volcano, and examples
+
+-   TWI RNA-seq PCA, volcano, and examples
+
+-   Twi RNA-seq GO analysis
+
+**Extended Data Figure 7: Chromatin features associated with Zld, Grh and Twi binding sites**
+
+-   analysis of published ChIP datasets not shown in main figure
+
+-   comparison of class I-III sites to show that class II-III sites are devoid of most histone mods
+
+-   Class III sites have higher levels of H1 and H3 than class II sites.
+
+**Extended Data Figure 8: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes**
+
+-   MNase data to show enrichment of nucleosomes around target motifs
+
+-   Clustering of MNase data to show that there is no preferred nucleosome position relative to motifs
+
+-   additional analysis of Fiber-seq data to show that nucleosomes at these regions are poorly positioned
+
+**Additional information to include in supplemental figures:**
 
@@
 -   Analysis of promoter motifs
+
+-   New binding sites at higher expression levels
+
+-   Zld DBD ChIP only at CAGGTAG sites

### 2022-12-14 workflow/scripts/fig4.R

@@ -1,0 +1,530 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(plotgardener))
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(org.Dm.eg.db))
+suppressPackageStartupMessages(library(TxDb.Dmelanogaster.UCSC.dm6.ensGene))
+# library(grImport)
+suppressPackageStartupMessages(library(grid))
+library(RColorBrewer)
+
+source("workflow/scripts/plot_heatmap.R")
+source("workflow/scripts/utils.R")
+
+# define input files ===========================================================
+zld_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv")
+grh_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/grh_tissue_classes.tsv")
+twi_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/twi_tissue_classes.tsv")
+
+S2_ZLD_ChIP_bw <-   "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_IP.bw"
+embryo_Zld_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/embryo-nc14_aZld.bw"
+brain_Zld_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/brain_aZld.bw"
+S2_H3K27me3_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE151983_S2_aH3K27me3_IP.bw"
+S2_H3K9me3_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE160855_aH3K9me3.bw"
+
+S2_Grh_ChIP_bw <-   "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_IP.bw"
+embryo_Grh_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/embryo-15-16H_aGrh.bw"
+wing_disc_Grh_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/wing-disc_aGrh.bw"
+
+S2_Twi_ChIP_bw <-  "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Twi_aTwi_IP.bw"
[…]

### 2022-12-14 workflow/scripts/get_tissue_classes.R

@@ -1,0 +1,96 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(GenomicRanges))
+suppressPackageStartupMessages(library(rtracklayer))
+suppressPackageStartupMessages(library(BSgenome.Dmelanogaster.UCSC.dm6))
+
+source("workflow/scripts/utils.R")
+
+
+# define input files ===========================================================
+# S2 ChIP peaks and classes
+S2_bound_regions <- c(
+  all_peaks = snakemake@input[["all_peaks"]],
+  class_I_peaks = snakemake@input[["class_I_peaks"]],
+  class_II_peaks = snakemake@input[["class_II_peaks"]],
+  class_III_peaks = snakemake@input[["class_III_peaks"]]
+)
+
+# other tissue ChIP peaks
+other_tissue_bound_regions <- c(
+  snakemake@input[["other_tissue_peaks"]]
+)
+names(other_tissue_bound_regions) <- str_replace(basename(other_tissue_bound_regions), ".bed", "")
+
+# Regions with repressive histone modifications
+repressed_regions <- c(
+  H3K27me3_peaks = snakemake@input[["H3K27me3_peaks"]],
+  H3K9me3_peaks = snakemake@input[["H3K9me3_peaks"]]
+)
+
+# nonspecific sites to exclude
+ns_sites <- c(
+  ns_sites = snakemake@input[["nonspecific_sites"]]
+)
+
+# combine all inputs
+input_files <- c(
+  S2_bound_regions,
+  other_tissue_bound_regions,
+  repressed_regions,
+  ns_sites
+)
+
+
+# Define potential binding sites ===============================================
+overlap_table <- input_files |>
+  map(rtracklayer::import) |> 
+  GRangesList()  |> 
+  peak_overlap_table()
+
+# annotate peaks ==============================================================
+other_tissue_names <- names(other_tissue_bound_regions)
+
+overlap_table <- overlap_table |>
+  rowwise() |> 
+  mutate(n_tissues_bound = sum(c_across(all_of(c("all_peaks",other_tissue_names))))) |>
+  ungroup() |> 
+  mutate(class = case_when(
+    (!all_peaks & n_tissues_bound >= 1 & H3K27me3_peaks) ~ "repressed_H3K27me3",
[…]

### 2022-12-19 manuscript/manuscript.qmd
update MS

@@
 
+Prior studies have extensively explored the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
+To better understand how cell-type specific variables affect pioneer-factor binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld and Grh binding.
+
 ## Repressive chromatin is a barrier to pioneer-factor binding
@@
 | Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|-----------------------|-----------------|-----------------|-----------------|
+|---------------------|-----------------|-----------------|-----------------|
 | pMT-puro                              | AddGene                   | 17923                |                         |

### 2022-12-19 workflow/scripts/fig3.R

@@ -19,14 +19,6 @@ signal_range <- function(x, extra = 0.05) {
   return(output)
 }
 
-# function to extract legend from ggplot ---------------------------------------
-get_legend <- function(plot){ 
-  tmp <- ggplot_gtable(ggplot_build(plot)) 
-  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
-  legend <- tmp$grobs[[leg]] 
-  legend
-} 
-
 # define input files ===========================================================
 class_I_bed_fn <- c(
   zld_class_I = snakemake@input[["zld_class_I_bed_fn"]],

### 2022-12-19 workflow/scripts/fig5.R

@@ -1,0 +1,444 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(plotgardener))
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(org.Dm.eg.db))
+suppressPackageStartupMessages(library(TxDb.Dmelanogaster.UCSC.dm6.ensGene))
+# library(grImport)
+suppressPackageStartupMessages(library(grid))
+library(RColorBrewer)
+
+source("workflow/scripts/plot_heatmap.R")
+source("workflow/scripts/utils.R")
+
+# define input files ===========================================================
+zld_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv")
+grh_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/grh_tissue_classes.tsv")
+twi_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/twi_tissue_classes.tsv")
+
+zld_0uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-0uM_aZld_IP.bw"
+zld_500uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-500uM_aZld_IP.bw"
+zld_1000uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-1000uM_aZld_IP.bw"
+zld_1500uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-1500uM_aZld_IP.bw"
+
+grh_0uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-0uM_aGrh_IP.bw"
+grh_25uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-25uM_aGrh_IP.bw"
+grh_100uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-100uM_aGrh_IP.bw"
+grh_400uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-400uM_aGrh_IP.bw"
+
[…]

### 2022-12-19 manuscript/manuscript.qmd
update MS figs 4 and 5

@@
 | Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|---------------------|-----------------|-----------------|-----------------|
+|---------------------------------------|---------------------------|----------------------|-------------------------|
 | pMT-puro                              | AddGene                   | 17923                |                         |

### 2022-12-19 workflow/scripts/define_titration_classes.R

@@ -1,0 +1,460 @@
+# setup ------------------------------------------------------------------------
+suppressPackageStartupMessages(library(rtracklayer))
+suppressPackageStartupMessages(library(GenomicRanges))
+suppressPackageStartupMessages(library(tidyverse))
+
+
+source("workflow/scripts/utils.R")
+
+## Zld 500 uM ==================================================================
+# get input file names ---------------------------------------------------------
+ChIP_peaks <- "ChIPseq/results/peaks/filtered/S2-Zld-500uM_aZld_IP.narrowPeak"
+wt_ATAC_peaks <- "ATACseq/results/peaks/merged_by_sample/S2-WT_1000uM_summits.bed"
+ATAC_results <- "ATACseq/results/DEseq2/S2-Zld-titration_ATACseq_S2-Zld-500uM-vs-0uM_results.tsv"
+
+# annotate ChIP classes --------------------------------------------------------
+# import ChIP peaks
+peaks_gr <- rtracklayer::import(ChIP_peaks)
+
+peaks_df <- peaks_gr %>% 
+  as.data.frame() %>% 
+  dplyr::rename(peak_chrom = seqnames, peak_start = start, peak_end = end, peak_id = name, MACS2_enrichment = signalValue, MACS2_pvalue = pValue, MACS2_qValue = qValue) %>% 
+  select(-c(width, strand,score,peak))
+
+# read in WT ATAC peaks
+wt_atac <- rtracklayer::import(wt_ATAC_peaks) %>% 
+  resize(width = 201, fix = "center")
+
+# read in diff ATAC results
+diff_atac_df <- read_tsv(ATAC_results)
+diff_atac_gr <- makeGRangesFromDataFrame(diff_atac_df, keep.extra.columns = TRUE)
+
+atac_increased <- diff_atac_df %>% 
+  filter(is_diff, log2FoldChange > 0) %>% 
+  makeGRangesFromDataFrame()
+
+# generate overlap table and annotate ChIP classes
+zld_500_ChIP_classes <- GRangesList(
+  ChIP_peak = peaks_gr,
+  wt_atac = wt_atac,
+  atac_increased = atac_increased
+) %>% 
+  
+  peak_overlap_table() %>% 
+  filter(ChIP_peak) %>% 
+  mutate(class = case_when(
+    ChIP_peak & wt_atac ~ "i",
[…]

### 2022-12-19 workflow/scripts/fig4.R

@@ -52,8 +52,8 @@ twi_heatmap_colors <- brewer.pal(9, "GnBu")
 # gb_height <- 0.3
 
 # set heatmap parameters
-hm_upstream <-  500
-hm_downstream <-  500
+hm_upstream <-  1000
+hm_downstream <-  1000
 
 # panel A ======================================================================
 # reference points for positioning figure components
@@ -298,7 +298,7 @@ plotText(
 # panel D ======================================================================
 # reference points for positioning figure components
 ref_x <- 0.5
-ref_y <- 6
+ref_y <- 6.5
 
 # panel label
 plotText(
@@ -311,8 +311,8 @@ plotText(
 bw <- c(
   S2_Zld_ChIP_DMSO =  "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_DMSO_aZld.bw",
   S2_Zld_ChIP_taz = "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_taz_aZld.bw",
-  S2_Zld_H3K27me3_DMSO = "CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_DMSO_aH3K27me3.bw",
-  S2_Zld_H3K27me3_taz = "CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_Taz_aH3K27me3.bw" 
+  S2_Zld_H3K27me3_DMSO = "CUTandRUN/results/bigwigs/spikeIn_normalized/merged/S2-Zld_DMSO_aH3K27me3.bw",
+  S2_Zld_H3K27me3_taz = "CUTandRUN/results/bigwigs/spikeIn_normalized/merged/S2-Zld_Taz_aH3K27me3.bw" 
 )
 
 
@@ -375,7 +375,7 @@ plotText(
 # panel E ======================================================================
 # reference points for positioning figure components
 ref_x <- 6.5
-ref_y <- 6
+ref_y <- 6.5
 
 # panel label
 plotText(
@@ -387,8 +387,8 @@ plotText(
 bw <- c(
   S2_Grh_ChIP_DMSO =  "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_DMSO_aGrh.bw",
[…]

### 2022-12-19 workflow/scripts/fig5.R

@@ -45,6 +45,10 @@ twi_500uM_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/titration
 twi_1000uM_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/titration_S2-HA-Twi_40uM_small.bw"
 twi_1500uM_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/titration_S2-HA-Twi_160uM_small.bw"
 
+zld_titration_classes_fn <- "results/ChIP_titration_classes/zld_titration_classes.tsv"
+grh_titration_classes_fn <- "results/ChIP_titration_classes/grh_titration_classes.tsv"
+twi_titration_classes_fn <- "results/ChIP_titration_classes/twi_titration_classes.tsv"
+
 
 ## create blank layout for plot =================================================
 # pdf(snakemake@output[[1]], useDingbats = FALSE)
@@ -94,33 +98,33 @@ bw <- c(
 
 
 regions <- zld_tissue_occupancy %>%
-  filter(class != "class I") |>
+  # filter(class != "class I") |>
   makeGRangesFromDataFrame(keep.extra.columns = TRUE)
 
 
-hm <- plot_heatmap_minimal(
-  bw, regions,
-  upstream = hm_upstream, downstream = hm_downstream,
-  colors  = zld_heatmap_colors,
-  row_split = regions$class,
-  # order_by_samples = 1,
-  individual_scales = FALSE,
-  use_raster = TRUE, raster_by_magick = TRUE, raster_magick_filter = "Triangle",
-  border = "black",
-  border_gp = gpar(lwd = 0.5),
-  show_heatmap_legend = FALSE,
-  return_heatmap_list = TRUE
-)
-
-b_hm <- grid.grabExpr(draw(hm, show_heatmap_legend = FALSE, padding = unit(c(0, 0, 0, 0), "mm")))
-
-# place heatmap on page
-plotGG(
-  plot = b_hm,
-  x = (ref_x + 0.25), y = (ref_y + 0.25),
-  width = 5, height = 5, just = c("left", "top"),
-  default.units = "cm"
-)
+# hm <- plot_heatmap_minimal(
+#   bw, regions,
+#   upstream = hm_upstream, downstream = hm_downstream,
+#   colors  = zld_heatmap_colors,
+#   row_split = regions$class,
+#   # order_by_samples = 1,
+#   individual_scales = FALSE,
[…]

### 2022-12-20 manuscript/manuscript.qmd
update MS results

--- 2022-12-14 manuscript/manuscript.qmd
@@
 ## Ectopically expressed pioneer factors bind and open closed chromatin
-
--   Gain of function system is necessary to understand mechanisms of PF activity
 
@@
 
--   Zelda binding is more similar to the brain than the embryo
+-   Z~~elda binding is more similar to the brain than the embryo~~
 
--   Grh binding is stable across tissues
+-   ~~Grh binding is stable across tissues~~
 
--   Twi sites in S2 cells have low level binding in embryo, but strongest peaks from embryo are not bound
+-   ~~Twi sites in S2 cells have low level binding in embryo, but strongest peaks from embryo are not bound~~
 
-## Repressive chromatin is a barrier to pioneer-factor binding
+-   ~~Enrichment of repressive marks at sites bound in other tissues~~
 
--   Enrichment of repressive marks at sites bound in other tissues
+-   ~~Results of Taz experiments~~
 
--   Results of Taz experiments
+-   Comment on class VI sites that are not bound and lack repressive chromatin.
 
--   Comment on sites that are not bound and lack repressive chromatin.
+Prior studies have extensively explored the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
+To better understand how cell-type specific variables affect pioneer-factor binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
+We compared Zld binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
+We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig.
+XX).
+Previous work has shown that regions resistant to pioneer-factor binding are enriched for the repressive histone modifications H3K9me3 and H3K27me3 [@Mayran2018a; @Soufi2012].
+To test if repressive marks were present at the resistant sites in S2 cells, we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
+Based on analysis of these datasets we divided the resistant sites into three additional classes of binding sites: class IV sites have high levels H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a).
+We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c).
+We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017], and to Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
+Despite the presence of class IV, V and VI binding sites for all three proteins, there were distinct patterns of binding across different tissues.
+Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
+For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig. 4b).
+Twi binding sites in S2 cells had low levels of binding in embryos, but many of the strongest embryonic binding sites were unbound.
+
+To test the functional relevance of repressive chromatin to pioneer-factor binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3.
+We treated our stable cell lines with DMSO or tazemetostat for 7 days, inducing Zld, Grh, or Twi expression on day 5.
+We confirmed loss of H3K27me3 by westerm blot and CUT&RUN [@Skene2017] (Extended Data Fig. XX).
+We used ChIP-seq to test if tazemetostat treatment resulted in changes to Zld, Grh, or Twi binding.
+Tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig. 4d-f).
+Despite a dramatic loss of H3K27me3, the majority of class IV binding sites remained unbound in tazemetostat-treated cells.
+We did identify some individual sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig. XX).
+Together, these data demonstrate that despite the correlation of H3K27me3 with resistance to pioneer factor binding, loss of PRC2 activity is insufficient to promote widespread binding of Zld, Grh, or Twi to repressed binding sites.
 
@@
 -   For Grh, increased concentration does allow binding and opening at repressed sites
+
+Previous studies have proposed that pioneer factor activity may depend on protein concentration [@Hansen2022].
+Considering that tazemetostat treatment had limited effects on pioneer factor binding, we asked if varying the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.
 
@@
 | Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|-----------------------|-----------------|-----------------|-----------------|
+|--------------------|------------------|------------------|------------------|
 | pMT-puro                              | AddGene                   | 17923                |                         |

### 2022-12-20 workflow/scripts/fig4.R

@@ -110,10 +110,40 @@ grid.xaxis(at = c(0, 1), label = c(paste0("-", hm_upstream / 1000, "KB"), paste0
 seekViewport(name = "page")
 
 # add heatmap labels
-# plotText(
-#   label = paste(length(regions), "overlapping", "\n", "class I sites"), params = small_text_params, fontface = "bold",
-#   x = (ref_x), y = (ref_y + 1.5), just = c("center"), default.units = "cm", rot = 90
-# )
+plotText(
+  label = "class:", params = small_text_params, fontface = "bold",
+  x = (ref_x - 0.1), y = (ref_y + 0.2), just = c("center"), default.units = "cm"
+)
+
+plotText(
+  label = "I", params = small_text_params, fontface = "bold",
+  x = (ref_x), y = (ref_y + 0.6), just = c("center"), default.units = "cm"
+)
+
+plotText(
+  label = "II", params = small_text_params, fontface = "bold",
+  x = (ref_x), y = (ref_y + 1.3), just = c("center"), default.units = "cm"
+)
+
+plotText(
+  label = "III", params = small_text_params, fontface = "bold",
+  x = (ref_x), y = (ref_y + 1.58), just = c("center"), default.units = "cm"
+)
+
+plotText(
+  label = "IV", params = small_text_params, fontface = "bold",
+  x = (ref_x), y = (ref_y + 1.9), just = c("center"), default.units = "cm"
+)
+
+plotText(
+  label = "V", params = small_text_params, fontface = "bold",
+  x = (ref_x), y = (ref_y + 2.5), just = c("center"), default.units = "cm"
+)
+
+plotText(
+  label = "VI", params = small_text_params, fontface = "bold",
+  x = (ref_x), y = (ref_y + 4), just = c("center"), default.units = "cm"
+)
 
 plotText(
   label = paste0("Zld", "\n", "S2 cells"), params = small_text_params, fontface = "bold",
@@ -190,11 +220,46 @@ plotGG(
 )
 
 # add axes to heatmaps
-# seekViewport(name = "matrix_1_heatmap_body_6_1")
-# grid.xaxis(at = c(0, 1), label = c(paste0("-", hm_upstream / 1000, "KB"), paste0("+",hm_downstream / 1000, "KB")), gp = gpar(lwd = 0.5, fontsize = small_text_params$fontsize))
-# seekViewport(name = "page")
[…]

### 2022-12-20 workflow/scripts/fig5.R

@@ -163,19 +163,12 @@ plotGG(
 
 # plot metaplot 3
 metaplot_3 <- plot_average(bw[3], regions = regions, row_split = regions$class, line_width = 0.2) +
-  # scale_color_brewer(palette = "Blues") +
   scale_color_manual(values = plot_colors) +
   theme(text = element_text(size = 5),
         line = element_line(size = 0.1),
         axis.title.y = element_blank(),
         plot.margin = margin(0,0,0,0),
         legend.position = "none"
-        # legend.key.size = unit(2, 'mm'),
-        # legend.title = element_blank(),
-        # legend.position = "bottom",
-        # legend.margin=margin(-5,-5,-5,-5),
-        # legend.box.margin=margin(0,0,0,0)
-        
   ) +
   ylim(plot_range)
 
@@ -188,19 +181,12 @@ plotGG(
 
 # plot metaplot 4
 metaplot_4 <- plot_average(bw[4], regions = regions, row_split = regions$class, line_width = 0.2) +
-  # scale_color_brewer(palette = "Blues") +
   scale_color_manual(values = plot_colors) +
   theme(text = element_text(size = 5),
         line = element_line(size = 0.1),
         axis.title.y = element_blank(),
         plot.margin = margin(0,0,0,0),
         legend.position = "none"
-        # legend.key.size = unit(2, 'mm'),
-        # legend.title = element_blank(),
-        # legend.position = "bottom",
-        # legend.margin=margin(-5,-5,-5,-5),
-        # legend.box.margin=margin(0,0,0,0)
-        
   ) +
   ylim(plot_range)
 
@@ -210,6 +196,33 @@ plotGG(
   width = 2, height = 2, just = c("left", "top"),
   default.units = "cm"
 )
+
+# add labels to plots
+plotText(
+  label = "ChIP-seq signal", params = small_text_params,
+  x = ref_x, y = (ref_y + 1.25), just = "center", default.units = "cm", rot = 90
+)
+
+plotText(
+  label = "0 µM", params = large_text_params,
+  x = (ref_x + 1.25), y = (ref_y), just = "center", default.units = "cm"
+)
+
+plotText(
+  label = "500 µM", params = large_text_params,
+  x = (ref_x + 3.75), y = (ref_y), just = "center", default.units = "cm"
+)
+
+plotText(
+  label = "1000 µM", params = large_text_params,
+  x = (ref_x + 6.25), y = (ref_y), just = "center", default.units = "cm"
+)
+
+plotText(
[…]

### 2022-12-22 manuscript/manuscript.qmd
update MS results

@@
 XX), where neither factor is endogenously expressed.
-We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers (Extended Data Fig. 1b) [@Cherbas2011].
-We confirmed that Zld and Grh RNA (Extended Data Fig. XX) and protein (Extended Data Fig. XX) were not present in WT S2 cells.
+We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers (Extended Data Fig.
+1b) [@Cherbas2011].
+We confirmed that Zld and Grh RNA (Extended Data Fig.
+XX) and protein (Extended Data Fig.
+XX) were not present in WT S2 cells.
 The stable cell lines express Zld or Grh under the control of the metallothionein promoter, which allows inducible and tunable expression with the addition of CuSO~4.~ To determine protein levels that would be approximately physiological, we compared Zld or Grh levels in cells with varying concentrations of CuSO~4~ to the levels of endogenous expression found in Drosophila embryos (Extended Data Fig.
@@
 We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
-To control for antibody specificity, we performed additional ChIP experiments in WT cells using IgG, anti-Zld, or anti-Grh (Extended Data Fig. XX).
+To control for antibody specificity, we performed additional ChIP experiments in WT cells using anti-Zld, or anti-Grh (Extended Data Fig.
+XX).
 We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh.
@@
 
-To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of binding sites (Fig. 1a,b).
+To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of binding sites (Fig.
+1a,b).
 Class I sites are accessible in WT cells and remain accessible when bound by ectopically-expressed Zld or Grh.
@@
 Class III sites are inaccessible in WT cells and Zld or Grh binding results in chromatin opening.
-The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig. 1c,f).
-While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
+The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig.
+1c,f).
+While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig.
+1d,g).
 Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
@@
 
+```{r}
+#| include: false
+
+zld_RNAseq_results <- here("RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results.tsv") |> 
+  read_tsv() |> 
+  mutate(direction_of_change = case_when(
+    is_diff & log2FoldChange > 0 ~ "increased",
+    is_diff & log2FoldChange < 0 ~ "decreased",
+    !is_diff ~ "ns"
+  ))
+
+grh_RNAseq_results <- here("RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results.tsv") |> 
+  read_tsv() |> 
+  mutate(direction_of_change = case_when(
+    is_diff & log2FoldChange > 0 ~ "increased",
+    is_diff & log2FoldChange < 0 ~ "decreased",
+    !is_diff ~ "ns"
+  ))
+
+zld_diff <- zld_RNAseq_results |> 
+  pull(is_diff) |> 
+  sum()
+zld_up <- (zld_RNAseq_results$direction_of_change == "increased") |> 
+  sum()
+zld_down <- (zld_RNAseq_results$direction_of_change == "decreased") |> 
+  sum()
+
+grh_diff <- grh_RNAseq_results |> 
+  pull(is_diff) |> 
+  sum()
+grh_up <- (grh_RNAseq_results$direction_of_change == "increased") |> 
+  sum()
+grh_down <- (grh_RNAseq_results$direction_of_change == "decreased") |> 
+  sum()
+
+```
+
 Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation.
 To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression.
-Differential expression analysis using DESeq2 [@love2014] identified XXX differentially expressed genes for Zld and XXX for Grh (Extended Data Fig. XX).
+Differential expression analysis using DESeq2 [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig.
+XX).
 Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig.
@@
 To better understand the relationship between chromatin opening and transcriptional activation, we directly compared our RNA-seq and ATAC-seq data.
-Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l).
-sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig. XX <!--# add statistics -->).
-When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
+Overall, chromatin opening was correlated with transcriptional activation (Fig.
+1i-l).
+sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig.
+XX <!--# add statistics -->).
+When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig.
+1j,l).
 This was particularly evident at class III promoters, which were most likely to correspond to upregulated genes.
@@
 In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012].
-To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. XX).
+To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig.
+XX).
 We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
@@
 Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
-Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
+Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig.
+2b).
 At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c).
-Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. XX).
-Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig. XX).
-GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. XX).
+Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig.
+XX).
+Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig.
+XX).
+GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig.
+XX).
 Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
-The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig. XX).
-Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig. XX).
+The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig.
+XX).
+Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig.
+XX).
 This was in contrast to Zld and Grh, where induction results in the activation of many genes that are not directly bound by Zld or Grh.
@@
 
-While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites are shared across Zld, Grh, and Twi (Fig. 3a).
-This was in contrast to class II and III peaks, which were largely specific to each individual factor (Extended Data Fig. XX).
-Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig. XX).
+While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites are shared across Zld, Grh, and Twi (Fig.
+3a).
+This was in contrast to class II and III peaks, which were largely specific to each individual factor (Extended Data Fig.
+XX).
+Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig.
+XX).
 We hypothesized that these overlapping class I binding sites may occur at regulatory regions that are active in WT cells.
 To test this, we analyzed published ChIP-seq data from WT S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV.
-Overlapping class I regions had high levels of chromatin marks associated with active regulatory regions (Fig. 3b).
+Overlapping class I regions had high levels of chromatin marks associated with active regulatory regions (Fig.
+3b).
 This enrichment of all three factors at highly active regions suggests that exogenously expressed factors may bind nonspecifically to active regions of the genome.
-This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig. XX).
+This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig.
+XX).
 
@@
 These data are in agreement with previous observations that Oct4, Sox2 and Klf4 access binding sites in naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
-Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. XX).
-Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig. XX).
+Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig.
+XX).
+Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig.
+XX).
 Together this suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin.
@@
 To further test the specificity of binding within the three classes of ChIP-seq peaks, we analyzed the enrichment for the canonical Zld, Grh or Twi motifs.
-We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig. 3c).
+We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig.
+3c).
 When compared to all accessible regions, class I regions had low levels of enrichment for the canonical motifs, supporting that binding to these regions is nonspecific.
-By contrast, class II and III sites were highly enriched for the canonical motifs (Fig. 3c).
+By contrast, class II and III sites were highly enriched for the canonical motifs (Fig.
+3c).
 Class III sites had the highest percentage of sites containing a canonical motif.
@@
 We therefore analyzed the number of motifs per peak and motif strength (when scored against the canonical motif) for class I-III regions.
-For Grh and Twi, but not for Zld, class III sites had more motifs per peak than class II sites (Fig. 3f,h,j).
-For all three factors, class III sites tended to have stronger motifs than class II sites (Fig. 3g,i,k).
+For Grh and Twi, but not for Zld, class III sites had more motifs per peak than class II sites (Fig.
+3f,h,j).
+For all three factors, class III sites tended to have stronger motifs than class II sites (Fig.
+3g,i,k).
 Together, this analysis suggests that a combination of more motifs and stronger motifs at class III sites is what allows Zld, Grh, and Twi to open chromatin at these regions.
-<!--# Add description of de novo motif enrichment, put data in supplement. -->
+<!--# Add description of de novo motif enrichment, put data in supplement.
+-->
 
@@
 We analyzed the distance from motifs at class II and regions to the nearest nucleosome dyad.
-This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
-Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
+This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig.
+XX).
+Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig.
+XX).
 Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
@@
 Despite the presence of class IV, V and VI binding sites for all three proteins, there were distinct patterns of binding across different tissues.
-Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
-For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig. 4b).
+Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig.
+4a).
+For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig.
+4b).
 Twi binding sites in S2 cells had low levels of binding in embryos, but many of the strongest embryonic binding sites were unbound.
@@
 We treated our stable cell lines with DMSO or tazemetostat for 7 days, inducing Zld, Grh, or Twi expression on day 5.
-We confirmed loss of H3K27me3 by westerm blot and CUT&RUN [@Skene2017] (Extended Data Fig. XX).
+We confirmed loss of H3K27me3 by westerm blot and CUT&RUN [@Skene2017] (Extended Data Fig.
+XX).
 We used ChIP-seq to test if tazemetostat treatment resulted in changes to Zld, Grh, or Twi binding.
-Tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig. 4d-f).
+Tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig.
+4d-f).
 Despite a dramatic loss of H3K27me3, the majority of class IV binding sites remained unbound in tazemetostat-treated cells.
-We did identify some individual sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig. XX).
+We did identify some individual sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig.
+XX).
 Together, these data demonstrate that despite the correlation of H3K27me3 with resistance to pioneer factor binding, loss of PRC2 activity is insufficient to promote widespread binding of Zld, Grh, or Twi to repressed binding sites.
@@
 
--   For all three factors, increased concentration results in increased binding and opening of class II and III sites.
+-   ~~For all three factors, increased concentration results in increased binding and opening of class II and III sites.~~
 
--   For all three factors, increased concentration leads to novel binding sites.
+-   ~~For all three factors, increased concentration leads to novel binding sites.~~
 
--   For Zld and Twi, increased concentration is insufficient to allow binding to repressed sites.
+-   ~~For Zld and Twi, increased concentration is insufficient to allow binding to repressed sites.~~
 
--   For Grh, increased concentration does allow binding and opening at repressed sites
+-   ~~For Grh, increased concentration does allow binding and opening at repressed sites~~
 
-Previous studies have proposed that pioneer factor activity may depend on protein concentration [@Hansen2022].
+Previous studies have proposed that pioneer factor activity may depend on protein concentration [@Hansen2022; @Blassberg2022].
 Considering that tazemetostat treatment had limited effects on pioneer factor binding, we asked if varying the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.
+We induced cells to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig.
+XX; see methods).
+We performed ChIP-seq and analyzed the data at our previously defined class I-VI binding sites.
+We found that binding and was concentration dependent (Fig.
+5).
+At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig.
+5a,e,i).
+Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig.
+5a,e,i).
+We also performed ATAC-seq to determine if concentration-dependent increases in binding led to changes in chromatin accessibility.
+Changes in chromatin accessibility were also concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig.
+5b,f,j).
+Intriguingly, many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels.
+For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites (Fig.
+5, Extended Data Fig.
+XX).
+To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration.
+For Zld and Twi, supraphysiological concentrations led to an increase in class III regions relative to class II regions (Fig.
+5c,k).
+For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with the majority of these sites being falling in III (Fig.
+5g).
+To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs.
+supraphysiological concentrations (Fig.
+5d,h,i).
+For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration.
+Together, these data demonstrate that both binding and chromatin opening by pioneer factors are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are restistant to binding at physiological levls.
 
@@
 However, little is known about how regions outside the DBD may contribute to pioneering function.
-<!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat. Genet. paper) is outside the DBD -->.
+<!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat.
+Genet.
+paper) is outside the DBD.
+See also new Zaret paper in NSMB looking at contribution of IDRs and activation domains to chromatin opening on nucleosome arrays in vitro -->.
 To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig.
@@
 This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required.
-Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
+Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig.
+6b,e).
 For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
@@
 | Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|--------------------|------------------|------------------|------------------|
+|-------------------|------------------|------------------|------------------|
 | pMT-puro                              | AddGene                   | 17923                |                         |
@@
 
-**Figure 1: Ectopic pioneer factors open chromatin and activate transcription.**
+**Figure 1: Ectopic pioneer factors open chromatin and activate transcription.
+a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
+ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
+**c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
+**d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
+**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h).
+i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
+**j,i,** Proportion of class I, II or III regions that are proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
+Promoter-proximal and promoter-distal binding sites are shown separately.
+The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene.
 
-**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.**
+**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.
+a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites.
+**c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions.
+**d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.
 
-**Figure 3: Motif content shapes pioneer-factor activity**
+**Figure 3: Motif content shapes pioneer-factor activity.
+a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
+**b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
+Zld, Grh, and Twi ChIP-seq data were generated in this study.
+Remaining ChIP-seq datasets were previously published (see Supplementary table XX).
+**c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
+Logos plot of the canonical motif are shown alongside the heatmap.
+Percentages are shown for class I, II or III binding sites, or all WT ATAC-seq peaks as a control for the background motif frequency.
+**f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
+**g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**.
+Motif scores were determined by comparing each identified motif instance to the canonical motif.
 
-**Figure 4: H3K27me3 and H3K9me3 are barriers to pioneer-factor binding.**
+**Figure 4: Many endogenous binding sites are resistant to ectopic pioneer-factor binding.
+a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
+ChIP-seq data for repressive histone modifications H3K27me3 and H3K9me3 is also shown.
+**d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
+Zld, Grh, and Twi ChIP-seq data are z-score normalized.
+CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).
 
-**Figure 5: Binding and opening of closed chromatin is concentration dependent.**
+**Figure 5: Binding and opening of closed chromatin is concentration dependent.
+a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
+Concetrations of CuSO~4~ used to induce protein expression are indicated above each plot.
+**b,f,j,** Metaplots of ATAC-seq signal at class I-VI regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
+**c,g,k,** Distribution of class I, II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
+Y-axis indicates the total number of peaks called at each concentration.
+Colors indicate the proportion of those peaks that are defined as class I, II, or III.
+**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
 
-**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function.**
+**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function.
+a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
+Shown for Zld **(a)** and Grh **(b)**.
+**c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**.
+**d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
 

### 2022-12-22 workflow/scripts/fig4.R

@@ -11,26 +11,58 @@ source("workflow/scripts/plot_heatmap.R")
 source("workflow/scripts/utils.R")
 
 # define input files ===========================================================
-zld_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv")
-grh_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/grh_tissue_classes.tsv")
-twi_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/twi_tissue_classes.tsv")
+# zld_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv")
+# grh_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/grh_tissue_classes.tsv")
+# twi_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/twi_tissue_classes.tsv")
+# 
+# S2_ZLD_ChIP_bw <-   "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_IP.bw"
+# embryo_Zld_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/embryo-nc14_aZld.bw"
+# brain_Zld_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/brain_aZld.bw"
+# S2_H3K27me3_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE151983_S2_aH3K27me3_IP.bw"
+# S2_H3K9me3_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE160855_aH3K9me3.bw"
+# 
+# S2_Grh_ChIP_bw <-   "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_IP.bw"
+# embryo_Grh_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/embryo-15-16H_aGrh.bw"
+# wing_disc_Grh_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/wing-disc_aGrh.bw"
+# 
[…]

### 2022-12-22 workflow/scripts/fig5.R

@@ -11,48 +11,85 @@ source("workflow/scripts/plot_heatmap.R")
 source("workflow/scripts/utils.R")
 
 # define input files ===========================================================
-zld_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv")
-grh_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/grh_tissue_classes.tsv")
-twi_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/twi_tissue_classes.tsv")
+# zld_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv")
+# grh_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/grh_tissue_classes.tsv")
+# twi_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/twi_tissue_classes.tsv")
+# 
+# zld_0uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-0uM_aZld_IP.bw"
+# zld_500uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-500uM_aZld_IP.bw"
+# zld_1000uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-1000uM_aZld_IP.bw"
+# zld_1500uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-1500uM_aZld_IP.bw"
+# 
+# grh_0uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-0uM_aGrh_IP.bw"
+# grh_25uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-25uM_aGrh_IP.bw"
+# grh_100uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-100uM_aGrh_IP.bw"
+# grh_400uM_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-400uM_aGrh_IP.bw"
+# 
[…]

### 2023-01-02 manuscript/manuscript.qmd
first draft of results section

@@
 We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
-To control for antibody specificity, we performed additional ChIP experiments in WT cells using anti-Zld, or anti-Grh (Extended Data Fig.
-XX).
+To control for antibody specificity, we performed additional ChIP experiments in WT cells using anti-Zld, or anti-Grh (Extended Data Fig. XX).
 We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh.
@@
 Class III sites are inaccessible in WT cells and Zld or Grh binding results in chromatin opening.
-The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig.
-1c,f).
-While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig.
-1d,g).
+The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig. 1c,f).
+While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
 Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
@@
 To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression.
-Differential expression analysis using DESeq2 [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig.
-XX).
+Differential expression analysis using DESeq2 [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. XX).
 Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig.
@@
 Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
-Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig.
-2b).
+Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
 At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c).
-Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig.
-XX).
-Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig.
-XX).
-GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig.
-XX).
+Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. XX).
+Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig. XX).
+GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. XX).
 Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
-The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig.
-XX).
-Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig.
-XX).
+The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig. XX).
+Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig. XX).
 This was in contrast to Zld and Grh, where induction results in the activation of many genes that are not directly bound by Zld or Grh.
@@
 
-While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites are shared across Zld, Grh, and Twi (Fig.
-3a).
-This was in contrast to class II and III peaks, which were largely specific to each individual factor (Extended Data Fig.
-XX).
-Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig.
-XX).
+While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites are shared across Zld, Grh, and Twi (Fig. 3a).
+This was in contrast to class II and III peaks, which were largely specific to each individual factor (Extended Data Fig. XX).
+Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig. XX).
 We hypothesized that these overlapping class I binding sites may occur at regulatory regions that are active in WT cells.
 To test this, we analyzed published ChIP-seq data from WT S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV.
-Overlapping class I regions had high levels of chromatin marks associated with active regulatory regions (Fig.
-3b).
+Overlapping class I regions had high levels of chromatin marks associated with active regulatory regions (Fig. 3b).
 This enrichment of all three factors at highly active regions suggests that exogenously expressed factors may bind nonspecifically to active regions of the genome.
-This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig.
-XX).
+This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig. XX).
 
@@
 These data are in agreement with previous observations that Oct4, Sox2 and Klf4 access binding sites in naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
-Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig.
-XX).
-Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig.
-XX).
+Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. XX).
+Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig. XX).
 Together this suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin.
@@
 To further test the specificity of binding within the three classes of ChIP-seq peaks, we analyzed the enrichment for the canonical Zld, Grh or Twi motifs.
-We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig.
-3c).
+We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig. 3c).
 When compared to all accessible regions, class I regions had low levels of enrichment for the canonical motifs, supporting that binding to these regions is nonspecific.
-By contrast, class II and III sites were highly enriched for the canonical motifs (Fig.
-3c).
+By contrast, class II and III sites were highly enriched for the canonical motifs (Fig. 3c).
 Class III sites had the highest percentage of sites containing a canonical motif.
@@
 We therefore analyzed the number of motifs per peak and motif strength (when scored against the canonical motif) for class I-III regions.
-For Grh and Twi, but not for Zld, class III sites had more motifs per peak than class II sites (Fig.
-3f,h,j).
-For all three factors, class III sites tended to have stronger motifs than class II sites (Fig.
-3g,i,k).
+For Grh and Twi, but not for Zld, class III sites had more motifs per peak than class II sites (Fig. 3f,h,j).
+For all three factors, class III sites tended to have stronger motifs than class II sites (Fig. 3g,i,k).
 Together, this analysis suggests that a combination of more motifs and stronger motifs at class III sites is what allows Zld, Grh, and Twi to open chromatin at these regions.
-<!--# Add description of de novo motif enrichment, put data in supplement.
--->
+<!--# Add description of de novo motif enrichment, put data in supplement. -->
 
@@
 We analyzed the distance from motifs at class II and regions to the nearest nucleosome dyad.
-This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig.
-XX).
-Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig.
-XX).
+This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
+Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
 Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
@@
 We compared Zld binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
-We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig.
-XX).
+We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig. XX).
 Previous work has shown that regions resistant to pioneer-factor binding are enriched for the repressive histone modifications H3K9me3 and H3K27me3 [@Mayran2018a; @Soufi2012].
@@
 Despite the presence of class IV, V and VI binding sites for all three proteins, there were distinct patterns of binding across different tissues.
-Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig.
-4a).
-For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig.
-4b).
+Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
+For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig. 4b).
 Twi binding sites in S2 cells had low levels of binding in embryos, but many of the strongest embryonic binding sites were unbound.
@@
 We treated our stable cell lines with DMSO or tazemetostat for 7 days, inducing Zld, Grh, or Twi expression on day 5.
-We confirmed loss of H3K27me3 by westerm blot and CUT&RUN [@Skene2017] (Extended Data Fig.
-XX).
+We confirmed loss of H3K27me3 by westerm blot and CUT&RUN [@Skene2017] (Extended Data Fig. XX).
 We used ChIP-seq to test if tazemetostat treatment resulted in changes to Zld, Grh, or Twi binding.
-Tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig.
-4d-f).
+Tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig. 4d-f).
 Despite a dramatic loss of H3K27me3, the majority of class IV binding sites remained unbound in tazemetostat-treated cells.
-We did identify some individual sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig.
-XX).
+We did identify some individual sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig. XX).
 Together, these data demonstrate that despite the correlation of H3K27me3 with resistance to pioneer factor binding, loss of PRC2 activity is insufficient to promote widespread binding of Zld, Grh, or Twi to repressed binding sites.
@@
 Considering that tazemetostat treatment had limited effects on pioneer factor binding, we asked if varying the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.
-We induced cells to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig.
-XX; see methods).
+We induced cells to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig. XX; see methods).
 We performed ChIP-seq and analyzed the data at our previously defined class I-VI binding sites.
-We found that binding and was concentration dependent (Fig.
-5).
-At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig.
-5a,e,i).
-Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig.
-5a,e,i).
+We found that binding and was concentration dependent (Fig. 5).
+At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i).
+Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i).
 We also performed ATAC-seq to determine if concentration-dependent increases in binding led to changes in chromatin accessibility.
-Changes in chromatin accessibility were also concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig.
-5b,f,j).
+Changes in chromatin accessibility were also concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig. 5b,f,j).
 Intriguingly, many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels.
-For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites (Fig.
-5, Extended Data Fig.
-XX).
+For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites (Fig. 5, Extended Data Fig. XX).
 To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration.
-For Zld and Twi, supraphysiological concentrations led to an increase in class III regions relative to class II regions (Fig.
-5c,k).
-For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with the majority of these sites being falling in III (Fig.
-5g).
+For Zld and Twi, supraphysiological concentrations led to an increase in class III regions relative to class II regions (Fig. 5c,k).
+For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with the majority of these sites being falling in III (Fig. 5g).
 To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs.
@@
 This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required.
-Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig.
-6b,e).
+Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
 For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
@@
 | Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|-------------------|------------------|------------------|------------------|
+|---------------------------------------|---------------------------|----------------------|-------------------------|
 | pMT-puro                              | AddGene                   | 17923                |                         |
@@
 
-**Figure 1: Ectopic pioneer factors open chromatin and activate transcription.
-a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
+**Figure 1: Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
 ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
@@
 
-**Figure 3: Motif content shapes pioneer-factor activity.
-a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
+**Figure 3: Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
 **b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.

### 2023-01-04 manuscript/manuscript.qmd
update MS results section

--- 2022-12-22 manuscript/manuscript.qmd
@@
 | Reagent                               | Source/vendor              | Catalog/accession \# | Notes                   |
-|-------------------|------------------|------------------|------------------|
+|---------------------|-----------------|-----------------|-----------------|
 | pMT-puro                              | AddGene                    | 17923                |                         |
@@
 | Anti-HA clone 12CA5                   | Roche                      |                      | ChIP: 7.5 µL            |
-| Anti-H3K27me3                         | Cell Signaling Technology |                      | WB: 1:1000, CR: 0.5 µL  |
+| Anti-H3K27me3                         | Cell Signaling Technology  | C36B11; Lot #19      | WB: 1:1000, CR: 0.5 µL  |
 |                                       |                            |                      |                         |
@@
 
-## Cloning
+##
+
+## Cell culture and generation of stable cell lines
 
@@
 For cloning of the Zld and Grh DNA-binding domains, DNA encoding amino acids 1114-1487 (Zld) or amino acids 603-1032 (Grh) was cloned into pMT-puro.
-
--   **TG2A- Zld RB**
-
--   **TG3A- Grh PH**
-
--   **TG4A**
-
--   **TG6A**
-
--   **TG8A**
-
--   **MF8A**
-
--   GFP-DBD lines
-
-## Cell culture and generation of stable cell lines
 
@@
 Reads mapping to unplaced scaffolds or the mitochondrial genome were discarded.
-
-Alignment
-
-Filtering
-
-peak calling
-
-peak filtering
-
-generation of bigwigs
-
-normalization of bigwigs
+Peak calling was performed using MACS2 version 2.2 with parameters -g dm \--call-summits.
+Only peaks that were detected in all replicates were considered in downstream analysis.
+bigWig files were generated using deepTools bamCoverage version 3.5.1 with parameters --binSize 10.
+bigWig files were z-score normalized by subtracting the genome-wide average of all bins from each bin value and dividing by the standard deviation of all bins.
 
@@
 
+ATAC-seq was performed as described previsouly [@Buenrostro2013].
+2×10^5^ cells were harvested for ATAC-seq by centrifugation at 600 × *g* for 3 minutes at RT.
+Cells were washed once in 100 µL 1X PBS and resuspended in 100 µL ATAC lysis buffer (10 mM Tris pH 7.5, 10 mM NaCl, 3 mM MgCl~2~,0.1% NP-40).
+Cells were centrifuged at 600 × *g* for 10 minutes at 4°C.
+Supernatant was removed and pellet was resuspended in 47.5 µL buffer TD (Illumina).
+2.5 uL of Tagment DNA Enzyme (Illumina) was added and samples were incubated at 37°C for 30 minutes.
+Tagmented DNA was purified using MinElute Reaction Cleanup Kit (Qiagen) and eluted in 10 µL of buffer EB.
+DNA was PCR amplified for 12 cycles with the following conditions:72°C for 5 minutes, 98°C for 30 seconds, then 12 cycles of 98°C for 10 seconds, 63°C for 30 seconds and 72°C for 1 minute.
+Amplified libraries were purified using a 1.2X ratio of Axygen paramagnetic beads.
+Sequencing was performed on the Illumina NovaSeq 6000 using 150bp paired-end reads.
+
 ## ATAC-seq analysis
+
+Raw ATAC-seq reads were trimmed to remove adapter sequences using NGmerge version 0.3 [@Gaspar2018].
+Trimmed reads were aligned to the *Drosophila melanogaster* genome (version dm6) using bowtie2 with paramters -k 2 \--very-sensitive \--no-mixed \--no-discordant -X 5000.
+Reads with a mapping quality \< 30 and reads aligning to the mitochondrial genome or scaffolds were discarded.
+As described previously, only fragments \< 100bp were considered for downstream analysis [@Buenrostro2013].
+Reads were combined across all replicates and peak calling was performed using MACS2 with parameters -f BAMPE \--keep-dup all -g dm \--call-summits.
+featureCounts was used to count the number of reads aligning within 200bp (100bp upstream or downstream) of peak summits.
+Differential accessibility analysis was performed using DESeq2.
+Regions with an adjusted p-value \< 0.05 were considered differentially accessible.
+
+## Treatment of S2 cells with tazemetostat
+
+Cells were plated at 1×10^6^ cells per mL and DMSO or tazemetostat was added to a final concentration of 10 µM.
+Cells were split on days 2 and 5 and tazemetostat was added to maintain a final concentration of 10 µM.
+On day 5, CuSO~4~ was added to achieve physiological expression of Zld, Grh or Twi (see above).
+On day 7, cells were harvested for immunoblotting, ChIP and CUT&RUN.
 
@@
 
+CUT&RUN was performed using EpiCypher reagents according to the manufacturer protocol.
+2×10^5^ cells were used for each CUT&RUN reaction.
+Overnight incubation with antibodies was performed in antibody buffer containing 0.05% digitonin.
+0.5 µL anti-H3K27me3 (Cell signaling technologies) or rabbit IgG were used for CUT&RUN.
+Prior to addition of antibody, a 1:5 dilution of K-MetStat Panel spike-in nucleosomes (EpiCypher, cat. \# 19-1002) was added to each reaction.
+Libraries were prepared using the NEBNext Ultra II kit (NEB).
+During library preparation, cleanup steps were performed using a 1.1X ratio of Axygen paramagnetic beads, as recommended by EpiCypher.
+PCR amplification was performed with the following conditions: 98°C for 45 seconds, then 14 cycles of 98°C for 15 seconds, 60°C for 10 seconds, and a final extension of 72°C for 1 minute.
+Sequencing was performed on the Illumina NovaSeq 6000 with 150bp paired-end reads.
+
 ## CUT&RUN analysis
 
-## Treatment of S2 cells with tazemetostat
+Trimming
+
+Alignment
+
+Filtering
+
+peak calling
+
+bigwigs
+
+z-score normalization
+
+spike-in normalization
 
@@
 
+Previously published datasets analyzed in this study can be found in supplemental table XXXX.
+All previously published data were analyzed in parallel with data from this study using analysis parameters described above.
+
 ## Data and code availability
+
+Sequencing data have been deposited in the Gene Expresion Omnibus under accession GSEXXXX.
+All analysis code is available on GitHub: https://github.com/tjgibson/S2_pioneers_manuscript.
+An archived version of the data processing and analysis pipeline is available on Zenodo: doiXXXX.
 

### 2023-01-04 workflow/scripts/extended_data_fig1.R

@@ -1,0 +1,82 @@
+# setup ========================================================================
+library(plotgardener)
+library(tidyverse)
+library(org.Dm.eg.db)
+library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
+library(grImport)
+library(grid)
+library(RColorBrewer)
+
+
+# define input files ===========================================================
+Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
+Zld_WT_ATAC_bw <- snakemake@input[["Zld_WT_ATAC_bw"]]
+Zld_Zld_ATAC_bw <- snakemake@input[["Zld_Zld_ATAC_bw"]]
+Zld_WT_RNAseq_bw <- snakemake@input[["Zld_WT_RNAseq_bw"]]
+Zld_Zld_RNAseq_bw <- snakemake@input[["Zld_Zld_RNAseq_bw"]]
+
+Grh_ChIP_bw <- snakemake@input[["Grh_ChIP_bw"]]
+Grh_WT_ATAC_bw <- snakemake@input[["Grh_WT_ATAC_bw"]]
+Grh_Grh_ATAC_bw <- snakemake@input[["Grh_Grh_ATAC_bw"]]
+Grh_WT_RNAseq_bw <- snakemake@input[["Grh_WT_RNAseq_bw"]]
+Grh_Grh_RNAseq_bw <- snakemake@input[["Grh_Grh_RNAseq_bw"]]
+
+zld_ChIP_classes_fn <- snakemake@input[["zld_ChIP_classes"]]
+grh_ChIP_classes_fn <- snakemake@input[["grh_ChIP_classes"]]
+
+# # create blank layout for plot ===============================================
+pdf(snakemake@output[[1]], useDingbats = FALSE)
+pageCreate(width = 18, height = 12.5, default.units = "cm", showGuides = TRUE)
+
+# general figure settings ======================================================
+# text parameters for Nature Genetics
+panel_label_params <- pgParams(fontsize = 8)
+large_text_params <- pgParams(fontsize = 7)
+small_text_params <- pgParams(fontsize = 5)
+
+# colors
+
+# model ========================================================================
+  PostScriptTrace(
+    "manuscript/figures/models/system_schematic.eps",
+    outfilename = "manuscript/figures/models/system_schematic.eps.xml",
+    charpath = FALSE,
+    charpos = FALSE
+  )
+schematic <- readPicture("manuscript/figures/models/system_schematic.eps.xml")
+# grid.picture(schematic)
[…]

### 2023-01-05 manuscript/manuscript.qmd
update outline of Extended Data figures

@@
 | Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|---------------------|-----------------|-----------------|-----------------|
+|--------------------|------------------|------------------|------------------|
 | pMT-puro                              | AddGene                   | 17923                |                         |
@@
 
--   RNA-seq data to show expression of S2 and not Kc cell markers.
-
 -   Western blots to determine physiological expression and lack of protein expression in WT cells.
@@
 
-**Extended Data Figure 2: Ectopically expressed Zld and Grh bind to thousands of sites in S2 cells.**
-
--   ChIP-seq replicate correlation
-
--   Genome browser and heatmaps for all ChIP peaks?
-
--   IgG and WT controls for antibody specificity
-
-**Extended Data Figure 3: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility.**
-
--   ATAC PCA plot
-
--   ATAC replicate correlation
-
--   volcano and/or MA plots with enrichment of binding
-
--   Example genome browsers?
-
--   Analysis of nonspecific changes?
-
-**Extended Data Figure 4: Zld and Grh bind to chromatin rapidly after induction of protein expression**
+**Extended Data Figure 2: Zld and Grh bind to chromatin rapidly after induction of protein expression**
 
@@
 
-**Extended Data Figure 5: Zld and Grh binding promotes transcriptional activation at a subset of binding sites**
+**Extended Data Figure 3: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.**
+
+-   volcano and/or MA plots with enrichment of binding
+
+-   Analysis of nonspecific changes?
 
 -   RNA-seq PCA
-
--   RNA-seq replicate correlation
 
@@
 
-**Extended Data Figure 6: Twist binding leads to chromatin opening and transcriptional activation**
+**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation**
 
--   Twi ATAC PCA, volcano, and examples
+-   Twi volcano and examples
 
--   TWI RNA-seq PCA, volcano, and examples
+-   TWI RNA-seq volcano and examples
 
@@
 
-**Extended Data Figure 7: Chromatin features associated with Zld, Grh and Twi binding sites**
+**Extended Data Figure 5: Chromatin features associated with Zld, Grh and Twi binding sites**
 
@@
 
-**Extended Data Figure 8: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes**
+**Extended Data Figure 6: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes**
+
+-   De novo motif enrichment for each factor
 

### 2023-01-05 manuscript/manuscript.qmd
update references to extended data Fig 1 in MS text

@@
 To investigate the mechanisms of chromatin binding and opening by pioneer factors, we generated stable cell lines inducibly expressing Zelda (Zld) and Grainy head (Grh) in Drosophila S2 cells (Extended Data Fig.
-XX), where neither factor is endogenously expressed.
-We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers (Extended Data Fig. 1b) [@Cherbas2011].
-We confirmed that Zld and Grh RNA (Extended Data Fig. XX) and protein (Extended Data Fig. XX) were not present in WT S2 cells.
+1a), where neither factor is endogenously expressed.
+We confirmed that Zld and Grh RNA (Extended Data Fig. 1b) and protein (Extended Data Fig. 1c,d) were not present in WT S2 cells.
 The stable cell lines express Zld or Grh under the control of the metallothionein promoter, which allows inducible and tunable expression with the addition of CuSO~4.~ To determine protein levels that would be approximately physiological, we compared Zld or Grh levels in cells with varying concentrations of CuSO~4~ to the levels of endogenous expression found in Drosophila embryos (Extended Data Fig.
-XX).
+1c,d).
 Physiological protein levels were achieved at 1000 mM or 100 mM CuSO~4~ for Zld or Grh, respectively.
@@
 We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
-To control for antibody specificity, we performed additional ChIP experiments in WT cells using anti-Zld, or anti-Grh (Extended Data Fig. XX).
+To control for antibody specificity, we performed additional ChIP experiments in WT cells using anti-Zld, or anti-Grh.
 We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh.
@@
 
+We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers and the absence of Kc cell markers [@Cherbas2011].
 Drosophila S2 cells were cultured at 27°C in Schneider's medium (Thermo Fisher Scientific) supplemented with 10% fetal bovine serum (Omega Scientific) and antibiotic/antimycotic (Thermo Fisher Scientific).

### 2023-01-06 manuscript/manuscript.qmd
break manuscript up into multiple qmd files

--- 2023-01-05 manuscript/manuscript.qmd
@@
 
-```{r}
-#| include: false
-library(here)
-library(tidyverse)
-```
-
 # Abstract
+{{< include manuscript_sections/_01_Abstract.qmd >}}
 
 # Introduction
+{{< include manuscript_sections/_02_Introduction.qmd >}}
 
 # Results
-
-## Ectopically expressed pioneer factors bind and open closed chromatin
-
--   ~~Establish system for ectopic expression of pioneer factors~~
-
--   ~~Identification of physiological expression levels~~
-
--   ~~Define binding classes I-III and describe characteristics (distribution, promoters/enhancers, etc.)~~
-
--   Time course and dynamics of binding to different classes (supplement)
-
-To investigate the mechanisms of chromatin binding and opening by pioneer factors, we generated stable cell lines inducibly expressing Zelda (Zld) and Grainy head (Grh) in Drosophila S2 cells (Extended Data Fig.
-XX), where neither factor is endogenously expressed.
-We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers (Extended Data Fig. 1b) [@Cherbas2011].
-We confirmed that Zld and Grh RNA (Extended Data Fig. XX) and protein (Extended Data Fig. XX) were not present in WT S2 cells.
-The stable cell lines express Zld or Grh under the control of the metallothionein promoter, which allows inducible and tunable expression with the addition of CuSO~4.~ To determine protein levels that would be approximately physiological, we compared Zld or Grh levels in cells with varying concentrations of CuSO~4~ to the levels of endogenous expression found in Drosophila embryos (Extended Data Fig.
-XX).
-Physiological protein levels were achieved at 1000 mM or 100 mM CuSO~4~ for Zld or Grh, respectively.
-
-```{r}
-#| include: false
-# zld_ChIP_classes <- snakemake@input[["zld_ChIP_classes"]]
-# grh_ChIP_classes <- snakemake@input[["grh_ChIP_classes"]]
-
-zld_ChIP_classes <- here("results/ChIP_peak_classes/zld_ChIP_classes.tsv") |> 
-  read_tsv()
-grh_ChIP_classes <- here("results/ChIP_peak_classes/grh_ChIP_classes.tsv") |> 
-  read_tsv()
-
-zld_n_ChIP_peaks <- nrow(zld_ChIP_classes)
-grh_n_ChIP_peaks <- nrow(grh_ChIP_classes)
-```
-
-To determine how exogenous Zld and Grh interact with preexisting chromatin in S2 cells, we expressed each factor at physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
-We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
-To control for antibody specificity, we performed additional ChIP experiments in WT cells using anti-Zld, or anti-Grh (Extended Data Fig. XX).
-We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh.
-<!--# Not sure the best way to discuss the control ChIP experiments.
-Should I mention how many peaks we detect in WT cells?
-What supplemental figures should I include?
-Should peak calling be done using IgG or WT cells as control instead of input?
--->
-
-```{r}
-#| include: false
-
-zld_ATAC_results <- here("ATACseq/results/DEseq2_results_filtered/S2-Zld_ATACseq_S2-Zld-FL-vs-S2-WT_results.tsv") |> 
-  read_tsv() |> 
-  mutate(direction_of_change = case_when(
-    is_diff & log2FoldChange > 0 ~ "increased",
-    is_diff & log2FoldChange < 0 ~ "decreased",
-    !is_diff ~ "ns"
-  ))
-
-grh_ATAC_results <- here("ATACseq/results/DEseq2_results_filtered/S2-Grh_ATACseq_S2-Grh-FL-vs-S2-WT_results.tsv") |> 
-  read_tsv() |> 
-  mutate(direction_of_change = case_when(
-    is_diff & log2FoldChange > 0 ~ "increased",
-    is_diff & log2FoldChange < 0 ~ "decreased",
-    !is_diff ~ "ns"
-  ))
-
-zld_diff <- zld_ATAC_results |> 
-  pull(is_diff) |> 
-  sum()
-zld_up <- (zld_ATAC_results$direction_of_change == "increased") |> 
-  sum()
-zld_down <- (zld_ATAC_results$direction_of_change == "decreased") |> 
-  sum()
-
-grh_diff <- grh_ATAC_results |> 
-  pull(is_diff) |> 
-  sum()
-grh_up <- (grh_ATAC_results$direction_of_change == "increased") |> 
-  sum()
-grh_down <- (grh_ATAC_results$direction_of_change == "decreased") |> 
-  sum()
-
-```
-
-Having established that Zld and Grh can bind to thousands of sites in S2 cells, we performed ATAC-seq on WT cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
-We performed differential accessibility analysis comparing Zld or Grh-expressing cells to WT cells.
-We detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
-XX).
-<!--# Add more information here about sites with differential accessibility.
-Include percentage of increased and decreased sites that are bound by Zld or Grh.
-Comment on how much is specific vs.
-secondary effect, and justify defining ChIP peak classes that include only a subset of DA sites.
--->
-
-To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of binding sites (Fig. 1a,b).
-Class I sites are accessible in WT cells and remain accessible when bound by ectopically-expressed Zld or Grh.
-Class II sites are inaccessible in WT cells, however binding of Zld or Grh does not lead to chromatin opening.
-Class III sites are inaccessible in WT cells and Zld or Grh binding results in chromatin opening.
-The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig. 1c,f).
-While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
-Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
-1e,h).
-Together, these data show that when expressed exogenously at physiolgical levels, Zld and Grh can bind and open closed chromatin at a subset of binding sites.
-
-<!--# Could potentially add the cut&run time course here? Think about where to put this, or if it is worth including at all. -->
-
-## Chromatin opening correlates with transcriptional activation
-
--   ~~Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.~~
-
--   ~~Comparison of RNA-seq and ATAC data: chromatin opening correlates with transcriptional activation.~~
-
--   ~~Class III promoters are most likely to be associated with upregulated genes~~
-
-```{r}
-#| include: false
-
-zld_RNAseq_results <- here("RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results.tsv") |> 
-  read_tsv() |> 
-  mutate(direction_of_change = case_when(
-    is_diff & log2FoldChange > 0 ~ "increased",
-    is_diff & log2FoldChange < 0 ~ "decreased",
-    !is_diff ~ "ns"
-  ))
-
-grh_RNAseq_results <- here("RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results.tsv") |> 
-  read_tsv() |> 
-  mutate(direction_of_change = case_when(
-    is_diff & log2FoldChange > 0 ~ "increased",
-    is_diff & log2FoldChange < 0 ~ "decreased",
-    !is_diff ~ "ns"
-  ))
-
-zld_diff <- zld_RNAseq_results |> 
-  pull(is_diff) |> 
-  sum()
-zld_up <- (zld_RNAseq_results$direction_of_change == "increased") |> 
-  sum()
-zld_down <- (zld_RNAseq_results$direction_of_change == "decreased") |> 
-  sum()
-
-grh_diff <- grh_RNAseq_results |> 
-  pull(is_diff) |> 
-  sum()
-grh_up <- (grh_RNAseq_results$direction_of_change == "increased") |> 
-  sum()
-grh_down <- (grh_RNAseq_results$direction_of_change == "decreased") |> 
-  sum()
-
-```
-
-Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation.
-To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression.
-Differential expression analysis using DESeq2 [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. XX).
-Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig.
-XX <!--# include statistics  -->), suggesting that binding of these factors mainly leads to transcriptional activation and not repression.
-<!--# Consider mentioning Grh as a repressor here and giving examples of genes where Grh may be repressing.
-Discuss if Grh-mediated transcriptional repression coincides with chromatin opening, chromatin closing, or neither --> Gene ontology (GO) analysis indicated that the genes activated by each factor are consistent with the endogenous functions of Zld and Grh (Extended Data Fig.
-XX).
-<!--# Add more detailed information on GO analysis here -->
-
-To better understand the relationship between chromatin opening and transcriptional activation, we directly compared our RNA-seq and ATAC-seq data.
-Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l).
-sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig. XX <!--# add statistics -->).
-When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
-This was particularly evident at class III promoters, which were most likely to correspond to upregulated genes.
-Together, we conclude that while binding and chromatin opening can happen in the absence of transcriptional changes, chromatin opening is correlated with transcriptional activation.
-
-## Twist binds closed chromatin extensively and opens chromatin at a subset of sites
-
--   ~~Introduce Twist and justify the comparison to Zld/Grh.~~
-
--   ~~Twi binds closed chromatin and opens at some sites.~~
-
--   Mention analysis of published Abd-A and Ubx cell lines.
-
--   ~~Class I-III sites, distribution, etc.~~
-
--   ~~Twi also promotes transcriptional activation~~
-
--   ~~Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.~~
-
--   ~~Comparison of RNA-seq and ATAC~~
-
--   ~~Mention differences between Twi vs. Zld and Grh~~
-
-Despite the differences between Zld and Grh during Drosophila development, both proteins behaved similarly when expressed in our exogenous S2 cell system.
-To provide additional context to our findings, we performed additional experiments with the transcription factor Twist (Twi).
-Although Twi is also essential for regulating Drosophila embryonic development, Twi has a number if distnict properties from Zld and Grh, making it an interesting point of comparison.
-Twi is a master regulator of mesodermal cell fate (REFS).
-Twi has many binding sites that are dynamic during early embryonic development [@Wilczynski2010a], in contrast to Zld and Grh.
-Although Twi was hypothesized to be a pioneer factor [@Sandmann2007], pioneering activity by Twi had not previously been tested.
-In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012].
-To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. XX).
-We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
-Comparison of Twi ChIP-seq and ATAC-seq data allowed for the identification of the same classes of binding sites that were found for Zld and Grh (Fig 2a,b).
-Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
-Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
-At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c).
-Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. XX).
-Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig. XX).
-GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. XX).
-Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
-The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig. XX).
-Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig. XX).
-This was in contrast to Zld and Grh, where induction results in the activation of many genes that are not directly bound by Zld or Grh.
-This suggests that Zld and Grh may activate many transcription factors, which go on to activate their own targets, leading to secondary effects in the RNA-seq data.
-Twi, by contrast, appears to directly activate genes required for mesodermal and muscle cell fate.
-<!--# Is this worth mentioning?
-Might require additional analysis of how many TFs are upregulated in RNA-seq data for each factor.
-Could also require additional motif analysis to identify/support secondary TFs and reason for nonspecific effects.
---> Similar to Zld and Grh, transcriptional activation was correlated with chromatin opening (Extended Data Fig.
-XX).
-Together, our data show that Twi can act as a pioneer factor.
-
-## Ectopically expressed transcription factors bind opportunistically to active chromatin
-
--   ~~Overlap of class I regions suggests nonspecific binding~~
-
--   ~~Analysis of published ChIP data shows that class I regions are within active chromatin~~
-
--   ~~RNA-seq and ATAC-seq for class I to show that the binding is not functional~~
-
--   ~~Analysis of chromatin at class II-III regions: mostly devoid of histone modifications~~
-
--   ~~Class III regions have higher H3 and H1 occupancy compared to class II~~
-
--   Where to put analysis of nucleosome positioning?
-
-While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites are shared across Zld, Grh, and Twi (Fig. 3a).
-This was in contrast to class II and III peaks, which were largely specific to each individual factor (Extended Data Fig. XX).
-Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig. XX).
-We hypothesized that these overlapping class I binding sites may occur at regulatory regions that are active in WT cells.
-To test this, we analyzed published ChIP-seq data from WT S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV.
-Overlapping class I regions had high levels of chromatin marks associated with active regulatory regions (Fig. 3b).
-This enrichment of all three factors at highly active regions suggests that exogenously expressed factors may bind nonspecifically to active regions of the genome.
-This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig. XX).
-
-Considering the enrichment of active chromatin at class I sites, we asked if there were any features of chromatin distinctive to class II or III regions.
-Analysis of published ChIP-seq data demonstrated that class II and III sites have low levels of many histone modifications, chromatin-associated proteins, and transcription factors (Extended Data Fig.
-XX).
-These data are in agreement with previous observations that Oct4, Sox2 and Klf4 access binding sites in naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
-Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. XX).
-Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig. XX).
-Together this suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin.
-Based on these observations, the inability to open class II regions is unlikely to be due to differences in preexisting features of chromatin.
-
-## Chromatin opening by pioneer factors is driven by motif content
-
--   De novo motif enrichment for ChIP peaks (supplemental)
-
--   ~~Motif analysis: percent with motif, average n motifs, motif quality~~
-
--   ~~Motif positioning around nucleosomes? (supplemental)~~
-
-To further test the specificity of binding within the three classes of ChIP-seq peaks, we analyzed the enrichment for the canonical Zld, Grh or Twi motifs.
-We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig. 3c).
-When compared to all accessible regions, class I regions had low levels of enrichment for the canonical motifs, supporting that binding to these regions is nonspecific.
-By contrast, class II and III sites were highly enriched for the canonical motifs (Fig. 3c).
-Class III sites had the highest percentage of sites containing a canonical motif.
-We hypothesized that differences in motif content could explain why binding to class II regions does not result in chromatin opening.
-We therefore analyzed the number of motifs per peak and motif strength (when scored against the canonical motif) for class I-III regions.
-For Grh and Twi, but not for Zld, class III sites had more motifs per peak than class II sites (Fig. 3f,h,j).
-For all three factors, class III sites tended to have stronger motifs than class II sites (Fig. 3g,i,k).
-Together, this analysis suggests that a combination of more motifs and stronger motifs at class III sites is what allows Zld, Grh, and Twi to open chromatin at these regions.
-<!--# Add description of de novo motif enrichment, put data in supplement. -->
-
-Previous work has shown that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (refs).
-<!--# Add a few sentences here with more details on past work --> To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
-Analysis of average MNase-seq signal at class I-III peaks indicated that class III sites have high levels of nucleosome occupancy that, on average, occur over the respective motif.
-Hierarchical clustering of MNase-seq data revealed that there was no preference for binding or chromatin opening at sites that had motifs in a particular orientation on nucleosomes (Extended Data Fig.
-XX).
-For class II and III peaks, some motifs were positioned close to the nuclesome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
-This analysis is complicated by the fact that MNase-seq is a bulk assay and may not accurately represent nucleosome positioning in individual cells, especially at regions with poorly positioned nucleosomes.
-To more directly assess the positioning of nucleosomes around motifs, we analyzed a previously published single-molecule dataset using methyltransferase footprinting to determine the locations of nucleosomes on individual chromatin fibers [@Stergachis2020].
-We analyzed the distance from motifs at class II and regions to the nearest nucleosome dyad.
-This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
-Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
-Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
-However, the lack of strong nucleosome positioning at these regions makes it difficult to fully assess the importance of motif positioning.
-
-## Ectopic Zld, Grh and Twi fail to access many endogenous binding sites
-
--   Z~~elda binding is more similar to the brain than the embryo~~
-
--   ~~Grh binding is stable across tissues~~
-
--   ~~Twi sites in S2 cells have low level binding in embryo, but strongest peaks from embryo are not bound~~
-
--   ~~Enrichment of repressive marks at sites bound in other tissues~~
-
--   ~~Results of Taz experiments~~
-
--   Comment on class VI sites that are not bound and lack repressive chromatin.
-
-Prior studies have extensively explored the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
-To better understand how cell-type specific variables affect pioneer-factor binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
-We compared Zld binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
-We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig. XX).
-Previous work has shown that regions resistant to pioneer-factor binding are enriched for the repressive histone modifications H3K9me3 and H3K27me3 [@Mayran2018a; @Soufi2012].
-To test if repressive marks were present at the resistant sites in S2 cells, we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
-Based on analysis of these datasets we divided the resistant sites into three additional classes of binding sites: class IV sites have high levels H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a).
-We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c).
-We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017], and to Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
-Despite the presence of class IV, V and VI binding sites for all three proteins, there were distinct patterns of binding across different tissues.
-Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
-For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig. 4b).
-Twi binding sites in S2 cells had low levels of binding in embryos, but many of the strongest embryonic binding sites were unbound.
-
-To test the functional relevance of repressive chromatin to pioneer-factor binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3.
-We treated our stable cell lines with DMSO or tazemetostat for 7 days, inducing Zld, Grh, or Twi expression on day 5.
-We confirmed loss of H3K27me3 by westerm blot and CUT&RUN [@Skene2017] (Extended Data Fig. XX).
-We used ChIP-seq to test if tazemetostat treatment resulted in changes to Zld, Grh, or Twi binding.
-Tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig. 4d-f).
-Despite a dramatic loss of H3K27me3, the majority of class IV binding sites remained unbound in tazemetostat-treated cells.
-We did identify some individual sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig. XX).
-Together, these data demonstrate that despite the correlation of H3K27me3 with resistance to pioneer factor binding, loss of PRC2 activity is insufficient to promote widespread binding of Zld, Grh, or Twi to repressed binding sites.
-
-## Pioneer-factor binding and chromatin opening is concentration dependent
-
--   ~~For all three factors, increased concentration results in increased binding and opening of class II and III sites.~~
-
--   ~~For all three factors, increased concentration leads to novel binding sites.~~
-
--   ~~For Zld and Twi, increased concentration is insufficient to allow binding to repressed sites.~~
-
--   ~~For Grh, increased concentration does allow binding and opening at repressed sites~~
-
-Previous studies have proposed that pioneer factor activity may depend on protein concentration [@Hansen2022; @Blassberg2022].
-Considering that tazemetostat treatment had limited effects on pioneer factor binding, we asked if varying the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.
-We induced cells to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig. XX; see methods).
-We performed ChIP-seq and analyzed the data at our previously defined class I-VI binding sites.
-We found that binding and was concentration dependent (Fig. 5).
-At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i).
-Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i).
-We also performed ATAC-seq to determine if concentration-dependent increases in binding led to changes in chromatin accessibility.
-Changes in chromatin accessibility were also concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig. 5b,f,j).
-Intriguingly, many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels.
-For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites (Fig. 5, Extended Data Fig. XX).
-To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration.
-For Zld and Twi, supraphysiological concentrations led to an increase in class III regions relative to class II regions (Fig. 5c,k).
-For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with the majority of these sites being falling in III (Fig. 5g).
-To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs.
-supraphysiological concentrations (Fig.
-5d,h,i).
-For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration.
-Together, these data demonstrate that both binding and chromatin opening by pioneer factors are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are restistant to binding at physiological levls.
-
-## Regions outside the DNA-binding domain are required for binding and opening of closed chromatin
-
--   ~~Generation of stable cell lines to express DBDs at same concentration as FL proteins~~
-
--   ~~Zld DBD is totally unable to bind to closed sites.~~
-
--   ~~In supplement, show Zld DBD binding specifically at class II and III sites with CAGGTAG to support that this is not due to a loss of ZnF2~~
-
--   ~~Grh DBD has reduced ability to access closed chromatin~~
-
--   ~~Neither DBD is sufficient for chromatin opening.~~
-
-Prior work has shown that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs).
-However, little is known about how regions outside the DBD may contribute to pioneering function.
-<!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat.
-Genet.
-paper) is outside the DBD.
-See also new Zaret paper in NSMB looking at contribution of IDRs and activation domains to chromatin opening on nucleosome arrays in vitro -->.
-To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig.
-XX).
-Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was totally unable to bind to class II or III regions (Fig.
-6a,c).
-We had previously shown that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
-Because many Zld binding sites in S2 cells also lack a canonical CAGGTA motif (Fig.
-3, Extended Data Fig XX), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
-To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
-This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required.
-Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
-For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
-Together, these experiments show that regions outside the DBD are essential for binding and opening of closed chromatin.
+{{< include manuscript_sections/_03_Results.qmd >}}
 
 # Discussion
+{{< include manuscript_sections/_04_Discussion.qmd >}}
 
 # Materials and Methods
-
-## Reagents
-
-| Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|--------------------|------------------|------------------|------------------|
-| pMT-puro                              | AddGene                   | 17923                |                         |
-| Gibson Cloning                        | New England Biolabs       | E2611                |                         |
-| Schneider's Medium                    | Thermo Fisher Scientific  | 21720001             |                         |
-| Fetal Bovine Serum                    | Omega Scientific          | FB-02                |                         |
-| Antibiotic/antimycotic                | Thermo Fisher Scientific  | 15240062             |                         |
-| Effectene Transfection Reagent        | Qiagen                    | 301427               |                         |
-| PVDF Membrane                         |                           |                      |                         |
-| Formaldehyde                          |                           |                      |                         |
-| Protease inhibitor tablets            |                           |                      |                         |
-| NEBNext Ultra II DNA library prep kit |                           |                      |                         |
-| TruSeq v2 RNA library prep kit        |                           |                      |                         |
-| Trizol                                |                           |                      |                         |
-| **Antibodies**                        |                           |                      |                         |
-| Anti-Zld                              | Harrison et al., 2010     |                      | ChIP: 6 µL , WB: 1:750  |
-| Anti-Grh                              | Harrison et al., 2010     |                      | ChIP: 8 µL, WB: 1:1000  |
-| Anti-Twi                              | He et al., 2015           |                      | ChIP: 10 µg, WB: 1:1000 |
-| Anti-HA-peroxidase clone 3F10         | Roche                     | 12013819001          | WB: 1:500               |
-| Anti-Tubulin DM1A                     | Sigma                     | T6199-200UL          | WB: 1:5000              |
-| goat anti-rabbit IgG-HRP conjugate    | Bio-Rad                   | 1706515              | WB: 1:3000              |
-| Anti-HA clone 12CA5                   | Roche                     |                      | ChIP: 7.5 µL            |
-| Anti-H3K27me3                         | Cell Signaling Technology | C36B11; Lot #19      | WB: 1:1000, CR: 0.5 µL  |
-|                                       |                           |                      |                         |
-|                                       |                           |                      |                         |
-
-##
-
-## Cell culture and generation of stable cell lines
-
-*zld*, *grh*, or *twi* cDNA (isoforms RB, RH, and RA, respectively) were cloned into pMT-puro plasmid (Addgene) via Gibson cloning (New England Biolabs).
-For cloning of the Zld and Grh DNA-binding domains, DNA encoding amino acids 1114-1487 (Zld) or amino acids 603-1032 (Grh) was cloned into pMT-puro.
-
-Drosophila S2 cells were cultured at 27°C in Schneider's medium (Thermo Fisher Scientific) supplemented with 10% fetal bovine serum (Omega Scientific) and antibiotic/antimycotic (Thermo Fisher Scientific).
-For the generation of stable cell lines, cells were plated at 5✕10^5^ cells per mL.
-After 24 hours, cells were transfected with 10 µg plasmid DNA using Effectene transfection reagent (Qiagen).
-After an additional 24 hours, puromycin was added to a final concentration of 2 µg/mL.
-Stable cell lines were recovered after 2-3 weeks of selection.
-Following recovery of stable cell lines, cells were cultured with 1 µg/mL puromycin.
-
-## Induction of transcription factor expression
-
-Transcription factor expression was induced by adding CuSO~4~ to the cell culture media.
-Expression of protein at physiological levels was achieved with 1000, 100, or 40 µM CuSO~4~ for Zld, Grh or Twi, respectively (Supplemental figure X).
-For experiments with sub- or supraphysiological protein levels (Fig.
-5), the following CuSO~4~ concentrations were used: 500 µM (Zld subphysiological), 1000 µM (Zld physiological) or 1500 µM (Zld supraphysiological); 25 µM (Grh subphysiological), 100 µM (Grh physiological) or 400 µM (Grh supraphysiological); 10 µM (Twi subphysiological), 40 µM (Twi physiological) or 160 µM (Twi supraphysiological).
-For all induction experiments, cells were plated at 1×10^6^ cells per mL.
-Unless otherwise specified, experiments were performed with cells harvested 48 hours after induction.
-
-## Immunoblotting
-
-Proteins were separated on denaturing polyacramide gels before transfer to a 0.45 μm Immobilon-P PVDF membrane (Millipore) in transfer buffer (25 mM Tris, 200 mM Glycine, 20% methanol) for 60 min (75 min for Zld) at 500mA at 4°C.
-The membranes were blocked with BLOTTO (2.5% non-fat dry milk, 0.5% BSA, 0.5% NP-40, in TBST) for 30 min at room temperature and then incubated with anti-Zld (1:750) (Harrison et al., 2010), anti-Grh (1:1000) (Harrison et al., 2010), anti-Twi (1:1000) (He et al.
-2015), anti-HA-peroxidase (1:500) (clone 3F10, Roche), or anti-Tubulin (DM1A, 1:5000) (Sigma), overnight at 4°C.
-The secondary incubation was performed with goat anti-rabbit IgG-HRP conjugate (1:3000) (Bio-Rad) for 1 hr at room temperature.
-Blots were treated with SuperSignal West Pico PLUS chemiluminescent substrate (Thermo Fisher Scientific) and visualized using the Azure Biosystems c600 or Kodak/Carestream BioMax Film (VWR).
-
-## Immunostaining
-
-## ChIP-seq
-
-For each ChIP experiment, \~50 million cells were crosslinked by adding methanol-free formaldehyde directly to the cell culture medium to a final concentration of 0.8%.
-Cells were rotated on a nutator for 7 minutes at room temperature (RT) before quenching of crosslinking by adding glycine to 125 mM.
-Crosslinked cells were pelleted by centrifugation at 600 × *g* for 3 minutes.
-Cells were washed twice with 1X PBS.
-Cells were lysed by resuspending in lysis buffer (50 mM HEPES pH 7.9, 140 mM NaCl, 1 mM EDTA, 10% glycerol, 0.5% NP40, 0.25% Triton-X 100) with protease inhibitors (Pierce) and incubating on ice for 10 minutes.
-The resulting chromatin was pelleted for 5 minutes at 1500 × *g* and resuspended in RIPA buffer.
-Chromatin was sonicated in a Covaris S220 Ultrasonicator (5 cycles of 120 seconds with 60 second delay, 170 peak power, 10% duty factor, 200 cycles/burst).
-Sonicated chromatin was centrifuged for 10 minutes at 10,000 × *g* to pellet insoluble material.
-5% of the supernatant was set aside as input, and the remainder was incubated at 4°C overnight with antibodies (6 µL anti-Zld, 8 µL anti-Grh, 10 µg anti-Twi, 7.5 µL anti-HA).
-20 µL protein A beads (Dynabeads Protein A, ThermoFisher Scientific) blocked with BSA were added and samples were incubated at 4°C for 4 hours.
-Beads were separated on a magnet and washed 3× with low salt wash buffer (10 mM Tris pH 7.6, 1 mM EDTA, 0.1% SDS, 0.1% Na-Deoxycholate, 1% Triton-X 100, 150 mM NaCl), 2× high salt wash buffer (10 mM Tris pH 7.6, 1 mM EDTA, 0.1% SDS, 0.1% sodium deoxycholate, 1% Triton-X 100, 300 mM NaC), 2× LiCl wash buffer (0.25 M LiCl, 0.5% NP40, 0.5% sodium deoxycholate), and 1× with TE buffer with NaCl (10 mM Tris-HCl pH 8.0, 1 mM EDTA, 50 mM NaCl).
-Beads were resuspended in elution buffer (50 mM Tris-HCl pH 8.0, 1% SDS, 10 mM EDTA) and incubated at 65°C for 10 minutes.
-The IP and input samples were treated with 4.5 µL RNAse A for 30 minutes.
-5 µL Proteinase K was added and samples were incubated overnight at 65°C to reverse crosslinking.
-DNA was isolated by phenol:chloroform extraction, precipitated with EtOH, and resuspended in 20 µL H~2~O.Preparation of sequencing libraries was performed using NEBNext Ultra II kit (NEB) with 7 PCR cycles for library amplification.
-Sequencing was performed on the Illumina Hi-Seq4000 using 50bp single-end reads, or on the Illumina NovaSeq 6000 using 150bp paired-end reads.
-
-## ChIP-seq analysis
-
-Bowtie2 version 2.4.4 [@Langmead2012] was used to align ChIP-seq reads to the *Drosophila melanogaster* genome (version dm6) with the following non-default parameters: -k 2 \--very-sensitive \--no-mixed \--no-discordant -X 5000.
-Aligned reads were filtered to include only reads with a mapping quality score \> 30.
-Reads mapping to unplaced scaffolds or the mitochondrial genome were discarded.
-Peak calling was performed using MACS2 version 2.2 with parameters -g dm \--call-summits.
-Only peaks that were detected in all replicates were considered in downstream analysis.
-bigWig files were generated using deepTools bamCoverage version 3.5.1 with parameters --binSize 10.
-bigWig files were z-score normalized by subtracting the genome-wide average of all bins from each bin value and dividing by the standard deviation of all bins.
-
-## RNA-seq
-
-5×10^6^ cells were harvested and resuspended in 800 µL Trizol (Invitrogen).
-RNA was purified using a chloroform extraction followed by a column-based Quick-RNA MiniPrep extraction (Zymo).
-Poly-A-selected RNA sequencing (RNA-seq) libraries were prepared using the TruSeq RNA sample prep kit v2 (Illumina).
-
-## RNA-seq analysis
-
-Raw reads were aligned to the Drosophila melanogaster (dm6) genome using HISAT2 version 2.1.0 with parameters -k 2 --very-sensitive.
-Reads with a mapping quality score \< 30 and reads aligning to the mitochondrial genome or unplaced scaffolds were discarded.
-FeatureCounts (from Subread version 2.0) was used to assign aligned reads to genes (FlyBase gene annotation release r6.45).
-The resultant table of read counts was used to perform differential expression analysis using DESeq2 [@love2014].
-Genes with an adjusted p-value \< 0.05 and a fold change \> 2 were considered differentially expressed.
-
-## ATAC-seq
-
-ATAC-seq was performed as described previsouly [@Buenrostro2013].
-2×10^5^ cells were harvested for ATAC-seq by centrifugation at 600 × *g* for 3 minutes at RT.
-Cells were washed once in 100 µL 1X PBS and resuspended in 100 µL ATAC lysis buffer (10 mM Tris pH 7.5, 10 mM NaCl, 3 mM MgCl~2~,0.1% NP-40).
-Cells were centrifuged at 600 × *g* for 10 minutes at 4°C.
-Supernatant was removed and pellet was resuspended in 47.5 µL buffer TD (Illumina).
-2.5 uL of Tagment DNA Enzyme (Illumina) was added and samples were incubated at 37°C for 30 minutes.
-Tagmented DNA was purified using MinElute Reaction Cleanup Kit (Qiagen) and eluted in 10 µL of buffer EB.
-DNA was PCR amplified for 12 cycles with the following conditions:72°C for 5 minutes, 98°C for 30 seconds, then 12 cycles of 98°C for 10 seconds, 63°C for 30 seconds and 72°C for 1 minute.
-Amplified libraries were purified using a 1.2X ratio of Axygen paramagnetic beads.
-Sequencing was performed on the Illumina NovaSeq 6000 using 150bp paired-end reads.
-
-## ATAC-seq analysis
-
-Raw ATAC-seq reads were trimmed to remove adapter sequences using NGmerge version 0.3 [@Gaspar2018].
-Trimmed reads were aligned to the *Drosophila melanogaster* genome (version dm6) using bowtie2 with paramters -k 2 \--very-sensitive \--no-mixed \--no-discordant -X 5000.
-Reads with a mapping quality \< 30 and reads aligning to the mitochondrial genome or scaffolds were discarded.
-As described previously, only fragments \< 100bp were considered for downstream analysis [@Buenrostro2013].
-Reads were combined across all replicates and peak calling was performed using MACS2 with parameters -f BAMPE \--keep-dup all -g dm \--call-summits.
-featureCounts was used to count the number of reads aligning within 200bp (100bp upstream or downstream) of peak summits.
-Differential accessibility analysis was performed using DESeq2.
-Regions with an adjusted p-value \< 0.05 were considered differentially accessible.
-
-## Treatment of S2 cells with tazemetostat
-
-Cells were plated at 1×10^6^ cells per mL and DMSO or tazemetostat was added to a final concentration of 10 µM.
-Cells were split on days 2 and 5 and tazemetostat was added to maintain a final concentration of 10 µM.
-On day 5, CuSO~4~ was added to achieve physiological expression of Zld, Grh or Twi (see above).
-On day 7, cells were harvested for immunoblotting, ChIP and CUT&RUN.
-
-## CUT&RUN
-
-CUT&RUN was performed using EpiCypher reagents according to the manufacturer protocol.
-2×10^5^ cells were used for each CUT&RUN reaction.
-Overnight incubation with antibodies was performed in antibody buffer containing 0.05% digitonin.
-0.5 µL anti-H3K27me3 (Cell signaling technologies) or rabbit IgG were used for CUT&RUN.
-Prior to addition of antibody, a 1:5 dilution of K-MetStat Panel spike-in nucleosomes (EpiCypher, cat. \# 19-1002) was added to each reaction.
-Libraries were prepared using the NEBNext Ultra II kit (NEB).
-During library preparation, cleanup steps were performed using a 1.1X ratio of Axygen paramagnetic beads, as recommended by EpiCypher.
-PCR amplification was performed with the following conditions: 98°C for 45 seconds, then 14 cycles of 98°C for 15 seconds, 60°C for 10 seconds, and a final extension of 72°C for 1 minute.
-Sequencing was performed on the Illumina NovaSeq 6000 with 150bp paired-end reads.
-
-## CUT&RUN analysis
-
-Trimming
-
-Alignment
-
-Filtering
-
-peak calling
-
-bigwigs
-
-z-score normalization
-
-spike-in normalization
-
-## Analysis of previously published data
-
-Previously published datasets analyzed in this study can be found in supplemental table XXXX.
-All previously published data were analyzed in parallel with data from this study using analysis parameters described above.
-
-## Data and code availability
-
-Sequencing data have been deposited in the Gene Expresion Omnibus under accession GSEXXXX.
-All analysis code is available on GitHub: https://github.com/tjgibson/S2_pioneers_manuscript.
-An archived version of the data processing and analysis pipeline is available on Zenodo: doiXXXX.
+{{< include manuscript_sections/_05_Discussion.qmd >}}
 
 # Figure Legends
-
-**Figure 1: Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
-ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
-**c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
-**d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
-**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h).
-i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
-**j,i,** Proportion of class I, II or III regions that are proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
-Promoter-proximal and promoter-distal binding sites are shown separately.
-The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene.
-
-**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.
-a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites.
-**c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions.
-**d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.
-
-**Figure 3: Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
-**b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
-Zld, Grh, and Twi ChIP-seq data were generated in this study.
-Remaining ChIP-seq datasets were previously published (see Supplementary table XX).
-**c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
-Logos plot of the canonical motif are shown alongside the heatmap.
-Percentages are shown for class I, II or III binding sites, or all WT ATAC-seq peaks as a control for the background motif frequency.
-**f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
-**g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**.
-Motif scores were determined by comparing each identified motif instance to the canonical motif.
-
-**Figure 4: Many endogenous binding sites are resistant to ectopic pioneer-factor binding.
-a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
-ChIP-seq data for repressive histone modifications H3K27me3 and H3K9me3 is also shown.
-**d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
-Zld, Grh, and Twi ChIP-seq data are z-score normalized.
-CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).
-
-**Figure 5: Binding and opening of closed chromatin is concentration dependent.
-a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
-Concetrations of CuSO~4~ used to induce protein expression are indicated above each plot.
-**b,f,j,** Metaplots of ATAC-seq signal at class I-VI regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
-**c,g,k,** Distribution of class I, II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
-Y-axis indicates the total number of peaks called at each concentration.
-Colors indicate the proportion of those peaks that are defined as class I, II, or III.
-**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
-
-**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function.
-a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
-Shown for Zld **(a)** and Grh **(b)**.
-**c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**.
-**d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
-
-**Figure 7: Pioneer-factor function requires stable chromatin occupancy.**
-
-**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.**
-
--   Schematic of cell line generation.
-
--   Western blots to determine physiological expression and lack of protein expression in WT cells.
-
--   RNA-seq data to show leaky expression and strong induction of RNA levels.
-
-**Extended Data Figure 2: Zld and Grh bind to chromatin rapidly after induction of protein expression**
-
--   Western blot time course of protein levels
-
--   CUT&RUN time course to show rapid binding
-
--   likely rules out dependence on DNA replication
-
-**Extended Data Figure 3: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.**
-
--   volcano and/or MA plots with enrichment of binding
-
--   Analysis of nonspecific changes?
-
--   RNA-seq PCA
-
--   RNA-seq volcano/MA plots
-
--   RNA-seq GO analysis
-
-**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation**
-
--   Twi volcano and examples
-
--   TWI RNA-seq volcano and examples
-
--   Twi RNA-seq GO analysis
-
-**Extended Data Figure 5: Chromatin features associated with Zld, Grh and Twi binding sites**
-
--   analysis of published ChIP datasets not shown in main figure
-
--   comparison of class I-III sites to show that class II-III sites are devoid of most histone mods
-
--   Class III sites have higher levels of H1 and H3 than class II sites.
-
-**Extended Data Figure 6: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes**
-
--   De novo motif enrichment for each factor
-
--   MNase data to show enrichment of nucleosomes around target motifs
-
--   Clustering of MNase data to show that there is no preferred nucleosome position relative to motifs
-
--   additional analysis of Fiber-seq data to show that nucleosomes at these regions are poorly positioned
-
-**Additional information to include in supplemental figures:**
-
--   De novo motif results
-
--   Analysis of promoter motifs
-
--   New binding sites at higher expression levels
-
--   Zld DBD ChIP only at CAGGTAG sites

### 2023-01-06 manuscript/manuscript_sections/_03_Results.qmd
break manuscript up into multiple qmd files

```{r}
#| include: false
library(here)
library(tidyverse)
```

## Ectopically expressed pioneer factors bind and open closed chromatin

-   ~~Establish system for ectopic expression of pioneer factors~~

-   ~~Identification of physiological expression levels~~

-   ~~Define binding classes I-III and describe characteristics (distribution, promoters/enhancers, etc.)~~

-   Time course and dynamics of binding to different classes (supplement)

To investigate the mechanisms of chromatin binding and opening by pioneer factors, we generated stable cell lines inducibly expressing Zelda (Zld) and Grainy head (Grh) in Drosophila S2 cells (Extended Data Fig. 1a), where neither factor is endogenously expressed.
We confirmed that Zld and Grh RNA (Extended Data Fig. 1b) and protein (Extended Data Fig. 1c,d) were not present in WT S2 cells.
The stable cell lines express Zld or Grh under the control of the metallothionein promoter, which allows inducible and tunable expression with the addition of CuSO~4.~ To determine protein levels that would be approximately physiological, we compared Zld or Grh levels in cells with varying concentrations of CuSO~4~ to the levels of endogenous expression found in Drosophila embryos (Extended Data Fig. 1c,d).
Physiological protein levels were achieved at 1000 mM or 100 mM CuSO~4~ for Zld or Grh, respectively.

```{r}
#| include: false
# zld_ChIP_classes <- snakemake@input[["zld_ChIP_classes"]]
# grh_ChIP_classes <- snakemake@input[["grh_ChIP_classes"]]

zld_ChIP_classes <- here("results/ChIP_peak_classes/zld_ChIP_classes.tsv") |> 
  read_tsv()
grh_ChIP_classes <- here("results/ChIP_peak_classes/grh_ChIP_classes.tsv") |> 
  read_tsv()

zld_n_ChIP_peaks <- nrow(zld_ChIP_classes)
grh_n_ChIP_peaks <- nrow(grh_ChIP_classes)
```

To determine how exogenous Zld and Grh interact with preexisting chromatin in S2 cells, we expressed each factor at physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
To control for antibody specificity, we performed additional ChIP experiments in WT cells using anti-Zld, or anti-Grh.
We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh.
<!--# Not sure the best way to discuss the control ChIP experiments. Should I mention how many peaks we detect in WT cells? What supplemental figures should I include? Should peak calling be done using IgG or WT cells as control instead of input? -->

```{r}
#| include: false

zld_ATAC_results <- here("ATACseq/results/DEseq2_results_filtered/S2-Zld_ATACseq_S2-Zld-FL-vs-S2-WT_results.tsv") |> 
  read_tsv() |> 
  mutate(direction_of_change = case_when(
    is_diff & log2FoldChange > 0 ~ "increased",
    is_diff & log2FoldChange < 0 ~ "decreased",
    !is_diff ~ "ns"
  ))

grh_ATAC_results <- here("ATACseq/results/DEseq2_results_filtered/S2-Grh_ATACseq_S2-Grh-FL-vs-S2-WT_results.tsv") |> 
  read_tsv() |> 
  mutate(direction_of_change = case_when(
    is_diff & log2FoldChange > 0 ~ "increased",
    is_diff & log2FoldChange < 0 ~ "decreased",
    !is_diff ~ "ns"
  ))

zld_diff <- zld_ATAC_results |> 
  pull(is_diff) |> 
  sum()
zld_up <- (zld_ATAC_results$direction_of_change == "increased") |> 
  sum()
zld_down <- (zld_ATAC_results$direction_of_change == "decreased") |> 
  sum()

grh_diff <- grh_ATAC_results |> 
  pull(is_diff) |> 
  sum()
grh_up <- (grh_ATAC_results$direction_of_change == "increased") |> 
  sum()
grh_down <- (grh_ATAC_results$direction_of_change == "decreased") |> 
  sum()

```

Having established that Zld and Grh can bind to thousands of sites in S2 cells, we performed ATAC-seq on WT cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
We performed differential accessibility analysis comparing Zld or Grh-expressing cells to WT cells.
We detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig. XX).
<!--# Add more information here about sites with differential accessibility. Include percentage of increased and decreased sites that are bound by Zld or Grh. Comment on how much is specific vs. secondary effect, and justify defining ChIP peak classes that include only a subset of DA sites. -->

To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of binding sites (Fig. 1a,b).
Class I sites are accessible in WT cells and remain accessible when bound by ectopically-expressed Zld or Grh.
Class II sites are inaccessible in WT cells, however binding of Zld or Grh does not lead to chromatin opening.
Class III sites are inaccessible in WT cells and Zld or Grh binding results in chromatin opening.
The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig. 1c,f).
While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig. 1e,h).
Together, these data show that when expressed exogenously at physiolgical levels, Zld and Grh can bind and open closed chromatin at a subset of binding sites.

<!--# Could potentially add the cut&run time course here? Think about where to put this, or if it is worth including at all. -->

## Chromatin opening correlates with transcriptional activation

-   ~~Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.~~

-   ~~Comparison of RNA-seq and ATAC data: chromatin opening correlates with transcriptional activation.~~

-   ~~Class III promoters are most likely to be associated with upregulated genes~~

```{r}
#| include: false

zld_RNAseq_results <- here("RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results.tsv") |> 
  read_tsv() |> 
  mutate(direction_of_change = case_when(
    is_diff & log2FoldChange > 0 ~ "increased",
    is_diff & log2FoldChange < 0 ~ "decreased",
    !is_diff ~ "ns"
  ))

grh_RNAseq_results <- here("RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results.tsv") |> 
  read_tsv() |> 
  mutate(direction_of_change = case_when(
    is_diff & log2FoldChange > 0 ~ "increased",
    is_diff & log2FoldChange < 0 ~ "decreased",
    !is_diff ~ "ns"
  ))

zld_diff <- zld_RNAseq_results |> 
  pull(is_diff) |> 
  sum()
zld_up <- (zld_RNAseq_results$direction_of_change == "increased") |> 
  sum()
zld_down <- (zld_RNAseq_results$direction_of_change == "decreased") |> 
  sum()

grh_diff <- grh_RNAseq_results |> 
  pull(is_diff) |> 
  sum()
grh_up <- (grh_RNAseq_results$direction_of_change == "increased") |> 
  sum()
grh_down <- (grh_RNAseq_results$direction_of_change == "decreased") |> 
  sum()

```

Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation.
To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression.
Differential expression analysis using DESeq2 [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. XX).
Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig. XX <!--# include statistics  -->), suggesting that binding of these factors mainly leads to transcriptional activation and not repression.
<!--# Consider mentioning Grh as a repressor here and giving examples of genes where Grh may be repressing. Discuss if Grh-mediated transcriptional repression coincides with chromatin opening, chromatin closing, or neither --> Gene ontology (GO) analysis indicated that the genes activated by each factor are consistent with the endogenous functions of Zld and Grh (Extended Data Fig. XX).
<!--# Add more detailed information on GO analysis here -->

To better understand the relationship between chromatin opening and transcriptional activation, we directly compared our RNA-seq and ATAC-seq data.
Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l).
sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig. XX <!--# add statistics -->).
When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
This was particularly evident at class III promoters, which were most likely to correspond to upregulated genes.
Together, we conclude that while binding and chromatin opening can happen in the absence of transcriptional changes, chromatin opening is correlated with transcriptional activation.

## Twist binds closed chromatin extensively and opens chromatin at a subset of sites

-   ~~Introduce Twist and justify the comparison to Zld/Grh.~~

-   ~~Twi binds closed chromatin and opens at some sites.~~

-   Mention analysis of published Abd-A and Ubx cell lines.

-   ~~Class I-III sites, distribution, etc.~~

-   ~~Twi also promotes transcriptional activation~~

-   ~~Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.~~

-   ~~Comparison of RNA-seq and ATAC~~

-   ~~Mention differences between Twi vs. Zld and Grh~~

Despite the differences between Zld and Grh during Drosophila development, both proteins behaved similarly when expressed in our exogenous S2 cell system.
To provide additional context to our findings, we performed additional experiments with the transcription factor Twist (Twi).
Although Twi is also essential for regulating Drosophila embryonic development, Twi has a number if distnict properties from Zld and Grh, making it an interesting point of comparison.
Twi is a master regulator of mesodermal cell fate (REFS).
Twi has many binding sites that are dynamic during early embryonic development [@Wilczynski2010a], in contrast to Zld and Grh.
Although Twi was hypothesized to be a pioneer factor [@Sandmann2007], pioneering activity by Twi had not previously been tested.
In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012].
To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. XX).
We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
Comparison of Twi ChIP-seq and ATAC-seq data allowed for the identification of the same classes of binding sites that were found for Zld and Grh (Fig 2a,b).
Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c).
Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. XX).
Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig. XX).
GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. XX).
Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig. XX).
Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig. XX).
This was in contrast to Zld and Grh, where induction results in the activation of many genes that are not directly bound by Zld or Grh.
This suggests that Zld and Grh may activate many transcription factors, which go on to activate their own targets, leading to secondary effects in the RNA-seq data.
Twi, by contrast, appears to directly activate genes required for mesodermal and muscle cell fate.
<!--# Is this worth mentioning? Might require additional analysis of how many TFs are upregulated in RNA-seq data for each factor. Could also require additional motif analysis to identify/support secondary TFs and reason for nonspecific effects. --> Similar to Zld and Grh, transcriptional activation was correlated with chromatin opening (Extended Data Fig. XX).
Together, our data show that Twi can act as a pioneer factor.

## Ectopically expressed transcription factors bind opportunistically to active chromatin

-   ~~Overlap of class I regions suggests nonspecific binding~~

-   ~~Analysis of published ChIP data shows that class I regions are within active chromatin~~

-   ~~RNA-seq and ATAC-seq for class I to show that the binding is not functional~~

-   ~~Analysis of chromatin at class II-III regions: mostly devoid of histone modifications~~

-   ~~Class III regions have higher H3 and H1 occupancy compared to class II~~

-   Where to put analysis of nucleosome positioning?

While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites are shared across Zld, Grh, and Twi (Fig. 3a).
This was in contrast to class II and III peaks, which were largely specific to each individual factor (Extended Data Fig. XX).
Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig. XX).
We hypothesized that these overlapping class I binding sites may occur at regulatory regions that are active in WT cells.
To test this, we analyzed published ChIP-seq data from WT S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV.
Overlapping class I regions had high levels of chromatin marks associated with active regulatory regions (Fig. 3b).
This enrichment of all three factors at highly active regions suggests that exogenously expressed factors may bind nonspecifically to active regions of the genome.
This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig. XX).

Considering the enrichment of active chromatin at class I sites, we asked if there were any features of chromatin distinctive to class II or III regions.
Analysis of published ChIP-seq data demonstrated that class II and III sites have low levels of many histone modifications, chromatin-associated proteins, and transcription factors (Extended Data Fig. XX).
These data are in agreement with previous observations that Oct4, Sox2 and Klf4 access binding sites in naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. XX).
Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig. XX).
Together this suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin.
Based on these observations, the inability to open class II regions is unlikely to be due to differences in preexisting features of chromatin.

## Chromatin opening by pioneer factors is driven by motif content

-   De novo motif enrichment for ChIP peaks (supplemental)

-   ~~Motif analysis: percent with motif, average n motifs, motif quality~~

-   ~~Motif positioning around nucleosomes? (supplemental)~~

To further test the specificity of binding within the three classes of ChIP-seq peaks, we analyzed the enrichment for the canonical Zld, Grh or Twi motifs.
We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig. 3c).
When compared to all accessible regions, class I regions had low levels of enrichment for the canonical motifs, supporting that binding to these regions is nonspecific.
By contrast, class II and III sites were highly enriched for the canonical motifs (Fig. 3c).
Class III sites had the highest percentage of sites containing a canonical motif.
We hypothesized that differences in motif content could explain why binding to class II regions does not result in chromatin opening.
We therefore analyzed the number of motifs per peak and motif strength (when scored against the canonical motif) for class I-III regions.
For Grh and Twi, but not for Zld, class III sites had more motifs per peak than class II sites (Fig. 3f,h,j).
For all three factors, class III sites tended to have stronger motifs than class II sites (Fig. 3g,i,k).
Together, this analysis suggests that a combination of more motifs and stronger motifs at class III sites is what allows Zld, Grh, and Twi to open chromatin at these regions.
<!--# Add description of de novo motif enrichment, put data in supplement. -->

Previous work has shown that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (refs).
<!--# Add a few sentences here with more details on past work --> To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
Analysis of average MNase-seq signal at class I-III peaks indicated that class III sites have high levels of nucleosome occupancy that, on average, occur over the respective motif.
Hierarchical clustering of MNase-seq data revealed that there was no preference for binding or chromatin opening at sites that had motifs in a particular orientation on nucleosomes (Extended Data Fig. XX).
For class II and III peaks, some motifs were positioned close to the nuclesome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
This analysis is complicated by the fact that MNase-seq is a bulk assay and may not accurately represent nucleosome positioning in individual cells, especially at regions with poorly positioned nucleosomes.
To more directly assess the positioning of nucleosomes around motifs, we analyzed a previously published single-molecule dataset using methyltransferase footprinting to determine the locations of nucleosomes on individual chromatin fibers [@Stergachis2020].
We analyzed the distance from motifs at class II and regions to the nearest nucleosome dyad.
This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
However, the lack of strong nucleosome positioning at these regions makes it difficult to fully assess the importance of motif positioning.

## Ectopic Zld, Grh and Twi fail to access many endogenous binding sites

-   Z~~elda binding is more similar to the brain than the embryo~~

-   ~~Grh binding is stable across tissues~~

-   ~~Twi sites in S2 cells have low level binding in embryo, but strongest peaks from embryo are not bound~~

-   ~~Enrichment of repressive marks at sites bound in other tissues~~

-   ~~Results of Taz experiments~~

-   Comment on class VI sites that are not bound and lack repressive chromatin.

Prior studies have extensively explored the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
To better understand how cell-type specific variables affect pioneer-factor binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
We compared Zld binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig. XX).
Previous work has shown that regions resistant to pioneer-factor binding are enriched for the repressive histone modifications H3K9me3 and H3K27me3 [@Mayran2018a; @Soufi2012].
To test if repressive marks were present at the resistant sites in S2 cells, we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
Based on analysis of these datasets we divided the resistant sites into three additional classes of binding sites: class IV sites have high levels H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a).
We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c).
We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017], and to Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
Despite the presence of class IV, V and VI binding sites for all three proteins, there were distinct patterns of binding across different tissues.
Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig. 4b).
Twi binding sites in S2 cells had low levels of binding in embryos, but many of the strongest embryonic binding sites were unbound.

To test the functional relevance of repressive chromatin to pioneer-factor binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3.
We treated our stable cell lines with DMSO or tazemetostat for 7 days, inducing Zld, Grh, or Twi expression on day 5.
We confirmed loss of H3K27me3 by westerm blot and CUT&RUN [@Skene2017] (Extended Data Fig. XX).
We used ChIP-seq to test if tazemetostat treatment resulted in changes to Zld, Grh, or Twi binding.
Tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig. 4d-f).
Despite a dramatic loss of H3K27me3, the majority of class IV binding sites remained unbound in tazemetostat-treated cells.
We did identify some individual sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig. XX).
Together, these data demonstrate that despite the correlation of H3K27me3 with resistance to pioneer factor binding, loss of PRC2 activity is insufficient to promote widespread binding of Zld, Grh, or Twi to repressed binding sites.

## Pioneer-factor binding and chromatin opening is concentration dependent

-   ~~For all three factors, increased concentration results in increased binding and opening of class II and III sites.~~

-   ~~For all three factors, increased concentration leads to novel binding sites.~~

-   ~~For Zld and Twi, increased concentration is insufficient to allow binding to repressed sites.~~

-   ~~For Grh, increased concentration does allow binding and opening at repressed sites~~

Previous studies have proposed that pioneer factor activity may depend on protein concentration [@Hansen2022; @Blassberg2022].
Considering that tazemetostat treatment had limited effects on pioneer factor binding, we asked if varying the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.
We induced cells to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig. XX; see methods).
We performed ChIP-seq and analyzed the data at our previously defined class I-VI binding sites.
We found that binding and was concentration dependent (Fig. 5).
At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i).
Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i).
We also performed ATAC-seq to determine if concentration-dependent increases in binding led to changes in chromatin accessibility.
Changes in chromatin accessibility were also concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig. 5b,f,j).
Intriguingly, many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels.
For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites (Fig. 5, Extended Data Fig. XX).
To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration.
For Zld and Twi, supraphysiological concentrations led to an increase in class III regions relative to class II regions (Fig. 5c,k).
For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with the majority of these sites being falling in III (Fig. 5g).
To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs. supraphysiological concentrations (Fig. 5d,h,i).
For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration.
Together, these data demonstrate that both binding and chromatin opening by pioneer factors are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are restistant to binding at physiological levls.

## Regions outside the DNA-binding domain are required for binding and opening of closed chromatin

-   ~~Generation of stable cell lines to express DBDs at same concentration as FL proteins~~

-   ~~Zld DBD is totally unable to bind to closed sites.~~

-   ~~In supplement, show Zld DBD binding specifically at class II and III sites with CAGGTAG to support that this is not due to a loss of ZnF2~~

-   ~~Grh DBD has reduced ability to access closed chromatin~~

-   ~~Neither DBD is sufficient for chromatin opening.~~

Prior work has shown that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs).
However, little is known about how regions outside the DBD may contribute to pioneering function.
<!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat. Genet. paper) is outside the DBD. See also new Zaret paper in NSMB looking at contribution of IDRs and activation domains to chromatin opening on nucleosome arrays in vitro -->.
To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig. XX).
Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was totally unable to bind to class II or III regions (Fig. 6a,c).
We had previously shown that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
Because many Zld binding sites in S2 cells also lack a canonical CAGGTA motif (Fig. 3, Extended Data Fig XX), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required.
Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
Together, these experiments show that regions outside the DBD are essential for binding and opening of closed chromatin.

### 2023-01-06 manuscript/manuscript_sections/_05_Methods.qmd
break manuscript up into multiple qmd files

## Reagents

| Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
|-------------------|------------------|------------------|------------------|
| pMT-puro                              | AddGene                   | 17923                |                         |
| Gibson Cloning                        | New England Biolabs       | E2611                |                         |
| Schneider's Medium                    | Thermo Fisher Scientific  | 21720001             |                         |
| Fetal Bovine Serum                    | Omega Scientific          | FB-02                |                         |
| Antibiotic/antimycotic                | Thermo Fisher Scientific  | 15240062             |                         |
| Effectene Transfection Reagent        | Qiagen                    | 301427               |                         |
| PVDF Membrane                         |                           |                      |                         |
| Formaldehyde                          |                           |                      |                         |
| Protease inhibitor tablets            |                           |                      |                         |
| NEBNext Ultra II DNA library prep kit |                           |                      |                         |
| TruSeq v2 RNA library prep kit        |                           |                      |                         |
| Trizol                                |                           |                      |                         |
| **Antibodies**                        |                           |                      |                         |
| Anti-Zld                              | Harrison et al., 2010     |                      | ChIP: 6 µL , WB: 1:750  |
| Anti-Grh                              | Harrison et al., 2010     |                      | ChIP: 8 µL, WB: 1:1000  |
| Anti-Twi                              | He et al., 2015           |                      | ChIP: 10 µg, WB: 1:1000 |
| Anti-HA-peroxidase clone 3F10         | Roche                     | 12013819001          | WB: 1:500               |
| Anti-Tubulin DM1A                     | Sigma                     | T6199-200UL          | WB: 1:5000              |
| goat anti-rabbit IgG-HRP conjugate    | Bio-Rad                   | 1706515              | WB: 1:3000              |
| Anti-HA clone 12CA5                   | Roche                     |                      | ChIP: 7.5 µL            |
| Anti-H3K27me3                         | Cell Signaling Technology | C36B11; Lot #19      | WB: 1:1000, CR: 0.5 µL  |
|                                       |                           |                      |                         |
|                                       |                           |                      |                         |

##

## Cell culture and generation of stable cell lines

*zld*, *grh*, or *twi* cDNA (isoforms RB, RH, and RA, respectively) were cloned into pMT-puro plasmid (Addgene) via Gibson cloning (New England Biolabs).
For cloning of the Zld and Grh DNA-binding domains, DNA encoding amino acids 1114-1487 (Zld) or amino acids 603-1032 (Grh) was cloned into pMT-puro.

We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers and the absence of Kc cell markers [@Cherbas2011].
Drosophila S2 cells were cultured at 27°C in Schneider's medium (Thermo Fisher Scientific) supplemented with 10% fetal bovine serum (Omega Scientific) and antibiotic/antimycotic (Thermo Fisher Scientific).
For the generation of stable cell lines, cells were plated at 5✕10^5^ cells per mL.
After 24 hours, cells were transfected with 10 µg plasmid DNA using Effectene transfection reagent (Qiagen).
After an additional 24 hours, puromycin was added to a final concentration of 2 µg/mL.
Stable cell lines were recovered after 2-3 weeks of selection.
Following recovery of stable cell lines, cells were cultured with 1 µg/mL puromycin.

## Induction of transcription factor expression

Transcription factor expression was induced by adding CuSO~4~ to the cell culture media.
Expression of protein at physiological levels was achieved with 1000, 100, or 40 µM CuSO~4~ for Zld, Grh or Twi, respectively (Supplemental figure X).
For experiments with sub- or supraphysiological protein levels (Fig. 5), the following CuSO~4~ concentrations were used: 500 µM (Zld subphysiological), 1000 µM (Zld physiological) or 1500 µM (Zld supraphysiological); 25 µM (Grh subphysiological), 100 µM (Grh physiological) or 400 µM (Grh supraphysiological); 10 µM (Twi subphysiological), 40 µM (Twi physiological) or 160 µM (Twi supraphysiological).
For all induction experiments, cells were plated at 1×10^6^ cells per mL.
Unless otherwise specified, experiments were performed with cells harvested 48 hours after induction.

## Immunoblotting

Proteins were separated on denaturing polyacramide gels before transfer to a 0.45 μm Immobilon-P PVDF membrane (Millipore) in transfer buffer (25 mM Tris, 200 mM Glycine, 20% methanol) for 60 min (75 min for Zld) at 500mA at 4°C.
The membranes were blocked with BLOTTO (2.5% non-fat dry milk, 0.5% BSA, 0.5% NP-40, in TBST) for 30 min at room temperature and then incubated with anti-Zld (1:750) (Harrison et al., 2010), anti-Grh (1:1000) (Harrison et al., 2010), anti-Twi (1:1000) (He et al. 2015), anti-HA-peroxidase (1:500) (clone 3F10, Roche), or anti-Tubulin (DM1A, 1:5000) (Sigma), overnight at 4°C.
The secondary incubation was performed with goat anti-rabbit IgG-HRP conjugate (1:3000) (Bio-Rad) for 1 hr at room temperature.
Blots were treated with SuperSignal West Pico PLUS chemiluminescent substrate (Thermo Fisher Scientific) and visualized using the Azure Biosystems c600 or Kodak/Carestream BioMax Film (VWR).

## Immunostaining

## ChIP-seq

For each ChIP experiment, \~50 million cells were crosslinked by adding methanol-free formaldehyde directly to the cell culture medium to a final concentration of 0.8%.
Cells were rotated on a nutator for 7 minutes at room temperature (RT) before quenching of crosslinking by adding glycine to 125 mM.
Crosslinked cells were pelleted by centrifugation at 600 × *g* for 3 minutes.
Cells were washed twice with 1X PBS.
Cells were lysed by resuspending in lysis buffer (50 mM HEPES pH 7.9, 140 mM NaCl, 1 mM EDTA, 10% glycerol, 0.5% NP40, 0.25% Triton-X 100) with protease inhibitors (Pierce) and incubating on ice for 10 minutes.
The resulting chromatin was pelleted for 5 minutes at 1500 × *g* and resuspended in RIPA buffer.
Chromatin was sonicated in a Covaris S220 Ultrasonicator (5 cycles of 120 seconds with 60 second delay, 170 peak power, 10% duty factor, 200 cycles/burst).
Sonicated chromatin was centrifuged for 10 minutes at 10,000 × *g* to pellet insoluble material.
5% of the supernatant was set aside as input, and the remainder was incubated at 4°C overnight with antibodies (6 µL anti-Zld, 8 µL anti-Grh, 10 µg anti-Twi, 7.5 µL anti-HA).
20 µL protein A beads (Dynabeads Protein A, ThermoFisher Scientific) blocked with BSA were added and samples were incubated at 4°C for 4 hours.
Beads were separated on a magnet and washed 3× with low salt wash buffer (10 mM Tris pH 7.6, 1 mM EDTA, 0.1% SDS, 0.1% Na-Deoxycholate, 1% Triton-X 100, 150 mM NaCl), 2× high salt wash buffer (10 mM Tris pH 7.6, 1 mM EDTA, 0.1% SDS, 0.1% sodium deoxycholate, 1% Triton-X 100, 300 mM NaC), 2× LiCl wash buffer (0.25 M LiCl, 0.5% NP40, 0.5% sodium deoxycholate), and 1× with TE buffer with NaCl (10 mM Tris-HCl pH 8.0, 1 mM EDTA, 50 mM NaCl).
Beads were resuspended in elution buffer (50 mM Tris-HCl pH 8.0, 1% SDS, 10 mM EDTA) and incubated at 65°C for 10 minutes.
The IP and input samples were treated with 4.5 µL RNAse A for 30 minutes.
5 µL Proteinase K was added and samples were incubated overnight at 65°C to reverse crosslinking.
DNA was isolated by phenol:chloroform extraction, precipitated with EtOH, and resuspended in 20 µL H~2~O.Preparation of sequencing libraries was performed using NEBNext Ultra II kit (NEB) with 7 PCR cycles for library amplification.
Sequencing was performed on the Illumina Hi-Seq4000 using 50bp single-end reads, or on the Illumina NovaSeq 6000 using 150bp paired-end reads.

## ChIP-seq analysis

Bowtie2 version 2.4.4 [@Langmead2012] was used to align ChIP-seq reads to the *Drosophila melanogaster* genome (version dm6) with the following non-default parameters: -k 2 \--very-sensitive \--no-mixed \--no-discordant -X 5000.
Aligned reads were filtered to include only reads with a mapping quality score \> 30.
Reads mapping to unplaced scaffolds or the mitochondrial genome were discarded.
Peak calling was performed using MACS2 version 2.2 with parameters -g dm \--call-summits.
Only peaks that were detected in all replicates were considered in downstream analysis.
bigWig files were generated using deepTools bamCoverage version 3.5.1 with parameters --binSize 10.
bigWig files were z-score normalized by subtracting the genome-wide average of all bins from each bin value and dividing by the standard deviation of all bins.

## RNA-seq

5×10^6^ cells were harvested and resuspended in 800 µL Trizol (Invitrogen).
RNA was purified using a chloroform extraction followed by a column-based Quick-RNA MiniPrep extraction (Zymo).
Poly-A-selected RNA sequencing (RNA-seq) libraries were prepared using the TruSeq RNA sample prep kit v2 (Illumina).

## RNA-seq analysis

Raw reads were aligned to the Drosophila melanogaster (dm6) genome using HISAT2 version 2.1.0 with parameters -k 2 --very-sensitive.
Reads with a mapping quality score \< 30 and reads aligning to the mitochondrial genome or unplaced scaffolds were discarded.
FeatureCounts (from Subread version 2.0) was used to assign aligned reads to genes (FlyBase gene annotation release r6.45).
The resultant table of read counts was used to perform differential expression analysis using DESeq2 [@love2014].
Genes with an adjusted p-value \< 0.05 and a fold change \> 2 were considered differentially expressed.

## ATAC-seq

ATAC-seq was performed as described previsouly [@Buenrostro2013].
2×10^5^ cells were harvested for ATAC-seq by centrifugation at 600 × *g* for 3 minutes at RT.
Cells were washed once in 100 µL 1X PBS and resuspended in 100 µL ATAC lysis buffer (10 mM Tris pH 7.5, 10 mM NaCl, 3 mM MgCl~2~,0.1% NP-40).
Cells were centrifuged at 600 × *g* for 10 minutes at 4°C.
Supernatant was removed and pellet was resuspended in 47.5 µL buffer TD (Illumina).
2.5 uL of Tagment DNA Enzyme (Illumina) was added and samples were incubated at 37°C for 30 minutes.
Tagmented DNA was purified using MinElute Reaction Cleanup Kit (Qiagen) and eluted in 10 µL of buffer EB.
DNA was PCR amplified for 12 cycles with the following conditions:72°C for 5 minutes, 98°C for 30 seconds, then 12 cycles of 98°C for 10 seconds, 63°C for 30 seconds and 72°C for 1 minute.
Amplified libraries were purified using a 1.2X ratio of Axygen paramagnetic beads.
Sequencing was performed on the Illumina NovaSeq 6000 using 150bp paired-end reads.

## ATAC-seq analysis

Raw ATAC-seq reads were trimmed to remove adapter sequences using NGmerge version 0.3 [@Gaspar2018].
Trimmed reads were aligned to the *Drosophila melanogaster* genome (version dm6) using bowtie2 with paramters -k 2 \--very-sensitive \--no-mixed \--no-discordant -X 5000.
Reads with a mapping quality \< 30 and reads aligning to the mitochondrial genome or scaffolds were discarded.
As described previously, only fragments \< 100bp were considered for downstream analysis [@Buenrostro2013].
Reads were combined across all replicates and peak calling was performed using MACS2 with parameters -f BAMPE \--keep-dup all -g dm \--call-summits.
featureCounts was used to count the number of reads aligning within 200bp (100bp upstream or downstream) of peak summits.
Differential accessibility analysis was performed using DESeq2.
Regions with an adjusted p-value \< 0.05 were considered differentially accessible.

## Treatment of S2 cells with tazemetostat

Cells were plated at 1×10^6^ cells per mL and DMSO or tazemetostat was added to a final concentration of 10 µM.
Cells were split on days 2 and 5 and tazemetostat was added to maintain a final concentration of 10 µM.
On day 5, CuSO~4~ was added to achieve physiological expression of Zld, Grh or Twi (see above).
On day 7, cells were harvested for immunoblotting, ChIP and CUT&RUN.

## CUT&RUN

CUT&RUN was performed using EpiCypher reagents according to the manufacturer protocol.
2×10^5^ cells were used for each CUT&RUN reaction.
Overnight incubation with antibodies was performed in antibody buffer containing 0.05% digitonin.
0.5 µL anti-H3K27me3 (Cell signaling technologies) or rabbit IgG were used for CUT&RUN.
Prior to addition of antibody, a 1:5 dilution of K-MetStat Panel spike-in nucleosomes (EpiCypher, cat. \# 19-1002) was added to each reaction.
Libraries were prepared using the NEBNext Ultra II kit (NEB).
During library preparation, cleanup steps were performed using a 1.1X ratio of Axygen paramagnetic beads, as recommended by EpiCypher.
PCR amplification was performed with the following conditions: 98°C for 45 seconds, then 14 cycles of 98°C for 15 seconds, 60°C for 10 seconds, and a final extension of 72°C for 1 minute.
Sequencing was performed on the Illumina NovaSeq 6000 with 150bp paired-end reads.

## CUT&RUN analysis

Trimming

Alignment

Filtering

peak calling

bigwigs

z-score normalization

spike-in normalization

## Analysis of previously published data

Previously published datasets analyzed in this study can be found in supplemental table XXXX.
All previously published data were analyzed in parallel with data from this study using analysis parameters described above.

## Data and code availability

Sequencing data have been deposited in the Gene Expresion Omnibus under accession GSEXXXX.
All analysis code is available on GitHub: https://github.com/tjgibson/S2_pioneers_manuscript.
An archived version of the data processing and analysis pipeline is available on Zenodo: doiXXXX.

### 2023-01-06 manuscript/manuscript_sections/_06_Figure_Legends.qmd
break manuscript up into multiple qmd files

**Figure 1: Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
**c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
**d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h). i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
**j,i,** Proportion of class I, II or III regions that are proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
Promoter-proximal and promoter-distal binding sites are shown separately.
The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene.

**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites. a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites.
**c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions.
**d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.

**Figure 3: Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
**b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
Zld, Grh, and Twi ChIP-seq data were generated in this study.
Remaining ChIP-seq datasets were previously published (see Supplementary table XX).
**c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
Logos plot of the canonical motif are shown alongside the heatmap.
Percentages are shown for class I, II or III binding sites, or all WT ATAC-seq peaks as a control for the background motif frequency.
**f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
**g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**.
Motif scores were determined by comparing each identified motif instance to the canonical motif.

**Figure 4: Many endogenous binding sites are resistant to ectopic pioneer-factor binding. a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
ChIP-seq data for repressive histone modifications H3K27me3 and H3K9me3 is also shown.
**d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
Zld, Grh, and Twi ChIP-seq data are z-score normalized.
CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).

**Figure 5: Binding and opening of closed chromatin is concentration dependent. a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
Concetrations of CuSO~4~ used to induce protein expression are indicated above each plot.
**b,f,j,** Metaplots of ATAC-seq signal at class I-VI regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
**c,g,k,** Distribution of class I, II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
Y-axis indicates the total number of peaks called at each concentration.
Colors indicate the proportion of those peaks that are defined as class I, II, or III.
**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.

**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function. a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
Shown for Zld **(a)** and Grh **(b)**.
**c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**.
**d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.

**Figure 7: Pioneer-factor function requires stable chromatin occupancy.**

**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.**

-   ~~Schematic of cell line generation.~~

-   ~~Western blots to determine physiological expression and lack of protein expression in WT cells.~~

-   ~~RNA-seq data to show leaky expression and strong induction of RNA levels.~~

**Extended Data Figure 2: Zld and Grh bind to chromatin rapidly after induction of protein expression**

-   Western blot time course of protein levels

-   CUT&RUN time course to show rapid binding

-   likely rules out dependence on DNA replication

**Extended Data Figure 3: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.**

-   volcano and/or MA plots with enrichment of binding

-   Analysis of nonspecific changes?

-   RNA-seq volcano/MA plots

-   RNA-seq GO analysis

**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation**

-   Twi volcano and examples

-   TWI RNA-seq volcano and examples

-   Twi RNA-seq GO analysis

**Extended Data Figure 5: Chromatin features associated with Zld, Grh and Twi binding sites**

-   analysis of published ChIP datasets not shown in main figure

-   comparison of class I-III sites to show that class II-III sites are devoid of most histone mods

-   Class III sites have higher levels of H1 and H3 than class II sites.

**Extended Data Figure 6: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes**

-   De novo motif enrichment for each factor

-   MNase data to show enrichment of nucleosomes around target motifs

-   Clustering of MNase data to show that there is no preferred nucleosome position relative to motifs

-   additional analysis of Fiber-seq data to show that nucleosomes at these regions are poorly positioned

**Additional information to include in supplemental figures:**

-   De novo motif results

-   Analysis of promoter motifs

-   New binding sites at higher expression levels

-   Zld DBD ChIP only at CAGGTAG sites

### 2023-01-10 manuscript/manuscript_sections/_06_Figure_Legends.qmd
add code for supp figs 2-4

@@
 
--   Western blot time course of protein levels
+-   ~~Western blot time course of protein levels~~
 
@@
 
--   RNA-seq volcano/MA plots
+-   ~~RNA-seq volcano/MA plots~~
 
--   RNA-seq GO analysis
+-   ~~RNA-seq GO analysis~~
+
+-   Genome browser examples of specific and nonspecific accessibility changes?
+
+-   motif enrichment for nonspecific changes?
+
+-   \% of TFs bound and activated by Zld/Grh as indicator of nonspecific changes?
+
+-   Should we include any analysis of decreased/downregulated sites for RNA-seq or ATAC?
+
+-   Violin plots to emphasize correlation between ATAC and RNA-seq?
 
@@
 
+-   Western blot of Twi induction
+
 -   Twi volcano and examples
 
--   TWI RNA-seq volcano and examples
+-   ~~TWI RNA-seq volcano and examples3~~
 
--   Twi RNA-seq GO analysis
+-   ~~Twi RNA-seq GO analysis~~
+
+-   Genome browser examples of specific and nonspecific accessibility changes?
+
+-   motif enrichment for nonspecific changes?
+
+-   \% of TFs bound and activated by Zld/Grh as indicator of nonspecific changes?
+
+-   Should we include any analysis of decreased/downregulated sites for RNA-seq or ATAC?
+
+-   Violin plots to emphasize correlation between ATAC and RNA-seq?
 

### 2023-01-10 workflow/scripts/extended_data_fig2.R

@@ -1,0 +1,169 @@
+# setup ========================================================================
+library(plotgardener)
+library(tidyverse)
+library(org.Dm.eg.db)
+library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
+library(RColorBrewer)
+
+source("workflow/scripts/plot_heatmap.R")
+
+# define input files ===========================================================
+# Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
+
+S2_Zld_CR_0H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Zld_aZld_0H_accessble_zscore.bw"
+S2_Zld_CR_4H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Zld_aZld_4H_accessble_zscore.bw"
+S2_Zld_CR_12H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Zld_aZld_12H_accessble_zscore.bw"
+S2_Zld_CR_24H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Zld_aZld_24H_accessble_zscore.bw"
+S2_Zld_CR_48H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Zld_aZld_48H_accessble_zscore.bw"
+
+S2_Grh_CR_0H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Grh_aGrh_0H_accessble_zscore.bw"
+S2_Grh_CR_4H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Grh_aGrh_4H_accessble_zscore.bw"
+S2_Grh_CR_12H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Grh_aGrh_12H_accessble_zscore.bw"
+S2_Grh_CR_24H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Grh_aGrh_24H_accessble_zscore.bw"
+S2_Grh_CR_48H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Grh_aGrh_48H_accessble_zscore.bw"
+
+zld_ChIP_classes_fn <- "results/ChIP_peak_classes/zld_ChIP_classes.tsv"
+grh_ChIP_classes_fn <- "results/ChIP_peak_classes/grh_ChIP_classes.tsv"
+
[…]

### 2023-01-10 workflow/scripts/extended_data_fig3.R

@@ -1,0 +1,218 @@
+# setup ========================================================================
+library(plotgardener)
+library(tidyverse)
+library(org.Dm.eg.db)
+library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
+library(RColorBrewer)
+library(clusterProfiler)
+library(enrichplot)
+
+
+# define input files ===========================================================
+# Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
+# Zld_WT_ATAC_bw <- snakemake@input[["Zld_WT_ATAC_bw"]]
+# Zld_Zld_ATAC_bw <- snakemake@input[["Zld_Zld_ATAC_bw"]]
+# Zld_WT_RNAseq_bw <- snakemake@input[["Zld_WT_RNAseq_bw"]]
+# Zld_Zld_RNAseq_bw <- snakemake@input[["Zld_Zld_RNAseq_bw"]]
+# 
+# Grh_ChIP_bw <- snakemake@input[["Grh_ChIP_bw"]]
+# Grh_WT_ATAC_bw <- snakemake@input[["Grh_WT_ATAC_bw"]]
+# Grh_Grh_ATAC_bw <- snakemake@input[["Grh_Grh_ATAC_bw"]]
+# Grh_WT_RNAseq_bw <- snakemake@input[["Grh_WT_RNAseq_bw"]]
+# Grh_Grh_RNAseq_bw <- snakemake@input[["Grh_Grh_RNAseq_bw"]]
+# 
+# zld_ChIP_classes_fn <- snakemake@input[["zld_ChIP_classes"]]
+# grh_ChIP_classes_fn <- snakemake@input[["grh_ChIP_classes"]]
+
+# zld_ATAC_results_fn <- "RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results_annotated.tsv"
+# grh_ATAC_results_fn  <- "RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results_annotated.tsv"
+
+
+zld_RNAseq_results_fn <- "RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results_annotated.tsv"
+grh_RNAseq_results_fn  <- "RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results_annotated.tsv"
+
[…]

### 2023-01-10 workflow/scripts/extended_data_fig4.R

@@ -1,0 +1,159 @@
+# setup ========================================================================
+library(plotgardener)
+library(tidyverse)
+library(org.Dm.eg.db)
+library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
+library(RColorBrewer)
+library(clusterProfiler)
+library(enrichplot)
+
+
+# define input files ===========================================================
+# Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
+# Zld_WT_ATAC_bw <- snakemake@input[["Zld_WT_ATAC_bw"]]
+# Zld_Zld_ATAC_bw <- snakemake@input[["Zld_Zld_ATAC_bw"]]
+# Zld_WT_RNAseq_bw <- snakemake@input[["Zld_WT_RNAseq_bw"]]
+# Zld_Zld_RNAseq_bw <- snakemake@input[["Zld_Zld_RNAseq_bw"]]
+# 
+# Grh_ChIP_bw <- snakemake@input[["Grh_ChIP_bw"]]
+# Grh_WT_ATAC_bw <- snakemake@input[["Grh_WT_ATAC_bw"]]
+# Grh_Grh_ATAC_bw <- snakemake@input[["Grh_Grh_ATAC_bw"]]
+# Grh_WT_RNAseq_bw <- snakemake@input[["Grh_WT_RNAseq_bw"]]
+# Grh_Grh_RNAseq_bw <- snakemake@input[["Grh_Grh_RNAseq_bw"]]
+# 
+# zld_ChIP_classes_fn <- snakemake@input[["zld_ChIP_classes"]]
+# grh_ChIP_classes_fn <- snakemake@input[["grh_ChIP_classes"]]
+
+# zld_ATAC_results_fn <- "RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results_annotated.tsv"
+# grh_ATAC_results_fn  <- "RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results_annotated.tsv"
+
+
+twi_RNAseq_results_fn <- "RNAseq/results/DEseq2/S2-Twi_RNAseq_S2-Twi-vs-S2-WT_results_annotated.tsv"
+
+# # create blank layout for plot ===============================================
+# pdf(snakemake@output[[1]], useDingbats = FALSE)
+pdf("manuscript/figures/extended_data_fig4.pdf", useDingbats = FALSE)
[…]

### 2023-01-10 workflow/scripts/fig1.R

@@ -3,8 +3,6 @@ library(plotgardener)
 library(tidyverse)
 library(org.Dm.eg.db)
 library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
-# library(grImport)
-library(grid)
 library(RColorBrewer)
 
 source("workflow/scripts/plot_heatmap.R")

### 2023-01-12 manuscript/manuscript.qmd
add code for supp fig6

@@
 # Abstract
+
 {{< include manuscript_sections/_01_Abstract.qmd >}}
@@
 # Introduction
+
 {{< include manuscript_sections/_02_Introduction.qmd >}}
@@
 # Results
+
 {{< include manuscript_sections/_03_Results.qmd >}}
@@
 # Discussion
+
 {{< include manuscript_sections/_04_Discussion.qmd >}}
@@
 # Materials and Methods
+
 {{< include manuscript_sections/_05_Discussion.qmd >}}
@@
 # Figure Legends
+
+{{< include manuscript_sections/_06_Figure_Legends.qmd >}}

### 2023-01-12 manuscript/manuscript_sections/_02_Introduction.qmd
add code for supp fig6

# Introduction

outline:

-   Introduction to pioneer factors: what are they and why do they matter

-   articulate the knowledge gap and questions: how do PFs select their binding sites?
    What controls context-dependent pioneer factor binding and activity?

-   Introduce Zld and Grh.

-   Summary paragraph.

### 2023-01-12 workflow/scripts/extended_data_fig2.R

@@ -50,7 +50,7 @@ hm_downstream <-  500
 zld_ChIP_classes <- zld_ChIP_classes_fn |> 
   read_tsv()
 
-grh_ChIP_classes <- grh_ChIP_classes |> 
+grh_ChIP_classes <- grh_ChIP_classes_fn |> 
   read_tsv()
 
 # panel A ======================================================================

### 2023-01-12 workflow/scripts/extended_data_fig6.R

@@ -1,0 +1,513 @@
+# setup ========================================================================
+library(plotgardener)
+library(tidyverse)
+library(org.Dm.eg.db)
+library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
+library(RColorBrewer)
+library(plyranges)
+library(rtracklayer)
+
+source("workflow/scripts/plot_heatmap.R")
+
+# define input files ===========================================================
+# Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
+
+S2_Zld_ChIP_bw <-   "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_IP.bw"
+S2_Grh_ChIP_bw <-   "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_IP.bw"
+S2_Twi_ChIP_bw <-   "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Twi_aTwi_IP.bw"
+MNase_seq_bw <-  "data/published_datasets/Chereji_2019/GSE128689_RAW/GSM3934479_S2_exp2_seq1_40min.bw"
+
+zld_ChIP_classes_fn <- "results/ChIP_peak_classes/zld_ChIP_classes.tsv"
+grh_ChIP_classes_fn <- "results/ChIP_peak_classes/grh_ChIP_classes.tsv"
+twi_ChIP_classes_fn <- "results/ChIP_peak_classes/twi_ChIP_classes.tsv"
+
+zld_motif_instances_fn <- "results/motif_instances/zld_motifs.bed"
+grh_motif_instances_fn <- "results/motif_instances/grh_motifs.bed"
+twi_motif_instances_fn <- "results/motif_instances/twi_motifs.bed"
+
+# impor ChIP classes ===========================================================
+zld_ChIP_classes <- zld_ChIP_classes_fn |> 
+  read_tsv()
+
+grh_ChIP_classes <- grh_ChIP_classes_fn |> 
+  read_tsv()
+
+twi_ChIP_classes <- twi_ChIP_classes_fn |> 
+  read_tsv()
+
+
+# # create blank layout for plot ===============================================
+# pdf(snakemake@output[[1]], useDingbats = FALSE)
+pdf("manuscript/figures/extended_data_fig6.pdf", useDingbats = FALSE)
+pageCreate(width = 18, height = 12, default.units = "cm", showGuides = FALSE)
+
[…]

### 2023-01-12 manuscript/manuscript_sections/_03_Results.qmd
add code for supp figs 2-4
add code for supp fig6
add code for supp fig7

@@
 To more directly assess the positioning of nucleosomes around motifs, we analyzed a previously published single-molecule dataset using methyltransferase footprinting to determine the locations of nucleosomes on individual chromatin fibers [@Stergachis2020].
-We analyzed the distance from motifs at class II and regions to the nearest nucleosome dyad.
+We analyzed the distance from motifs at class II and III regions to the nearest nucleosome dyad.
 This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
@@
 To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
-This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required.
+This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required (Extended Data Fig. XX).
 Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).

### 2023-01-12 manuscript/manuscript_sections/_06_Figure_Legends.qmd
add code for supp fig6
add code for supp fig7

@@
 
+**Extended Data Figure 7: Zld, Grh and Twi display cell-type specific binding**
+
+-   comparison to other tissues
+
+-   analysis of motifs across tissues?
+
+**Extended Data Figure 8: PRC2 inhibition is insufficient to promote Zld, Grh, and Twi binding to repressed targets**
+
+-   Examples of class III regions with H3K27me3
+
+-   examples of class IV-VII regions
+
+-   Spike-in validation of H3K27me3 antibody
+
+-   Genome browsers showing loss of H3K27me3 in taz-treated cells without changes in binding.
+
+**Extended Data Figure 9: Expression of Zld, Grh, or Twi at supraphysiological levels results in novel binding sites**
+
+-   Western blot showing protein levels for titration experiments
+
+-   Examples of novel binding sites for each factor
+
+-   motif enrichment at novel binding sites
+
+-   Chromatin features at novel binding sites.
+
+**Extended Data Figure 10:**
+
+-   Analysis of Zld DBD ChIP only at CAGGTAG sites
+
+-   Examples of any novel DBD binding sites not detected for full-length proteins
+
 **Additional information to include in supplemental figures:**

### 2023-01-12 workflow/scripts/extended_data_fig7.R

@@ -1,0 +1,831 @@
+# setup ------------------------------------------------------------------------
+library(rtracklayer)
+library(GenomicRanges)
+library(tidyverse)
+library(eulerr)
+library(RColorBrewer)
+library(plotgardener)
+
+source("workflow/scripts/utils.R")
+source("workflow/scripts/plot_heatmap.R")
+source("../useful_functions/peak_processing.R")
+
+# define input files ===========================================================
+# Zld input files
+S2_Zld_ChIP_fn <- "ChIPseq/results/peaks/final/S2-Zld_aZld_IP.bed"
+nc14_Zld_ChIP_fn <- "published_ChIPseq/results/peaks/individual/narrow/embryo-nc14_aZld_summits.bed"
+brain_Zld_ChIP_fn <-  "../S2_pioneers/results/bed_files/peaks/brain_aZld.bed"
+
+S2_ZLD_ChIP_bw <-   "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_IP.bw"
+embryo_Zld_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/embryo-nc14_aZld.bw"
+brain_Zld_ChIP_bw <-  "../S2_pioneers/results/bigwigs/normalized/Zld_NB_merged_zscore.bw"
+zld_motif_bw <-  "results/motif_instances/zld_motifs_pvalue.bw"
+
+zld_motif_instances_fn <- "results/motif_instances/zld_motifs.tsv"
+
+# Grh input files
+S2_Grh_ChIP_fn <- "ChIPseq/results/peaks/final/S2-Grh_aGrh_IP.bed"
+embryo_Grh_ChIP_fn <- "published_ChIPseq/results/peaks/individual/narrow/embryo-15-16H_aGrh_summits.bed"
+wing_Grh_ChIP_fn <- "published_ChIPseq/results/peaks/individual/narrow/wing-disc_aGrh_summits.bed"
+
+S2_Grh_ChIP_bw <-   "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_IP.bw"
+embryo_Grh_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/embryo-15-16H_aGrh.bw"
[…]

### 2023-01-18 manuscript/manuscript_sections/_02_Introduction.qmd
add introduction draft

@@
+---
+editor: 
+  markdown: 
+    wrap: 72
+---
+
 # Introduction
 
-outline:
+Transcription factors activate transcription by binding to specific DNA
+motifs and recruiting transcriptional machinery. The packaging of DNA
+into chromatin obstructs TFs from accessing their target motifs. Thus,
+many TFs can only bind to motifs that fall within nucleosome-free
+regions (NFRs). One mechanism for how NFRs are established is through
+the action of pioneer transcription factors (PFs). PFs are defined by
+their ability to recognize target motifs on nucleosomes. PFs bind to
+nucleosomal targets, promote chromatin opening, and create NFRs that
+allow additional TFs to bind and activate transcription. Through this
+activity, PFs drive cell-type specific gene expression programs that
+determine cell fate. Due to these properties, PFs are essential players
+in development, induced reprogramming, and disease.
 
--   Introduction to pioneer factors: what are they and why do they matter
+Despite the importance of PFs to these diverse biological processes, how
+PFs select their binding sites in different biological contexts remains
+incompletely understood. Despite their ability to target inaccessible,
+nucleosomal binding sites, PFs do not bind to all genomic instances of
+their target motifs. Many PFs have also been shown to have cell-type
+specific patterns of binding and activity. How PF binding is restricted
+in different tissues or cellular contexts remains unclear. Many PFs bind
+primarily to naïve chromatin that is inaccessible, inactive, and lacks
+most histone modifications. Cell-type specific binding is often
+associated with cell-type specific chromatin features, however many of
+these observations have been correlative, and this has not been
+extensively tested functionally. In some cases, inhibition of various
+chromatin modifiers can enhance the ability of PFs to promote cellular
+reprogramming, but the molecular basis of how this relates to PF binding
+is unclear. Thus, a more systematic exploration of how PF binding and
+activity is regulated is essential to understanding how these factors
+function in development and disease.
 
--   articulate the knowledge gap and questions: how do PFs select their binding sites?
-    What controls context-dependent pioneer factor binding and activity?
+Two well studied PFs that display context-dependent activity and Zelda
+(Zld) and Grainy head (Grh). Zld is essential for driving early
+embryonic development in Drosophila. Immediately following
+fertilization, the zygotic genome is transcriptionally silent and
+development is controlled by maternally deposited mRNAs and proteins.
+Maternal products are gradually degraded as the zygotic genome is
+activated during a process known as the maternal-to-zygotic transition
+(MZT). In Drosophila, the pioneering activity Zld is essential for
+driving the MZT. Zld was the first discovered activator of the zygotic
+genome, and work on Zld paved the way for the identification of
+analogous proteins in other species. Our lab recently identified an
+additional function of Zld in promoting neural stem cell (NSC) fate in
+the developing larval brain. The majority of Zld binding sites are
+unique to either the embryo or NSCs, with only a subset of regions being
+bound in both tissues. How Zld is directed to distinct binding sites in
+the embryo and NSCs in unknown.
 
--   Introduce Zld and Grh.
+Grh is a master regulator of epithelial cell fate that is conserved
+among metazoans. In Drosophila, Grh acts as a pioneer factor in late
+embryos and larval tissues. Unlike Zld, Grh binding is similar across
+different tissues and developmental stages. Despite this stable binding,
+Grh is dispensible for chromatin accessibility in the early embryo, and
+only displays pioneering activity in late embryos or larval tissues.
+This intriguing combination of stable binding and cell-type specific
+activity raises questions about how pioneering by Grh is regulated
+during development. Additionally, overexpression of Grh is a hallmark of
+many epithelial cancers. The role of Grh in cancer is not well
+understood, as Grh can either inhibit or promote tumorigenesis,
+depending on the context.
 
--   Summary paragraph.
+The essential roles of Zld and Grh in diverse biological processes and
+the well documented context-dependent binding and activity of these
+factors makes them excellent models to explore how PF binding and
+activity are regulated. Because many previous studies have relied on
+loss-of-function experiments, it has been difficult to unravel causal
+relationships between chromatin structure and PF activity. We therefore
+expressed Zld and Grh in Drosophila S2 cells, where neither factor is
+endogenously expressed. We demonstrate that when expressed ectopically,
+Zld and Grh bind to closed chromatin, establish NFRs, and promote
+transcriptional activation. We show that an additional TF, Twist (Twi),
+can act as a pioneer factor. Pioneering by Zld, Grh, and Twi depends on
+a combination of local chromatin structure, sequence context, and PF
+concentration and requires protein regions outside the DNA-binding
+domain. We propose that pioneering activity is not a binary feature
+intrinsic to a protein, but occurs on a spectrum and can be regulated by
+a variety of protein-intrinsic and cell-type specific features.
+
+<!--# Should I discuss what is known about DBDs, structural features, and how PFs engage chromatin? -->

### 2023-01-18 manuscript/manuscript_sections/_04_Discussion.qmd
add introduction draft

# Discussion

Major conclusions to emphasize:

-   Twist is a pioneer factor

-   PF activity occurs on a spectrum and is context-dependent.

-   Context-dependent PF activity can be regulated by both protein-intrinsic and extrinsic features.

-   Preexisting active chromatin attracts TF binding, even with weaker motifs.

-   Most functional binding occurs at canonical motifs in naïve chromatin

-   Sequence composition is a main determinant of PF binding and activity

-   Regions outside the DBD are essential for pioneering

### 2023-01-18 manuscript/manuscript_sections/_06_Figure_Legends.qmd

@@
 
--   examples of class IV-VII regions
+-   examples of class IV-VI regions
 
@@
 -   Genome browsers showing loss of H3K27me3 in taz-treated cells without changes in binding.
+
+-   Individual examples of sites that are gained in taz-treated cells
 
@@
 
+-   FL and DBD preference for motifs that are in open vs.
+closed chromatin: % of closed vs.
+open motifs bound, motif scores for accessible bound, accessible unbound, inaccessible bound, and inaccessible unbound.
+
 **Additional information to include in supplemental figures:**
@@
 
--   New binding sites at higher expression levels
-
--   Zld DBD ChIP only at CAGGTAG sites
+-   Analysis of deep learning models

### 2023-01-19 manuscript/manuscript_sections/_02_Introduction.qmd
update MS text

@@
 ---
-editor: 
-  markdown: 
-    wrap: 72
+bibliography: references.bib
 ---
@@
 
-Transcription factors activate transcription by binding to specific DNA
-motifs and recruiting transcriptional machinery. The packaging of DNA
-into chromatin obstructs TFs from accessing their target motifs. Thus,
-many TFs can only bind to motifs that fall within nucleosome-free
-regions (NFRs). One mechanism for how NFRs are established is through
-the action of pioneer transcription factors (PFs). PFs are defined by
-their ability to recognize target motifs on nucleosomes. PFs bind to
-nucleosomal targets, promote chromatin opening, and create NFRs that
-allow additional TFs to bind and activate transcription. Through this
-activity, PFs drive cell-type specific gene expression programs that
-determine cell fate. Due to these properties, PFs are essential players
-in development, induced reprogramming, and disease.
+Transcription factors activate transcription by binding to specific DNA motifs and recruiting transcriptional machinery.
+The packaging of DNA into chromatin obstructs TFs from accessing their target motifs.
+Thus, many TFs can only bind to motifs that fall within nucleosome-free regions (NFRs).
+One mechanism for how NFRs are established is through the action of pioneer transcription factors (PFs).
+PFs are defined by their ability to recognize target motifs on nucleosomes.
+PFs bind to nucleosomal targets, promote chromatin opening, and create NFRs that allow additional TFs to bind and activate transcription[@Iwafuchi-Doi2016].
+Through this activity, PFs drive cell-type specific gene expression programs that determine cell fate.
+Due to these properties, PFs are essential players in development, induced reprogramming, and disease [@Larson2021].
 
-Despite the importance of PFs to these diverse biological processes, how
-PFs select their binding sites in different biological contexts remains
-incompletely understood. Despite their ability to target inaccessible,
-nucleosomal binding sites, PFs do not bind to all genomic instances of
-their target motifs. Many PFs have also been shown to have cell-type
-specific patterns of binding and activity. How PF binding is restricted
-in different tissues or cellular contexts remains unclear. Many PFs bind
-primarily to naïve chromatin that is inaccessible, inactive, and lacks
-most histone modifications. Cell-type specific binding is often
-associated with cell-type specific chromatin features, however many of
-these observations have been correlative, and this has not been
-extensively tested functionally. In some cases, inhibition of various
-chromatin modifiers can enhance the ability of PFs to promote cellular
-reprogramming, but the molecular basis of how this relates to PF binding
-is unclear. Thus, a more systematic exploration of how PF binding and
-activity is regulated is essential to understanding how these factors
-function in development and disease.
+Despite the importance of PFs to these diverse biological processes, how PFs select their binding sites in different biological contexts remains incompletely understood.
+Despite their ability to target inaccessible, nucleosomal binding sites, PFs do not bind to all genomic instances of their target motifs.
+Many PFs have also been shown to have cell-type specific patterns of binding and activity [@Soufi2012; @Donaghey2018].
+How PF binding is restricted in different tissues or cellular contexts remains unclear.
+Many PFs bind primarily to naïve chromatin that is inaccessible, inactive, and lacks most histone modifications [@Soufi2012; @Donaghey2018].
+Cell-type specific binding is often associated with cell-type specific chromatin features [@Mayran2019], however many of these observations have been correlative, and this has not been extensively tested functionally.
+In some cases, inhibition of various chromatin modifiers can enhance the ability of PFs to promote cellular reprogramming [@Larson2021a; @Soufi2012], but the molecular basis of how this relates to PF binding is unclear.
+Thus, a more systematic exploration of how PF binding and activity is regulated is essential to understanding how these factors function in development and disease.
 
-Two well studied PFs that display context-dependent activity and Zelda
-(Zld) and Grainy head (Grh). Zld is essential for driving early
-embryonic development in Drosophila. Immediately following
-fertilization, the zygotic genome is transcriptionally silent and
-development is controlled by maternally deposited mRNAs and proteins.
-Maternal products are gradually degraded as the zygotic genome is
-activated during a process known as the maternal-to-zygotic transition
-(MZT). In Drosophila, the pioneering activity Zld is essential for
-driving the MZT. Zld was the first discovered activator of the zygotic
-genome, and work on Zld paved the way for the identification of
-analogous proteins in other species. Our lab recently identified an
-additional function of Zld in promoting neural stem cell (NSC) fate in
-the developing larval brain. The majority of Zld binding sites are
-unique to either the embryo or NSCs, with only a subset of regions being
-bound in both tissues. How Zld is directed to distinct binding sites in
-the embryo and NSCs in unknown.
+Two well studied PFs that display context-dependent activity and Zelda (Zld) and Grainy head (Grh).
+Zld is essential for driving early embryonic development in Drosophila [@Liang2008; @Staudt2006].
+Immediately following fertilization, the zygotic genome is transcriptionally silent and development is controlled by maternally deposited mRNAs and proteins.
+Maternal products are gradually degraded as the zygotic genome is activated during a process known as the maternal-to-zygotic transition (MZT) [@Schulz2019; @Vastenhouw2019].
+In Drosophila, the pioneering activity Zld is essential for driving the MZT[@Liang2008; @Schulz2015; @Sun2015; @McDaniel2019a].
+Zld was the first discovered activator of the zygotic genome, and work on Zld paved the way for the identification of analogous proteins in other species[@Lee2013; @Leichsenring2013; @Gassler2022].
+Our lab recently identified an additional function of Zld in promoting neural stem cell (NSC) fate in the developing larval brain[@Larson2021a].
+The majority of Zld binding sites are unique to either the embryo or NSCs, with only a subset of regions being bound in both tissues.
+How Zld is directed to distinct binding sites in the embryo and NSCs in unknown.
 
-Grh is a master regulator of epithelial cell fate that is conserved
-among metazoans. In Drosophila, Grh acts as a pioneer factor in late
-embryos and larval tissues. Unlike Zld, Grh binding is similar across
-different tissues and developmental stages. Despite this stable binding,
-Grh is dispensible for chromatin accessibility in the early embryo, and
-only displays pioneering activity in late embryos or larval tissues.
-This intriguing combination of stable binding and cell-type specific
-activity raises questions about how pioneering by Grh is regulated
-during development. Additionally, overexpression of Grh is a hallmark of
-many epithelial cancers. The role of Grh in cancer is not well
-understood, as Grh can either inhibit or promote tumorigenesis,
-depending on the context.
+Grh is a master regulator of epithelial cell fate that is conserved among metazoans [@Wang2012].
+In Drosophila, Grh acts as a pioneer factor in late embryos and larval tissues [@Jacobs2018; @Nevil2020].
+Unlike Zld, Grh binding is similar across different tissues and developmental stages[@Nevil2017].
+Despite this stable binding, Grh is dispensible for chromatin accessibility in the early embryo, and only displays pioneering activity in late embryos or larval tissues.
+This intriguing combination of stable binding and cell-type specific activity raises questions about how pioneering by Grh is regulated during development.
+Additionally, overexpression of Grh is a hallmark of many epithelial cancers.
+The role of Grh in cancer is not well understood, as Grh can either inhibit or promote tumorigenesis, depending on the context[@Reese2019].
 
-The essential roles of Zld and Grh in diverse biological processes and
-the well documented context-dependent binding and activity of these
-factors makes them excellent models to explore how PF binding and
-activity are regulated. Because many previous studies have relied on
-loss-of-function experiments, it has been difficult to unravel causal
-relationships between chromatin structure and PF activity. We therefore
-expressed Zld and Grh in Drosophila S2 cells, where neither factor is
-endogenously expressed. We demonstrate that when expressed ectopically,
-Zld and Grh bind to closed chromatin, establish NFRs, and promote
-transcriptional activation. We show that an additional TF, Twist (Twi),
-can act as a pioneer factor. Pioneering by Zld, Grh, and Twi depends on
-a combination of local chromatin structure, sequence context, and PF
-concentration and requires protein regions outside the DNA-binding
-domain. We propose that pioneering activity is not a binary feature
-intrinsic to a protein, but occurs on a spectrum and can be regulated by
-a variety of protein-intrinsic and cell-type specific features.
+The essential roles of Zld and Grh in diverse biological processes and the well documented context-dependent binding and activity of these factors makes them excellent models to explore how PF binding and activity are regulated.
+Because many previous studies have relied on loss-of-function experiments, it has been difficult to unravel causal relationships between chromatin structure and PF activity.
+We therefore expressed Zld and Grh in Drosophila S2 cells, where neither factor is endogenously expressed.
+We demonstrate that when expressed ectopically, Zld and Grh bind to closed chromatin, establish NFRs, and promote transcriptional activation.
+We show that an additional TF, Twist (Twi), can act as a pioneer factor.
+Pioneering by Zld, Grh, and Twi depends on a combination of local chromatin structure, sequence context, and PF concentration and requires protein regions outside the DNA-binding domain.
+We propose that pioneering activity is not a binary feature intrinsic to a protein, but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type specific features.
 

### 2023-01-19 manuscript/manuscript_sections/_03_Results.qmd
update MS text

@@
 ## Ectopically expressed pioneer factors bind and open closed chromatin
-
--   ~~Establish system for ectopic expression of pioneer factors~~
-
--   ~~Identification of physiological expression levels~~
-
--   ~~Define binding classes I-III and describe characteristics (distribution, promoters/enhancers, etc.)~~
-
--   Time course and dynamics of binding to different classes (supplement)
 
@@
 We detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
-XX).
+3a,d).
 <!--# Add more information here about sites with differential accessibility.
@@
 To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression.
-Differential expression analysis using DESeq2 [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. XX).
-Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig.
-XX <!--# include statistics  -->), suggesting that binding of these factors mainly leads to transcriptional activation and not repression.
-<!--# Consider mentioning Grh as a repressor here and giving examples of genes where Grh may be repressing.
-Discuss if Grh-mediated transcriptional repression coincides with chromatin opening, chromatin closing, or neither --> Gene ontology (GO) analysis indicated that the genes activated by each factor are consistent with the endogenous functions of Zld and Grh (Extended Data Fig.
-XX).
+Differential expression analysis using DESeq2 [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. 3b,e).
+<!--# Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig.
+3b,e  include statistics), suggesting that binding of these factors mainly leads to transcriptional activation and not repression.
+--> <!--# Consider mentioning Grh as a repressor here and giving examples of genes where Grh may be repressing.
+Discuss if Grh-mediated transcriptional repression coincides with chromatin opening, chromatin closing, or neither --> The genes activated by Grh were enriched for gene ontology (GO) terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig.
+3c).
+Genes activated by Zld expression were enriched for embryonic morphogenesis, consistent with the known function of Zld, as well as imaginal disc development, in which Zld has not been studied (Extended Data Fig.
+3f).
 <!--# Add more detailed information on GO analysis here -->
@@
 Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l).
-sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig. XX <!--# add statistics -->).
-When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
+<!--# Sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig.
+XX).
+--> When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig.
+1j,l).
 This was particularly evident at class III promoters, which were most likely to correspond to upregulated genes.
@@
 
--   ~~Introduce Twist and justify the comparison to Zld/Grh.~~
+```{r}
+twi_ChIP_classes <- here("results/ChIP_peak_classes/twi_ChIP_classes.tsv") |> 
+  read_tsv()
 
--   ~~Twi binds closed chromatin and opens at some sites.~~
+twi_closed <- twi_ChIP_classes |> 
+  filter(class == "ii" | class == "iii") |> 
+  nrow()
+```
 
--   Mention analysis of published Abd-A and Ubx cell lines.
-
--   ~~Class I-III sites, distribution, etc.~~
-
--   ~~Twi also promotes transcriptional activation~~
-
--   ~~Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.~~
-
--   ~~Comparison of RNA-seq and ATAC~~
-
--   ~~Mention differences between Twi vs. Zld and Grh~~
+<!--# To do: Mention analysis of published Abd-A and Ubx cell lines. -->
 
@@
 Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
-Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
+Twi bound extensively to closed chromatin, with `r twi_closed`% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
 At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c).
-Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. XX).
-Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig. XX).
-GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. XX).
-Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
-The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig. XX).
-Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig. XX).
+Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4a).
+<!--# Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig.
+XX).
+--> GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig.
+4b).
+<!--# Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
+The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig.
+XX).
+Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig.
+XX).
 This was in contrast to Zld and Grh, where induction results in the activation of many genes that are not directly bound by Zld or Grh.
@@
 Twi, by contrast, appears to directly activate genes required for mesodermal and muscle cell fate.
-<!--# Is this worth mentioning?
+--> <!--# Is this worth mentioning?
 Might require additional analysis of how many TFs are upregulated in RNA-seq data for each factor.
 Could also require additional motif analysis to identify/support secondary TFs and reason for nonspecific effects.
---> Similar to Zld and Grh, transcriptional activation was correlated with chromatin opening (Extended Data Fig.
+--> <!--# Similar to Zld and Grh, transcriptional activation was correlated with chromatin opening (Extended Data Fig.
 XX).
-Together, our data show that Twi can act as a pioneer factor.
+--> Together, our data show that Twi can act as a pioneer factor.
 
@@
 
--   ~~Overlap of class I regions suggests nonspecific binding~~
-
--   ~~Analysis of published ChIP data shows that class I regions are within active chromatin~~
-
--   ~~RNA-seq and ATAC-seq for class I to show that the binding is not functional~~
-
--   ~~Analysis of chromatin at class II-III regions: mostly devoid of histone modifications~~
-
--   ~~Class III regions have higher H3 and H1 occupancy compared to class II~~
-
--   Where to put analysis of nucleosome positioning?
-
 While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites are shared across Zld, Grh, and Twi (Fig. 3a).
-This was in contrast to class II and III peaks, which were largely specific to each individual factor (Extended Data Fig. XX).
-Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig. XX).
-We hypothesized that these overlapping class I binding sites may occur at regulatory regions that are active in WT cells.
+This was in contrast to class II and III peaks, which were largely specific to each individual factor.
+<!--# Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig.
+XX).
+--> We hypothesized that these overlapping class I binding sites may occur at regulatory regions that are active in WT cells.
 To test this, we analyzed published ChIP-seq data from WT S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV.
@@
 This enrichment of all three factors at highly active regions suggests that exogenously expressed factors may bind nonspecifically to active regions of the genome.
-This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig. XX).
+<!--# This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig.
+XX).
+-->
 
@@
 Analysis of published ChIP-seq data demonstrated that class II and III sites have low levels of many histone modifications, chromatin-associated proteins, and transcription factors (Extended Data Fig.
-XX).
+5).
 These data are in agreement with previous observations that Oct4, Sox2 and Klf4 access binding sites in naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
-Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. XX).
-Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig. XX).
+Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. 5).
+Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig. 8a-c).
 Together this suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin.
@@
 ## Chromatin opening by pioneer factors is driven by motif content
-
--   De novo motif enrichment for ChIP peaks (supplemental)
-
--   ~~Motif analysis: percent with motif, average n motifs, motif quality~~
-
--   ~~Motif positioning around nucleosomes? (supplemental)~~
 
@@
 
-Previous work has shown that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (refs).
+Previous work has shown that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (REFs).
 <!--# Add a few sentences here with more details on past work --> To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
-Analysis of average MNase-seq signal at class I-III peaks indicated that class III sites have high levels of nucleosome occupancy that, on average, occur over the respective motif.
+Analysis of average MNase-seq signal at class I-III peaks indicated that class III sites have high levels of nucleosome occupancy that, on average, occur over the respective motif (Extended Data Fig.
+6a-c).
 Hierarchical clustering of MNase-seq data revealed that there was no preference for binding or chromatin opening at sites that had motifs in a particular orientation on nucleosomes (Extended Data Fig.
-XX).
+6d-f).
 For class II and III peaks, some motifs were positioned close to the nuclesome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
@@
 
--   Z~~elda binding is more similar to the brain than the embryo~~
-
--   ~~Grh binding is stable across tissues~~
-
--   ~~Twi sites in S2 cells have low level binding in embryo, but strongest peaks from embryo are not bound~~
-
--   ~~Enrichment of repressive marks at sites bound in other tissues~~
-
--   ~~Results of Taz experiments~~
-
--   Comment on class VI sites that are not bound and lack repressive chromatin.
-
 Prior studies have extensively explored the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
@@
 We compared Zld binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
-We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig. XX).
+We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig. 7a,b).
 Previous work has shown that regions resistant to pioneer-factor binding are enriched for the repressive histone modifications H3K9me3 and H3K27me3 [@Mayran2018a; @Soufi2012].
 To test if repressive marks were present at the resistant sites in S2 cells, we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
-Based on analysis of these datasets we divided the resistant sites into three additional classes of binding sites: class IV sites have high levels H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a).
-We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c).
+Based on analysis of these datasets we divided the resistant sites into three additional classes of binding sites: class IV sites have high levels H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a; Extended Data Fig.
+8d-f).
+We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c; Extended Data Fig. 8e-l).
 We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017], and to Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
@@
 ## Pioneer-factor binding and chromatin opening is concentration dependent
-
--   ~~For all three factors, increased concentration results in increased binding and opening of class II and III sites.~~
-
--   ~~For all three factors, increased concentration leads to novel binding sites.~~
-
--   ~~For Zld and Twi, increased concentration is insufficient to allow binding to repressed sites.~~
-
--   ~~For Grh, increased concentration does allow binding and opening at repressed sites~~
 
@@
 For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration.
-Together, these data demonstrate that both binding and chromatin opening by pioneer factors are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are restistant to binding at physiological levls.
+Together, these data demonstrate that both binding and chromatin opening by pioneer factors are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are restistant to binding at physiological levels.
 
 ## Regions outside the DNA-binding domain are required for binding and opening of closed chromatin
-
--   ~~Generation of stable cell lines to express DBDs at same concentration as FL proteins~~
-
--   ~~Zld DBD is totally unable to bind to closed sites.~~
-
--   ~~In supplement, show Zld DBD binding specifically at class II and III sites with CAGGTAG to support that this is not due to a loss of ZnF2~~
-
--   ~~Grh DBD has reduced ability to access closed chromatin~~
-
--   ~~Neither DBD is sufficient for chromatin opening.~~
 
@@
 Because many Zld binding sites in S2 cells also lack a canonical CAGGTA motif (Fig.
-3, Extended Data Fig XX), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
+3), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
 To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
-This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required (Extended Data Fig. XX).
+This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required (Extended Data Fig. 10a).
 Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).

### 2023-01-19 workflow/scripts/extended_data_fig10.R
update manuscript draft

@@ -34,8 +34,8 @@ grh_ChIP_classes <- grh_FL_ChIP_classes_fn |>
 
 # # create blank layout for plot ===============================================
 # pdf(snakemake@output[[1]], useDingbats = FALSE)
-# pdf("manuscript/figures/extended_data_fig10.pdf", useDingbats = FALSE)
-pageCreate(width = 18, height = 12, default.units = "cm", showGuides = TRUE)
+pdf("manuscript/figures/extended_data_fig10.pdf", useDingbats = FALSE)
+pageCreate(width = 18, height = 12, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics

### 2023-01-19 workflow/scripts/extended_data_fig4.R

@@ -6,6 +6,7 @@ library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
 library(RColorBrewer)
 library(clusterProfiler)
 library(enrichplot)
+library(rtracklayer)
 
 
 # define input files ===========================================================
@@ -29,6 +30,10 @@ library(enrichplot)
 
 
 twi_RNAseq_results_fn <- "RNAseq/results/DEseq2/S2-Twi_RNAseq_S2-Twi-vs-S2-WT_results_annotated.tsv"
+twi_ATAC_results_fn <- "ATACseq/results/DEseq2_results_filtered/S2-Twi_ATACseq_S2-Twi-vs-S2-WT-40uM_results.tsv"
+
+twi_ChIP_peaks_fn <- "ChIPseq/results/peaks/final/S2-Twi_aTwi_IP.narrowPeak"
+
 
 # # create blank layout for plot ===============================================
 # pdf(snakemake@output[[1]], useDingbats = FALSE)
@@ -46,11 +51,69 @@ small_text_params <- pgParams(fontsize = 5)
 twi_color <- "#00A08A"
   
 
-
-
 # panel A ======================================================================
 # reference points for positioning figure components
 ref_x <- 0.5
+ref_y <- 0.5
+
+
+# read in ATAC-seq results
+twi_ATAC_results <- read_tsv(twi_ATAC_results_fn)
+
+# annotate ATAC peaks as overlapping a ChIP-seq peak
+twi_ChIP_peaks <- import(twi_ChIP_peaks_fn)
+
+twi_ATAC_gr <- twi_ATAC_results |> 
+  makeGRangesFromDataFrame()
+
+twi_ATAC_results$has_ChIP_peak <- FALSE
+overlaps <- findOverlaps(twi_ATAC_gr, twi_ChIP_peaks)@from
+twi_ATAC_results$has_ChIP_peak[overlaps] <- TRUE
+
+# define categories for coloring points
+twi_ATAC_results <- twi_ATAC_results |> 
+  mutate(diff_class = case_when(
+    is_diff & has_ChIP_peak ~ "twi bound",
+    is_diff & !has_ChIP_peak ~ "not twi bound",
+    !is_diff ~ "ns"
+  ))
+
+# volcano plot
+a_plot <- twi_ATAC_results |>
+  mutate(diff_class = fct_relevel(diff_class, c("ns", "not twi bound", "twi bound"))) |>
[…]

### 2023-01-19 workflow/scripts/extended_data_fig5.R

@@ -92,6 +92,22 @@ zld_chip_classes <- read_tsv("results/ChIP_peak_classes/zld_ChIP_classes.tsv")
 grh_chip_classes <- read_tsv("results/ChIP_peak_classes/grh_ChIP_classes.tsv")
 twi_chip_classes <- read_tsv("results/ChIP_peak_classes/twi_ChIP_classes.tsv")
 
+Zld_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_IP.bw"
+Zld_WT_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/S2-WT_1000uM_small.bw"
+Zld_Zld_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/S2-Zld_1000uM_small.bw"
+
+Grh_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_IP.bw"
+Grh_WT_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/FL_ATAC_S2-WT_100uM_small.bw"
+Grh_Grh_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/S2-Grh_100uM_small.bw"
+
+Twi_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Twi_aTwi_IP.bw"
+Twi_WT_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/Twi_ATAC_S2-WT_40uM_small.bw"
+Twi_Twi_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/S2-Twi_40uM_small.bw"
+
+H3K27me3_ChIP_bw <- "published_ChIPseq/results/bigwigs/zscore_normalized/individual/GSE151983_S2_aH3K27me3_IP.bw"
+H3K9me3_ChIP_bw <- "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE160855_aH3K9me3.bw"
+
+
 
 # # create blank layout for plot ===============================================
 # pdf(snakemake@output[[1]], useDingbats = FALSE)

### 2023-01-19 workflow/scripts/extended_data_fig6.R

@@ -119,6 +119,17 @@ plotGG(
   default.units = "cm"
 )
 
+plotText(
+  label = "a", params = panel_label_params, fontface = "bold",
+  x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
+)
+
+plotText(
+  label = "MNase-seq", params = large_text_params, fontface = "bold",
+  x = ref_x + 2.5, y = ref_y, just = "bottom", default.units = "cm"
+)
+
+
 # panel B ======================================================================
 # reference points for positioning figure components
 ref_x <- 5.5
@@ -178,6 +189,11 @@ plotGG(
   default.units = "cm"
 )
 
+plotText(
+  label = "MNase-seq", params = large_text_params, fontface = "bold",
+  x = ref_x + 2.5, y = ref_y, just = "bottom", default.units = "cm"
+)
+
 # panel C ======================================================================
 # reference points for positioning figure components
 ref_x <- 10.5
@@ -235,6 +251,11 @@ plotGG(
   x = (ref_x + 0.25), y = (ref_y + 0.25),
   width = 4, height = 3, just = c("left", "top"),
   default.units = "cm"
+)
+
+plotText(
+  label = "MNase-seq", params = large_text_params, fontface = "bold",
+  x = ref_x + 2.5, y = ref_y, just = "bottom", default.units = "cm"
 )
 
 # panel D ======================================================================
@@ -328,6 +349,23 @@ plotGG(
   default.units = "cm"
 )
 
+plotText(
+  label = "MNase-seq", params = large_text_params, fontface = "bold",
+  x = ref_x + 2.5, y = ref_y, just = "bottom", default.units = "cm"
+)
+
+plotText(
+  label = "class II", params = large_text_params, fontface = "bold",
+  x = ref_x + 0.5, y = ref_y + 1.5, just = "bottom", default.units = "cm", rot = 90,
+)
+
+
+plotText(
+  label = "class III", params = large_text_params, fontface = "bold",
+  x = ref_x + 0.5, y = ref_y + 5, just = "bottom", default.units = "cm", rot = 90,
+)
+
+
 # panel E ======================================================================
 # reference points for positioning figure components
 ref_x <- 5.5
@@ -418,6 +456,23 @@ plotGG(
   default.units = "cm"
 )
 
+plotText(
+  label = "MNase-seq", params = large_text_params, fontface = "bold",
[…]

### 2023-01-19 manuscript/manuscript_sections/_03_Results.qmd
update manuscript draft
update supp fig5

@@
 Twi bound extensively to closed chromatin, with `r twi_closed`% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
-At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c).
-Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4a).
+At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c; Extended Data Fig. 4a).
+Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4b).
 <!--# Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig.
@@
 --> GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig.
-4b).
+4C).
 <!--# Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
@@
 Analysis of published ChIP-seq data demonstrated that class II and III sites have low levels of many histone modifications, chromatin-associated proteins, and transcription factors (Extended Data Fig.
-5).
+5a,c,e).
 These data are in agreement with previous observations that Oct4, Sox2 and Klf4 access binding sites in naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
 Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. 5).
-Some class III sites also were within regions marked with the repressive histone modifications H3K27me3 or H3K9me3 (Extended Data Fig. 8a-c).
+Some class III sites also were within regions marked with the repressive histone modification H3K27me3 (Extended Data Fig. 5b,d,f).
 Together this suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin.

### 2023-01-19 workflow/scripts/extended_data_fig5.R

@@ -7,6 +7,7 @@ library(RColorBrewer)
 library(rtracklayer)
 
 source("workflow/scripts/plot_heatmap.R")
+source("workflow/scripts/utils.R")
 
 # define input files ===========================================================
 # Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
@@ -135,7 +136,18 @@ zld_color <- "#5BBCD6"
       # twi_heatmap_colors <- c("white","#00A08A", "#154734")
     twi_heatmap_colors <- brewer.pal(9, "GnBu")
     
-  
+    H3K27me3_color <- "gray40"      
+      
+    # reference points for positioning figure components
+    x_offset_class_label <- 0.25
+    x_offset_browser_label <- 1
+    x_offset_browser <- 1.5
+    
+    # set genome browser height
+    gb_height <- 0.3
+    
+    
+    
     # panel A ======================================================================
     # reference points for positioning figure components
     ref_x <- 0.5
@@ -188,55 +200,59 @@ zld_color <- "#5BBCD6"
       theme_minimal(base_size = small_text_params$fontsize) +
       theme(axis.text.x = element_text(angle = 45, hjust=1),
             axis.title = element_blank(),
-                  panel.grid.major = element_blank(),
-                  panel.grid.minor = element_blank(),
-                  panel.border = element_blank(),
-                  panel.background = element_blank(),
-            legend.key.size = unit(2, 'mm')
-            ) +
+            panel.grid.major = element_blank(),
+            panel.grid.minor = element_blank(),
+            panel.border = element_blank(),
+            panel.background = element_blank(),
+            legend.key.size = unit(2, 'mm'),
+            legend.position = "top",
+            legend.justification = c(0.9),
+            legend.margin=margin(0,0,4,0),
+            legend.box.margin=margin(-10,-10,-10,-10)
+      ) +
       scale_fill_distiller(palette = "Blues", direction = 1)
     
     
     plotGG(
       plot = a_plot,
-      x = (ref_x), y = ref_y,
-      width = 12, height = 2.5, just = c("left", "top"),
[…]

### 2023-01-19 workflow/scripts/extended_data_fig8.R

@@ -83,483 +83,7 @@ zld_color <- "#5BBCD6"
       hm_upstream <-  500
       hm_downstream <-  500
       
-      # panel A ======================================================================
-      # panel label
-      ref_x <- 0.5
-      ref_y <- 0.5
-      
-      plotText(
-        label = "a", params = panel_label_params, fontface = "bold",
-        x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
-      )
-      
-      plotText(
-        label = "class III", params = large_text_params, fontface = "bold",
-        x = ref_x + 3, y = ref_y, just = "bottom", default.units = "cm"
-      )
-      
-      # add broad H3K27me3 region
-      plotText(
-        label = paste("H3K27me3"),
-        params = small_text_params,
-        fontface = "bold",
-        x = (ref_x + x_offset_browser_label),
-        y = (ref_y + 0.25),
-        just = c("right", "center"),
-        default.units = "cm"
-      )
-      
-      region <- pgParams(
-        chrom = "chr3R",
-        chromstart = 6647050, chromend = 6823266,
-        assembly = "dm6"
-      )
-      
-      `H3K27me3_ChIP` <- readBigwig(file = H3K27me3_ChIP_bw, 
-                                    params = region
-      )
-      
-      ChIP_range <- signal_range(c(H3K27me3_ChIP$score))
-      
-      s1 <- plotSignal(
-        data = `H3K27me3_ChIP`, params = region,
-        x = (ref_x + x_offset_browser), y = (ref_y + 0.25), width = 3.75, height = gb_height,
-        just = c("left", "center"), default.units = "cm",
-        linecolor = H3K27me3_color, fill = H3K27me3_color, baseline = FALSE, baseline.lwd = 0, baseline.color = H3K27me3_color,
-        range = ChIP_range
-      )
-      
-      annoYaxis(
-        plot = s1, at = round(ChIP_range, 1),
-        axisLine = TRUE, fontsize = 5, lwd = 0.5, 
-      )
-      
-      # add zoomed in view of ChIP and ATAC
-      zoomRegion <- pgParams(
-        chrom = "chr3R",
-        chromstart = 6751637, chromend = 6760890,
-        assembly = "dm6"
-      )
-      
-      annoZoomLines(
-        plot = s1, params = zoomRegion,
[…]

### 2023-01-23 manuscript/manuscript_sections/_03_Results.qmd
update supplemental figs

@@
 
--   ~~Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.~~
-
--   ~~Comparison of RNA-seq and ATAC data: chromatin opening correlates with transcriptional activation.~~
-
--   ~~Class III promoters are most likely to be associated with upregulated genes~~
-
 ```{r}
@@
 ```{r}
+#| include: false
 twi_ChIP_classes <- here("results/ChIP_peak_classes/twi_ChIP_classes.tsv") |> 
@@
 In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012].
-To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. XX).
+To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. 4a).
 We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
@@
 Twi bound extensively to closed chromatin, with `r twi_closed`% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
-At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c; Extended Data Fig. 4a).
-Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4b).
+At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c; Extended Data Fig. 4b).
+Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4c).
 <!--# Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig.
@@
 --> GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig.
-4C).
+4d).
 <!--# Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
@@
 Based on analysis of these datasets we divided the resistant sites into three additional classes of binding sites: class IV sites have high levels H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a; Extended Data Fig.
-8d-f).
-We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c; Extended Data Fig. 8e-l).
+7d-f).
+We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c; Extended Data Fig. 7e-l).
 We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017], and to Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].

### 2023-01-23 workflow/scripts/extended_data_fig4.R

@@ -7,7 +7,7 @@ library(RColorBrewer)
 library(clusterProfiler)
 library(enrichplot)
 library(rtracklayer)
-
+library(EBImage)
 
 # define input files ===========================================================
 # Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
@@ -47,7 +47,6 @@ large_text_params <- pgParams(fontsize = 7)
 small_text_params <- pgParams(fontsize = 5)
 
 # colors
-# colors
 twi_color <- "#00A08A"
   
 
@@ -56,6 +55,154 @@ twi_color <- "#00A08A"
 ref_x <- 0.5
 ref_y <- 0.5
 
+# panel label
+plotText(
+  label = "a", params = panel_label_params, fontface = "bold",
+  x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
+)
+
+# read in tiff of western blot
+blot_image <- readImage("data/immunoblot_raw_images/2021-9-13_Twi-vs-embryo/Twi-vs-embryos.tif")
+
+# rotate image
+blot_image <- blot_image |> 
+  rotate(89, bg.col = "white") |> 
+  flop()
+
+# crop image
+blot_image <- blot_image[1335:3074,830:1211]
+
+# place blot on page
+plotRaster(
+  blot_image,
+  x = ref_x + 1,
+  y = ref_y + 1.5,
+  width = 10,
+  height = 2.195402,
+  default.units = "cm",
+  just = c("left, top")
+  
+  )
+
+# add labels to western blot
+plotSegments(
+  x0 = ref_x + 1.1, y0 = ref_y + 0.25, x1 = ref_x + 4.7, y1 = ref_y + 0.25,
+  default.units = "cm",
+  lwd = 1
+)
+
+plotSegments(
+  x0 = ref_x + 4.9, y0 = ref_y + 0.25, x1 = ref_x + 8.3, y1 = ref_y + 0.25,
+  default.units = "cm",
+  lwd = 1
+)
+
+plotSegments(
+  x0 = ref_x + 8.6, y0 = ref_y + 0.25, x1 = ref_x + 10.6, y1 = ref_y + 0.25,
+  default.units = "cm",
+  lwd = 1
+)
+
+# panel label
+plotText(
+  label = "S2 Twi", params = large_text_params, fontface = "bold",
+  x = ref_x + 2.75, y = ref_y, just = "top", default.units = "cm"
+)
+
+plotText(
[…]

### 2023-01-24 workflow/scripts/extended_data_fig1.R
update results

@@ -6,6 +6,7 @@ library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
 library(grImport)
 library(grid)
 library(RColorBrewer)
+library(EBImage)
 
 
 # define input files ===========================================================
@@ -26,9 +27,12 @@ library(RColorBrewer)
 
 RPKM_table_fn <- "RNAseq/results/count_tables/S2-Grh_RNAseq_RPKM.tsv"
 
+zld_blot_image <- "data/immunoblot_raw_images/2018-10-17_Zld_induction/2018-1018-144408_pub.tif"
+grh_blot_image <- "data/immunoblot_raw_images/2018-11-08_Grh_induction/2018-1108-132834_pub.tif"
+
 # # create blank layout for plot ===============================================
 # pdf(snakemake@output[[1]], useDingbats = FALSE)
-pdf("manuscript/figures/extended_data_fig1.pdf", useDingbats = FALSE)
+# pdf("manuscript/figures/extended_data_fig1.pdf", useDingbats = FALSE)
 pageCreate(width = 18, height = 18.5, default.units = "cm", showGuides = TRUE)
 
 # general figure settings ======================================================
@@ -227,8 +231,33 @@ plotText(
   x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
 )
 
-# placeholder for western blot
-plotRect(x = (ref_x + 0.25), y = (ref_y + 0.25), width = 8, height = 3, default.units = "cm", just = c("top","left"))
+# read in tiff of western blot in grayscale
+blot_image <- readImage(zld_blot_image) |> 
+  channel("gray")
+
+# crop image
+blot_image <- blot_image[511:1154,421:580]
+
+# get blot aspect ratio
+blot_dim <- dim(blot_image)
+blot_aspect_ratio <- blot_dim[2] / blot_dim[1]
+
+
+
+# place blot on page
+plot_width <- 7
+
+plotRaster(
+  blot_image,
+  x = ref_x + 1,
+  y = ref_y + 1,
+  width = plot_width,
+  height = plot_width * blot_aspect_ratio,
+  default.units = "cm",
+  just = c("left, top")
+  
+)
+
 
 # panel D ======================================================================
 # reference points for positioning figure components
@@ -241,33 +270,33 @@ plotText(
   x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
 )
 
-# placeholder for western blot
[…]

### 2023-01-30 manuscript/manuscript_sections/_03_Results.qmd
update manuscript
update results
add MMH changes

@@
 
-To investigate the mechanisms of chromatin binding and opening by pioneer factors, we generated stable cell lines inducibly expressing Zelda (Zld) and Grainy head (Grh) in Drosophila S2 cells (Extended Data Fig.
-1a), where neither factor is endogenously expressed.
-We confirmed that Zld and Grh RNA (Extended Data Fig. 1b) and protein (Extended Data Fig. 1c,d) were not present in WT S2 cells.
-The stable cell lines express Zld or Grh under the control of the metallothionein promoter, which allows inducible and tunable expression with the addition of CuSO~4.~ To determine protein levels that would be approximately physiological, we compared Zld or Grh levels in cells with varying concentrations of CuSO~4~ to the levels of endogenous expression found in Drosophila embryos (Extended Data Fig.
+To investigate the mechanisms by which pioneer factors bind and open chromatin, we generated S2 cell lines to enable the individual expression of Zld or Grh (Extended Data Fig. 1a).
+Neither pioneer factor is normally expressed in WT S2 cells (Extended Data Fig. 1b,c,d).
+As these cell lines allowed the inducible and tunable expression of each factor, we determined the concentration of CuSO4 required to induce expression at levels similar to the endogenous embryonic expression levels of each factor (Extended Data Fig.
 1c,d).
-Physiological protein levels were achieved at 1000 mM or 100 mM CuSO~4~ for Zld or Grh, respectively.
+Approximately physiological protein levels were achieved with 1000 mM or 100 mM CuSO4 for Zld or Grh, respectively.
 
@@
 
-To determine how exogenous Zld and Grh interact with preexisting chromatin in S2 cells, we expressed each factor at physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
+To determine how exogenous Zld and Grh interact with preexisting chromatin, we expressed each factor at approximately physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
 We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
-To control for antibody specificity, we performed additional ChIP experiments in WT cells using anti-Zld, or anti-Grh.
-We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh.
+Additional ChIP-seq experiments were performed in WT cells using anti-Zld, or anti-Grh to control for antibody specificity.
+By excluding peaks that were called in these datasets, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh.
 <!--# Not sure the best way to discuss the control ChIP experiments.
@@
 
-Having established that Zld and Grh can bind to thousands of sites in S2 cells, we performed ATAC-seq on WT cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
-We performed differential accessibility analysis comparing Zld or Grh-expressing cells to WT cells.
-We detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
+Having established that Zld and Grh can bind thousands of sites in S2 cells, we performed assay for transposase-accessible chromatin (ATAC)-seq on WT cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
+Using differential accessibility analysis comparing Zld or Grh-expressing cells to WT cells, we detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
 3a,d).
@@
 
-To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of binding sites (Fig. 1a,b).
-Class I sites are accessible in WT cells and remain accessible when bound by ectopically-expressed Zld or Grh.
-Class II sites are inaccessible in WT cells, however binding of Zld or Grh does not lead to chromatin opening.
-Class III sites are inaccessible in WT cells and Zld or Grh binding results in chromatin opening.
-The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig. 1c,f).
-While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
+To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of pioneer factor bound sites (Fig. 1a,b).
+Class I sites are accessible in WT cells and remain accessible when bound by the ectopically expressed factor.
+Class II sites are inaccessible in WT cells.
+However, binding of Zld or Grh does not lead to chromatin opening.
+Class III sites are inaccessible in WT cells, but become accessible upon pioneer-factor binding.
+The majority of ChIP-seq peaks for Zld and Grh were class I sites.
+For each factor, only a subset of binding sites were closed chromatin, and at only a subset of these did the chromatin become accessible (Fig. 1c,f).
+While class I binding sites occured predominantly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
 Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
@@
 Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation.
-To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression.
-Differential expression analysis using DESeq2 [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. 3b,e).
+To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq.
+Differential expression analysis using DESeq2 to compare induced cells to WT cells without Zld or Grh expression [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig.
+3b,e).
 <!--# Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig.
@@
 3c).
-Genes activated by Zld expression were enriched for embryonic morphogenesis, consistent with the known function of Zld, as well as imaginal disc development, in which Zld has not been studied (Extended Data Fig.
+Genes activated by Zld expression were enriched for embryonic morphogenesis, consistent with the known function of Zld, as well as imaginal disc development, in which Zld-target genes have not been well characterized (Extended Data Fig.
 3f).
@@
 XX).
---> When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig.
+--> When we individually analyzed our three classes of binding sites for changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig.
 1j,l).
 This was particularly evident at class III promoters, which were most likely to correspond to upregulated genes.
-Together, we conclude that while binding and chromatin opening can happen in the absence of transcriptional changes, chromatin opening is correlated with transcriptional activation.
+Together, we conclude that while binding and chromatin opening were not always linked to changes in transcription, increases in chromatin accessibility are correlated with transcriptional activation.
 
-## Twist binds closed chromatin extensively and opens chromatin at a subset of sites
+## Twist binds extensively to closed chromatin and opens chromatin at a subset of sites
 
@@
   nrow()
+
+perc_twi_closed <- (twi_closed / nrow(twi_ChIP_classes) * 100) |> 
+  round(1)
 ```
@@
 
-Despite the differences between Zld and Grh during Drosophila development, both proteins behaved similarly when expressed in our exogenous S2 cell system.
-To provide additional context to our findings, we performed additional experiments with the transcription factor Twist (Twi).
-Although Twi is also essential for regulating Drosophila embryonic development, Twi has a number if distnict properties from Zld and Grh, making it an interesting point of comparison.
-Twi is a master regulator of mesodermal cell fate (REFS).
-Twi has many binding sites that are dynamic during early embryonic development [@Wilczynski2010a], in contrast to Zld and Grh.
-Although Twi was hypothesized to be a pioneer factor [@Sandmann2007], pioneering activity by Twi had not previously been tested.
-In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012].
-To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. 4a).
+Because both Zld and Grh had previously been shown to have features of pioneer factors, we sought to test an additional transcription factor with distinctive features.
+For this purpose, we tested the effects of the expression of the master regulator of mesodermal cell fate, Twist (Twi), which was not expressed in our WT S2 cells.
+Although like Zld and Grh Twi is also essential for regulating Drosophila embryonic development, Twi has a number of properties that are distinct.
+For example, many Twi-binding sites are developmentally dynamic and are not maintained through embryonic development [@Wilczynski2010a].
+Furthermore, Zld pioneering activity is required for Twi binding in the early embryo as mutations to Zld-binding sites disrupted Twi binding to the *cactus* enhancer [@Yáñez-Cuna2012].
+To test the properties of Twi in our system, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. 4a).
 We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
-Comparison of Twi ChIP-seq and ATAC-seq data allowed for the identification of the same classes of binding sites that were found for Zld and Grh (Fig 2a,b).
+. Using these data, we identified the same three classes of binding sites that we described for Zld and Grh (Fig 2a,b).
 Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
-Twi bound extensively to closed chromatin, with `r twi_closed`% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
-At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c; Extended Data Fig. 4b).
-Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4c).
+Twi bound extensively to closed chromatin, with `r perc_twi_closed`% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
+At a small subset of Twi-binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c; Extended Data Fig. 4b).
+RNA-seq revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4c).
 <!--# Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig.
@@
 XX).
---> Together, our data show that Twi can act as a pioneer factor.
+--> Thus, despite Twi possessing features distinct from Zld and Grh in vivo, ectopically expressed Twi could bind closed chromatin and promote accessibility.
 
@@
 
-While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites are shared across Zld, Grh, and Twi (Fig. 3a).
-This was in contrast to class II and III peaks, which were largely specific to each individual factor.
+While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites were shared across all three factors (Fig. 3a).
+This contrasted with class II and III peaks, which were largely specific to each individual factor.
 <!--# Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig.
 XX).
---> We hypothesized that these overlapping class I binding sites may occur at regulatory regions that are active in WT cells.
+--> We hypothesized that the overlapping class I binding sites may occur at active cis-regulatory regions that are active in WT cells.
 To test this, we analyzed published ChIP-seq data from WT S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV.
-Overlapping class I regions had high levels of chromatin marks associated with active regulatory regions (Fig. 3b).
-This enrichment of all three factors at highly active regions suggests that exogenously expressed factors may bind nonspecifically to active regions of the genome.
+Overlapping class I regions showed high levels of these chromatin marks (Fig. 3b), suggesting that exogenously expressed factors may bind nonspecifically to active regions of the genome.
 <!--# This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig.
@@
 
-Considering the enrichment of active chromatin at class I sites, we asked if there were any features of chromatin distinctive to class II or III regions.
+Considering the enrichment of marks of active chromatin at class I sites, we asked if there were any features of chromatin distinctive to class II or III regions.
 Analysis of published ChIP-seq data demonstrated that class II and III sites have low levels of many histone modifications, chromatin-associated proteins, and transcription factors (Extended Data Fig.
 5a,c,e).
-These data are in agreement with previous observations that Oct4, Sox2 and Klf4 access binding sites in naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
-Interestingly, levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. 5).
-Some class III sites also were within regions marked with the repressive histone modification H3K27me3 (Extended Data Fig. 5b,d,f).
-Together this suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin.
-Based on these observations, the inability to open class II regions is unlikely to be due to differences in preexisting features of chromatin.
+This supports the hypothesis that pioneer factors preferentially bind naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
+The ability of transcription-factor binding to promote chromatin accessibility at class III sites and not class II sites was not explained by a pre-existing lack of histones.
+In fact, the levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. 5).
+While most class III sites were regions of naïve chromatin, a subset were within regions marked with the repressive histone modification H3K27me3 (Extended Data Fig. 5b,d,f).
+Together this analysis suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin, and that the inability of these factors to promote chromatin accessibility at class II regions is unlikely to be due to differences in preexisting features of chromatin.
 
@@
 
-To further test the specificity of binding within the three classes of ChIP-seq peaks, we analyzed the enrichment for the canonical Zld, Grh or Twi motifs.
+To further determine the features that drive the different outcomes at pioneer-bound regions, we analyzed the enrichment for the canonical Zld, Grh or Twi motifs for each of the three bound classes.
 We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig. 3c).
-When compared to all accessible regions, class I regions had low levels of enrichment for the canonical motifs, supporting that binding to these regions is nonspecific.
+When compared to all accessible regions, class I regions had low levels of enrichment for the canonical motifs, supporting our prior analysis suggesting that binding to these regions is nonspecific.
 By contrast, class II and III sites were highly enriched for the canonical motifs (Fig. 3c).
@@
 For all three factors, class III sites tended to have stronger motifs than class II sites (Fig. 3g,i,k).
-Together, this analysis suggests that a combination of more motifs and stronger motifs at class III sites is what allows Zld, Grh, and Twi to open chromatin at these regions.
+Together, this analysis suggests that a combination of both a higher number of motifs and stronger motifs at class III sites enables Zld, Grh, and Twi to open chromatin at these regions.
 <!--# Add description of de novo motif enrichment, put data in supplement. -->
 
-Previous work has shown that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (REFs).
+Previous work showed that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (REFs).
 <!--# Add a few sentences here with more details on past work --> To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
@@
 This analysis is complicated by the fact that MNase-seq is a bulk assay and may not accurately represent nucleosome positioning in individual cells, especially at regions with poorly positioned nucleosomes.
-To more directly assess the positioning of nucleosomes around motifs, we analyzed a previously published single-molecule dataset using methyltransferase footprinting to determine the locations of nucleosomes on individual chromatin fibers [@Stergachis2020].
-We analyzed the distance from motifs at class II and III regions to the nearest nucleosome dyad.
-This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
-Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
+To more directly assess the positioning of nucleosomes around motifs, we analyzed a previously published single-molecule methyltransferase footprinting dataset to determine the locations of nucleosomes on individual chromatin fibers [@Stergachis2020].
+We analyzed the distance from motifs to the nearest nucleosome dyad at class II and III regions.
+It was clear that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
+Comparing class II and III regions did not identify any clear differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
 Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
-However, the lack of strong nucleosome positioning at these regions makes it difficult to fully assess the importance of motif positioning.
+However, the lack of strong nucleosome positioning at these regions made it challenging to fully assess the importance of motif positioning.
 
@@
 
-Prior studies have extensively explored the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
+Prior studies have extensively characterized the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
 To better understand how cell-type specific variables affect pioneer-factor binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
-We compared Zld binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
-We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig. 7a,b).
-Previous work has shown that regions resistant to pioneer-factor binding are enriched for the repressive histone modifications H3K9me3 and H3K27me3 [@Mayran2018a; @Soufi2012].
-To test if repressive marks were present at the resistant sites in S2 cells, we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
-Based on analysis of these datasets we divided the resistant sites into three additional classes of binding sites: class IV sites have high levels H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a; Extended Data Fig.
-7d-f).
-We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c; Extended Data Fig. 7e-l).
-We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017], and to Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
+We compared Zld-binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
+We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs (Nevil et al. 2017), and to Twi ChIP-seq data from early embryos (1-3 hours) (Ozdemir et al. 2011).
+For each factor, we identified many binding sites that were bound *in vivo*, but were not bound upon ectopic expression in S2 cells (Extended Data Fig. 7a,b).
+Chromatin enriched for repressive histone modifications H3K9me3 and H3K27me3 has been shown to be resistant to pioneer-factor binding [@Mayran2018a; @Soufi2012].
+To test if these repressive marks were present in S2 cells at the resistant sites (bound in vivo but not upon ectopic expression), we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
+Using these data, we divided the resistant sites into three classes: class IV sites have high levels of H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a-c; Extended Data Fig.
+7d-l).
 Despite the presence of class IV, V and VI binding sites for all three proteins, there were distinct patterns of binding across different tissues.
 Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
-For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig. 4b).
-Twi binding sites in S2 cells had low levels of binding in embryos, but many of the strongest embryonic binding sites were unbound.
+This differed for Grh, in which many of the strongest binding sites were shared across the three tissues (Fig. 4b).
+Twi binding was largely distinct in S2 cells and the embryo, with only a small subset of binding sites shared across the two tissues (Fig. 4c; Extended Data Fig. 7c).
 
 To test the functional relevance of repressive chromatin to pioneer-factor binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3.
-We treated our stable cell lines with DMSO or tazemetostat for 7 days, inducing Zld, Grh, or Twi expression on day 5.
-We confirmed loss of H3K27me3 by westerm blot and CUT&RUN [@Skene2017] (Extended Data Fig. XX).
-We used ChIP-seq to test if tazemetostat treatment resulted in changes to Zld, Grh, or Twi binding.
-Tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig. 4d-f).
-Despite a dramatic loss of H3K27me3, the majority of class IV binding sites remained unbound in tazemetostat-treated cells.
-We did identify some individual sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig. XX).
+Despite causing dramtic reductions in H3K27me3 levels, tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig.
+4d-f; Extended Data Fig.
+8a).
+The majority of class IV binding sites remained unbound in tazemetostat-treated cells, demonstrating that loss of PRC2 activity is insufficient to promoted widespread binding.
+Thus, H3K27me3 is broadly restricting binding in S2 cells.
+However, we identified a few sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig.
+XX).
 Together, these data demonstrate that despite the correlation of H3K27me3 with resistance to pioneer factor binding, loss of PRC2 activity is insufficient to promote widespread binding of Zld, Grh, or Twi to repressed binding sites.
@@
 
-Previous studies have proposed that pioneer factor activity may depend on protein concentration [@Hansen2022; @Blassberg2022].
+Previous studies proposed that pioneer-factor activity depends on protein concentration [@Hansen2022; @Blassberg2022].
 Considering that tazemetostat treatment had limited effects on pioneer factor binding, we asked if varying the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.
-We induced cells to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig. XX; see methods).
-We performed ChIP-seq and analyzed the data at our previously defined class I-VI binding sites.
-We found that binding and was concentration dependent (Fig. 5).
+We performed ChIP-seq on cells induced to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig.
+9; see methods).
+Analysis of our previously defined class I-VI binding sites showed that binding and was concentration dependent (Fig. 5).
 At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i).
-Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i).
-We also performed ATAC-seq to determine if concentration-dependent increases in binding led to changes in chromatin accessibility.
-Changes in chromatin accessibility were also concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig. 5b,f,j).
-Intriguingly, many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels.
-For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites (Fig. 5, Extended Data Fig. XX).
+Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig.
+5a,e,i).
+Thus, increased expression allowed pioneer factors access to previously unbound regions.
+To identify the relationship between pioneer-factor binding and chromatin accessibility, we performed ATAC-seq on cells induced with the same range in expression.
+Changes in chromatin accessibility were similarly concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig. 5b,f,j).
+Many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels, suggesting that increased chromatin occupancy, as assayed by ChIP-seq, could result in chromatin accessibility.
+For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites, which were not bound at lower concentrations (Fig. 5, Extended Data Fig. XX).
 To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration.
 For Zld and Twi, supraphysiological concentrations led to an increase in class III regions relative to class II regions (Fig. 5c,k).
-For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with the majority of these sites being falling in III (Fig. 5g).
+For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with most of these sites falling in III (Fig. 5g).
 To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs.
@@
 For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration.
-Together, these data demonstrate that both binding and chromatin opening by pioneer factors are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are restistant to binding at physiological levels.
+Together, these data demonstrate that both binding and chromatin opening by pioneer factors are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are resistant to binding at physiological levels.
 
@@
 
-Prior work has shown that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs).
+Prior work demonstrated that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs).
 However, little is known about how regions outside the DBD may contribute to pioneering function.
@@
 paper) is outside the DBD.
-See also new Zaret paper in NSMB looking at contribution of IDRs and activation domains to chromatin opening on nucleosome arrays in vitro -->.
+See also new Zaret paper in NSMB looking at contribution of IDRs and activation domains to chromatin opening on nucleosome arrays in vitro -->
 To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig.
-XX).
-Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was totally unable to bind to class II or III regions (Fig.
+10a,b).
+Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was unable to bind closed chromatin at class II or III regions (Fig.
 6a,c).
-We had previously shown that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
-Because many Zld binding sites in S2 cells also lack a canonical CAGGTA motif (Fig.
+We previously showed that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
+Because many Zld-binding sites in S2 cells also lack a canonical CAGGTA motif (Fig.
 3), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
@@
 For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
-Together, these experiments show that regions outside the DBD are essential for binding and opening of closed chromatin.
+Together, these experiments show that regions outside the DBD are essential for binding and opening closed chromatin.

### 2023-01-30 manuscript/manuscript_sections/_03_Results.qmd
add MMH changes

@@
 We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
-Additional ChIP-seq experiments were performed in WT cells using anti-Zld, or anti-Grh to control for antibody specificity.
+Additional ChIP experiments were performed in WT cells using anti-Zld, or anti-Grh to control for antibody specificity.
 By excluding peaks that were called in these datasets, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh.
@@
 The majority of ChIP-seq peaks for Zld and Grh were class I sites.
-For each factor, only a subset of binding sites were closed chromatin, and at only a subset of these did the chromatin become accessible (Fig. 1c,f).
+For each factor, only a subset of binding sites were within closed chromatin, and at only a subset of these did the chromatin become accessible (Fig. 1c,f).
 While class I binding sites occured predominantly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
@@
 1j,l).
-This was particularly evident at class III promoters, which were most likely to correspond to upregulated genes.
+This was particularly evident at class III promoters.
 Together, we conclude that while binding and chromatin opening were not always linked to changes in transcription, increases in chromatin accessibility are correlated with transcriptional activation.
@@
 For this purpose, we tested the effects of the expression of the master regulator of mesodermal cell fate, Twist (Twi), which was not expressed in our WT S2 cells.
-Although like Zld and Grh Twi is also essential for regulating Drosophila embryonic development, Twi has a number of properties that are distinct.
+Although like Zld and Grh, Twi is essential for regulating Drosophila embryonic development, Twi has a number of properties that are distinct.
 For example, many Twi-binding sites are developmentally dynamic and are not maintained through embryonic development [@Wilczynski2010a].
@@
 We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
-. Using these data, we identified the same three classes of binding sites that we described for Zld and Grh (Fig 2a,b).
+Using these data, we identified the same three classes of binding sites that we described for Zld and Grh (Fig 2a,b).
 Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).

### 2023-02-03 manuscript/manuscript_sections/_04_Discussion.qmd
update MS text

@@
 
-Major conclusions to emphasize:
+## outline:
 
--   Twist is a pioneer factor
+### Summary of study
 
--   PF activity occurs on a spectrum and is context-dependent.
+### Pioneer-factor activity occurs on a spectrum and is context dependent
 
--   Context-dependent PF activity can be regulated by both protein-intrinsic and extrinsic features.
+-   Twi is a PF
 
--   Preexisting active chromatin attracts TF binding, even with weaker motifs.
+-   In our study, we see a range of pioneering activities: Twi\<Zld\<Grh
 
--   Most functional binding occurs at canonical motifs in naïve chromatin
+### Pioneer factors are regulated by both protein-intrinsic and protein-extrinsic features Local chromatin.
 
--   Sequence composition is a main determinant of PF binding and activity
+-   **Active chromatin:**
 
--   Regions outside the DBD are essential for pioneering
+    -   promotes PF binding in the absence fo strong motifs.
+
+    -   H3K27me3, despite correlation, does not affect binding and pioneering
+
+-   **Protein concentration**:
+
+    -   Strong determinant of binding and chromatin opening in our study.
+
+    -   Different binding sites have different thresholds for binding and opening.
+        Some remain unbound even at high levels.
+
+-   **Cofactors**?
+
+-   Not explored extensively in our study.
+    How to discuss without inviting more experiments from reviewers?
+
+-   **Motif content**:
+
+    -   Strong predictor of binding and opening.
+
+-   **Protein domains**:
+
+    -   Required for stable binding and opening.
+
+    -   Allows uncoupling of binding from opening.
+
+### Model:
+
+-   PF activity requires high chromatin occupancy which can be achieved in a variety of ways.
+
+### Regions outside the DBD are essential for pioneering
+
+-   This and other recent evidence complicates the simple modular model of TF function
+
+-   Multiple possibilities for how IDRs or activation domains could impact pioneering:
+
+    -   multivalent interactions with nucleosomes stabilize binding (supported by in vitro evidence).
+
+    -   Recruitment/interaction with cofactors that stabilizes binding (chromatin opening via cofactors could provide positive feedback that stabilizes binding)
+
+    -   Allowing broader exploration of nuclear space, including repressed domains (supported by imaging data)
+
+    -   Clustering or phase separation that increases local concentration
+
+### Future directions we could mention:
+
+-   How do Zld, Grh, Twi and other PFs interact with nucleosomal targets?
+    TFs likely have diverse mechanisms of interaction, making it important to characterize for many factors.
+
+-   What are the mechanisms of chromatin opening for these factors?
+    Do they recruit remodelers?
+    Does binding alone destabilize nucleosomes?
+
+-   How do regions outside the DBD contribute to binding and opening?

### 2023-02-03 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update MS text
update outline of supplemental figures

@@
 
-**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.**
-
--   ~~Schematic of cell line generation.~~
-
--   ~~Western blots to determine physiological expression and lack of protein expression in WT cells.~~
-
--   ~~RNA-seq data to show leaky expression and strong induction of RNA levels.~~
+**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.
+a,** Schematic of generation of stable cell lines and induction of protein expression.
+**b,** mRNA levels of Zld and Grh in S2 cells.
+Top histogram shows the distribution of mRNA levels for all Drosophila genes.
+Vertical dashed line indicates a log~2~ RPKM value of 0 as a threshold for considering a gene to be expressed.
+**c-d,** Immunoblots showing titration of Zld **(c)** or Grh **(d)** protein levels in stable cell lines.
+Two independently generated cell lines are shown and compared to 2-3H embryos.
 
@@
 
+-   Western blots for determining DBD expression levels
+
+-   IF to confirm nuclear localization of DBDs.
+
 -   Analysis of Zld DBD ChIP only at CAGGTAG sites
@@
 open motifs bound, motif scores for accessible bound, accessible unbound, inaccessible bound, and inaccessible unbound.
-
-**Additional information to include in supplemental figures:**
-
--   De novo motif results
-
--   Analysis of promoter motifs
-
--   Analysis of deep learning models

### 2023-02-09 manuscript/manuscript_sections/_05_Methods.qmd
add CUT&RUN analysis to methods

@@
+---
+editor: 
+  markdown: 
+    wrap: sentence
+---
+
 ## Reagents
@@
 featureCounts was used to count the number of reads aligning within 200bp (100bp upstream or downstream) of peak summits.
+Generation of z-score normalized bigWig files was performed as described above for ChIP-seq data.
 Differential accessibility analysis was performed using DESeq2.
@@
 
-Trimming
-
-Alignment
-
-Filtering
-
-peak calling
-
-bigwigs
-
-z-score normalization
-
-spike-in normalization
+Trimming, alignment, and quality filtering of CUT&RUN reads was performed as described above for ATAC-seq data.
+Peak calling was performed using MACS2 with parameters --broad -f BAMPE --keep-dup all -g dm.
+Generation of z-score normalized bigWig files was performed as described above for ChIP-seq data.
+The percentage of reads containing barcoded Epicypher nucleosomes was calculated and 1 divided by this percentage was used as the scaling factor for spike-in normalization.
+The z-score normalized read depth was multiplied by these scaling factors to generate spike-in normalized bigWigs.
 

### 2023-02-10 manuscript/manuscript_sections/_02_Introduction.qmd
update MS discussion

@@
 
-Two well studied PFs that display context-dependent activity and Zelda (Zld) and Grainy head (Grh).
+Two well studied PFs that display context-dependent activity are Zelda (Zld) and Grainy head (Grh).
 Zld is essential for driving early embryonic development in Drosophila [@Liang2008; @Staudt2006].
@@
 The majority of Zld binding sites are unique to either the embryo or NSCs, with only a subset of regions being bound in both tissues.
-How Zld is directed to distinct binding sites in the embryo and NSCs in unknown.
+How Zld is directed to distinct binding sites in the embryo and NSCs is unknown.
 

### 2023-02-10 manuscript/manuscript_sections/_04_Discussion.qmd
update MS discussion

@@
+---
+format: html
+editor: visual
+bibliography: ../references.bib
+#csl: nature.csl
+---
+
 # Discussion
 
-## outline:
+In this study, we systematically explore chromatin binding and opening by the transcription factors Zld, Grh and Twi.
+Unlike many mammalian TFs, which have mostly been studied in cell culture, Zld, Grh and Twi have been extensively studied during normal Drosophila development.
+By comparing the results of TF induction in S2 cells to Zld, Grh and Twi activity during normal development, we are able to provide mechanistic insights that are relevant to the endogenous function of pioneer factors.
+Zld, Grh, and Twi displayed a range of abilities to bind and open closed chromatin, suggesting that pioneering activity occurs on a spectrum and is context dependent.
+Binding is separable from chromatin opening through the identification of class II sites that are bound by PFs but remain closed, or by the expression of the Grh DBD, which retains some binding but cannot promote chromatin opening.
+The separability of binding and opening suggests that additional downstream steps such as recruitment of chromatin remodelers are required for chromatin opening [@Judd2021; @King2017].
+Local restructuring of nucleosome-DNA interactions [@Michael2020; @Dodonova2020], if it occurs in the case of Zld, Grh, or Twi, may be insufficient for establishing NFRs and may instead stabilize PF binding to allow recruitment of cofactors.
+We propose that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features (Fig 7) such as local chromatin features, protein concentration, DNA motif content, and protein domains outside the DBD.
 
-### Summary of study
+Despite the ability of PFs to bind to closed chromatin, most PFs occupy only a subset of target motifs in the genome and PF binding and activity is often cell-type specific [@Larson2021].
+In many cases, the presence of active or repressive chromatin is an important determinant of transcription factor binding, but the extent to which this informs PF binding is unclear.
+For the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM), high levels of the repressive histone modification H3K9me3 mark regions that are resistant to OSKM binding.
+Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions [@Soufi2012].
+In the case of Pax7, a PF important for specification of pituitary melanotropes, H3K27me3 is correlated with cell-type specific Pax7 binding [@Mayran2019].
+Zld, Grh and Twi binding showed similar correlations with local chromatin structure.
+Even when expressed at low levels, all three proteins bind to active, accessible regions with degenerate motifs or no detectable motifs at all.
+At approximately physiological levels, these factors bind to canonical motifs in naïve chromatin, and are largely excluded from repressed chromatin (bearing H3K9me3 or H3K27me3).
+When we depleted H3K27me3 with the polycomb inhibitor tazemetostat, we observed no effect on binding within these repressed regions.
+In contrast to taz treatment, expression of Zld and Grh, but not Twi, at higher levels allowed at least some binding to these repressed regions.
+These results suggest that despite the correlation with H3K27me3, protein concentration and affinity for target motifs may be more important than the presence of polycomb silencing for tissue-specific binding and emphasize the importance of functionally testing the importance of different chromatin marks to PF binding.
+Cooperative binding with cell-type specific cofactors can also stabilize PF binding to tissue-specific binding sites [@Donaghey2018].
+This may also be important for Zld, Grh, and Twi, and may explain the presence of class VI sites that are resistant to binding and lack any repressive histone modifications.
+Further exploration of how chromatin structure and cofactor expression influence PF binding and activity will be essential to developing a more complete understanding of how these factors function in development and disease.
 
-### Pioneer-factor activity occurs on a spectrum and is context dependent
+The strong concentration dependence of chromatin opening by Zld, Grh and Twi provides important context for the *in vivo* biological functions of these proteins.
+Zld, along with many other TFs, forms high concentration foci within the nucleus [@Mir2018].
+These foci may facilitate chromatin opening by pioneer factors by creating a high protein concentration at specific genomic regions or locations within the nucleus.
+Grh has pioneering activity in late drosophila embryos and larval tissues, but not in early embryos [@Nevil2020; @Jacobs2018].
+This could, in part, be accounted for by differences in Grh expression between tissues.
+Additionally, GRHL2, one the mammalian Grh orthologs, can act as either a tumor suppressor or an oncogene, depending on the context [@Reese2019].
+Considering Grh expression levels in different tissues, cell types, and cancer types will be important for determining these contradictory effects.
+This is especially relevant considering that GRHL2 is frequently subject to copy number amplification in various cancers, leading to overexpression [@Reese2019].
+Pioneering activity by other mammalian factors depends on protein concentration [@Hansen2022; @Blassberg2022], suggesting that this is likely a general feature of PFs.
 
--   Twi is a PF
+In addition to DBDs that provide sequence specificity, many transcription factors contain long intrinsically disordered regions (IDRs) that are frequently required for transcriptional activation [@Minezaki2006].
+Fusing different IDRs to the same DBD can dramatically change which genomic regions are bound by transcription factors [@Brodsky2020].
+This observation complicates the textbook model of modular transcription factors in which the DBD determines sequence specificity and additional domains carry out transcriptional activation.
+Both Zld and Grh contain long disordered domains.
+The requirement of the Zld and Grh DBDs for robust binding to close chromatin demonstrates that regions outside the DBD, potentially the IDRs, are also important for binding site selection by pioneer factors.
+Others have suggested that IDRs may contribute to transcription-factor binding through protein-protein interactions with other transcription factors, nonspecific interactions with DNA, or by driving condensate formation [@Brodsky2021; @Staller2022].
+In the case of pioneer factors, another possibility is that IDRs form multivalent interactions with nucleosomes that stabilize binding.
+This model is supported by the observation that *in vitro*, full-length Zelda binds nucleosomes with a higher affinity than the DBD alone [@McDaniel2019a].
+Recent work showed that like Zld and Grh, the DBDs of human pioneer factors FOXA1 and SOX2 are required for robust binding of closed chromatin [@lerner2022].
+Additionally, single-molecule tracking indicated that the DBDs alone showed reduced scanning of silent chromatin within the nucleus.
+Zld and Grh DBDs still bound to class I regions, despite the weaker motifs present at these sites.
+This suggests that the less specific binding to class I regions is not due to interactions between IDRs and other proteins present at active chromatin.
+This binding may instead be driven by the high affinity of the DBDs for nucleosome-free DNA.
+Further exploration of how IDRs contribute to PF binding and chromatin accessibility will be essential for a more complete understanding of how *cis-*regulatory regions are established.
 
--   In our study, we see a range of pioneering activities: Twi\<Zld\<Grh
+-   discuss Zld and Grh data
 
-### Pioneer factors are regulated by both protein-intrinsic and protein-extrinsic features Local chromatin.
+-   mention Barkai and Staller reviews and models of IDR specificity: interactions with DNA, condensate formation, protein-protein interactions
 
--   **Active chromatin:**
+-   propose that interactions with nucleosomes is another model for how IDRs affect specificty.
+    Support this with Kath's EMSA data for Zld.
+    Cite Zaret papers as additional examples.
+    Could mention Giraldez data as well.
 
-    -   promotes PF binding in the absence fo strong motifs.
+-   Wrap up by mentioning that future directions could more fully explore how IDRs contribute to binding and specificity of PFs and non-PFs.
 
-    -   H3K27me3, despite correlation, does not affect binding and pioneering
+**P1: advantages of our study, summary of findings and our model**
 
--   **Protein concentration**:
+**P2: The effect of chromatin on binding: contextualize with references to past work by Zaret, Drouin, Mahony, and Giraldez and Onitchouk.
+Emphasize the importance of functionally testing correlative relationships.
+Mention potential importance of cofactors.**
 
-    -   Strong determinant of binding and chromatin opening in our study.
+**P3: the effect of concetration on binding.
+Mention that this could explain differences in Grh pioneering during development.
+Mention that Grh is overexpressed in cancers, so this has disease relevance.
+Mention Cohen studies.**
 
-    -   Different binding sites have different thresholds for binding and opening.
-        Some remain unbound even at high levels.
+P4: Discussion of DBD and conclusion.
+Discuss potential models and mention importance for understanding disease and reprogramming.
+Mention Zaret, Barkai and Scholer reprogramming paper.
+Concluding statement and any comments on future directions.
 
--   **Cofactors**?
-
--   Not explored extensively in our study.
-    How to discuss without inviting more experiments from reviewers?
-
--   **Motif content**:
-
-    -   Strong predictor of binding and opening.
-
--   **Protein domains**:
-
-    -   Required for stable binding and opening.
-
-    -   Allows uncoupling of binding from opening.
-
-### Model:
-
--   PF activity requires high chromatin occupancy which can be achieved in a variety of ways.
+##
 

### 2023-02-10 manuscript/manuscript_sections/_04_Discussion.qmd
update MS discussion

@@
 Further exploration of how IDRs contribute to PF binding and chromatin accessibility will be essential for a more complete understanding of how *cis-*regulatory regions are established.
-
--   discuss Zld and Grh data
-
--   mention Barkai and Staller reviews and models of IDR specificity: interactions with DNA, condensate formation, protein-protein interactions
-
--   propose that interactions with nucleosomes is another model for how IDRs affect specificty.
-    Support this with Kath's EMSA data for Zld.
-    Cite Zaret papers as additional examples.
-    Could mention Giraldez data as well.
-
--   Wrap up by mentioning that future directions could more fully explore how IDRs contribute to binding and specificity of PFs and non-PFs.
-
-**P1: advantages of our study, summary of findings and our model**
-
-**P2: The effect of chromatin on binding: contextualize with references to past work by Zaret, Drouin, Mahony, and Giraldez and Onitchouk.
-Emphasize the importance of functionally testing correlative relationships.
-Mention potential importance of cofactors.**
-
-**P3: the effect of concetration on binding.
-Mention that this could explain differences in Grh pioneering during development.
-Mention that Grh is overexpressed in cancers, so this has disease relevance.
-Mention Cohen studies.**
-
-P4: Discussion of DBD and conclusion.
-Discuss potential models and mention importance for understanding disease and reprogramming.
-Mention Zaret, Barkai and Scholer reprogramming paper.
-Concluding statement and any comments on future directions.
-
-##
-
-### Regions outside the DBD are essential for pioneering
-
--   This and other recent evidence complicates the simple modular model of TF function
-
--   Multiple possibilities for how IDRs or activation domains could impact pioneering:
-
-    -   multivalent interactions with nucleosomes stabilize binding (supported by in vitro evidence).
-
-    -   Recruitment/interaction with cofactors that stabilizes binding (chromatin opening via cofactors could provide positive feedback that stabilizes binding)
-
-    -   Allowing broader exploration of nuclear space, including repressed domains (supported by imaging data)
-
-    -   Clustering or phase separation that increases local concentration
-
-### Future directions we could mention:
-
--   How do Zld, Grh, Twi and other PFs interact with nucleosomal targets?
-    TFs likely have diverse mechanisms of interaction, making it important to characterize for many factors.
-
--   What are the mechanisms of chromatin opening for these factors?
-    Do they recruit remodelers?
-    Does binding alone destabilize nucleosomes?
-
--   How do regions outside the DBD contribute to binding and opening?

### 2023-02-10 manuscript/manuscript.qmd
update MS text
update manuscript draft
update manuscript

@@
 title: "Defining properties regulating pioneer-factor binding and function"
-format: html
+author:
+  - Tyler J. Gibson, Melissa M. Harrison
+  # - Melissa M. Harrison
+affiliations:
+  - name: Department of Biomolecular Chemistry, University of Wisconsin School of Medicine and Public Health
+    city: Madison
+    state: WI
+format:
+  docx:
+    reference-doc: custom-reference-doc.docx
 editor: visual
 bibliography: references.bib
-#csl: nature.csl
+csl: nsmb.csl
 ---
 
-# Abstract
+{{< pagebreak >}}
 
-{{< include manuscript_sections/_01_Abstract.qmd >}}
+<!-- # Abstract -->
+
+<!-- {{< include manuscript_sections/_01_Abstract.qmd >}} -->
+
+<!-- {{< pagebreak >}} -->
 
@@
 
+{{< pagebreak >}}
+
 # Materials and Methods
 
-{{< include manuscript_sections/_05_Discussion.qmd >}}
+{{< include manuscript_sections/_05_Methods.qmd >}}
+
+{{< pagebreak >}}
+
+# References
+
+::: {#refs}
+:::
+
+{{< pagebreak >}}
 

### 2023-02-10 manuscript/manuscript_sections/_03_Results.qmd
update MS text
update manuscript

@@
+---
+#bibliography: ../references.bib
+---
+
 ```{r}
@@
 By excluding peaks that were called in these datasets, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh.
-<!--# Not sure the best way to discuss the control ChIP experiments.
-Should I mention how many peaks we detect in WT cells?
-What supplemental figures should I include?
-Should peak calling be done using IgG or WT cells as control instead of input?
--->
 
@@
 3a,d).
-<!--# Add more information here about sites with differential accessibility.
-Include percentage of increased and decreased sites that are bound by Zld or Grh.
-Comment on how much is specific vs.
-secondary effect, and justify defining ChIP peak classes that include only a subset of DA sites.
--->
 
@@
 
-<!--# Could potentially add the cut&run time course here? Think about where to put this, or if it is worth including at all. -->
+Next, we sought to determine the binding dynamics of Zld and Grh.
+Zld and Grh protein was detectable at 4 or 12 hours following induction, respectively (Extended Data Fig. 2a-b).
+We used CUT&RUN [@Skene2017] to measure Zld and Grh binding to chromatin at 0, 4, 12, 24 or 48 hours following induction.
+Both proteins were bound to class I regions even in uninduced cells (0H), suggesting that these sites can be occupied even when protein levels are below the limit of detection of immunoblotting (Extended Data Fig 2c-d).
+Class II and III regions were bound at 4 hours, with slight increases in occupancy at later time points.
+These data show that Zld and Grh bind to closed chromatin rapidly following induction of protein expression.
 
@@
 3b,e).
-<!--# Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig.
-3b,e  include statistics), suggesting that binding of these factors mainly leads to transcriptional activation and not repression.
---> <!--# Consider mentioning Grh as a repressor here and giving examples of genes where Grh may be repressing.
-Discuss if Grh-mediated transcriptional repression coincides with chromatin opening, chromatin closing, or neither --> The genes activated by Grh were enriched for gene ontology (GO) terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig.
-3c).
+The genes activated by Grh were enriched for gene ontology (GO) terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig. 3c).
 Genes activated by Zld expression were enriched for embryonic morphogenesis, consistent with the known function of Zld, as well as imaginal disc development, in which Zld-target genes have not been well characterized (Extended Data Fig.
 3f).
-<!--# Add more detailed information on GO analysis here -->
 
@@
 Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l).
-<!--# Sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig.
-XX).
---> When we individually analyzed our three classes of binding sites for changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig.
-1j,l).
+When we individually analyzed our three classes of binding sites for changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
 This was particularly evident at class III promoters.
@@
 
-<!--# To do: Mention analysis of published Abd-A and Ubx cell lines. -->
-
 Because both Zld and Grh had previously been shown to have features of pioneer factors, we sought to test an additional transcription factor with distinctive features.
@@
 RNA-seq revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4c).
-<!--# Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig.
-XX).
---> GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig.
-4d).
-<!--# Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences.
-The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig.
-XX).
-Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig.
-XX).
-This was in contrast to Zld and Grh, where induction results in the activation of many genes that are not directly bound by Zld or Grh.
-This suggests that Zld and Grh may activate many transcription factors, which go on to activate their own targets, leading to secondary effects in the RNA-seq data.
-Twi, by contrast, appears to directly activate genes required for mesodermal and muscle cell fate.
---> <!--# Is this worth mentioning?
-Might require additional analysis of how many TFs are upregulated in RNA-seq data for each factor.
-Could also require additional motif analysis to identify/support secondary TFs and reason for nonspecific effects.
---> <!--# Similar to Zld and Grh, transcriptional activation was correlated with chromatin opening (Extended Data Fig.
-XX).
---> Thus, despite Twi possessing features distinct from Zld and Grh in vivo, ectopically expressed Twi could bind closed chromatin and promote accessibility.
+GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. 4d).
+Thus, despite Twi possessing features distinct from Zld and Grh in vivo, ectopically expressed Twi could bind closed chromatin and promote accessibility.
 
@@
 Together, this analysis suggests that a combination of both a higher number of motifs and stronger motifs at class III sites enables Zld, Grh, and Twi to open chromatin at these regions.
-<!--# Add description of de novo motif enrichment, put data in supplement. -->
 
-Previous work showed that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (REFs).
-<!--# Add a few sentences here with more details on past work --> To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
+Previous work showed that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes [@Zhu2018; @Michael2020].
+To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
 Analysis of average MNase-seq signal at class I-III peaks indicated that class III sites have high levels of nucleosome occupancy that, on average, occur over the respective motif (Extended Data Fig.
@@
 For class II and III peaks, some motifs were positioned close to the nuclesome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
-This analysis is complicated by the fact that MNase-seq is a bulk assay and may not accurately represent nucleosome positioning in individual cells, especially at regions with poorly positioned nucleosomes.
-To more directly assess the positioning of nucleosomes around motifs, we analyzed a previously published single-molecule methyltransferase footprinting dataset to determine the locations of nucleosomes on individual chromatin fibers [@Stergachis2020].
-We analyzed the distance from motifs to the nearest nucleosome dyad at class II and III regions.
-It was clear that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
-Comparing class II and III regions did not identify any clear differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
+Additionally, the MNase signal in many regions did not show strongly positioned nucleosomes, likely due to the fact that nucleosomes at silent regions often occur at different positions in different cells [@Stergachis2020; @Abdulhay2020].
 Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
-However, the lack of strong nucleosome positioning at these regions made it challenging to fully assess the importance of motif positioning.
 
@@
 Thus, H3K27me3 is broadly restricting binding in S2 cells.
-However, we identified a few sites where tazemetostat treatment lead to a loss of K27me3 and a concomitant gain of Zld, Grh or Twi, binding (Extended Data Fig.
-XX).
-Together, these data demonstrate that despite the correlation of H3K27me3 with resistance to pioneer factor binding, loss of PRC2 activity is insufficient to promote widespread binding of Zld, Grh, or Twi to repressed binding sites.
+These data demonstrate that despite the correlation of H3K27me3 with resistance to pioneer factor binding, loss of PRC2 activity is insufficient to promote widespread binding of Zld, Grh, or Twi to repressed binding sites.
 
@@
 At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i).
-Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig.
-5a,e,i).
+Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i).
 Thus, increased expression allowed pioneer factors access to previously unbound regions.
@@
 Many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels, suggesting that increased chromatin occupancy, as assayed by ChIP-seq, could result in chromatin accessibility.
-For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites, which were not bound at lower concentrations (Fig. 5, Extended Data Fig. XX).
+For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites, which were not bound at lower concentrations (Fig.
+5, Extended Data Fig.
+9d-f).
 To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration.
@@
 
-Prior work demonstrated that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs).
+Binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) [@Zaret2020].
 However, little is known about how regions outside the DBD may contribute to pioneering function.
-<!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat.
-Genet.
-paper) is outside the DBD.
-See also new Zaret paper in NSMB looking at contribution of IDRs and activation domains to chromatin opening on nucleosome arrays in vitro -->
 To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig.

### 2023-02-13 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update MS text

@@
 
-**Extended Data Figure 2: Zld and Grh bind to chromatin rapidly after induction of protein expression**
+**Extended Data Figure 2: Zld and Grh bind to chromatin rapidly after induction of protein expression.
+a-b,** Immunoblots showing time course of Zld **(a)** or Grh **(b)** protein expression following induction of stable cell lines.
+**c-d,** Metaplots showing average z-score normalized CUT&RUN signal at class I, II or III sites at different time points following induction.
 
--   ~~Western blot time course of protein levels~~
-
--   CUT&RUN time course to show rapid binding
-
--   likely rules out dependence on DNA replication
-
-**Extended Data Figure 3: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.**
-
--   volcano and/or MA plots with enrichment of binding
-
--   Analysis of nonspecific changes?
-
--   ~~RNA-seq volcano/MA plots~~
-
--   ~~RNA-seq GO analysis~~
-
--   Genome browser examples of specific and nonspecific accessibility changes?
-
--   motif enrichment for nonspecific changes?
-
--   \% of TFs bound and activated by Zld/Grh as indicator of nonspecific changes?
-
--   Should we include any analysis of decreased/downregulated sites for RNA-seq or ATAC?
-
--   Violin plots to emphasize correlation between ATAC and RNA-seq?
+**Extended Data Figure 3: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.
+a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
+**b,e,** RNA-seq volcano plots showing gene expression changes in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
+**c,f,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(c)** or Grh **(f)**.
 

### 2023-02-13 workflow/scripts/generate_fig5_RPKM_tables.R

@@ -318,7 +318,7 @@ rpkm_counts |>
   rownames_to_column("peak_id") |> 
   left_join(peak_class) |> 
   select(-peak_id) |> 
-  write_tsv("results/ChIP_tissue_classes/grh_classes_titration_ATAC_rpkm.tsv")
+  write_tsv("results/ChIP_tissue_classes/twi_classes_titration_ChIP_rpkm.tsv")
 
 # Twi ATAC =====================================================================
 # import unit names
@@ -373,18 +373,11 @@ rpkm_counts <- rpkm(select(read_counts, -c("peak_width")), widths = read_counts$
 peak_class <- annotation |> 
   select(GeneID,Chr, Start, End) |> 
   dplyr::rename(peak_id = "GeneID", seqnames = "Chr", start = "Start", end = "End") |> 
-  left_join(twi_tissue_occupancy) |> 
-  select(peak_id, class)
+  left_join(twi_tissue_occupancy)
 
-# plot boxplot
+# prepare output table
 rpkm_counts |> 
   rownames_to_column("peak_id") |> 
   left_join(peak_class) |> 
-  pivot_longer(-c("peak_id", "class"), names_to = "sample_name", values_to = "RPKM") |> 
-  mutate(sample_name = factor(sample_name, levels = c("titration_S2-HA-Twi_0uM","titration_S2-HA-Twi_10uM", "titration_S2-HA-Twi_40uM", "titration_S2-HA-Twi_160uM"))) |> 
-  ggplot(aes(x = sample_name, y = log2(RPKM))) +
-  geom_boxplot(fill = twi_color) +
-  theme_bw(base_size = 24) +
-  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
-  facet_grid(~class)
-
+  select(-peak_id) |> 
+  write_tsv("results/ChIP_tissue_classes/twi_classes_titration_ATAC_rpkm.tsv")

### 2023-02-20 manuscript/manuscript_sections/_04_Discussion.qmd
update manuscript
update manuscript draft

@@
 editor: visual
-bibliography: ../references.bib
-#csl: nature.csl
+#bibliography: ../references.bib
 ---
-
-# Discussion
 
@@
 Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions [@Soufi2012].
-In the case of Pax7, a PF important for specification of pituitary melanotropes, H3K27me3 is correlated with cell-type specific Pax7 binding [@Mayran2019].
+In the case of Pax7, H3K27me3 is correlated with cell-type specific Pax7 binding [@Mayran2019].
 Zld, Grh and Twi binding showed similar correlations with local chromatin structure.

### 2023-02-20 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update manuscript draft

@@
 
-**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation**
-
--   Western blot of Twi induction
-
--   Twi volcano and examples
-
--   ~~TWI RNA-seq volcano and examples3~~
-
--   ~~Twi RNA-seq GO analysis~~
-
--   Genome browser examples of specific and nonspecific accessibility changes?
-
--   motif enrichment for nonspecific changes?
-
--   \% of TFs bound and activated by Zld/Grh as indicator of nonspecific changes?
-
--   Should we include any analysis of decreased/downregulated sites for RNA-seq or ATAC?
-
--   Violin plots to emphasize correlation between ATAC and RNA-seq?
+**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
 

### 2023-02-21 manuscript/manuscript_sections/_05_Methods.qmd
update manuscript
update supp fig legends and add supp figs to snakemake pipeline

@@
 All analysis code is available on GitHub: https://github.com/tjgibson/S2_pioneers_manuscript.
-An archived version of the data processing and analysis pipeline is available on Zenodo: doiXXXX.

### 2023-02-21 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update supp fig legends and add supp figs to snakemake pipeline

@@
 **Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
+Protein levels in stable cell lines are compared to 3-4 hour old embryos.
+**b,** Volcano plots showing changes in ATAC-seq signal in cells expressing Twi when compared to WT cells treated with the same concentration of CuSO~4~.
+**c,** RNA-seq volcano plots showing gene expression changes in cells expressing Twi when compared to WT cells treated with the same concentration of CuSO~4~.
+**d,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Twi.
 
-**Extended Data Figure 5: Chromatin features associated with Zld, Grh and Twi binding sites**
+**Extended Data Figure 5: Chromatin features associated with Zld, Grh and Twi binding sites.
+a,c,e,** Heatmaps showing the levels of different chromatin marks in class I, II or III regions for Zld (a), Grh (c) or Twi (e).
+The color represents the average z-score normalized read depth across a 1 KB region surrounding the center of class I, II or III ChIP-seq peaks.
+**b,d,f,** Example genome browser tracks for class III regions with high levels of H3K27me3 for Zld (b), Grh (d), or Twi (f).
 
--   analysis of published ChIP datasets not shown in main figure
+**Extended Data Figure 6: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes.
+a-c,** Metaplots showing average MNase signal at motifs within class I, II, or III regions for Zld (a), Grh (b), or Twi (c).
+**d-f,** Heatmaps showing MNase signal at motifs within class II and III regions for Zld (d), Grh (e), or Twi (f).
+Rows are ordered based on hierarchical clustering to highlight the various patterns of MNase signal around motifs.
 
--   comparison of class I-III sites to show that class II-III sites are devoid of most histone mods
+**Extended Data Figure 7: Zld, Grh and Twi display cell-type specific binding.
+a-c,** Venn diagrams showing overlap between ChIP-seq peaks identified in different tissues for Zld (a), Grh (b) or Twi (c).
+**d-l,** Genome browser tracks showing examples of class IV, V, and VI regions for Zld (d-f), Grh (g-i) or Twi (j-l).
+For each example, the top tracks show H3K27me3 and H3K9me3 signal over a larger region.
+Dashed gray lines indicate a zoomed-in region where Zld, Grh or Twi ChIP-seq signal is shown in S2 cells or in embryos.
 
--   Class III sites have higher levels of H1 and H3 than class II sites.
-
-**Extended Data Figure 6: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes**
-
--   De novo motif enrichment for each factor
-
--   MNase data to show enrichment of nucleosomes around target motifs
-
--   Clustering of MNase data to show that there is no preferred nucleosome position relative to motifs
-
--   additional analysis of Fiber-seq data to show that nucleosomes at these regions are poorly positioned
-
-**Extended Data Figure 7: Zld, Grh and Twi display cell-type specific binding**
-
--   comparison to other tissues
-
--   analysis of motifs across tissues?
-
-**Extended Data Figure 8: PRC2 inhibition is insufficient to promote Zld, Grh, and Twi binding to repressed targets**
-
--   Examples of class III regions with H3K27me3
-
--   examples of class IV-VI regions
-
--   Spike-in validation of H3K27me3 antibody
-
--   Genome browsers showing loss of H3K27me3 in taz-treated cells without changes in binding.
-
--   Individual examples of sites that are gained in taz-treated cells
+**Extended Data Figure 8: PRC2 inhibition is insufficient to promote Zld, Grh, and Twi binding to repressed targets.
+a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
+Tubulin levels are shown as a loading control.
 

### 2023-02-22 manuscript/manuscript_sections/_06_Figure_Legends.qmd
Add legends for Supp figs 9 and 10

@@
-**Figure 1: Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
+**Figure 1: Ectopic pioneer factors open chromatin and activate transcription.
+a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
 ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
@@
 
-**Figure 3: Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
+**Figure 3: Motif content shapes pioneer-factor activity.
+a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
 **b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
@@
 
-**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
+**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation.
+a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
 Protein levels in stable cell lines are compared to 3-4 hour old embryos.
@@
 
-**Extended Data Figure 9: Expression of Zld, Grh, or Twi at supraphysiological levels results in novel binding sites**
+**Extended Data Figure 9: Expression of Zld, Grh, or Twi at supraphysiological levels results in chromatin opening at a small number of novel binding sites.
+a-c,** Immunoblots showing Zld (a), Grh (b) or Twi (c) protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
+**d-f,** Bar plots showing the percentage of previously defined class I-VI binding sites that are bound by Zld (d), Grh (e), or Twi (f) when expressed at varying concentrations.
+**g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld (g), Grh (h), or Twi (i) are expressed at varying concentrations.
 
--   Western blot showing protein levels for titration experiments
-
--   Examples of novel binding sites for each factor
-
--   motif enrichment at novel binding sites
-
--   Chromatin features at novel binding sites.
-
-**Extended Data Figure 10:**
-
--   Western blots for determining DBD expression levels
-
--   IF to confirm nuclear localization of DBDs.
-
--   Analysis of Zld DBD ChIP only at CAGGTAG sites
-
--   Examples of any novel DBD binding sites not detected for full-length proteins
-
--   FL and DBD preference for motifs that are in open vs.
-closed chromatin: % of closed vs.
-open motifs bound, motif scores for accessible bound, accessible unbound, inaccessible bound, and inaccessible unbound.
+**Extended Data Figure 10: Expression of Zld and Grh DNA-binding domains at protein levels comparable to the full-length proteins.
+a-b,** Immunoblots showing titration of protein levels for Zld (a) or Grh (b) DNA-binding domains to match expression of the full-length proteins.
+DNA-binding domain protein levels are shown at a range of CuSO~4~ concentrations and compared to an equivalent number of cells expressing full-length Zld or Grh at approximately physiological levels.
+**c,** Metaplots showing the average ChIP-seq signal for full-length Zld or the DNA-binding domain alone at the subset of previously identified class I, II or III binding sites that contain a canonical Zelda CAGGTAG motif.

### 2023-02-22 manuscript/manuscript_sections/_02_Introduction.qmd
update manuscript
update MS draft

@@
 ---
-bibliography: references.bib
+bibliography: ../references.bib
 ---
 
-# Introduction
-
-Transcription factors activate transcription by binding to specific DNA motifs and recruiting transcriptional machinery.
-The packaging of DNA into chromatin obstructs TFs from accessing their target motifs.
-Thus, many TFs can only bind to motifs that fall within nucleosome-free regions (NFRs).
-One mechanism for how NFRs are established is through the action of pioneer transcription factors (PFs).
-PFs are defined by their ability to recognize target motifs on nucleosomes.
-PFs bind to nucleosomal targets, promote chromatin opening, and create NFRs that allow additional TFs to bind and activate transcription[@Iwafuchi-Doi2016].
+During animal development, the totipotent embryo divides and differentiates to give rise to the diverse cell and tissue types of the adult organism.
+The establishment of cell-type specific gene expression programs is driven by transcription factors that bind sequence-specifically to *cis*-regulatory elements and recruiting transcriptional machinery.
+The packaging of DNA into chromatin obstructs TFs from accessing their target motifs [@Luger1997].
+Thus, many TFs only bind the small subset of genomic motifs that fall within nucleosome-free regions (NFRs) [@Li2011].
+A specialized subset of transcription factors, termed pioneer factors (PFs), are defined by their ability to bind target motifs on nucleosomes, promote chromatin opening, and create NFRs that allow additional TFs to bind and activate transcription [@Iwafuchi-Doi2016; @Zaret2020].
 Through this activity, PFs drive cell-type specific gene expression programs that determine cell fate.
@@
 Despite their ability to target inaccessible, nucleosomal binding sites, PFs do not bind to all genomic instances of their target motifs.
-Many PFs have also been shown to have cell-type specific patterns of binding and activity [@Soufi2012; @Donaghey2018].
+Many PFs have also been shown to have cell-type specific patterns of binding and activity [@Soufi2012; @Donaghey2018; @Chronis2017; @Buecker2014; @Cernilogar2019; @Lupien2008].
 How PF binding is restricted in different tissues or cellular contexts remains unclear.
 Many PFs bind primarily to naïve chromatin that is inaccessible, inactive, and lacks most histone modifications [@Soufi2012; @Donaghey2018].
-Cell-type specific binding is often associated with cell-type specific chromatin features [@Mayran2019], however many of these observations have been correlative, and this has not been extensively tested functionally.
+Cell-type specific binding is often associated with cell-type specific chromatin features [@Mayran2018a; @Lupien2008; @Cernilogar2019], however many of these observations have been correlative, and this has not been extensively tested functionally.
 In some cases, inhibition of various chromatin modifiers can enhance the ability of PFs to promote cellular reprogramming [@Larson2021a; @Soufi2012], but the molecular basis of how this relates to PF binding is unclear.
 Thus, a more systematic exploration of how PF binding and activity is regulated is essential to understanding how these factors function in development and disease.
+
+In addition to cell-type specific PF binding, the chromatin opening activity of PFs is also subject to regulation.
+Although PFs are defined by their ability to promote chromatin opening, many PFs are only required for chromatin accessibility at a subset of their binding sites.
+Upon depletion of the PFs Zelda, GAGA-factor, Oct4, or Sox2, chromatin accessibility is only lost at a subset of the binding sites for the respective factor [@Schulz2015; @Gaskill2021; @maresca2022].
+This suggests that PF binding does not necessarily translate to chromatin opening.
+Pax7, a PF important for specification of pituitary melanotropes, binds to thousands of regions where binding does not result in chromatin opening, which the authors term "pioneer primed" sites [@Mayran2018a].
+Additionally, a time course of Pax7 binding and chromatin accessibility indicated that there is a delay between Pax7 binding and chromatin opening.
+For GAGA-factor and Oct4, establishment of chromatin accessibility requires the activity of a nucleosome remodeler [@Judd2021; @King2017].
+In the case of Oct4, however, binding to nucleosomes does result in local remodeling of histone-DNA interactions, especially at the nucleosome entry-exit site [@Michael2020].
+Together, these observations highlight that the relationship between PF binding and activity is well understood.
 
@@
 
-The essential roles of Zld and Grh in diverse biological processes and the well documented context-dependent binding and activity of these factors makes them excellent models to explore how PF binding and activity are regulated.
+The essential roles of Zld and Grh in diverse biological processes and the well documented context-dependent binding and activity of these factors makes them excellent models to explore how PF binding is regulated during development, and how PF binding relates to chromatin opening.
 Because many previous studies have relied on loss-of-function experiments, it has been difficult to unravel causal relationships between chromatin structure and PF activity.
@@
 We propose that pioneering activity is not a binary feature intrinsic to a protein, but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type specific features.
-
-<!--# Should I discuss what is known about DBDs, structural features, and how PFs engage chromatin? -->

### 2023-02-23 manuscript/manuscript_sections/_01_Abstract.qmd
add abstract from EMBL txn meeting as draft abstract

Chromatin represents a substantial barrier to transcription-factor function.
While many transcription factors cannot bind to target DNA in the context of nucleosomes, pioneer transcription factors access nucleosomal targets and promote chromatin opening.
How pioneer factors navigate complex chromatin substrates is not well understood.
To better understand pioneer-factor function, we expressed the Drosophila pioneer factors Zelda (Zld) and Grainy head (Grh) in a cell type in which these factors are not normally expressed.
Zld and Grh bind both open and closed chromatin.
At many sites where they bind closed chromatin, binding leads to chromatin opening and transcriptional activation of proximal genes.
Analysis of sequence motifs, genomic features and histone modifications surrounding these binding sites revealed that Zld and Grh act as pioneer factors by binding to canonical motifs at sites that are mostly devoid of histone modifications.
Expression of the Zld or Grh DNA-binding domain (DBD) alone was insufficient to bind and open chromatin, suggesting that nonspecific interactions with nucleosomes mediated by domains outside of the canonical DBD are essential for pioneer function.
We additionally investigated properties of the putative non-pioneer factor Twist (Twi).
Contrary to prediction, Twi bound to thousands of inaccessible sites, but resulted in chromatin opening at a very small subset.
These results complicate the definition of a pioneer factor but suggest that factors likely employ separable strategies for binding and opening closed chromatin.
Finally, we examined sites that are bound by Zld, Grh and Twi in Drosophila embryos, but not upon exogenous expression and discovered that repressive histone modifications are enriched at these regions.
Our data identify unique features that shape pioneer-factor function and binding and suggest that pioneering is better defined by the capacity to open chromatin than by simply the ability to bind closed chromatin.

### 2023-02-23 workflow/scripts/extended_data_fig5.R

@@ -143,7 +143,7 @@ bw <- c(
   CTCF =snakemake@input[["CTCF_bw"]],
   CP190 = snakemake@input[["CP190_bw"]],
   Su_hw = snakemake@input[["Su_hw_bw"]],
-  Mod_mdg4 = snakemake@input[["Mod_mdg4_bw"]],
+  Mod_mdg4 = snakemake@input[["Mod_mdg4_bw"]]
   
 )
 

### 2023-02-24 manuscript/manuscript_sections/_02_Introduction.qmd
add abstract from EMBL txn meeting as draft abstract
incorporate MMH changes from 2023-02-23

@@
 During animal development, the totipotent embryo divides and differentiates to give rise to the diverse cell and tissue types of the adult organism.
-The establishment of cell-type specific gene expression programs is driven by transcription factors that bind sequence-specifically to *cis*-regulatory elements and recruiting transcriptional machinery.
+The establishment of cell-type specific gene expression programs is driven by transcription factors (TFs) that bind sequence-specifically to *cis*-regulatory elements.
 The packaging of DNA into chromatin obstructs TFs from accessing their target motifs [@Luger1997].
-Thus, many TFs only bind the small subset of genomic motifs that fall within nucleosome-free regions (NFRs) [@Li2011].
-A specialized subset of transcription factors, termed pioneer factors (PFs), are defined by their ability to bind target motifs on nucleosomes, promote chromatin opening, and create NFRs that allow additional TFs to bind and activate transcription [@Iwafuchi-Doi2016; @Zaret2020].
+Thus, many TFs only bind the small subset of genomic motifs that fall within regions of open chromatin [@Li2011].
+A specialized subset of transcription factors, termed pioneer factors (PFs), are defined by their ability to bind target motifs on nucleosomes and promote chromatin opening, allowing additional TFs to bind and activate transcription [@Iwafuchi-Doi2016; @Zaret2020].
 Through this activity, PFs drive cell-type specific gene expression programs that determine cell fate.
@@
 
-Despite the importance of PFs to these diverse biological processes, how PFs select their binding sites in different biological contexts remains incompletely understood.
-Despite their ability to target inaccessible, nucleosomal binding sites, PFs do not bind to all genomic instances of their target motifs.
-Many PFs have also been shown to have cell-type specific patterns of binding and activity [@Soufi2012; @Donaghey2018; @Chronis2017; @Buecker2014; @Cernilogar2019; @Lupien2008].
+Despite their ability to target closed, nucleosomally occupied binding sites, PFs do not bind to all genomic instances of their target motifs.
+Many PFs bind primarily to naïve chromatin that is inaccessible, inactive, and lacks most histone modifications [@Soufi2012; @Donaghey2018].
+In addition, PFs have also been shown to have cell-type specific patterns of binding and activity [@Soufi2012; @Donaghey2018; @Chronis2017; @Buecker2014; @Cernilogar2019; @Lupien2008].
 How PF binding is restricted in different tissues or cellular contexts remains unclear.
 Many PFs bind primarily to naïve chromatin that is inaccessible, inactive, and lacks most histone modifications [@Soufi2012; @Donaghey2018].
-Cell-type specific binding is often associated with cell-type specific chromatin features [@Mayran2018a; @Lupien2008; @Cernilogar2019], however many of these observations have been correlative, and this has not been extensively tested functionally.
-In some cases, inhibition of various chromatin modifiers can enhance the ability of PFs to promote cellular reprogramming [@Larson2021a; @Soufi2012], but the molecular basis of how this relates to PF binding is unclear.
-Thus, a more systematic exploration of how PF binding and activity is regulated is essential to understanding how these factors function in development and disease.
+Cell-type specific binding is often associated with distinct chromatin features [@Mayran2018a; @Lupien2008; @Cernilogar2019].
+Together, these studies indicate that PF occupancy is regulated by both genomic context and cell-type specific features.
+Nonetheless, the mechanisms that govern PF occupancy and activity remain largely unknown.
 
-In addition to cell-type specific PF binding, the chromatin opening activity of PFs is also subject to regulation.
 Although PFs are defined by their ability to promote chromatin opening, many PFs are only required for chromatin accessibility at a subset of their binding sites.
@@
 In the case of Oct4, however, binding to nucleosomes does result in local remodeling of histone-DNA interactions, especially at the nucleosome entry-exit site [@Michael2020].
-Together, these observations highlight that the relationship between PF binding and activity is well understood.
+Together, these observations highlight that the relationship between PF binding and activity is not well understood.
 
-Two well studied PFs that display context-dependent activity are Zelda (Zld) and Grainy head (Grh).
+To begin to elucidate the factors that reuglate PF binding and activity, we focused on two well-studied PFs that control conserved developmental transitions, Zelda (Zld) and Grainy head (Grh).
 Zld is essential for driving early embryonic development in Drosophila [@Liang2008; @Staudt2006].
-Immediately following fertilization, the zygotic genome is transcriptionally silent and development is controlled by maternally deposited mRNAs and proteins.
+Immediately following fertilization, the zygotic genome is transcriptionally silent, and development is controlled by maternally deposited mRNAs and proteins.
 Maternal products are gradually degraded as the zygotic genome is activated during a process known as the maternal-to-zygotic transition (MZT) [@Schulz2019; @Vastenhouw2019].
-In Drosophila, the pioneering activity Zld is essential for driving the MZT[@Liang2008; @Schulz2015; @Sun2015; @McDaniel2019a].
-Zld was the first discovered activator of the zygotic genome, and work on Zld paved the way for the identification of analogous proteins in other species[@Lee2013; @Leichsenring2013; @Gassler2022].
-Our lab recently identified an additional function of Zld in promoting neural stem cell (NSC) fate in the developing larval brain[@Larson2021a].
-The majority of Zld binding sites are unique to either the embryo or NSCs, with only a subset of regions being bound in both tissues.
-How Zld is directed to distinct binding sites in the embryo and NSCs is unknown.
+In Drosophila, the pioneering activity of Zld is essential for driving activation of the zygotic genome [@Liang2008; @Schulz2015; @Sun2015; @McDaniel2019a].
+Since the discovery of Zld as an activator of the initial wave of zygotic transcription, pioneer factors have been shown to drive the transition in all other species studied to date [@Lee2013; @Leichsenring2013; @Gassler2022; @Gentsch2019; @Charney2017].
+Similar to the reprogramming role of Zld in the early embryo, Zld promotes the neural stem cell (NSC) fate in the developing larval brain[@Larson2021a].
+Despite the shared ability to promote the undifferentiated fate in both tissues, the majority of Zld-binding sites are unique to either the embryo or NSCs, with only a subset of regions being bound in both tissues.
+By contrast, Grh-binding sites are largely shared between tissues [@Nevil2017].
+Grh is a master regulator of epithelial cell fate that is conserved among metazoans and functions as a pioeer factor in both *Drosophila* larva and mammalian cell culture [@Wang2012; @Jacobs2018; @Nevil2020].
+However, in the early *Drosophila* embryo Grh is dispensible for chromatin accessibility, and only displays pioneering activity in late embryos or larval tissues.
+Thus, Grh pioneering activity, but not Grh binding is regulated over development.
 
-Grh is a master regulator of epithelial cell fate that is conserved among metazoans [@Wang2012].
-In Drosophila, Grh acts as a pioneer factor in late embryos and larval tissues [@Jacobs2018; @Nevil2020].
-Unlike Zld, Grh binding is similar across different tissues and developmental stages[@Nevil2017].
-Despite this stable binding, Grh is dispensible for chromatin accessibility in the early embryo, and only displays pioneering activity in late embryos or larval tissues.
-This intriguing combination of stable binding and cell-type specific activity raises questions about how pioneering by Grh is regulated during development.
-Additionally, overexpression of Grh is a hallmark of many epithelial cancers.
-The role of Grh in cancer is not well understood, as Grh can either inhibit or promote tumorigenesis, depending on the context[@Reese2019].
-
-The essential roles of Zld and Grh in diverse biological processes and the well documented context-dependent binding and activity of these factors makes them excellent models to explore how PF binding is regulated during development, and how PF binding relates to chromatin opening.
+The essential roles of Zld and Grh in diverse biological processes and the well-documented context-dependent binding and activity of these factors makes them excellent models to explore how PF binding is regulated during development, and how PF occupancy relates to chromatin opening.
 Because many previous studies have relied on loss-of-function experiments, it has been difficult to unravel causal relationships between chromatin structure and PF activity.
 We therefore expressed Zld and Grh in Drosophila S2 cells, where neither factor is endogenously expressed.
-We demonstrate that when expressed ectopically, Zld and Grh bind to closed chromatin, establish NFRs, and promote transcriptional activation.
-We show that an additional TF, Twist (Twi), can act as a pioneer factor.
-Pioneering by Zld, Grh, and Twi depends on a combination of local chromatin structure, sequence context, and PF concentration and requires protein regions outside the DNA-binding domain.
+When ectopically expressed, Zld and Grh bind to closed chromatin, establish open chromatin, and promote transcriptional activation at a subset of binding sites.
+We show that an additional TF, Twist (Twi), can function as a pioneer factor.
+By comparing the features of these three factors and relating exogenous activity to *in vivo* functionality, we demonstrate that pioneering depends on a combination of local chromatin structure, sequence context, and PF concentration.
 We propose that pioneering activity is not a binary feature intrinsic to a protein, but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type specific features.

### 2023-02-24 workflow/scripts/extended_data_fig2.R

@@ -9,29 +9,48 @@ library(EBImage)
 source("workflow/scripts/plot_heatmap.R")
 
 # define input files ===========================================================
-# Zld_ChIP_bw <- snakemake@input[["Zld_ChIP_bw"]]
+# specify input files explicitly for testing script interactively
+# S2_Zld_CR_0H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_0H_small.bw"
+# S2_Zld_CR_4H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_4H_small.bw"
+# S2_Zld_CR_12H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_12H_small.bw"
+# S2_Zld_CR_24H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_24H_small.bw"
+# S2_Zld_CR_48H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_48H_small.bw"
+# 
+# S2_Grh_CR_0H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_0H_small.bw"
+# S2_Grh_CR_4H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_4H_small.bw"
+# S2_Grh_CR_12H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_12H_small.bw"
+# S2_Grh_CR_24H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_24H_small.bw"
+# S2_Grh_CR_48H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_48H_small.bw"
 
-S2_Zld_CR_0H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Zld_aZld_0H_accessble_zscore.bw"
-S2_Zld_CR_4H_bw <- "data/2021-03_CR/results/bigwigs/z_score_normalized/S2-Zld_aZld_4H_accessble_zscore.bw"
[…]

### 2023-02-24 workflow/scripts/extended_data_fig4.R

@@ -10,15 +10,17 @@ library(rtracklayer)
 library(EBImage)
 
 # define input files ===========================================================
-# specify input files explicitly for testing 
+# specify input files explicitly for testing script interactively 
 # twi_RNAseq_results_fn <- "RNAseq/results/DEseq2/S2-Twi_RNAseq_S2-Twi-vs-S2-WT_results_annotated.tsv"
 # twi_ATAC_results_fn <- "ATACseq/results/DEseq2_results_filtered/S2-Twi_ATACseq_S2-Twi-vs-S2-WT-40uM_results.tsv"
 # twi_ChIP_peaks_fn <- "ChIPseq/results/peaks/final/S2-Twi_aTwi_IP.narrowPeak"
+# twi_blot_image <- "data/immunoblot_raw_images/2021-9-13_Twi-vs-embryo/Twi-vs-embryos.tif"
 
 # get input files from snakemake
 twi_RNAseq_results_fn <- snakemake@input[["twi_RNAseq_results_fn"]]
-twi_ATAC_results_fn <- snakemake@input[["twi_RNAseq_results_fn"]]
+twi_ATAC_results_fn <- snakemake@input[["twi_ATAC_results_fn"]]
 twi_ChIP_peaks_fn <- snakemake@input[["twi_ChIP_peaks_fn"]]
+twi_blot_image <- snakemake@input[["twi_blot_image"]]
 
 
 # # create blank layout for plot ===============================================
@@ -48,7 +50,7 @@ plotText(
 )
 
 # read in tiff of western blot
-blot_image <- readImage("data/immunoblot_raw_images/2021-9-13_Twi-vs-embryo/Twi-vs-embryos.tif")
+blot_image <- readImage(twi_blot_image)
 
 # rotate image
 blot_image <- blot_image |> 

### 2023-03-01 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update supp figures

@@
-**Figure 1: Ectopic pioneer factors open chromatin and activate transcription.
-a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
+**Figure 1: Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
 ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
@@
 
-**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation.
-a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
+**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
 Protein levels in stable cell lines are compared to 3-4 hour old embryos.
@@
 DNA-binding domain protein levels are shown at a range of CuSO~4~ concentrations and compared to an equivalent number of cells expressing full-length Zld or Grh at approximately physiological levels.
-**c,** Metaplots showing the average ChIP-seq signal for full-length Zld or the DNA-binding domain alone at the subset of previously identified class I, II or III binding sites that contain a canonical Zelda CAGGTAG motif.
+**c-d,** Immunofluorescent microscopy images of stable cell lines expressing full-length protein or DBD only for Zld (c) or (Grh).
+Stable cell lines are compared to WT cells.
+**e,** Metaplots showing the average ChIP-seq signal for full-length Zld or the DNA-binding domain alone at the subset of previously identified class I, II or III binding sites that contain a canonical Zelda CAGGTAG motif.

### 2023-03-01 workflow/scripts/extended_data_fig1.R

@@ -8,19 +8,56 @@ library(grid)
 library(RColorBrewer)
 library(EBImage)
 
+source("workflow/scripts/plot_heatmap.R")
 
 # define input files ===========================================================
-# for testing script
+# define input files explicitly for interactve testing
 # RPKM_table_fn <- "RNAseq/results/count_tables/S2-Grh_RNAseq_RPKM.tsv"
 # 
 # zld_blot_image <- "data/immunoblot_raw_images/2018-10-17_Zld_induction/2018-1018-144408_pub.tif"
 # grh_blot_image <- "data/immunoblot_raw_images/2018-11-08_Grh_induction/2018-1108-132834_pub.tif"
+
+# S2_Zld_aZld_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_IP.bw"
+# S2_WT_aZld_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-WT_aZld_IP.bw"
+# S2_Zld_IgG_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2_Zld_aIgG_IP.bw"
+# 
+# zld_ChIP_classes_fn <- "results/ChIP_peak_classes/zld_ChIP_classes.tsv"
+# 
+# S2_Grh_aGrh_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_IP.bw"
+# S2_WT_aGrh_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-WT_aGrh_IP.bw"
+# S2_Grh_IgG_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2_Grh_aIgG_IP.bw"
+# 
+# grh_ChIP_classes_fn <- "results/ChIP_peak_classes/grh_ChIP_classes.tsv"
+# 
+# S2_Twi_aTwi_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Twi_aTwi_IP.bw"
+# S2_WT_aTwi_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-WT_aTwi.bw"
[…]

### 2023-03-01 workflow/scripts/fig4.R

@@ -45,20 +45,20 @@ wing_disc_Grh_ChIP_bw <-  snakemake@input[["wing_disc_Grh_ChIP_bw"]]
 S2_Twi_ChIP_bw <-  snakemake@input[["S2_Twi_ChIP_bw"]]
 embryo_Twi_ChIP_bw <-  snakemake@input[["embryo_Twi_ChIP_bw"]]
 
-S2_Zld_ChIP_DMSO <- snakemake@input[["S2_Zld_ChIP_DMSO_bw"]]
-S2_Zld_ChIP_taz  <- snakemake@input[["S2_Zld_ChIP_taz_bw"]]
-S2_Zld_H3K27me3_DMSO <-  snakemake@input[["S2_Zld_H3K27me3_DMSO_bw"]]
-S2_Zld_H3K27me3_taz <-  snakemake@input[["S2_Zld_H3K27me3_taz_bw"]]
+S2_Zld_ChIP_DMSO_bw <- snakemake@input[["S2_Zld_ChIP_DMSO_bw"]]
+S2_Zld_ChIP_taz_bw  <- snakemake@input[["S2_Zld_ChIP_taz_bw"]]
+S2_Zld_H3K27me3_DMSO_bw <-  snakemake@input[["S2_Zld_H3K27me3_DMSO_bw"]]
+S2_Zld_H3K27me3_taz_bw <-  snakemake@input[["S2_Zld_H3K27me3_taz_bw"]]
 
-S2_Grh_ChIP_DMSO <-   snakemake@input[["S2_Grh_ChIP_DMSO_bw"]]
-S2_Grh_ChIP_taz <-  snakemake@input[["S2_Grh_ChIP_taz_bw"]]
-S2_Grh_H3K27me3_DMSO <-  snakemake@input[["S2_Grh_H3K27me3_DMSO_bw"]]
-S2_Grh_H3K27me3_taz <-  snakemake@input[["S2_Grh_H3K27me3_taz_bw"]]
+S2_Grh_ChIP_DMSO_bw <-   snakemake@input[["S2_Grh_ChIP_DMSO_bw"]]
+S2_Grh_ChIP_taz_bw <-  snakemake@input[["S2_Grh_ChIP_taz_bw"]]
+S2_Grh_H3K27me3_DMSO_bw <-  snakemake@input[["S2_Grh_H3K27me3_DMSO_bw"]]
+S2_Grh_H3K27me3_taz_bw <-  snakemake@input[["S2_Grh_H3K27me3_taz_bw"]]
[…]

### 2023-03-01 workflow/scripts/utils.R

@@ -175,3 +175,54 @@ rpkm <- function(count_table, widths) {
   return(all_rpkm)
   
 }
+
+# define function for processing IF images =====================================
+# x: named list of EBImage Image objects corresponding to fluorescence channels
+# coordinates: named list of x,y coordinates for cropping images. e.g coordinates = list(x=1:100, y=1:100)
+# normalize: if true, will apply linear histogram normalization
+# intensity range: list with same names as x. Each element should be a numeric vector that will be passed to EBImage::normalize as the inputRange parameter
+process_IF_images <- function(x, coordinates = NULL, normalize = FALSE, intensity_range = NULL) {
+  require(EBImage)
+  # check input files
+  if (is.null(names(x))) stop("x should be a named list, set names for x")
+  
+  if (!all(unlist(lapply(images,class)) == "Image")) stop("x should be a list of Image objects")
+  
+  if (!is.null(coordinates)) {
+    if (!(identical(names(coordinates),c("x","y")) | identical(names(coordinates),c("y","x")))) stop("coordinates should be a named list with elements 'x' and 'y'")
+  }
+  
+  if (!is.null(intensity_range)) {
+    if (!identical(names(x), names(intensity_range))) stop("`intensity_range` should have same names as 'x'")
+  }
+  
+  # crop images
+  if (!is.null(coordinates)) {
+    for (channel in names(x)) {
+      # crop image to show cell of interest
+      x[[channel]] <- x[[channel]][coordinates$x,coordinates$y,]
+    }
+  }
+  
+  # convert images to grayscale
+  if (!is.null(coordinates)) {
+    for (channel in names(x)) {
+      # crop image to show cell of interest
+      x[[channel]] <- channel(x[[channel]], mode = "gray")
+    }
+  }
+  
+  # perform histogram normalization
+  if (normalize) {
+    if(is.null(intensity_range)) {
+      for (channel in names(x)) {
+        x[[channel]] <- normalize(x[[channel]])
+      }
+    } else {
+      for (channel in names(x)) {
+        x[[channel]] <- normalize(x[[channel]], inputRange = intensity_range[[channel]])
+      }
+    }
+  }  
+  return(x) 
+}

### 2023-03-02 manuscript/manuscript_sections/_01_Abstract.qmd
update abstract draft

@@
-Chromatin represents a substantial barrier to transcription-factor function.
+During development, transcription factors establish cell-type specific gene expression programs.
 While many transcription factors cannot bind to target DNA in the context of nucleosomes, pioneer transcription factors access nucleosomal targets and promote chromatin opening.
-How pioneer factors navigate complex chromatin substrates is not well understood.
-To better understand pioneer-factor function, we expressed the Drosophila pioneer factors Zelda (Zld) and Grainy head (Grh) in a cell type in which these factors are not normally expressed.
-Zld and Grh bind both open and closed chromatin.
-At many sites where they bind closed chromatin, binding leads to chromatin opening and transcriptional activation of proximal genes.
-Analysis of sequence motifs, genomic features and histone modifications surrounding these binding sites revealed that Zld and Grh act as pioneer factors by binding to canonical motifs at sites that are mostly devoid of histone modifications.
-Expression of the Zld or Grh DNA-binding domain (DBD) alone was insufficient to bind and open chromatin, suggesting that nonspecific interactions with nucleosomes mediated by domains outside of the canonical DBD are essential for pioneer function.
-We additionally investigated properties of the putative non-pioneer factor Twist (Twi).
-Contrary to prediction, Twi bound to thousands of inaccessible sites, but resulted in chromatin opening at a very small subset.
-These results complicate the definition of a pioneer factor but suggest that factors likely employ separable strategies for binding and opening closed chromatin.
-Finally, we examined sites that are bound by Zld, Grh and Twi in Drosophila embryos, but not upon exogenous expression and discovered that repressive histone modifications are enriched at these regions.
-Our data identify unique features that shape pioneer-factor function and binding and suggest that pioneering is better defined by the capacity to open chromatin than by simply the ability to bind closed chromatin.
+Despite the ability to bind to nucleosomal target motifs, many pioneer factors display cell-type specific binding and activity.
+It is unclear how pioneer factor occupancy is regulated in different cell types during development and how chromatin occupancy relates to chromatin opening activity.
+To determine how pioneer-factor binding and activity are regulated, we systematically studied three Drosophila transcription factors with distinct DNA-binding domains and biological functions: Zelda, Grainy head, and Twist.
+We demonstrated that the level of chromatin occupancy is a key determinant of pioneer-factor activity.
+Zld, Grh and Twi occupancy are regulated by a numerous factors including motif content at binding sites, local chromatin context, and protein concentration.
+We found that regions outside the DNA-binding domain are required for pioneer-factor binding and chromatin opening.
+Together, our results demonstrate that both protein-intrinsic and cell-type specific factors regulate pioneer-factor occupancy and activity.
+These findings have important implications for how pioneer factors drive development and how misexpression can lead to disease.

### 2023-03-06 manuscript/manuscript.qmd
update MS draft
update abstract draft

@@
 ---
-title: "Defining properties regulating pioneer-factor binding and function"
+title: "Protein-intrinsic properties and epigenetic effects regulate pioneer-factor binding and function"
 author:
@@
 
+Department of Biomolecular Chemistry, University of Wisconsin School of Medicine and Public Health, Madison, WI
+
 {{< pagebreak >}}
 
-<!-- # Abstract -->
+# Abstract
+{{< include manuscript_sections/_01_Abstract.qmd >}}
 
-<!-- {{< include manuscript_sections/_01_Abstract.qmd >}} -->
-
-<!-- {{< pagebreak >}} -->
+{{< pagebreak >}}
 

### 2023-03-06 manuscript/manuscript_sections/_01_Abstract.qmd
update MS draft

@@
-During development, transcription factors establish cell-type specific gene expression programs.
-While many transcription factors cannot bind to target DNA in the context of nucleosomes, pioneer transcription factors access nucleosomal targets and promote chromatin opening.
-Despite the ability to bind to nucleosomal target motifs, many pioneer factors display cell-type specific binding and activity.
-It is unclear how pioneer factor occupancy is regulated in different cell types during development and how chromatin occupancy relates to chromatin opening activity.
+Transcription factors establish cell-type specific gene expression programs to drive developmental transitions.
+While chromatin presents a barrier to the binding of many transcription factors, pioneer factors access nucleosomal targets and promote chromatin opening.
+Despite the ability to bind to target motifs in closed chromatin, many pioneer factors display cell-type specific binding and activity.
+The mechanisms that govern pioneer-factor occupancy in different cell types and the relationship between chromatin occupancy and chromatin opening remain unclear.
 To determine how pioneer-factor binding and activity are regulated, we systematically studied three Drosophila transcription factors with distinct DNA-binding domains and biological functions: Zelda, Grainy head, and Twist.
 We demonstrated that the level of chromatin occupancy is a key determinant of pioneer-factor activity.
-Zld, Grh and Twi occupancy are regulated by a numerous factors including motif content at binding sites, local chromatin context, and protein concentration.
-We found that regions outside the DNA-binding domain are required for pioneer-factor binding and chromatin opening.
-Together, our results demonstrate that both protein-intrinsic and cell-type specific factors regulate pioneer-factor occupancy and activity.
+Zelda, Grainy head and Twist occupancy is regulated by multiple factors, including motif content at binding sites, local chromatin context, and protein concentration.
+We found that regions outside the DNA-binding domain are required for binding and chromatin opening.
+Together, our results demonstrate that pioneering activity is not a binary feature intrinsic to a protein but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type-specific features.
 These findings have important implications for how pioneer factors drive development and how misexpression can lead to disease.

### 2023-03-06 manuscript/manuscript_sections/_03_Results.qmd
update MS text
update MS draft

@@
 ---
-#bibliography: ../references.bib
+# bibliography: ../references.bib
 ---
@@
 Additional ChIP experiments were performed in WT cells using anti-Zld, or anti-Grh to control for antibody specificity.
-By excluding peaks that were called in these datasets, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh.
+By using these datasets as controls for peak calling, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh (Extended Data Fig.
+1e-g).
 
@@
 Together, these data show that when expressed exogenously at physiolgical levels, Zld and Grh can bind and open closed chromatin at a subset of binding sites.
+
+::: {#fig-1}
+![](figures/fig1.pdf)
+
+**Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
+ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
+**c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
+**d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
+**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h).
+i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
+**j,i,** Proportion of class I, II or III regions that are proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
+Promoter-proximal and promoter-distal binding sites are shown separately.
+The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene.
+:::
 
@@
 
+::: {#fig-2}
+![](figures/fig2.pdf)
+
+**Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.
+a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites.
+**c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions.
+**d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.
+:::
+
 ## Ectopically expressed transcription factors bind opportunistically to active chromatin
@@
 -->
+
+::: {#fig-3}
+![](figures/fig3.pdf)
+
+**Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
+**b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
+Zld, Grh, and Twi ChIP-seq data were generated in this study.
+Remaining ChIP-seq datasets were previously published (see Supplementary table XX).
+**c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
+Logos plot of the canonical motif are shown alongside the heatmap.
+Percentages are shown for class I, II or III binding sites, or all WT ATAC-seq peaks as a control for the background motif frequency.
+**f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
+**g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**.
+Motif scores were determined by comparing each identified motif instance to the canonical motif.
+:::
 
@@
 
+::: {#fig-4}
+![](figures/fig4.pdf)
+
+**Many endogenous binding sites are resistant to ectopic pioneer-factor binding. a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
+ChIP-seq data for repressive histone modifications H3K27me3 and H3K9me3 is also shown.
+**d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
+Zld, Grh, and Twi ChIP-seq data are z-score normalized.
+CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).
+:::
+
 To test the functional relevance of repressive chromatin to pioneer-factor binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3.
@@
 
+::: {#fig-5}
+![](figures/fig5.pdf)
+
+**Binding and opening of closed chromatin is concentration dependent.
+a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
+Concetrations of CuSO~4~ used to induce protein expression are indicated above each plot.
+**b,f,j,** Metaplots of ATAC-seq signal at class I-VI regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
+**c,g,k,** Distribution of class I, II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
+Y-axis indicates the total number of peaks called at each concentration.
+Colors indicate the proportion of those peaks that are defined as class I, II, or III.
+**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
+:::
+
 ## Regions outside the DNA-binding domain are required for binding and opening of closed chromatin
@@
 However, little is known about how regions outside the DBD may contribute to pioneering function.
-To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig.
-10a,b).
+We generated stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig. 10a,b).
+Immunostaining confirmed that the each DBD alone was properly localized to the nucleus (Extended Data Fig.
+10c-d) To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq.
 Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was unable to bind closed chromatin at class II or III regions (Fig.
@@
 Together, these experiments show that regions outside the DBD are essential for binding and opening closed chromatin.
+
+::: {#fig-6}
+![](figures/fig6.pdf)
+
+**Regions outside the DNA-binding domain are required for pioneer-factor function.
+a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
+Shown for Zld **(a)** and Grh **(b)**.
+**c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**.
+**d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
+:::

### 2023-03-06 manuscript/manuscript_sections/_04_Discussion.qmd
update MS draft

@@
 editor: visual
-#bibliography: ../references.bib
+# bibliography: ../references.bib
 ---
@@
 Unlike many mammalian TFs, which have mostly been studied in cell culture, Zld, Grh and Twi have been extensively studied during normal Drosophila development.
-By comparing the results of TF induction in S2 cells to Zld, Grh and Twi activity during normal development, we are able to provide mechanistic insights that are relevant to the endogenous function of pioneer factors.
+By comparing the results of TF induction in S2 cells to Zld, Grh and Twi activity during normal development, we are able to provide mechanistic insights into how PFs are regulated in different cell types, and how PF binding relates to function.
 Zld, Grh, and Twi displayed a range of abilities to bind and open closed chromatin, suggesting that pioneering activity occurs on a spectrum and is context dependent.
@@
 The separability of binding and opening suggests that additional downstream steps such as recruitment of chromatin remodelers are required for chromatin opening [@Judd2021; @King2017].
-Local restructuring of nucleosome-DNA interactions [@Michael2020; @Dodonova2020], if it occurs in the case of Zld, Grh, or Twi, may be insufficient for establishing NFRs and may instead stabilize PF binding to allow recruitment of cofactors.
-We propose that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features (Fig 7) such as local chromatin features, protein concentration, DNA motif content, and protein domains outside the DBD.
+Local restructuring of nucleosome-DNA interactions [@Michael2020; @Dodonova2020], if it occurs in the case of Zld, Grh, or Twi, may be insufficient for establishing accessible chromatin and may instead stabilize PF binding to allow recruitment of cofactors.
+We find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin features, protein concentration, DNA motif content, and protein domains outside the DBD.
 
@@
 In contrast to taz treatment, expression of Zld and Grh, but not Twi, at higher levels allowed at least some binding to these repressed regions.
-These results suggest that despite the correlation with H3K27me3, protein concentration and affinity for target motifs may be more important than the presence of polycomb silencing for tissue-specific binding and emphasize the importance of functionally testing the importance of different chromatin marks to PF binding.
+These results suggest that despite the correlation with H3K27me3, protein concentration and affinity for target motifs may be more important than the presence of polycomb silencing for tissue-specific.
 Cooperative binding with cell-type specific cofactors can also stabilize PF binding to tissue-specific binding sites [@Donaghey2018].
@@
 Further exploration of how chromatin structure and cofactor expression influence PF binding and activity will be essential to developing a more complete understanding of how these factors function in development and disease.
-
-The strong concentration dependence of chromatin opening by Zld, Grh and Twi provides important context for the *in vivo* biological functions of these proteins.
-Zld, along with many other TFs, forms high concentration foci within the nucleus [@Mir2018].
-These foci may facilitate chromatin opening by pioneer factors by creating a high protein concentration at specific genomic regions or locations within the nucleus.
-Grh has pioneering activity in late drosophila embryos and larval tissues, but not in early embryos [@Nevil2020; @Jacobs2018].
-This could, in part, be accounted for by differences in Grh expression between tissues.
-Additionally, GRHL2, one the mammalian Grh orthologs, can act as either a tumor suppressor or an oncogene, depending on the context [@Reese2019].
-Considering Grh expression levels in different tissues, cell types, and cancer types will be important for determining these contradictory effects.
-This is especially relevant considering that GRHL2 is frequently subject to copy number amplification in various cancers, leading to overexpression [@Reese2019].
-Pioneering activity by other mammalian factors depends on protein concentration [@Hansen2022; @Blassberg2022], suggesting that this is likely a general feature of PFs.
 
@@
 Further exploration of how IDRs contribute to PF binding and chromatin accessibility will be essential for a more complete understanding of how *cis-*regulatory regions are established.
+
+The strong concentration dependence of chromatin opening by Zld, Grh and Twi provides important context for the *in vivo* biological functions of these proteins.
+Zld, along with many other TFs, forms high concentration foci within the nucleus [@Mir2018].
+These foci may facilitate chromatin opening by pioneer factors by creating a high protein concentration at specific genomic regions or locations within the nucleus.
+Grh has pioneering activity in late drosophila embryos and larval tissues, but not in early embryos [@Nevil2020; @Jacobs2018].
+This could, in part, be accounted for by differences in Grh expression between tissues.
+Pioneering activity by other mammalian factors depends on protein concentration [@Hansen2022; @Blassberg2022], suggesting that this is likely a general feature of PFs.
+
+Due to the essential role of PFs in regulating cell fate, their misexpression frequently results in disease [@Larson2021; @Dobersch2019].
+GRHL2, one the mammalian Grh orthologs, is frequently subject to copy number amplification in various cancers, leading to overexpression, and can act as either a tumor suppressor or an oncogene, depending on the context [@Reese2019].
+Misexpression of the PF DUX4 results in the muscular atrophy disease facioscapulohumeral muscular dystrophy (FSHD) [@Campbell2018].
+Our data provide evidence for how misexpression and overexpression of PFs can lead to ectopic activation of gene-regulatory programs.
+A better understanding of how PF binding and activity is regulated will be essential for understanding and treating diseases caused by these proteins.

### 2023-03-06 workflow/scripts/fig2.R

@@ -37,12 +37,19 @@ zld_RNAseq_results_fn <- snakemake@input[["zld_RNAseq_results"]]
 grh_RNAseq_results_fn <- snakemake@input[["grh_RNAseq_results"]]
 
 # create blank layout for plot =================================================
-pdf(snakemake@output[[1]], useDingbats = FALSE)
-pageCreate(width = 18.3, height = 12.5, default.units = "cm", showGuides = FALSE)
+# define figure dimensions in cm
+fig_width <-  13
+fig_height <- 6.5
+
+# open pdf
+pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
+
+# generate plotGardener page
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics
-panel_label_params <- pgParams(fontsize = 8)
+panel_label_params <- pgParams(fontsize = 7)
 large_text_params <- pgParams(fontsize = 7)
 small_text_params <- pgParams(fontsize = 5)
 
@@ -831,3 +838,4 @@ plotText(
 
 # close graphics device ========================================================
 dev.off()
+

### 2023-03-06 workflow/scripts/fig3.R

@@ -79,9 +79,15 @@ twi_WT_atac_fn <- snakemake@input[["twi_WT_atac_fn"]]
 # twi_WT_atac_fn <- "ATACseq/results/peaks/merged_by_sample/Twi_ATAC_S2-WT_40uM_summits.bed"
 
 ## create blank layout for plot =================================================
-pdf(snakemake@output[[1]], useDingbats = FALSE)
-# pdf("manuscript/figures/fig3.pdf")
-pageCreate(width = 12.0, height = 12.5, default.units = "cm", showGuides = FALSE)
+# define figure dimensions in cm
+fig_width <-  12
+fig_height <- 7.5
+
+# open pdf
+pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
+
+# generate plotGardener page
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics

### 2023-03-06 workflow/scripts/fig5.R

@@ -43,9 +43,15 @@ grh_titration_classes_fn <- snakemake@input[["grh_titration_classes_fn"]]
 twi_titration_classes_fn <- snakemake@input[["twi_titration_classes_fn"]]
 
 ## create blank layout for plot =================================================
-pdf(snakemake@output[[1]], useDingbats = FALSE)
-# pdf("manuscript/figures/fig5.pdf")
-pageCreate(width = 18, height = 16, default.units = "cm", showGuides = FALSE)
+# define figure dimensions in cm
+fig_width <-  18
+fig_height <- 16
+
+# open pdf
+pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
+
+# generate plotGardener page
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics

### 2023-03-09 manuscript/manuscript_sections/_04_Discussion.qmd
incorporate MMH discussion changes

--- 2023-02-20 manuscript/manuscript_sections/_04_Discussion.qmd
@@
 editor: visual
-#bibliography: ../references.bib
+# bibliography: ../references.bib
 ---
 
-In this study, we systematically explore chromatin binding and opening by the transcription factors Zld, Grh and Twi.
-Unlike many mammalian TFs, which have mostly been studied in cell culture, Zld, Grh and Twi have been extensively studied during normal Drosophila development.
-By comparing the results of TF induction in S2 cells to Zld, Grh and Twi activity during normal development, we are able to provide mechanistic insights that are relevant to the endogenous function of pioneer factors.
-Zld, Grh, and Twi displayed a range of abilities to bind and open closed chromatin, suggesting that pioneering activity occurs on a spectrum and is context dependent.
-Binding is separable from chromatin opening through the identification of class II sites that are bound by PFs but remain closed, or by the expression of the Grh DBD, which retains some binding but cannot promote chromatin opening.
-The separability of binding and opening suggests that additional downstream steps such as recruitment of chromatin remodelers are required for chromatin opening [@Judd2021; @King2017].
-Local restructuring of nucleosome-DNA interactions [@Michael2020; @Dodonova2020], if it occurs in the case of Zld, Grh, or Twi, may be insufficient for establishing NFRs and may instead stabilize PF binding to allow recruitment of cofactors.
-We propose that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features (Fig 7) such as local chromatin features, protein concentration, DNA motif content, and protein domains outside the DBD.
-
-Despite the ability of PFs to bind to closed chromatin, most PFs occupy only a subset of target motifs in the genome and PF binding and activity is often cell-type specific [@Larson2021].
-In many cases, the presence of active or repressive chromatin is an important determinant of transcription factor binding, but the extent to which this informs PF binding is unclear.
-For the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM), high levels of the repressive histone modification H3K9me3 mark regions that are resistant to OSKM binding.
-Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions [@Soufi2012].
-In the case of Pax7, H3K27me3 is correlated with cell-type specific Pax7 binding [@Mayran2019].
-Zld, Grh and Twi binding showed similar correlations with local chromatin structure.
-Even when expressed at low levels, all three proteins bind to active, accessible regions with degenerate motifs or no detectable motifs at all.
-At approximately physiological levels, these factors bind to canonical motifs in naïve chromatin, and are largely excluded from repressed chromatin (bearing H3K9me3 or H3K27me3).
-When we depleted H3K27me3 with the polycomb inhibitor tazemetostat, we observed no effect on binding within these repressed regions.
-In contrast to taz treatment, expression of Zld and Grh, but not Twi, at higher levels allowed at least some binding to these repressed regions.
-These results suggest that despite the correlation with H3K27me3, protein concentration and affinity for target motifs may be more important than the presence of polycomb silencing for tissue-specific binding and emphasize the importance of functionally testing the importance of different chromatin marks to PF binding.
-Cooperative binding with cell-type specific cofactors can also stabilize PF binding to tissue-specific binding sites [@Donaghey2018].
-This may also be important for Zld, Grh, and Twi, and may explain the presence of class VI sites that are resistant to binding and lack any repressive histone modifications.
-Further exploration of how chromatin structure and cofactor expression influence PF binding and activity will be essential to developing a more complete understanding of how these factors function in development and disease.
-
-The strong concentration dependence of chromatin opening by Zld, Grh and Twi provides important context for the *in vivo* biological functions of these proteins.
-Zld, along with many other TFs, forms high concentration foci within the nucleus [@Mir2018].
-These foci may facilitate chromatin opening by pioneer factors by creating a high protein concentration at specific genomic regions or locations within the nucleus.
-Grh has pioneering activity in late drosophila embryos and larval tissues, but not in early embryos [@Nevil2020; @Jacobs2018].
-This could, in part, be accounted for by differences in Grh expression between tissues.
-Additionally, GRHL2, one the mammalian Grh orthologs, can act as either a tumor suppressor or an oncogene, depending on the context [@Reese2019].
-Considering Grh expression levels in different tissues, cell types, and cancer types will be important for determining these contradictory effects.
-This is especially relevant considering that GRHL2 is frequently subject to copy number amplification in various cancers, leading to overexpression [@Reese2019].
-Pioneering activity by other mammalian factors depends on protein concentration [@Hansen2022; @Blassberg2022], suggesting that this is likely a general feature of PFs.
-
-In addition to DBDs that provide sequence specificity, many transcription factors contain long intrinsically disordered regions (IDRs) that are frequently required for transcriptional activation [@Minezaki2006].
-Fusing different IDRs to the same DBD can dramatically change which genomic regions are bound by transcription factors [@Brodsky2020].
-This observation complicates the textbook model of modular transcription factors in which the DBD determines sequence specificity and additional domains carry out transcriptional activation.
-Both Zld and Grh contain long disordered domains.
-The requirement of the Zld and Grh DBDs for robust binding to close chromatin demonstrates that regions outside the DBD, potentially the IDRs, are also important for binding site selection by pioneer factors.
-Others have suggested that IDRs may contribute to transcription-factor binding through protein-protein interactions with other transcription factors, nonspecific interactions with DNA, or by driving condensate formation [@Brodsky2021; @Staller2022].
-In the case of pioneer factors, another possibility is that IDRs form multivalent interactions with nucleosomes that stabilize binding.
-This model is supported by the observation that *in vitro*, full-length Zelda binds nucleosomes with a higher affinity than the DBD alone [@McDaniel2019a].
-Recent work showed that like Zld and Grh, the DBDs of human pioneer factors FOXA1 and SOX2 are required for robust binding of closed chromatin [@lerner2022].
-Additionally, single-molecule tracking indicated that the DBDs alone showed reduced scanning of silent chromatin within the nucleus.
-Zld and Grh DBDs still bound to class I regions, despite the weaker motifs present at these sites.
-This suggests that the less specific binding to class I regions is not due to interactions between IDRs and other proteins present at active chromatin.
-This binding may instead be driven by the high affinity of the DBDs for nucleosome-free DNA.
-Further exploration of how IDRs contribute to PF binding and chromatin accessibility will be essential for a more complete understanding of how *cis-*regulatory regions are established.
+By exogenously expressing Zld, Grh, and Twi in S2 cells where they are not normally expressed, leveraging prior data identifying binding sites for these factors in vivo, and analyzing the wealth of data on chromatin structure in S2 cells, we identified features that govern PF activity.
+Studying the two well-defined PFs Zld and Grh, along with Twi, which possess features distinct from these two factors, allowed us to determine general properties regulating PF activity.
+Furthermore, our analysis of both binding and accessibility allowed us to separate PF occupancy from activity.
+We find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin structure, protein concentration, DNA motif content, and protein domains outside the DBD.
+Even when expressed at low levels Zld, Grh and Twi bind promiscuously to active, accessible regions with degenerate motifs or no detectable motifs at all.
+When expressed at approximately physiological levels these factors all possess the capacity to bind to canonical motifs in closed chromatin.
+However, these factors are largely excluded from repressed chromatin enriched for H3K9me3 or H3K27me3.
+This binding to naïve chromatin is similar to what has previously been reported for other pioneer factors.
+Regions with high levels of H3K9me3 are resistant to binding by the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM).
+Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions, suggesting that this repressive chromatin state is a barrier to PF binding [@Soufi2012].
+Similarly, cell-type specific binding of the PF Pax7 is anti-correlated with levels of H3K27me3 [@Mayran2019].
+Here, we specifically tested whether H3K27me3 regulates cell-type specific occupancy of Zld, Grh and Twi using the PRC2 inhibitor tazmetostat.
+While H3K27me3 levels were dramatically decreased in our treatment, PF binding was largely unchanged.
+Thus, unlike H3K9me3 in the case of OSKM, H3K27me3 is not a barrier to binding by the three factors assayed and suggest other features are restricting binding to these regions.
+While depleting H3K27me3 levels did not result in Zld, Grh and Twi binding to additional regions, increased expression of both Zld and Grh resulted in binding to regions that were unbound at lower expression levels.
+These results suggest that protein concentration and affinity for the target motif drives tissue-specific PF binding rather than the presence or absence of Polycomb-mediated silencing.
+Tissue-specific PF occupancy is also likely regulated by the collection of cofactors expressed in a given tissue.
+This may partially explain the presence of class VI sites that lack repressive histone modifications in culture and are bound in vivo but not when expressed in S2 cells.
+For example, in mammalian cell culture FOXA2 occupancy at a subset of tissue-specific binding sites was increased when GATA4 was present [@Donaghey2018].
+Similarly, PARP-1 stabilizes Sox2 binding at subset of physiologically relevant sites in mouse embryonic stem cells [@Liu2017].
+Unlike Zld and Grh, increased expression of Twi did not result in binding to regions bound in other tissues.
+It is probable that additional cofactors, such as Zld itself, may be required for Twi occupancy at these regions.
+Prior studies demonstrated an important role for Zld in promoting Twi binding in the early embryo [@Yáñez-Cuna2012; @brennan2022].
+Along with prior studies, our data suggest that PF occupancy is regulated by tissue-intrinsic features, including levels of PF expression, the complement of cofactors expressed and chromatin structure.
+In addition to cell-type specific factors, we demonstrate that protein-intrinsic features control PF binding.
+We show that regions outside the DBD of Zld and Grh are important for binding-site selection.
+Similarly, regions outside the DBDs of human PFs FOXA1 and SOX2 are required for robust binding of closed chromatin [@lerner2022].
+Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and have demonstrated the role that intrinsically disordered regions (IDRs) play in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
+For Zld, this disordered domain is required for transcriptional activation and promotes nucleosome binding in vitro [@Hamm2015; @McDaniel2019].
+IDRs may contribute to transcription-factor binding through protein-protein interactions with other transcription factors, nonspecific interactions with DNA, or by driving condensate formation [@Brodsky2021; @Staller2022].
+Indeed, Zld is visualized in hubs within the nucleus, and these promote binding of additional transcription factors [@Mir2018; @Yamada2019].
+Zld and Grh DBDs retain the capacity to bind to class I regions, despite the weaker motifs present at these sites.
+Thus, the binding to class I regions is not due to interactions between IDRs and other proteins present at active chromatin.
+Instead, this binding may be driven by the high affinity of the DBDs for nucleosome-free DNA.
+To provide a more complete understanding of how cis-regulatory regions are established, we must further investigate how IDRs contribute to PF binding and chromatin accessibility.
+The failure of the DBD alone to promote chromatin accessibility implicates a separation between the ability to bind nucleosomes and the capacity to promote chromatin accessibility.
+Local restructuring of nucleosome-DNA interactions [@Michael2020; @Dodonova2020] may be insufficient for establishing accessible chromatin and, instead, may stabilize PF binding to allow recruitment of cofactors, such as chromatin remodelers [@Judd2021; @King2017].
+Furthermore, our identification of hundreds of class II regions, which are bound but not opened by exogenous PF expression, further separates chromatin binding and PF activity.
+Increased motif content at the sites in which PFs can promote accessibility suggests a model whereby PF chromatin occupancy is correlated with PF activity.
+Supporting this model, we identified a strong concentration dependence of chromatin opening by Zld and Grh.
+This model would suggest that PF expression levels must be tightly controlled during development.
+Indeed, misexpression of the PFs Grh and DUX4 lead to diseases such as epithelial cancer or facioscapulohumeral muscular dystrophy (FSHD), respectively [@Campbell2018; @Reese2019].
+Together our data identify protein intrinsic and extrinsic features that govern PF binding and activity, providing insights into how these factors define novel cis-regulatory elements and drive gene-regulatory networks to modulate cell fate.
+A deeper understanding of how PF activity is regulated will have important implications for determining how diseases are caused by their misexpression and our ability to use these powerful factors for cellular reprogramming.

### 2023-03-09 manuscript/manuscript_sections/_05_Methods.qmd
incorporate MMH discussion changes

@@
 | Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|---------------------------------------|---------------------------|----------------------|-------------------------|
+|-----------------------|-----------------|-----------------|-----------------|
 | pMT-puro                              | AddGene                   | 17923                |                         |
@@
 Transcription factor expression was induced by adding CuSO~4~ to the cell culture media.
-Expression of protein at physiological levels was achieved with 1000, 100, or 40 µM CuSO~4~ for Zld, Grh or Twi, respectively (Supplemental figure X).
+Expression of protein at physiological levels was achieved with 1000, 100, or 40 µM CuSO~4~ for Zld, Grh or Twi, respectively (Extended Data Fig. 1c-d; Extended Data Fig. 4a).
 For experiments with sub- or supraphysiological protein levels (Fig.
-5), the following CuSO~4~ concentrations were used: 500 µM (Zld subphysiological), 1000 µM (Zld physiological) or 1500 µM (Zld supraphysiological); 25 µM (Grh subphysiological), 100 µM (Grh physiological) or 400 µM (Grh supraphysiological); 10 µM (Twi subphysiological), 40 µM (Twi physiological) or 160 µM (Twi supraphysiological).
+5; Extended Data Fig.
+9a-c), the following CuSO~4~ concentrations were used: 500 µM (Zld subphysiological), 1000 µM (Zld physiological) or 1500 µM (Zld supraphysiological); 25 µM (Grh subphysiological), 100 µM (Grh physiological) or 400 µM (Grh supraphysiological); 10 µM (Twi subphysiological), 40 µM (Twi physiological) or 160 µM (Twi supraphysiological).
 For all induction experiments, cells were plated at 1×10^6^ cells per mL.

### 2023-03-10 manuscript/supplemental_figures.qmd
update figs

--- 2023-03-01 manuscript/manuscript_sections/_06_Figure_Legends.qmd
@@
-**Figure 1: Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
-ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
-**c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
-**d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
-**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h).
-i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
-**j,i,** Proportion of class I, II or III regions that are proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
-Promoter-proximal and promoter-distal binding sites are shown separately.
-The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene.
+---
+format: 
+  pdf:
+    include-in-header:
+      text: |
+        \usepackage[labelfont=bf,format=plain]{caption}
+        \pagenumbering{gobble}
+    mainfont: Arial
+  docx:
+    reference-doc: custom-reference-doc.docx
+crossref:
+  fig-title: '**Extended Data Figure**'
+  title-delim: "**.**"
 
-**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.
-a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites.
-**c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions.
-**d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.
+---
 
-**Figure 3: Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
-**b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
-Zld, Grh, and Twi ChIP-seq data were generated in this study.
-Remaining ChIP-seq datasets were previously published (see Supplementary table XX).
-**c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
-Logos plot of the canonical motif are shown alongside the heatmap.
-Percentages are shown for class I, II or III binding sites, or all WT ATAC-seq peaks as a control for the background motif frequency.
-**f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
-**g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**.
-Motif scores were determined by comparing each identified motif instance to the canonical motif.
+::: {#fig-1}
+![](figures/extended_data_fig1.pdf)
 
-**Figure 4: Many endogenous binding sites are resistant to ectopic pioneer-factor binding.
-a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
-ChIP-seq data for repressive histone modifications H3K27me3 and H3K9me3 is also shown.
-**d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
-Zld, Grh, and Twi ChIP-seq data are z-score normalized.
-CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).
-
-**Figure 5: Binding and opening of closed chromatin is concentration dependent.
-a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
-Concetrations of CuSO~4~ used to induce protein expression are indicated above each plot.
-**b,f,j,** Metaplots of ATAC-seq signal at class I-VI regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
-**c,g,k,** Distribution of class I, II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
-Y-axis indicates the total number of peaks called at each concentration.
-Colors indicate the proportion of those peaks that are defined as class I, II, or III.
-**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
-
-**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function.
-a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
-Shown for Zld **(a)** and Grh **(b)**.
-**c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**.
-**d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
-
-**Figure 7: Pioneer-factor function requires stable chromatin occupancy.**
-
-**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.
-a,** Schematic of generation of stable cell lines and induction of protein expression.
+**Stable cell lines allow inducible expression of transcription factors at physiological concentrations. a,** Schematic of generation of stable cell lines and induction of protein expression.
 **b,** mRNA levels of Zld and Grh in S2 cells.
@@
 Two independently generated cell lines are shown and compared to 2-3H embryos.
+:::
 
-**Extended Data Figure 2: Zld and Grh bind to chromatin rapidly after induction of protein expression.
+::: {#fig-2}
+![](figures/extended_data_fig2.pdf)
+
+**Zld and Grh bind to chromatin rapidly after induction of protein expression.
 a-b,** Immunoblots showing time course of Zld **(a)** or Grh **(b)** protein expression following induction of stable cell lines.
 **c-d,** Metaplots showing average z-score normalized CUT&RUN signal at class I, II or III sites at different time points following induction.
+:::
 
-**Extended Data Figure 3: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.
+::: {#fig-3}
+![](figures/extended_data_fig3.pdf)
+
+**Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.
 a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
@@
 **c,f,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(c)** or Grh **(f)**.
+:::
 
-**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
+::: {#fig-4}
+![](figures/extended_data_fig4.pdf)
+
+**Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
 Protein levels in stable cell lines are compared to 3-4 hour old embryos.
@@
 **d,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Twi.
+:::
 
-**Extended Data Figure 5: Chromatin features associated with Zld, Grh and Twi binding sites.
-a,c,e,** Heatmaps showing the levels of different chromatin marks in class I, II or III regions for Zld (a), Grh (c) or Twi (e).
+::: {#fig-5}
+![](figures/extended_data_fig5.pdf)
+
+**Chromatin features associated with Zld, Grh and Twi binding sites. a,c,e,** Heatmaps showing the levels of different chromatin marks in class I, II or III regions for Zld (a), Grh (c) or Twi (e).
 The color represents the average z-score normalized read depth across a 1 KB region surrounding the center of class I, II or III ChIP-seq peaks.
 **b,d,f,** Example genome browser tracks for class III regions with high levels of H3K27me3 for Zld (b), Grh (d), or Twi (f).
+:::
 
-**Extended Data Figure 6: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes.
+::: {#fig-6}
+![](figures/extended_data_fig6.pdf)
+
+**Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes.
 a-c,** Metaplots showing average MNase signal at motifs within class I, II, or III regions for Zld (a), Grh (b), or Twi (c).
@@
 Rows are ordered based on hierarchical clustering to highlight the various patterns of MNase signal around motifs.
+:::
 
-**Extended Data Figure 7: Zld, Grh and Twi display cell-type specific binding.
-a-c,** Venn diagrams showing overlap between ChIP-seq peaks identified in different tissues for Zld (a), Grh (b) or Twi (c).
+::: {#fig-7}
+![](figures/extended_data_fig7.pdf)
+
+**Zld, Grh and Twi display cell-type specific binding. a-c,** Venn diagrams showing overlap between ChIP-seq peaks identified in different tissues for Zld (a), Grh (b) or Twi (c).
 **d-l,** Genome browser tracks showing examples of class IV, V, and VI regions for Zld (d-f), Grh (g-i) or Twi (j-l).
@@
 Dashed gray lines indicate a zoomed-in region where Zld, Grh or Twi ChIP-seq signal is shown in S2 cells or in embryos.
+:::
 
-**Extended Data Figure 8: PRC2 inhibition is insufficient to promote Zld, Grh, and Twi binding to repressed targets.
-a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
+::: {#fig-8}
+![](figures/extended_data_fig8.pdf)
+
+**PRC2 inhibition is insufficient to promote Zld, Grh, and Twi binding to repressed targets. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
 Tubulin levels are shown as a loading control.
+:::
 
-**Extended Data Figure 9: Expression of Zld, Grh, or Twi at supraphysiological levels results in chromatin opening at a small number of novel binding sites.
+::: {#fig-9}
+![](figures/extended_data_fig9.pdf)
+
+**Expression of Zld, Grh, or Twi at supraphysiological levels results in chromatin opening at a small number of novel binding sites.
 a-c,** Immunoblots showing Zld (a), Grh (b) or Twi (c) protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
@@
 **g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld (g), Grh (h), or Twi (i) are expressed at varying concentrations.
+:::
 
-**Extended Data Figure 10: Expression of Zld and Grh DNA-binding domains at protein levels comparable to the full-length proteins.
+::: {#fig-10}
+![](figures/extended_data_fig10.pdf)
+
+**Expression of Zld and Grh DNA-binding domains at protein levels comparable to the full-length proteins.
 a-b,** Immunoblots showing titration of protein levels for Zld (a) or Grh (b) DNA-binding domains to match expression of the full-length proteins.
@@
 **e,** Metaplots showing the average ChIP-seq signal for full-length Zld or the DNA-binding domain alone at the subset of previously identified class I, II or III binding sites that contain a canonical Zelda CAGGTAG motif.
+:::

### 2023-03-10 workflow/scripts/extended_data_fig3.R

@@ -31,9 +31,17 @@ zld_ChIP_peaks_fn <- snakemake@input[["zld_ChIP_peaks_fn"]]
 grh_ChIP_peaks_fn <- snakemake@input[["grh_ChIP_peaks_fn"]]
 
 # # create blank layout for plot ===============================================
-pdf(snakemake@output[[1]], useDingbats = FALSE)
+# # create blank layout for plot ===============================================
+fig_width <-  18
+fig_height <- 12
+
+# open pdf
+pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
+
 # pdf("manuscript/figures/extended_data_fig3.pdf", useDingbats = FALSE)
-pageCreate(width = 18, height = 12, default.units = "cm", showGuides = FALSE)
+
+# generate plotGardener page
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics
@@ -87,7 +95,9 @@ a_plot <- zld_ATAC_results |>
         legend.position = "bottom",
         legend.title = element_blank(),
         legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(-10,-10,-10,-10))
+        legend.box.margin=margin(0,0,0,0)) +
+  xlab(bquote(log[2]("fold change"))) +
+  ylab(bquote(-log[10]("adj. p-value")))
 
 # place plots on plotGardener page
 plotGG(
@@ -138,7 +148,9 @@ b_plot <- zld_RNAseq_results |>
         legend.position = "bottom",
         legend.title = element_blank(),
         legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(-10,-10,-10,-10))
+        legend.box.margin=margin(0,0,0,0)) +
+  xlab(bquote(log[2]("fold change"))) +
+  ylab(bquote(-log[10]("adj. p-value")))
 
 # place plots on plotGardener page
 plotGG(
@@ -235,7 +247,9 @@ d_plot <- grh_ATAC_results |>
         legend.position = "bottom",
         legend.title = element_blank(),
         legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(-10,-10,-10,-10))
+        legend.box.margin=margin(0,0,0,0)) +
+  xlab(bquote(log[2]("fold change"))) +
[…]

### 2023-03-10 workflow/scripts/extended_data_fig4.R

@@ -23,10 +23,17 @@ twi_ChIP_peaks_fn <- snakemake@input[["twi_ChIP_peaks_fn"]]
 twi_blot_image <- snakemake@input[["twi_blot_image"]]
 
 
-# # create blank layout for plot ===============================================
-pdf(snakemake@output[[1]], useDingbats = FALSE)
+# create blank layout for plot ===============================================
+# create blank layout for plot ===============================================
+fig_width <-  18
+fig_height <- 12
+
+# open pdf
+pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
 # pdf("manuscript/figures/extended_data_fig4.pdf", useDingbats = FALSE)
-pageCreate(width = 18, height = 12, default.units = "cm", showGuides = FALSE)
+
+# generate plotGardener page
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics
@@ -232,7 +239,9 @@ b_plot <- twi_ATAC_results |>
         legend.position = "bottom",
         legend.title = element_blank(),
         legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(-10,-10,-10,-10))
+        legend.box.margin=margin(0,0,0,0)) +
+  xlab(bquote(log[2]("fold change"))) +
+  ylab(bquote(-log[10]("adj. p-value")))
 
 # place plots on plotGardener page
 plotGG(
@@ -289,7 +298,9 @@ c_plot <- twi_RNAseq_results |>
         legend.position = "bottom",
         legend.title = element_blank(),
         legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(-10,-10,-10,-10))
+        legend.box.margin=margin(0,0,0,0)) +
+  xlab(bquote(log[2]("fold change"))) +
+  ylab(bquote(-log[10]("adj. p-value")))
 
 # place plots on plotGardener page
 plotGG(

### 2023-03-10 workflow/scripts/fig1.R

@@ -24,7 +24,7 @@ Grh_Grh_RNAseq_bw <- snakemake@input[["Grh_Grh_RNAseq_bw"]]
 zld_ChIP_classes_fn <- snakemake@input[["zld_ChIP_classes"]]
 grh_ChIP_classes_fn <- snakemake@input[["grh_ChIP_classes"]]
 
-# # create blank layout for plot ===============================================
+# create blank layout for plot ===============================================
 fig_width <-  18
 fig_height <- 12.5
 
@@ -32,7 +32,7 @@ fig_height <- 12.5
 pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
 
 # generate plotGardener page
-pageCreate(width = 18, height = 12.5, default.units = "cm", showGuides = FALSE)
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics
@@ -760,12 +760,12 @@ plotText(
 )
 
 # read in ChIP classes
-zld_chip_classes <- read_tsv(zld_ChIP_classes_fn)
+zld_chip_classes <- read_tsv(zld_ChIP_classes_fn) |> 
+  mutate(class = str_to_upper(class))
 
 # generate pie charts
-zld_class_plot <- zld_chip_classes  %>% 
-  # mutate(zygotic_class = factor(zygotic_class, levels = c("i", "ii", "iii"))) %>%
-  group_by(class) %>% summarise(n = n()) %>%
+zld_class_plot <- zld_chip_classes  |> 
+  group_by(class) |> summarise(n = n()) |>
   ggplot(aes(x='', y = n, fill = class)) + 
   geom_bar(stat = "identity", position = "fill") + 
   coord_polar("y", direction = 1, start = pi / 2) + 
@@ -785,17 +785,17 @@ plotGG(
 
 # add labels
 plotText(
-  label = paste("class I", "\n", "n =", sum(zld_chip_classes$class == "i")), fontsize = small_text_params$fontsize,
+  label = paste("class I", "\n", "n =", sum(zld_chip_classes$class == "I")), fontsize = small_text_params$fontsize,
[…]

### 2023-03-10 workflow/scripts/fig2.R

@@ -425,12 +425,13 @@ plotText(
 )
 
 # read in ChIP classes
-twi_chip_classes <- read_tsv(twi_ChIP_classes_fn)
+twi_chip_classes <- read_tsv(twi_ChIP_classes_fn) |> 
+  mutate(class = str_to_upper(class))
 
 # generate pie charts
-twi_class_plot <- twi_chip_classes  %>% 
-  # mutate(zygotic_class = factor(zygotic_class, levels = c("i", "ii", "iii"))) %>%
-  group_by(class) %>% summarise(n = n()) %>%
+twi_class_plot <- twi_chip_classes  |> 
+  # mutate(zygotic_class = factor(zygotic_class, levels = c("I", "II", "III"))) |>
+  group_by(class) |> summarise(n = n()) |>
   ggplot(aes(x='', y = n, fill = class)) + 
   geom_bar(stat = "identity", position = "fill") + 
   coord_polar("y", direction = 1, start = pi / 2) + 
@@ -450,17 +451,17 @@ plotGG(
 
 # add labels
 plotText(
-  label = paste("class I", "\n", "n =", sum(twi_chip_classes$class == "i")), fontsize = small_text_params$fontsize,
+  label = paste("class I", "\n", "n =", sum(twi_chip_classes$class == "I")), fontsize = small_text_params$fontsize,
   x = (ref_x + 1), y = (ref_y + 0.5), just = "center", default.units = "cm"
 )
 
 plotText(
-  label = paste("class II", "\n", "n =", sum(twi_chip_classes$class == "ii")), fontsize = small_text_params$fontsize,
+  label = paste("class II", "\n", "n =", sum(twi_chip_classes$class == "II")), fontsize = small_text_params$fontsize,
   x = (ref_x + 2.2), y = (ref_y + 1.5), just = "center", default.units = "cm"
 )
 
 plotText(
-  label = paste("class III", "\n", "n =", sum(twi_chip_classes$class == "iii")), fontsize = small_text_params$fontsize,
+  label = paste("class III", "\n", "n =", sum(twi_chip_classes$class == "III")), fontsize = small_text_params$fontsize,
   x = (ref_x + 2.2), y = (ref_y + 1), just = "center", default.units = "cm"
 )
 
[…]

### 2023-03-10 workflow/scripts/fig3.R

@@ -9,15 +9,6 @@ suppressPackageStartupMessages(library(grid))
 source("workflow/scripts/plot_heatmap.R")
 source("workflow/scripts/utils.R")
 
-# function to get range for browser tracks -------------------------------------
-signal_range <- function(x, extra = 0.05) {
-  min <- floor(min(x))
-  max <- ceiling(max(x))
-  range <- max - min
-  margin <- range * extra
-  output <- c(min - margin, max + margin)
-  return(output)
-}
 
 # define input files ===========================================================
 class_I_bed_fn <- c(
@@ -80,14 +71,14 @@ twi_WT_atac_fn <- snakemake@input[["twi_WT_atac_fn"]]
 
 ## create blank layout for plot =================================================
 # define figure dimensions in cm
-fig_width <-  12
-fig_height <- 7.5
+fig_width <-  18
+fig_height <- 12
 
 # open pdf
 pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
 
 # generate plotGardener page
-pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = TRUE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics
@@ -281,7 +272,8 @@ plotText(
 )
 
 # import ChIP classes
-zld_ChIP_classes <- read_tsv(zld_ChIP_classes_fn)
+zld_ChIP_classes <- read_tsv(zld_ChIP_classes_fn) |> 
+  mutate(class = str_to_upper(class))
 
 # get background motif frequency
 zld_motifs_gr <- 
@@ -293,7 +285,7 @@ zld_ATAC_peaks <-
   zld_WT_atac_fn |> 
   rtracklayer::import() |> 
   resize(width = 201, fix = "center")
-zld_ATAC_peaks$class <- "all_ATAC_peaks"
+zld_ATAC_peaks$class <- "all ATAC peaks"
 
[…]

### 2023-03-10 workflow/scripts/fig5.R

@@ -152,14 +152,16 @@ ref_y <- 0.5
 
 # generate plot
 zld_class_plot <- zld_titration_classes_fn |> 
-  read_tsv() |> 
-  filter(class != "i") |> 
+  read_tsv()  |> 
+  mutate(class = str_to_upper(class)) |> 
+  filter(class != "I") |> 
     ggplot(aes(x = CuSO4, fill = class)) + 
   geom_bar(color = "black", linewidth = 0.1) +
   theme_classic(base_size = small_text_params$fontsize) +
   theme(legend.key.size = unit(2, 'mm'),
         legend.position = "right") +
   scale_fill_manual(values = brewer.pal(3, "Blues")[2:3]) +
+  xlab(bquote("[" ~ CuSO[4] ~ "]")) +
   ylab("n peaks") 
 
 # place chart on page
@@ -198,16 +200,22 @@ input_files <- c(
   
 )
 
-peak_overlaps <- input_files %>%
-  map(rtracklayer::import) %>%
-  GRangesList() %>%
-  peak_overlap_table()
+peak_overlaps <- input_files |>
+  map(rtracklayer::import) |>
+  GRangesList() |>
+  peak_overlap_table() |> 
+  dplyr::rename(
+    `Zld 0` = zld_0uM,
+    `Zld 500` = zld_500uM,
+    `Zld 1000` = zld_1000uM,
+    `Zld 1500` = zld_1500uM
+  )
 
 
-zld_euler_plot <- peak_overlaps %>% 
-  dplyr::select(8,9) %>% 
-  as.matrix() %>% 
-  euler() %>% 
+zld_euler_plot <- peak_overlaps |> 
+  dplyr::select(8,9) |> 
+  as.matrix() |> 
+  euler() |> 
   plot(quantities = list(fontsize = small_text_params$fontsize), labels = list(fontsize = small_text_params$fontsize))
 
 plotGG(
@@ -304,12 +312,14 @@ ref_y <- 5.5
 # generate plot
 grh_class_plot <- grh_titration_classes_fn |> 
   read_tsv() |> 
-  filter(class != "i") |> 
+  mutate(class = str_to_upper(class)) |> 
+  filter(class != "I") |> 
   ggplot(aes(x = as.factor(CuSO4), fill = class)) + 
   geom_bar(color = "black", linewidth = 0.1) +
   theme_classic(base_size = small_text_params$fontsize) +
[…]

### 2023-03-10 workflow/scripts/fig6.R

@@ -3,7 +3,7 @@ suppressPackageStartupMessages(library(plotgardener))
 suppressPackageStartupMessages(library(tidyverse))
 suppressPackageStartupMessages(library(org.Dm.eg.db))
 suppressPackageStartupMessages(library(TxDb.Dmelanogaster.UCSC.dm6.ensGene))
-# library(grImport)
+library(grImport)
 suppressPackageStartupMessages(library(grid))
 suppressPackageStartupMessages(library(RColorBrewer))
 
@@ -32,8 +32,8 @@ grh_DBD_atac_results_fn <- snakemake@input[["grh_DBD_atac_results"]]
 
 ## create blank layout for plot =================================================
 # define figure dimensions in cm
-fig_width <-  16
-fig_height <- 6
+fig_width <-  18
+fig_height <- 14
 
 # open pdf
 pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
@@ -124,7 +124,7 @@ ATAC_range <- signal_range(c(`S2-WT_ATAC`$score, `S2-Zld_ATAC`$score))
 
 s1 <- plotSignal(
   data = `Zld_FL_ChIP`, params = region,
-  x = (ref_x + x_offset_browser), y = (ref_y + 0.25), width = 5, height = gb_height,
+  x = (ref_x + x_offset_browser), y = (ref_y + 0.25), width = 6, height = gb_height,
   just = c("left", "center"), default.units = "cm",
   linecolor = zld_color, fill = zld_color, baseline = FALSE, baseline.lwd = 0, baseline.color = zld_color,
   range = ChIP_range
@@ -138,7 +138,7 @@ annoYaxis(
 
 s2 <- plotSignal(
   data = `Zld_DBD_ChIP`, params = region,
-  x = (ref_x + x_offset_browser), y = (ref_y + 0.75), width = 5, height = gb_height,
+  x = (ref_x + x_offset_browser), y = (ref_y + 0.75), width = 6, height = gb_height,
   just = c("left", "center"), default.units = "cm",
   linecolor = zld_color, fill = zld_color, baseline = FALSE, baseline.lwd = 0, baseline.color = zld_color,
   range = ChIP_range
@@ -152,7 +152,7 @@ annoYaxis(
 
 s3 <- plotSignal(
   data = `S2-WT_ATAC`, params = region,
-  x = (ref_x + x_offset_browser), y = (ref_y + 1.25), width = 5, height = gb_height,
[…]

### 2023-03-15 manuscript/manuscript_sections/_01_Abstract.qmd
incorporate MMH comments from 2023-03-13

@@
-Transcription factors establish cell-type specific gene expression programs to drive developmental transitions.
 While chromatin presents a barrier to the binding of many transcription factors, pioneer factors access nucleosomal targets and promote chromatin opening.
-Despite the ability to bind to target motifs in closed chromatin, many pioneer factors display cell-type specific binding and activity.
-The mechanisms that govern pioneer-factor occupancy in different cell types and the relationship between chromatin occupancy and chromatin opening remain unclear.
-To determine how pioneer-factor binding and activity are regulated, we systematically studied three Drosophila transcription factors with distinct DNA-binding domains and biological functions: Zelda, Grainy head, and Twist.
+Despite binding to target motifs in closed chromatin, many pioneer factors display cell-type specific binding and activity.
+The mechanisms that govern pioneer-factor occupancy and the relationship between chromatin occupancy and opening remain unclear.
+We studied three Drosophila transcription factors with distinct DNA-binding domains and biological functions: Zelda, Grainy head, and Twist.
 We demonstrated that the level of chromatin occupancy is a key determinant of pioneer-factor activity.
-Zelda, Grainy head and Twist occupancy is regulated by multiple factors, including motif content at binding sites, local chromatin context, and protein concentration.
-We found that regions outside the DNA-binding domain are required for binding and chromatin opening.
-Together, our results demonstrate that pioneering activity is not a binary feature intrinsic to a protein but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type-specific features.
-These findings have important implications for how pioneer factors drive development and how misexpression can lead to disease.
+Occupancy is regulated by multiple factors, including motif content at binding sites, local chromatin context, and protein concentration.
+Regions outside the DNA-binding domain are required for pioneer-factor binding and chromatin opening.
+Together, our results show that pioneering activity is not a binary feature intrinsic to a protein but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type-specific features.

### 2023-03-15 manuscript/manuscript_sections/_02_Introduction.qmd
update supp figures
update abstract draft
incorporate MMH comments from 2023-03-13

@@
 ---
-bibliography: ../references.bib
+# bibliography: ../references.bib
 ---
@@
 Despite their ability to target closed, nucleosomally occupied binding sites, PFs do not bind to all genomic instances of their target motifs.
+PFs also have cell-type specific patterns of binding and activity, suggesting that features beyond sequence motif govern PF function [@Soufi2012; @Donaghey2018; @Chronis2017; @Buecker2014; @Cernilogar2019; @Lupien2008].
+Chromatin structure may influence PF binding as cell-type specific occupancy is often associated with distinct chromatin features [@Mayran2018a; @Lupien2008; @Cernilogar2019].
 Many PFs bind primarily to naïve chromatin that is inaccessible, inactive, and lacks most histone modifications [@Soufi2012; @Donaghey2018].
-In addition, PFs have also been shown to have cell-type specific patterns of binding and activity [@Soufi2012; @Donaghey2018; @Chronis2017; @Buecker2014; @Cernilogar2019; @Lupien2008].
-How PF binding is restricted in different tissues or cellular contexts remains unclear.
-Many PFs bind primarily to naïve chromatin that is inaccessible, inactive, and lacks most histone modifications [@Soufi2012; @Donaghey2018].
-Cell-type specific binding is often associated with distinct chromatin features [@Mayran2018a; @Lupien2008; @Cernilogar2019].
 Together, these studies indicate that PF occupancy is regulated by both genomic context and cell-type specific features.
@@
 Although PFs are defined by their ability to promote chromatin opening, many PFs are only required for chromatin accessibility at a subset of their binding sites.
-Upon depletion of the PFs Zelda, GAGA-factor, Oct4, or Sox2, chromatin accessibility is only lost at a subset of the binding sites for the respective factor [@Schulz2015; @Gaskill2021; @maresca2022].
+Thus, PF activity is regulated at a step other than chromatin binding.
+For example, upon depletion of the PFs Zelda, GAGA-factor, Oct4, or Sox2, chromatin accessibility is lost only at a subset of the binding sites for the respective factor [@Schulz2015; @Gaskill2021; @maresca2022].
 This suggests that PF binding does not necessarily translate to chromatin opening.
-Pax7, a PF important for specification of pituitary melanotropes, binds to thousands of regions where binding does not result in chromatin opening, which the authors term "pioneer primed" sites [@Mayran2018a].
-Additionally, a time course of Pax7 binding and chromatin accessibility indicated that there is a delay between Pax7 binding and chromatin opening.
-For GAGA-factor and Oct4, establishment of chromatin accessibility requires the activity of a nucleosome remodeler [@Judd2021; @King2017].
-In the case of Oct4, however, binding to nucleosomes does result in local remodeling of histone-DNA interactions, especially at the nucleosome entry-exit site [@Michael2020].
+Pax7, a PF important for specification of pituitary melanotropes, binds to thousands of "pioneer primed" regions where binding does not result in chromatin opening [@Mayran2018a].
+Further separating PF binding from opening, at some Pax7-bound loci there is a temporal delay between binding and chromatin opening.
+This distinction between binding and opening may be due to PFs recruiting additional factors to drive chromatin accessibility.
+For example, both GAGA factor and Oct4 require nucleosome remodelers to promote chromatin accessibility [@Judd2021; @King2017].
+Nonetheless, PF binding can also perturb nucleosome structure.
+Oct4 binding to nucleosomes can result in local remodeling of histone-DNA interactions, especially at the nucleosome entry-exit site [@Michael2020].
 Together, these observations highlight that the relationship between PF binding and activity is not well understood.
 
-To begin to elucidate the factors that reuglate PF binding and activity, we focused on two well-studied PFs that control conserved developmental transitions, Zelda (Zld) and Grainy head (Grh).
-Zld is essential for driving early embryonic development in Drosophila [@Liang2008; @Staudt2006].
+To begin to elucidate the factors that regulate PF binding and activity, we focused on two well-studied PFs that control conserved developmental transitions, Zelda (Zld) and Grainy head (Grh).
+Zld is essential for driving early embryonic development in *Drosophila* [@Liang2008; @Staudt2006].
 Immediately following fertilization, the zygotic genome is transcriptionally silent, and development is controlled by maternally deposited mRNAs and proteins.
 Maternal products are gradually degraded as the zygotic genome is activated during a process known as the maternal-to-zygotic transition (MZT) [@Schulz2019; @Vastenhouw2019].
-In Drosophila, the pioneering activity of Zld is essential for driving activation of the zygotic genome [@Liang2008; @Schulz2015; @Sun2015; @McDaniel2019a].
-Since the discovery of Zld as an activator of the initial wave of zygotic transcription, pioneer factors have been shown to drive the transition in all other species studied to date [@Lee2013; @Leichsenring2013; @Gassler2022; @Gentsch2019; @Charney2017].
-Similar to the reprogramming role of Zld in the early embryo, Zld promotes the neural stem cell (NSC) fate in the developing larval brain[@Larson2021a].
-Despite the shared ability to promote the undifferentiated fate in both tissues, the majority of Zld-binding sites are unique to either the embryo or NSCs, with only a subset of regions being bound in both tissues.
-By contrast, Grh-binding sites are largely shared between tissues [@Nevil2017].
+In *Drosophila*, the pioneering activity of Zld is essential for activating transcription from the zygotic genome [@Liang2008; @Schulz2015; @Sun2015; @McDaniel2019a].
+Since the discovery of Zld as an activator of the initial wave of zygotic transcription, pioneer factors have been shown to drive zygotic genome activation in all other species studied to date [@Lee2013; @Leichsenring2013; @Gassler2022; @Gentsch2019; @Charney2017].
+Similar to the role of Zld in genomic reprogramming in the early embryo, Zld promotes the neural stem cell fate in the developing larval brain [@Larson2021a].
+Despite the shared ability to promote the undifferentiated fate in both tissues, the majority of Zld-binding sites are unique to either tissue.
+In contrast to the tissue-specific chromatin occupancy of Zld, Grh-binding sites are largely shared between tissues [@Nevil2017].
 Grh is a master regulator of epithelial cell fate that is conserved among metazoans and functions as a pioeer factor in both *Drosophila* larva and mammalian cell culture [@Wang2012; @Jacobs2018; @Nevil2020].
@@
 
-The essential roles of Zld and Grh in diverse biological processes and the well-documented context-dependent binding and activity of these factors makes them excellent models to explore how PF binding is regulated during development, and how PF occupancy relates to chromatin opening.
-Because many previous studies have relied on loss-of-function experiments, it has been difficult to unravel causal relationships between chromatin structure and PF activity.
-We therefore expressed Zld and Grh in Drosophila S2 cells, where neither factor is endogenously expressed.
-When ectopically expressed, Zld and Grh bind to closed chromatin, establish open chromatin, and promote transcriptional activation at a subset of binding sites.
+The essential roles of Zld and Grh in diverse biological processes and the well-documented context-dependent binding and activity of these factors makes them excellent models to explore how PF binding is regulated and how PF occupancy relates to chromatin opening.
+Because many previous studies have relied on loss-of-function experiments, it has been difficult to disentangle causal relationships between chromatin structure and PF activity.
+We therefore expressed Zld and Grh in Drosophila Schneider 2 (S2) cells, where neither factor is endogenously expressed.
+When ectopically expressed, Zld and Grh bind to closed chromatin, establish chromatin accessibility, and promote transcriptional activation at a subset of binding sites.
 We show that an additional TF, Twist (Twi), can function as a pioneer factor.

### 2023-03-15 manuscript/manuscript_sections/_03_Results.qmd
incorporate MMH comments from 2023-03-13

@@
 ## Ectopically expressed pioneer factors bind and open closed chromatin
-
-To investigate the mechanisms by which pioneer factors bind and open chromatin, we generated S2 cell lines to enable the individual expression of Zld or Grh (Extended Data Fig. 1a).
-Neither pioneer factor is normally expressed in WT S2 cells (Extended Data Fig. 1b,c,d).
-As these cell lines allowed the inducible and tunable expression of each factor, we determined the concentration of CuSO4 required to induce expression at levels similar to the endogenous embryonic expression levels of each factor (Extended Data Fig.
-1c,d).
-Approximately physiological protein levels were achieved with 1000 mM or 100 mM CuSO4 for Zld or Grh, respectively.
 
@@
 
-To determine how exogenous Zld and Grh interact with preexisting chromatin, we expressed each factor at approximately physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
-We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
-Additional ChIP experiments were performed in WT cells using anti-Zld, or anti-Grh to control for antibody specificity.
+To investigate the mechanisms by which pioneer factors bind and open chromatin, we generated S2 cell lines to enable the individual expression of Zld or Grh (Extended Data Fig. 1a).
+These cell lines allowed the inducible and tunable expression of each factor, as neither pioneer factor is normally expressed in WT S2 cells (Extended Data Fig. 1b,c,d).
+We expressed each factor at approximately physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
+Approximately physiological protein levels were achieved with 1000 mM or 100 mM CuSO4 for Zld or Grh, respectively (Extended Data Fig.
+1c,d).We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
+Additional ChIP experiments were performed in wild-type cells using anti-Zld, or anti-Grh to control for antibody specificity.
 By using these datasets as controls for peak calling, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh (Extended Data Fig.
@@
 
-Having established that Zld and Grh can bind thousands of sites in S2 cells, we performed assay for transposase-accessible chromatin (ATAC)-seq on WT cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
-Using differential accessibility analysis comparing Zld or Grh-expressing cells to WT cells, we detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
+Having established that Zld and Grh bind thousands of sites in S2 cells, we performed assay for transposase-accessible chromatin (ATAC)-seq on wild-type cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
+Using differential accessibility analysis comparing Zld or Grh-expressing cells to wild-type cells, we detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
 3a,d).
-
-To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of pioneer factor bound sites (Fig. 1a,b).
-Class I sites are accessible in WT cells and remain accessible when bound by the ectopically expressed factor.
-Class II sites are inaccessible in WT cells.
+By integrating our ChIP-seq and ATAC-seq data, we identified three classes of PF-bound sites (Fig. 1a,b).
+Class I sites are accessible in wild-type cells and remain accessible when bound by the ectopically expressed factor.
+Class II sites are inaccessible in wild-type cells.
 However, binding of Zld or Grh does not lead to chromatin opening.
-Class III sites are inaccessible in WT cells, but become accessible upon pioneer-factor binding.
+Class III sites are inaccessible in wild-type cells, but become accessible upon PF binding.
 The majority of ChIP-seq peaks for Zld and Grh were class I sites.
@@
 While class I binding sites occured predominantly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
-Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
+Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in wild-type cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
 1e,h).
@@
 Zld and Grh protein was detectable at 4 or 12 hours following induction, respectively (Extended Data Fig. 2a-b).
-We used CUT&RUN [@Skene2017] to measure Zld and Grh binding to chromatin at 0, 4, 12, 24 or 48 hours following induction.
+We used CUT&RUN to measure Zld and Grh binding to chromatin at 0, 4, 12, 24 or 48 hours following induction[@Skene2017].
 Both proteins were bound to class I regions even in uninduced cells (0H), suggesting that these sites can be occupied even when protein levels are below the limit of detection of immunoblotting (Extended Data Fig 2c-d).
-Class II and III regions were bound at 4 hours, with slight increases in occupancy at later time points.
-These data show that Zld and Grh bind to closed chromatin rapidly following induction of protein expression.
+Class II and III regions were bound at 4 hours, with slight increases in occupancy at later time points, demonstrating that Zld and Grh bind closed chromatin rapidly following induction of protein expression.
 
@@
 
-Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation.
-To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq.
-Differential expression analysis using DESeq2 to compare induced cells to WT cells without Zld or Grh expression [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig.
-3b,e).
-The genes activated by Grh were enriched for gene ontology (GO) terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig. 3c).
-Genes activated by Zld expression were enriched for embryonic morphogenesis, consistent with the known function of Zld, as well as imaginal disc development, in which Zld-target genes have not been well characterized (Extended Data Fig.
+To identify whether Zld- and Grh-mediated chromatin accessibility led to transcriptional activation, we performed RNA-seq.
+Differential expression analysis using DESeq2 to compare induced cells to wild-type cells identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig.
+3b,e)[@love2014].
+The genes activated by Grh were enriched for gene ontology (GO) terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig. 3c)[@Nevil2017].
+Genes activated by Zld expression were enriched for embryonic morphogenesis, consistent with the known function of Zld[@Nien2011], as well as imaginal disc development, in which Zld-target genes have not been well characterized (Extended Data Fig.
 3f).
 
-To better understand the relationship between chromatin opening and transcriptional activation, we directly compared our RNA-seq and ATAC-seq data.
+We compared our RNA-seq and ATAC-seq data, allowing us to investigate the relationship between chromatin opening and transcriptional activation.
 Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l).
-When we individually analyzed our three classes of binding sites for changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
+When we individually analyzed the three classes of binding sites for changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
 This was particularly evident at class III promoters.
-Together, we conclude that while binding and chromatin opening were not always linked to changes in transcription, increases in chromatin accessibility are correlated with transcriptional activation.
+Thus, increases in chromatin accessibility were correlated with transcriptional activation, although binding and chromatin opening did not always result in gene activation.
 
@@
 
-Because both Zld and Grh had previously been shown to have features of pioneer factors, we sought to test an additional transcription factor with distinctive features.
-For this purpose, we tested the effects of the expression of the master regulator of mesodermal cell fate, Twist (Twi), which was not expressed in our WT S2 cells.
-Although like Zld and Grh, Twi is essential for regulating Drosophila embryonic development, Twi has a number of properties that are distinct.
-For example, many Twi-binding sites are developmentally dynamic and are not maintained through embryonic development [@Wilczynski2010a].
-Furthermore, Zld pioneering activity is required for Twi binding in the early embryo as mutations to Zld-binding sites disrupted Twi binding to the *cactus* enhancer [@Yáñez-Cuna2012].
+Zld and Grh had previously been shown to have features of PFs.
+To determine whether these factors had unique properties, we tested an additional transcription factor with distinctive features, Twist (Twi).
+Twi is a master regulator of mesodermal cell fate and, importantly, was not expressed in our wild-type S2 cells.
+Although like Zld and Grh, Twi is essential for regulating *Drosophila* embryonic development, Twi has properties that are distinct from these PFs.
+Unlike Grh, many Twi-binding sites are developmentally dynamic and are not maintained through embryonic development [@Wilczynski2010a].
+Additionally, Zld pioneering activity is required for Twi binding in the early embryo, as mutations to Zld-binding sites disrupted Twi binding to the *cactus* enhancer [@Yáñez-Cuna2012].
+Deep learning models trained on ChIP-seq and ATAC-seq data for Zld and Twi suggest that Zld motifs contribute to chromatin accessibility and Twi binding, while Twi motifs are not predictive of chromatin accessibility [@brennan2022].
 To test the properties of Twi in our system, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. 4a).
-We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
+As we had done for Zld and Grh, we determined Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
 Using these data, we identified the same three classes of binding sites that we described for Zld and Grh (Fig 2a,b).
-Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
+Class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
 Twi bound extensively to closed chromatin, with `r perc_twi_closed`% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
@@
 RNA-seq revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4c).
-GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. 4d).
-Thus, despite Twi possessing features distinct from Zld and Grh in vivo, ectopically expressed Twi could bind closed chromatin and promote accessibility.
+GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal genes, consistent with the endogenous function of Twi (Extended Data Fig. 4d) [@Wilczynski2010a].
+Despite Twi possessing features distinct from Zld and Grh *in vivo*, ectopically expressed Twi binds closed chromatin and promotes accessibility.
 
@@
 
-While comparing ChIP-seq data for Zld, Grh, and Twi, we observed that many class I binding sites were shared across all three factors (Fig. 3a).
+Comparing binding sites for Zld, Grh, and Twi showed that many class I sites were common to all three factors (Fig. 3a).
 This contrasted with class II and III peaks, which were largely specific to each individual factor.
-<!--# Control IP experiments in WT S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig.
-XX).
---> We hypothesized that the overlapping class I binding sites may occur at active cis-regulatory regions that are active in WT cells.
-To test this, we analyzed published ChIP-seq data from WT S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV.
+Control IP experiments in wild-type S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig. 1e-f).
+We hypothesized that the overlapping class I binding sites may occur at active *cis*-regulatory regions in wild-type cells.
+To test this, we analyzed published ChIP-seq data from S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV[@Henriques2018; @Philip2015; @Tettey2019].
 Overlapping class I regions showed high levels of these chromatin marks (Fig. 3b), suggesting that exogenously expressed factors may bind nonspecifically to active regions of the genome.
-<!--# This model is supported by comparison to RNA-seq and ATAC-seq data, showing that binding to class I regions leads to few changes in chromatin accessibility and gene expression (Extended Data Fig.
-XX).
--->
 
@@
 
-Considering the enrichment of marks of active chromatin at class I sites, we asked if there were any features of chromatin distinctive to class II or III regions.
+Considering the enrichment of marks of active chromatin at class I sites, we tested if there were any features of chromatin distinctive to class II or III regions.
 Analysis of published ChIP-seq data demonstrated that class II and III sites have low levels of many histone modifications, chromatin-associated proteins, and transcription factors (Extended Data Fig.
-5a,c,e).
-This supports the hypothesis that pioneer factors preferentially bind naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
-The ability of transcription-factor binding to promote chromatin accessibility at class III sites and not class II sites was not explained by a pre-existing lack of histones.
-In fact, the levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig. 5).
+5a,c,e) [@Henriques2018; @Philip2015; @Tettey2019; @Straub2013; @Zouaz2017; @Climent-Cantó2020; @Liu2017a; @Enderle2011; @Jain2020; @Eastwood2021; @Alekseyenko2014; @Liang2014; @Ong2013].
+This supports the hypothesis that PFs preferentially bind naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
+The ability of TF binding to promote chromatin accessibility at class III sites and not class II sites was not explained by a pre-existing lack of histones; the levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig.
+5).
 While most class III sites were regions of naïve chromatin, a subset were within regions marked with the repressive histone modification H3K27me3 (Extended Data Fig. 5b,d,f).
@@
 
-To further determine the features that drive the different outcomes at pioneer-bound regions, we analyzed the enrichment for the canonical Zld, Grh or Twi motifs for each of the three bound classes.
+To further determine the features that drive the different outcomes on chromatin accessibility at PF-bound regions, we analyzed enrichment of the canonical Zld, Grh or Twi motifs for each of the three bound classes.
 We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig. 3c).
@@
 For all three factors, class III sites tended to have stronger motifs than class II sites (Fig. 3g,i,k).
-Together, this analysis suggests that a combination of both a higher number of motifs and stronger motifs at class III sites enables Zld, Grh, and Twi to open chromatin at these regions.
+These data suggest that a more and stronger motifs at class III sites as compared to class II sites facilitates the ability of Zld, Grh, and Twi to open chromatin.
 
-Previous work showed that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes [@Zhu2018; @Michael2020].
-To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
+Previous work showed that binding of some PFs depends on how target motifs are positioned on nucleosomes [@Zhu2018; @Michael2020].
+To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from S2 cells [@Chereji2019].
 Analysis of average MNase-seq signal at class I-III peaks indicated that class III sites have high levels of nucleosome occupancy that, on average, occur over the respective motif (Extended Data Fig.
@@
 For class II and III peaks, some motifs were positioned close to the nuclesome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
-Additionally, the MNase signal in many regions did not show strongly positioned nucleosomes, likely due to the fact that nucleosomes at silent regions often occur at different positions in different cells [@Stergachis2020; @Abdulhay2020].
-Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
+Additionally, the MNase signal in many regions did not show strongly positioned nucleosomes, likely because nucleosomes at silent regions often are not strongly positioned and show variable distributions in different cell types [@Stergachis2020; @Abdulhay2020].
+While some PFs may recognize their motif when specifically positioned on a nucleosome, Zld, Grh and Twi do not have a strong preference for motif positioning when exogenously expressed.
 
@@
 
-Prior studies have extensively characterized the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
-To better understand how cell-type specific variables affect pioneer-factor binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
-We compared Zld-binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
-We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs (Nevil et al. 2017), and to Twi ChIP-seq data from early embryos (1-3 hours) (Ozdemir et al. 2011).
-For each factor, we identified many binding sites that were bound *in vivo*, but were not bound upon ectopic expression in S2 cells (Extended Data Fig. 7a,b).
-Chromatin enriched for repressive histone modifications H3K9me3 and H3K27me3 has been shown to be resistant to pioneer-factor binding [@Mayran2018a; @Soufi2012].
-To test if these repressive marks were present in S2 cells at the resistant sites (bound in vivo but not upon ectopic expression), we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
-Using these data, we divided the resistant sites into three classes: class IV sites have high levels of H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a-c; Extended Data Fig.
+Prior studies have extensively characterized the *in vivo* binding sites of Zld, Grh, and Twi during normal *Drosophila* development.
+To better understand how cell-type specific variables affect PF binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
+We analyzed Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a], Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs (Nevil et al.
+2017), and Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
+We identified many binding sites that were bound *in vivo*, but were not bound upon ectopic expression in S2 cells (Extended Data Fig. 7a,b).
+Chromatin enriched for repressive histone modifications H3K9me3 and H3K27me3 is resistant to binding of some PFs [@Mayran2018a; @Soufi2012].
+To test if these repressive marks were present in S2 cells at the resistant sites (bound *in vivo* but not upon ectopic expression), we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
+We divided the sites bound *in vivo* but not in our S2 cell system into three classes: class IV sites have high levels of H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a-c; Extended Data Fig.
 7d-l).
 Despite the presence of class IV, V and VI binding sites for all three proteins, there were distinct patterns of binding across different tissues.
-Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
+Zld binding in S2 cells was more similar to that of neural stem cells (NSC) than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
 This differed for Grh, in which many of the strongest binding sites were shared across the three tissues (Fig. 4b).
@@
 
-To test the functional relevance of repressive chromatin to pioneer-factor binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3.
-Despite causing dramtic reductions in H3K27me3 levels, tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig.
+To test the functional relevance of repressive chromatin to PF binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3.
+Despite causing dramatic reductions in H3K27me3 levels, tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig.
 4d-f; Extended Data Fig.
 8a).
-The majority of class IV binding sites remained unbound in tazemetostat-treated cells, demonstrating that loss of PRC2 activity is insufficient to promoted widespread binding.
-Thus, H3K27me3 is broadly restricting binding in S2 cells.
-These data demonstrate that despite the correlation of H3K27me3 with resistance to pioneer factor binding, loss of PRC2 activity is insufficient to promote widespread binding of Zld, Grh, or Twi to repressed binding sites.
+The majority of class IV binding sites remained unbound in tazemetostat-treated cells, demonstrating that loss of PRC2 activity is insufficient to promote widespread binding of any of the three factors studied.
 
@@
 
-Previous studies proposed that pioneer-factor activity depends on protein concentration [@Hansen2022; @Blassberg2022].
-Considering that tazemetostat treatment had limited effects on pioneer factor binding, we asked if varying the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.
+Previous studies proposed that PF activity depends on protein concentration [@Hansen2022; @Blassberg2022].
+Therefore, we tested if differences in the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.
 We performed ChIP-seq on cells induced to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig.
@@
 Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i).
-Thus, increased expression allowed pioneer factors access to previously unbound regions.
-To identify the relationship between pioneer-factor binding and chromatin accessibility, we performed ATAC-seq on cells induced with the same range in expression.
-Changes in chromatin accessibility were similarly concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig. 5b,f,j).
+To identify the relationship between PF binding and chromatin accessibility, we performed ATAC-seq on cells induced with the same range in expression.
+As we identified for PF binding, changes in chromatin accessibility were concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig.
+5b,f,j).
 Many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels, suggesting that increased chromatin occupancy, as assayed by ChIP-seq, could result in chromatin accessibility.
@@
 9d-f).
+
 To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration.
@@
 For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration.
-Together, these data demonstrate that both binding and chromatin opening by pioneer factors are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are resistant to binding at physiological levels.
+Together, these data demonstrate that both binding and chromatin opening by PFs are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are resistant to binding at physiological levels.
 
@@
 
-Binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) [@Zaret2020].
+Binding to nucleosomes by PFs requires recognition of target motifs by DNA-binding domains (DBDs) [@Zaret2020].
 However, little is known about how regions outside the DBD may contribute to pioneering function.
-We generated stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig. 10a,b).
-Immunostaining confirmed that the each DBD alone was properly localized to the nucleus (Extended Data Fig.
-10c-d) To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq.
+We generated stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length protein (Extended Data Fig. 10a,b).
+Immunostaining confirmed that the each DBD alone was properly localized to the nucleus (Extended Data Fig. 10c-d).
+We performed ChIP-seq and ATAC-seq on cells expressing approximately physiological levels of the DBDs alone.
 Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was unable to bind closed chromatin at class II or III regions (Fig.
 6a,c).
-We previously showed that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
-Because many Zld-binding sites in S2 cells also lack a canonical CAGGTA motif (Fig.
-3), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
-To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
-This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required (Extended Data Fig. 10a).
-Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
-For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
+Binding of the Grh DBD to class II-III regions was strongly reduced compared to the full-length protein, although some closed regions retained low levels of binding (Fig. 6b,e).
+For both Zld and Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
 Together, these experiments show that regions outside the DBD are essential for binding and opening closed chromatin.
@@
 **d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
+**g,** Model for the regulation of PFs.
+A high level of chromatin occupancy is required for PFs to initiate chromatin opening.
+Chromatin occupancy can be regulated by protein-intrinsic features such as motif affinity and protein domains, and by cell-intrinsic properties such as histone modifications, PF concentration, and cell-type specific cofactors.
 :::

### 2023-03-15 manuscript/manuscript_sections/_04_Discussion.qmd
update figs
incorporate MMH comments from 2023-03-13

@@
 
-By exogenously expressing Zld, Grh, and Twi in S2 cells where they are not normally expressed, leveraging prior data identifying binding sites for these factors in vivo, and analyzing the wealth of data on chromatin structure in S2 cells, we identified features that govern PF activity.
+By exogenously expressing Zld, Grh, and Twi in S2 cells where they are not normally expressed, leveraging prior data identifying binding sites for these factors *in vivo*, and analyzing the wealth of data on chromatin structure in S2 cells, we identified features that govern PF activity.
 Studying the two well-defined PFs Zld and Grh, along with Twi, which possess features distinct from these two factors, allowed us to determine general properties regulating PF activity.
 Furthermore, our analysis of both binding and accessibility allowed us to separate PF occupancy from activity.
-We find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin structure, protein concentration, DNA motif content, and protein domains outside the DBD.
+We find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin structure, protein concentration, DNA motif content, and protein domains outside the DBD (Fig.
+6g).
+
 Even when expressed at low levels Zld, Grh and Twi bind promiscuously to active, accessible regions with degenerate motifs or no detectable motifs at all.
@@
 Here, we specifically tested whether H3K27me3 regulates cell-type specific occupancy of Zld, Grh and Twi using the PRC2 inhibitor tazmetostat.
-While H3K27me3 levels were dramatically decreased in our treatment, PF binding was largely unchanged.
+While H3K27me3 levels were decreased in our treatment, PF binding was largely unchanged.
 Thus, unlike H3K9me3 in the case of OSKM, H3K27me3 is not a barrier to binding by the three factors assayed and suggest other features are restricting binding to these regions.
+
 While depleting H3K27me3 levels did not result in Zld, Grh and Twi binding to additional regions, increased expression of both Zld and Grh resulted in binding to regions that were unbound at lower expression levels.
@@
 Tissue-specific PF occupancy is also likely regulated by the collection of cofactors expressed in a given tissue.
-This may partially explain the presence of class VI sites that lack repressive histone modifications in culture and are bound in vivo but not when expressed in S2 cells.
+This may partially explain the presence of class VI sites that lack repressive histone modifications in culture and are bound *in vivo* but not when expressed in S2 cells.
 For example, in mammalian cell culture FOXA2 occupancy at a subset of tissue-specific binding sites was increased when GATA4 was present [@Donaghey2018].
@@
 Along with prior studies, our data suggest that PF occupancy is regulated by tissue-intrinsic features, including levels of PF expression, the complement of cofactors expressed and chromatin structure.
+
 In addition to cell-type specific factors, we demonstrate that protein-intrinsic features control PF binding.
@@
 Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and have demonstrated the role that intrinsically disordered regions (IDRs) play in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
-For Zld, this disordered domain is required for transcriptional activation and promotes nucleosome binding in vitro [@Hamm2015; @McDaniel2019].
-IDRs may contribute to transcription-factor binding through protein-protein interactions with other transcription factors, nonspecific interactions with DNA, or by driving condensate formation [@Brodsky2021; @Staller2022].
+For Zld, this disordered domain is required for transcriptional activation and promotes nucleosome binding *in vitro* [@Hamm2015; @McDaniel2019a].
+IDRs may contribute to TF binding through protein-protein interactions with other transcription factors, nonspecific interactions with DNA, or by driving condensate formation [@Brodsky2021; @Staller2022].
 Indeed, Zld is visualized in hubs within the nucleus, and these promote binding of additional transcription factors [@Mir2018; @Yamada2019].
@@
 To provide a more complete understanding of how cis-regulatory regions are established, we must further investigate how IDRs contribute to PF binding and chromatin accessibility.
+
 The failure of the DBD alone to promote chromatin accessibility implicates a separation between the ability to bind nucleosomes and the capacity to promote chromatin accessibility.

### 2023-03-15 manuscript/manuscript_sections/_05_Methods.qmd
incorporate MMH comments from 2023-03-13

@@
 ---
-editor: 
-  markdown: 
-    wrap: sentence
+# bibliography: ../references.bib
 ---
-
-## Reagents
-
-| Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
-|-----------------------|-----------------|-----------------|-----------------|
-| pMT-puro                              | AddGene                   | 17923                |                         |
-| Gibson Cloning                        | New England Biolabs       | E2611                |                         |
-| Schneider's Medium                    | Thermo Fisher Scientific  | 21720001             |                         |
-| Fetal Bovine Serum                    | Omega Scientific          | FB-02                |                         |
-| Antibiotic/antimycotic                | Thermo Fisher Scientific  | 15240062             |                         |
-| Effectene Transfection Reagent        | Qiagen                    | 301427               |                         |
-| PVDF Membrane                         |                           |                      |                         |
-| Formaldehyde                          |                           |                      |                         |
-| Protease inhibitor tablets            |                           |                      |                         |
-| NEBNext Ultra II DNA library prep kit |                           |                      |                         |
-| TruSeq v2 RNA library prep kit        |                           |                      |                         |
-| Trizol                                |                           |                      |                         |
-| **Antibodies**                        |                           |                      |                         |
-| Anti-Zld                              | Harrison et al., 2010     |                      | ChIP: 6 µL , WB: 1:750  |
-| Anti-Grh                              | Harrison et al., 2010     |                      | ChIP: 8 µL, WB: 1:1000  |
-| Anti-Twi                              | He et al., 2015           |                      | ChIP: 10 µg, WB: 1:1000 |
-| Anti-HA-peroxidase clone 3F10         | Roche                     | 12013819001          | WB: 1:500               |
-| Anti-Tubulin DM1A                     | Sigma                     | T6199-200UL          | WB: 1:5000              |
-| goat anti-rabbit IgG-HRP conjugate    | Bio-Rad                   | 1706515              | WB: 1:3000              |
-| Anti-HA clone 12CA5                   | Roche                     |                      | ChIP: 7.5 µL            |
-| Anti-H3K27me3                         | Cell Signaling Technology | C36B11; Lot #19      | WB: 1:1000, CR: 0.5 µL  |
-|                                       |                           |                      |                         |
-|                                       |                           |                      |                         |
-
-##
 
@@
 
-We validated our wild-type (WT) S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers and the absence of Kc cell markers [@Cherbas2011].
-Drosophila S2 cells were cultured at 27°C in Schneider's medium (Thermo Fisher Scientific) supplemented with 10% fetal bovine serum (Omega Scientific) and antibiotic/antimycotic (Thermo Fisher Scientific).
+We validated our wild-type S2 cells by performing RNA-seq and verifying the expression of known S2 cell markers and the absence of Kc cell markers [@Cherbas2011].
+S2 cells were cultured at 27°C in Schneider's medium (Thermo Fisher Scientific) supplemented with 10% fetal bovine serum (Omega Scientific) and antibiotic/antimycotic (Thermo Fisher Scientific).
 For the generation of stable cell lines, cells were plated at 5✕10^5^ cells per mL.
@@
 Proteins were separated on denaturing polyacramide gels before transfer to a 0.45 μm Immobilon-P PVDF membrane (Millipore) in transfer buffer (25 mM Tris, 200 mM Glycine, 20% methanol) for 60 min (75 min for Zld) at 500mA at 4°C.
-The membranes were blocked with BLOTTO (2.5% non-fat dry milk, 0.5% BSA, 0.5% NP-40, in TBST) for 30 min at room temperature and then incubated with anti-Zld (1:750) (Harrison et al., 2010), anti-Grh (1:1000) (Harrison et al., 2010), anti-Twi (1:1000) (He et al.
-2015), anti-HA-peroxidase (1:500) (clone 3F10, Roche), or anti-Tubulin (DM1A, 1:5000) (Sigma), overnight at 4°C.
-The secondary incubation was performed with goat anti-rabbit IgG-HRP conjugate (1:3000) (Bio-Rad) for 1 hr at room temperature.
+The membranes were blocked with BLOTTO (2.5% non-fat dry milk, 0.5% BSA, 0.5% NP-40, in TBST) for 30 min at room temperature and then incubated with anti-Zld (1:750) [@Harrison2010], anti-Grh (1:1000) [@Harrison2010], anti-Twi (1:1000) [@He2015], anti-HA-peroxidase (1:500) (clone 3F10, Roche), or anti-Tubulin (DM1A, 1:5000) (Sigma), overnight at 4°C.
+The secondary incubation was performed with goat anti-rabbit IgG-HRP conjugate (1:3000) or goat anti-mouse IgG-HRP conjugate (1:3000) (Bio-Rad) for 1 hr at room temperature.
 Blots were treated with SuperSignal West Pico PLUS chemiluminescent substrate (Thermo Fisher Scientific) and visualized using the Azure Biosystems c600 or Kodak/Carestream BioMax Film (VWR).
@@
 
+Immunostaining was performed as described previously[@angshuman2007].
+Briefly, 1×10^6^ cells in 200 µL PBS were applied to a coverslip pre-coated with 10% poly-L-lysine.
+Cells were fixed with 4% methanol-free formaldehyde for 30 minutes at 37°C, permeabilized with 0.2% Triton-X 100 for 10 minutes at room temperature, and blocked with 1% BSA for 1 hour.
+Cells were incubated with a 1:500 dilution of anti-Zld or anti-Grh primary antibody overnight at 4°C.
+After washing with PBS, cells were incubated with a 1:500 dilution of goat anti-rabbit IgG DyLight 488 conjugated secondary antibody (Thermo Fisher Scientific #35552) for 1 hour at RT.
+Imaging was performed at 60X on a Nikon Ti2 epi-fluorescent microscope.
+
 ## ChIP-seq
@@
 For each ChIP experiment, \~50 million cells were crosslinked by adding methanol-free formaldehyde directly to the cell culture medium to a final concentration of 0.8%.
-Cells were rotated on a nutator for 7 minutes at room temperature (RT) before quenching of crosslinking by adding glycine to 125 mM.
+Cells were rotated on a nutator for 7 minutes at room temperature before quenching of crosslinking by adding glycine to 125 mM.
 Crosslinked cells were pelleted by centrifugation at 600 × *g* for 3 minutes.
@@
 
-Bowtie2 version 2.4.4 [@Langmead2012] was used to align ChIP-seq reads to the *Drosophila melanogaster* genome (version dm6) with the following non-default parameters: -k 2 \--very-sensitive \--no-mixed \--no-discordant -X 5000.
+Bowtie2 version 2.4.4 [@Langmead2012] was used to align ChIP-seq reads to the *Drosophila melanogaster* genome (version dm6) with the following non-default parameters: `-k 2 --very-sensitive --no-mixed --no-discordant -X 5000`.
 Aligned reads were filtered to include only reads with a mapping quality score \> 30.
 Reads mapping to unplaced scaffolds or the mitochondrial genome were discarded.
-Peak calling was performed using MACS2 version 2.2 with parameters -g dm \--call-summits.
+Peak calling was performed using MACS2 version 2.2 with parameters `-g dm --call-summits`.
+Data from ChIP experiments performed with the same antibody in WT cells were used as a control for peak calling.
 Only peaks that were detected in all replicates were considered in downstream analysis.
-bigWig files were generated using deepTools bamCoverage version 3.5.1 with parameters --binSize 10.
+bigWig files were generated using deepTools bamCoverage version 3.5.1 with parameters `–binSize 10`.
 bigWig files were z-score normalized by subtracting the genome-wide average of all bins from each bin value and dividing by the standard deviation of all bins.
@@
 
-Raw reads were aligned to the Drosophila melanogaster (dm6) genome using HISAT2 version 2.1.0 with parameters -k 2 --very-sensitive.
+Raw reads were aligned to the Drosophila melanogaster (dm6) genome using HISAT2 version 2.1.0 with parameters `-k 2 –very-sensitive`.
 Reads with a mapping quality score \< 30 and reads aligning to the mitochondrial genome or unplaced scaffolds were discarded.
@@
 
-ATAC-seq was performed as described previsouly [@Buenrostro2013].
+ATAC-seq was performed as described previously [@Buenrostro2013].
 2×10^5^ cells were harvested for ATAC-seq by centrifugation at 600 × *g* for 3 minutes at RT.
@@
 Raw ATAC-seq reads were trimmed to remove adapter sequences using NGmerge version 0.3 [@Gaspar2018].
-Trimmed reads were aligned to the *Drosophila melanogaster* genome (version dm6) using bowtie2 with paramters -k 2 \--very-sensitive \--no-mixed \--no-discordant -X 5000.
+Trimmed reads were aligned to the *Drosophila melanogaster* genome (version dm6) using bowtie2 with paramters `-k 2 --very-sensitive --no-mixed --no-discordant -X 5000`.
 Reads with a mapping quality \< 30 and reads aligning to the mitochondrial genome or scaffolds were discarded.
 As described previously, only fragments \< 100bp were considered for downstream analysis [@Buenrostro2013].
-Reads were combined across all replicates and peak calling was performed using MACS2 with parameters -f BAMPE \--keep-dup all -g dm \--call-summits.
+Reads were combined across all replicates and peak calling was performed using MACS2 with parameters `-f BAMPE --keep-dup all -g dm --call-summits`.
 featureCounts was used to count the number of reads aligning within 200bp (100bp upstream or downstream) of peak summits.
@@
 Overnight incubation with antibodies was performed in antibody buffer containing 0.05% digitonin.
-0.5 µL anti-H3K27me3 (Cell signaling technologies) or rabbit IgG were used for CUT&RUN.
-Prior to addition of antibody, a 1:5 dilution of K-MetStat Panel spike-in nucleosomes (EpiCypher, cat. \# 19-1002) was added to each reaction.
+0.5 µL anti-H3K27me3 (Cell signaling technologies), anti-Zld, anti-Grh, or rabbit IgG were used for CUT&RUN.
+For H3K27me3 CUT&RUN, a 1:5 dilution of K-MetStat Panel spike-in nucleosomes (EpiCypher, cat. \# 19-1002) was added to each reaction prior to addition of antibody.
 Libraries were prepared using the NEBNext Ultra II kit (NEB).
@@
 Trimming, alignment, and quality filtering of CUT&RUN reads was performed as described above for ATAC-seq data.
-Peak calling was performed using MACS2 with parameters --broad -f BAMPE --keep-dup all -g dm.
+Peak calling was performed using MACS2 with parameters `–broad -f BAMPE –keep-dup all -g dm` for H3K27me3 or parameters `-f BAMPE --keep-dup all -g dm --call-summits` for Zld and Grh.
 Generation of z-score normalized bigWig files was performed as described above for ChIP-seq data.
-The percentage of reads containing barcoded Epicypher nucleosomes was calculated and 1 divided by this percentage was used as the scaling factor for spike-in normalization.
+For H3K27me3 data, the percentage of reads containing barcoded Epicypher nucleosomes was calculated and 1 divided by this percentage was used as the scaling factor for spike-in normalization.
 The z-score normalized read depth was multiplied by these scaling factors to generate spike-in normalized bigWigs.
@@
 
-Previously published datasets analyzed in this study can be found in supplemental table XXXX.
+Previously published datasets analyzed in this study can be found in supplemental table 1.
 All previously published data were analyzed in parallel with data from this study using analysis parameters described above.

### 2023-03-15 manuscript/manuscript_sections/_06_Figure_Legends.qmd
incorporate MMH comments from 2023-03-13

--- 2023-02-22 manuscript/manuscript_sections/_06_Figure_Legends.qmd
@@
-**Figure 1: Ectopic pioneer factors open chromatin and activate transcription.
+**Figure 1: Ectopically expressed pioneer factors open chromatin and activate transcription.
 a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
-ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
+ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for wild-type (WT) cells lacking Zld or Grh expression.
 **c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
@@
 i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
-**j,i,** Proportion of class I, II or III regions that are proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
+**j,i,** Proportion of class I, II or III regions that is proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
 Promoter-proximal and promoter-distal binding sites are shown separately.
@@
 
-**Figure 3: Motif content shapes pioneer-factor activity.
-a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
+**Figure 3: Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
 **b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
 Zld, Grh, and Twi ChIP-seq data were generated in this study.
-Remaining ChIP-seq datasets were previously published (see Supplementary table XX).
+Remaining ChIP-seq datasets were previously published (see Supplementary table 1)[@Henriques2018;@Philip2015;@Tettey2019].
 **c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
 Logos plot of the canonical motif are shown alongside the heatmap.
-Percentages are shown for class I, II or III binding sites, or all WT ATAC-seq peaks as a control for the background motif frequency.
+Percentages are shown for class I, II or III binding sites, or all wild-type ATAC-seq peaks as a control for the background motif frequency.
 **f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
@@
 a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
-ChIP-seq data for repressive histone modifications H3K27me3 and H3K9me3 is also shown.
+ChIP-seq data for repressive histone modifications H3K27me3 [@Jain2020] and H3K9me3[@Eastwood2021] in WT S2 cells is also shown.
 **d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
@@
 
-**Figure 5: Binding and opening of closed chromatin is concentration dependent.
+**Figure 5: Pioneer-factor binding and opening of closed chromatin is concentration dependent.
 a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
@@
 a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
-Shown for Zld **(a)** and Grh **(b)**.
+Adjacent class I regions are shown where the DBD alone can bind.
+Tracks shown for Zld **(a)** and Grh **(b)**.
 **c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**.
 **d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
-
-**Figure 7: Pioneer-factor function requires stable chromatin occupancy.**
+**g,** Model for the regulation of PFs.
+A high level of chromatin occupancy is required for PFs to initiate chromatin opening.
+Chromatin occupancy can be regulated by protein-intrinsic features such as motif affinity and protein domains, and by cell-intrinsic properties such as histone modifications, PF concentration, and cell-type specific cofactors (outlined in blue).
 
@@
 Two independently generated cell lines are shown and compared to 2-3H embryos.
+Black arrowheads indicate Zld **(c)** or Grh **(d)**.
+Gray arrowheads indicate background bands used to assess loading.
+**e-f,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in WT cells.
 
@@
 a-b,** Immunoblots showing time course of Zld **(a)** or Grh **(b)** protein expression following induction of stable cell lines.
+Black arrowheads indicate Zld **(a)** or Grh **(b)**.
+Gray arrowheads indicate background bands used to assess loading.
 **c-d,** Metaplots showing average z-score normalized CUT&RUN signal at class I, II or III sites at different time points following induction.
@@
 a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
-**b,e,** RNA-seq volcano plots showing gene expression changes in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
+**b,e,** RNA-seq volcano plots showing gene expression changes in cells expressing Zld **(b)** or Grh **(e)** when compared to WT cells treated with the same concentration of CuSO~4~.
 **c,f,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(c)** or Grh **(f)**.
 
-**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation.
-a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
+**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
 Protein levels in stable cell lines are compared to 3-4 hour old embryos.
@@
 **Extended Data Figure 5: Chromatin features associated with Zld, Grh and Twi binding sites.
-a,c,e,** Heatmaps showing the levels of different chromatin marks in class I, II or III regions for Zld (a), Grh (c) or Twi (e).
+a,c,e,** Heatmaps showing the levels of different chromatin marks in class I, II or III regions for Zld **(a)**, Grh **(c)** or Twi **(e)**.
 The color represents the average z-score normalized read depth across a 1 KB region surrounding the center of class I, II or III ChIP-seq peaks.
-**b,d,f,** Example genome browser tracks for class III regions with high levels of H3K27me3 for Zld (b), Grh (d), or Twi (f).
+**b,d,f,** Example genome browser tracks for class III regions with high levels of H3K27me3 for Zld **(b)**, Grh **(d)**, or Twi **(f)**.
 
 **Extended Data Figure 6: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes.
-a-c,** Metaplots showing average MNase signal at motifs within class I, II, or III regions for Zld (a), Grh (b), or Twi (c).
-**d-f,** Heatmaps showing MNase signal at motifs within class II and III regions for Zld (d), Grh (e), or Twi (f).
+a-c,** Metaplots showing average MNase signal from WT cells centered on motifs within class I, II, or III regions for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
+**d-f,** Heatmaps showing MNase signal centered on motifs within class II and III regions for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
 Rows are ordered based on hierarchical clustering to highlight the various patterns of MNase signal around motifs.
@@
 **Extended Data Figure 7: Zld, Grh and Twi display cell-type specific binding.
-a-c,** Venn diagrams showing overlap between ChIP-seq peaks identified in different tissues for Zld (a), Grh (b) or Twi (c).
-**d-l,** Genome browser tracks showing examples of class IV, V, and VI regions for Zld (d-f), Grh (g-i) or Twi (j-l).
+a-c,** Venn diagrams showing overlap between ChIP-seq peaks identified in different tissues for Zld **(a)**, Grh **(b)** or Twi **(c)**.
+**d-l,** Genome browser tracks showing examples of class IV, V, and VI regions for Zld **(d-f)**, Grh **(g-i)** or Twi **(j-l)**.
 For each example, the top tracks show H3K27me3 and H3K9me3 signal over a larger region.
@@
 
-**Extended Data Figure 8: PRC2 inhibition is insufficient to promote Zld, Grh, and Twi binding to repressed targets.
-a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
+**Extended Data Figure 8: Specific detection of H3K27me3 in tazemtostat-treated cells. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
 Tubulin levels are shown as a loading control.
+**b,** Heatmap showing specificity of anti-H327me3 antibody in CUT&RUN reactions.
+A panel of barcoded spike-in nucleosomes bearing different modifications was added to each CUT&RUN reaction (see methods).
+For each sample, the heatmap displays the percentage of barcode reads for each sample and histone modification relative to the total number of barcode reads for all modifications.
 
 **Extended Data Figure 9: Expression of Zld, Grh, or Twi at supraphysiological levels results in chromatin opening at a small number of novel binding sites.
-a-c,** Immunoblots showing Zld (a), Grh (b) or Twi (c) protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
-**d-f,** Bar plots showing the percentage of previously defined class I-VI binding sites that are bound by Zld (d), Grh (e), or Twi (f) when expressed at varying concentrations.
-**g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld (g), Grh (h), or Twi (i) are expressed at varying concentrations.
+a-c,** Immunoblots showing Zld **(a)**, Grh **(b)** or Twi **(c)** protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
+**d-f,** Bar plots showing the percentage of previously defined class I-VI binding sites that are bound by Zld **(d)**, Grh **(e)**, or Twi **(f)** when expressed at varying concentrations.
+**g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld **(g)**, Grh **(h)**, or Twi **(i)** are expressed at varying concentrations.
 
 **Extended Data Figure 10: Expression of Zld and Grh DNA-binding domains at protein levels comparable to the full-length proteins.
-a-b,** Immunoblots showing titration of protein levels for Zld (a) or Grh (b) DNA-binding domains to match expression of the full-length proteins.
+a-b,** Immunoblots showing titration of protein levels for Zld **(a)** or Grh **(b)** DNA-binding domains to match expression of the full-length proteins.
 DNA-binding domain protein levels are shown at a range of CuSO~4~ concentrations and compared to an equivalent number of cells expressing full-length Zld or Grh at approximately physiological levels.
-**c,** Metaplots showing the average ChIP-seq signal for full-length Zld or the DNA-binding domain alone at the subset of previously identified class I, II or III binding sites that contain a canonical Zelda CAGGTAG motif.
+**c-d,** Immunofluorescent microscopy images of stable cell lines expressing full-length protein or DBD only for Zld **(c)** or (Grh).
+Stable cell lines are compared to WT cells.
+
+**Supplementary table 1:** List of previously published genomics datasets that were analyzed in this work.

### 2023-03-15 workflow/scripts/extended_data_fig1.R

@@ -363,6 +363,10 @@ plotText(
   x = ref_x + 7.8, y = ref_y + 1.25, just = c("center"), default.units = "cm"
 )
 
+# add arrowheads to blot
+plotPolygon(x = c(1, 1.25, 1), y = c(7.2,7.3,7.4), default.units = "cm", fill = "black")
+
+plotPolygon(x = c(1, 1.25, 1), y = c(8.35,8.45,8.55), default.units = "cm", fill = "grey")
 
 # panel D ======================================================================
 # reference points for positioning figure components
@@ -519,6 +523,12 @@ plotText(
   x = ref_x + 7.8, y = ref_y + 1.25, just = c("center"), default.units = "cm"
 )
 
+# add arrowheads to blot
+plotPolygon(x = c(10, 10.25, 10), y = c(7.3,7.4,7.5), default.units = "cm", fill = "black")
+
+plotPolygon(x = c(10, 10.25, 10), y = c(8.9,9,9.1), default.units = "cm", fill = "grey")
+
+
 # panel E ======================================================================
 # reference points for positioning figure components
 ref_x <- 0.5
@@ -538,8 +548,9 @@ bw <- c(
 )
 
 # define regions to use for heatmap
-regions <- zld_ChIP_classes_fn %>%
+regions <- zld_ChIP_classes_fn |>
   read_tsv() |> 
+  mutate(class = str_to_upper(class)) |> 
   makeGRangesFromDataFrame(keep.extra.columns = TRUE)
 
 
@@ -627,8 +638,9 @@ bw <- c(
 )
 
 # define regions for heatmap
-regions <- grh_ChIP_classes_fn %>%
+regions <- grh_ChIP_classes_fn |>
   read_tsv() |> 
+  mutate(class = str_to_upper(class)) |> 
   makeGRangesFromDataFrame(keep.extra.columns = TRUE)
 
 
@@ -716,8 +728,9 @@ bw <- c(
 )
 
 # define regions to use for heatmap
-regions <- twi_ChIP_classes_fn %>%
+regions <- twi_ChIP_classes_fn |>
   read_tsv() |> 
+  mutate(class = str_to_upper(class)) |> 
   makeGRangesFromDataFrame(keep.extra.columns = TRUE)
 
 

### 2023-03-15 workflow/scripts/extended_data_fig2.R

@@ -21,8 +21,8 @@ source("workflow/scripts/plot_heatmap.R")
 # S2_Grh_CR_12H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_12H_small.bw"
 # S2_Grh_CR_24H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_24H_small.bw"
 # S2_Grh_CR_48H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_48H_small.bw"
-
-
+# 
+# 
 # zld_ChIP_classes_fn <- "results/ChIP_peak_classes/zld_ChIP_classes.tsv"
 # grh_ChIP_classes_fn <- "results/ChIP_peak_classes/grh_ChIP_classes.tsv"
 # 
@@ -48,10 +48,16 @@ grh_ChIP_classes_fn <- snakemake@input[["grh_ChIP_classes_fn"]]
 zld_blot_image <- snakemake@input[["zld_blot_image"]]
 grh_blot_image <- snakemake@input[["grh_blot_image"]]
 
-# # create blank layout for plot ===============================================
-pdf(snakemake@output[[1]], useDingbats = FALSE)
+# create blank layout for plot ===============================================
+fig_width <-  18
+fig_height <- 14
+
+# open pdf
+pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
 # pdf("manuscript/figures/extended_data_fig2.pdf", useDingbats = FALSE)
-pageCreate(width = 18.3, height = 15, default.units = "cm", showGuides = FALSE)
+
+# generate plotGardener page
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics
@@ -141,6 +147,12 @@ plotText(
   x = ref_x + 7.3, y = ref_y + 0.25, just = c("center"), default.units = "cm"
 )
 
+# add arrowheads to blot
+plotPolygon(x = c(1, 1.25, 1), y = c(1.4,1.5,1.6), default.units = "cm", fill = "black")
+
+plotPolygon(x = c(1, 1.25, 1), y = c(3.6,3.7,3.8), default.units = "cm", fill = "grey")
+
+
[…]

### 2023-03-15 workflow/scripts/extended_data_fig3.R

@@ -92,10 +92,12 @@ a_plot <- zld_ATAC_results |>
   scale_color_manual(values = c("grey", "black", zld_color)) +
   theme_classic(base_size = small_text_params$fontsize) +
   theme(legend.key.size = unit(2, 'mm'),
+        legend.text = element_text(size=5),
         legend.position = "bottom",
         legend.title = element_blank(),
         legend.margin=margin(0,0,0,0),
         legend.box.margin=margin(0,0,0,0)) +
+  guides(colour = guide_legend(override.aes = list(size=0.6))) +
   xlab(bquote(log[2]("fold change"))) +
   ylab(bquote(-log[10]("adj. p-value")))
 
@@ -103,7 +105,7 @@ a_plot <- zld_ATAC_results |>
 plotGG(
   plot = a_plot,
   x = (ref_x), y = (ref_y),
-  width = 4.5, height = 4, just = c("left", "top"),
+  width = 4.5, height = 5, just = c("left", "top"),
   default.units = "cm"
 )
 
@@ -145,10 +147,12 @@ b_plot <- zld_RNAseq_results |>
   scale_color_manual(values = c("grey", "black", zld_color)) +
   theme_classic(base_size = small_text_params$fontsize) +
   theme(legend.key.size = unit(2, 'mm'),
+        legend.text = element_text(size=5),
         legend.position = "bottom",
         legend.title = element_blank(),
         legend.margin=margin(0,0,0,0),
         legend.box.margin=margin(0,0,0,0)) +
+  guides(colour = guide_legend(override.aes = list(size=0.6))) +
   xlab(bquote(log[2]("fold change"))) +
   ylab(bquote(-log[10]("adj. p-value")))
 
@@ -156,7 +160,7 @@ b_plot <- zld_RNAseq_results |>
 plotGG(
   plot = b_plot,
   x = (ref_x), y = (ref_y),
-  width = 4.5, height = 4, just = c("left", "top"),
+  width = 4.5, height = 5, just = c("left", "top"),
   default.units = "cm"
 )
 
@@ -204,14 +208,14 @@ c_plot <- barplot(ego, showCategory=10) +
 plotGG(
   plot = c_plot,
   x = (ref_x), y = (ref_y),
-  width = 6, height = 4, just = c("left", "top"),
+  width = 7, height = 5, just = c("left", "top"),
   default.units = "cm"
 )
 
 # panel D ======================================================================
[…]

### 2023-03-15 workflow/scripts/extended_data_fig4.R

@@ -24,9 +24,8 @@ twi_blot_image <- snakemake@input[["twi_blot_image"]]
 
 
 # create blank layout for plot ===============================================
-# create blank layout for plot ===============================================
 fig_width <-  18
-fig_height <- 12
+fig_height <- 9.5
 
 # open pdf
 pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)

### 2023-03-15 workflow/scripts/extended_data_fig5.R

@@ -18,16 +18,16 @@ source("workflow/scripts/utils.R")
 #   H3K4me3 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE85191_aH3K4me3.bw",
 #   H4K16ac = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE37865_aH4K16ac.bw",
 #   H2AV = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_H2Av_IP.bw",
-#   
+# 
 #   Rpb3 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_Rpb3_IP.bw",
 #   `PolII-pS2` = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE101557_S2_PolII_phosphoSer2_IP.bw",
 #   H3K36me3 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_H3K36me3_IP.bw",
 #   SSRP1 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_SSRP1_IP.bw",
 #   Spt16 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_Spt16_IP.bw",
-#   
+# 
 #   H3 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_H3_IP.bw",
 #   H1 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE127227_aH1.bw",
-#   
+# 
 #   Pho = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE84502_aPho_IP.bw",
 #   Ez = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE101557_S2_Ez_IP.bw",
 #   Pc = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE24521_Pc_IP.bw",
@@ -35,20 +35,20 @@ source("workflow/scripts/utils.R")
 #   Ph = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE24521_Ph_IP.bw",
 #   dRing = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE101557_S2_dRing_IP.bw",
[…]

### 2023-03-15 workflow/scripts/extended_data_fig7.R

@@ -60,10 +60,16 @@ H3K27me3_ChIP_bw <- snakemake@input[["H3K27me3_ChIP_bw"]]
 H3K9me3_ChIP_bw <- snakemake@input[["H3K9me3_ChIP_bw"]]
 
 
-# # create blank layout for plot ===============================================
-pdf(snakemake@output[[1]], useDingbats = FALSE)
+# create blank layout for plot ===============================================
+fig_width <-  18
+fig_height <- 18
+
+# open pdf
+pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
 # pdf("manuscript/figures/extended_data_fig7.pdf", useDingbats = FALSE)
-pageCreate(width = 18, height = 18, default.units = "cm", showGuides = FALSE)
+
+# generate plotGardener page
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics
@@ -84,12 +90,12 @@ twi_heatmap_colors <- brewer.pal(9, "GnBu")
 H3K27me3_color <- "gray40"      
   
   # reference points for positioning figure components
-  x_offset_class_label <- 0.25
-  x_offset_browser_label <- 1
-  x_offset_browser <- 1.5
-  
-  # set genome browser height
-  gb_height <- 0.3
+x_offset_class_label <- 0.25
+x_offset_browser_label <- 1
+x_offset_browser <- 1.5
+
+# set genome browser height
+gb_height <- 0.3
   
 # panel A ======================================================================
 # reference points for positioning figure components ---------------------------
@@ -146,18 +152,18 @@ plotGG(
 
 # add labels for different groups
 plotText(
-  label = "Brain Zld ChIP", params = small_text_params, fontface = "bold",
-  x = (ref_x + 2), y = (ref_y + 0.25), just = "top", default.units = "cm"
+  label = "Zld NSC", params = small_text_params, fontface = "bold",
+  x = (ref_x + 2), y = (ref_y), just = "top", default.units = "cm"
 )
 
 plotText(
-  label = paste0("S2 Zld", "\n", "ChIP"), params = small_text_params, fontface = "bold",
[…]

### 2023-03-15 workflow/scripts/fig2.R

@@ -9,18 +9,23 @@ library(RColorBrewer)
 
 
 source("workflow/scripts/plot_heatmap.R")
-
-# function to get range for browser tracks -------------------------------------
-signal_range <- function(x, extra = 0.05) {
-  min <- floor(min(x))
-  max <- ceiling(max(x))
-  range <- max - min
-  margin <- range * extra
-  output <- c(min - margin, max + margin)
-  return(output)
-}
+source("workflow/scripts/utils.R")
 
 # define input files ===========================================================
+# define input files explictly for interactive testing
+# Twi_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Twi_aTwi_IP.bw"
+# Twi_WT_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/Twi_ATAC_S2-WT_40uM_small.bw"
+# Twi_Twi_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/S2-Twi_40uM_small.bw"
+# Zld_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_IP.bw"
+# Grh_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_IP.bw"
+# twi_ChIP_classes_fn <- "results/ChIP_peak_classes/twi_ChIP_classes.tsv"
+# zld_ChIP_classes_fn <- "results/ChIP_peak_classes/zld_ChIP_classes.tsv"
+# grh_ChIP_classes_fn <- "results/ChIP_peak_classes/grh_ChIP_classes.tsv"
+# twi_RNAseq_results_fn <- "RNAseq/results/DEseq2/S2-Twi_RNAseq_S2-Twi-vs-S2-WT_results.tsv"
+# zld_RNAseq_results_fn <- "RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results.tsv"
+# grh_RNAseq_results_fn <- "RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results.tsv"
+
+# get input files from snakemake
 Twi_ChIP_bw <- snakemake@input[["Twi_ChIP_bw"]]
[…]

### 2023-03-15 workflow/scripts/fig3.R

@@ -71,14 +71,14 @@ twi_WT_atac_fn <- snakemake@input[["twi_WT_atac_fn"]]
 
 ## create blank layout for plot =================================================
 # define figure dimensions in cm
-fig_width <-  18
-fig_height <- 12
+fig_width <-  12
+fig_height <- 7.5
 
 # open pdf
 pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
 
 # generate plotGardener page
-pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = TRUE)
+pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
 
 # general figure settings ======================================================
 # text parameters for Nature Genetics

### 2023-03-15 workflow/scripts/fig4.R

@@ -204,7 +204,7 @@ plotText(
 )
 
 plotText(
-  label = paste0("Zld", "\n", "brain"), params = small_text_params, fontface = "bold",
+  label = paste0("Zld", "\n", "NSC"), params = small_text_params, fontface = "bold",
   x = (ref_x + 2.71), y = (ref_y), just = c("center"), default.units = "cm"
 )
 

### 2023-03-15 workflow/scripts/fig5.R

@@ -216,13 +216,24 @@ zld_euler_plot <- peak_overlaps |>
   dplyr::select(8,9) |> 
   as.matrix() |> 
   euler() |> 
-  plot(quantities = list(fontsize = small_text_params$fontsize), labels = list(fontsize = small_text_params$fontsize))
+  plot(quantities = list(fontsize = small_text_params$fontsize), labels = NULL)
 
 plotGG(
   plot = zld_euler_plot,
   x = (ref_x + 0.5), y = (ref_y + 0.25),
   width = 2, height = 2, just = c("left", "top"),
   default.units = "cm"
+)
+
+# add labels for different groups
+plotText(
+  label = paste0("Zld", "\n", "1000"), params = small_text_params, fontface = "bold",
+  x = (ref_x + 0.2), y = (ref_y + 1), just = "top", default.units = "cm"
+)
+
+plotText(
+  label = paste0("Zld", "\n", "1500"), params = small_text_params, fontface = "bold",
+  x = (ref_x + 2.8), y = (ref_y + 1), just = "top", default.units = "cm"
 )
 
 # panel E ======================================================================
@@ -377,13 +388,23 @@ grh_euler_plot <- peak_overlaps |>
   dplyr::select(8,9) |> 
   as.matrix() |> 
   euler() |> 
-  plot(quantities = list(fontsize = small_text_params$fontsize), labels = list(fontsize = small_text_params$fontsize))
+  plot(quantities = list(fontsize = small_text_params$fontsize), labels = NULL)
 
 plotGG(
   plot = grh_euler_plot,
   x = (ref_x + 0.5), y = (ref_y + 0.25),
   width = 2, height = 2, just = c("left", "top"),
   default.units = "cm"
+)
+
+plotText(
+  label = paste0("Grh", "\n", "100"), params = small_text_params, fontface = "bold",
+  x = (ref_x + 0.2), y = (ref_y + 1), just = "top", default.units = "cm"
+)
+
+plotText(
+  label = paste0("Grh", "\n", "400"), params = small_text_params, fontface = "bold",
+  x = (ref_x + 2.8), y = (ref_y + 1), just = "top", default.units = "cm"
 )
 
 # panel I ======================================================================
@@ -535,7 +556,7 @@ twi_euler_plot <- peak_overlaps |>
   dplyr::select(8,9) |> 
   as.matrix() |> 
   euler() |> 
[…]

### 2023-03-20 manuscript/manuscript_sections/_01_Abstract.qmd
update MS

@@
 Despite binding to target motifs in closed chromatin, many pioneer factors display cell-type specific binding and activity.
-The mechanisms that govern pioneer-factor occupancy and the relationship between chromatin occupancy and opening remain unclear.
-We studied three Drosophila transcription factors with distinct DNA-binding domains and biological functions: Zelda, Grainy head, and Twist.
-We demonstrated that the level of chromatin occupancy is a key determinant of pioneer-factor activity.
-Occupancy is regulated by multiple factors, including motif content at binding sites, local chromatin context, and protein concentration.
-Regions outside the DNA-binding domain are required for pioneer-factor binding and chromatin opening.
-Together, our results show that pioneering activity is not a binary feature intrinsic to a protein but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type-specific features.
+The mechanisms governing pioneer-factor occupancy and the relationship between chromatin occupancy and opening remain unclear.
+We studied three *Drosophila* transcription factors with distinct DNA-binding domains and biological functions: Zelda, Grainy head, and Twist.
+We demonstrated that the level of chromatin occupancy is a key determinant of activity.
+Multiple factors regulate occupancy, including motif content, local chromatin, and protein concentration.
+Regions outside the DNA-binding domain are required for binding and chromatin opening.
+Our results show that pioneering activity is not a binary feature intrinsic to a protein but occurs on a spectrum and is regulated by a variety of protein-intrinsic and cell-type-specific features.

### 2023-03-20 manuscript/manuscript_sections/_02_Introduction.qmd
update MS

@@
 In contrast to the tissue-specific chromatin occupancy of Zld, Grh-binding sites are largely shared between tissues [@Nevil2017].
-Grh is a master regulator of epithelial cell fate that is conserved among metazoans and functions as a pioeer factor in both *Drosophila* larva and mammalian cell culture [@Wang2012; @Jacobs2018; @Nevil2020].
+Grh is a master regulator of epithelial cell fate that is conserved among metazoans and functions as a pioneer factor in both *Drosophila* larva and mammalian cell culture [@Wang2012; @Jacobs2018; @Nevil2020].
 However, in the early *Drosophila* embryo Grh is dispensible for chromatin accessibility, and only displays pioneering activity in late embryos or larval tissues.
@@
 
-The essential roles of Zld and Grh in diverse biological processes and the well-documented context-dependent binding and activity of these factors makes them excellent models to explore how PF binding is regulated and how PF occupancy relates to chromatin opening.
+The essential roles of Zld and Grh in diverse biological processes and the context-dependent binding and activity of these factors makes them excellent models to explore how PF binding is regulated and how PF occupancy relates to chromatin opening.
 Because many previous studies have relied on loss-of-function experiments, it has been difficult to disentangle causal relationships between chromatin structure and PF activity.

### 2023-03-20 manuscript/manuscript_sections/_03_Results.qmd
update MS

@@
 To investigate the mechanisms by which pioneer factors bind and open chromatin, we generated S2 cell lines to enable the individual expression of Zld or Grh (Extended Data Fig. 1a).
-These cell lines allowed the inducible and tunable expression of each factor, as neither pioneer factor is normally expressed in WT S2 cells (Extended Data Fig. 1b,c,d).
+These cell lines allowed the inducible and tunable expression of each factor, as neither pioneer factor is normally expressed in wild-type S2 cells (Extended Data Fig. 1b,c,d).
 We expressed each factor at approximately physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
@@
 By using these datasets as controls for peak calling, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh (Extended Data Fig.
-1e-g).
+1e-f).
 
@@
 Having established that Zld and Grh bind thousands of sites in S2 cells, we performed assay for transposase-accessible chromatin (ATAC)-seq on wild-type cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
-Using differential accessibility analysis comparing Zld or Grh-expressing cells to wild-type cells, we detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
-3a,d).
-By integrating our ChIP-seq and ATAC-seq data, we identified three classes of PF-bound sites (Fig. 1a,b).
+Differential accessibility analysis using DESeq2 to compare Zld or Grh-expressing cells to wild-type cells detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
+2a,d)[@love2014].
+By integrating our ChIP-seq and ATAC-seq data, we identified three classes of PF-bound sites (Fig. 1a,b; Supplementary Table 1).
 Class I sites are accessible in wild-type cells and remain accessible when bound by the ectopically expressed factor.
@@
 For each factor, only a subset of binding sites were within closed chromatin, and at only a subset of these did the chromatin become accessible (Fig. 1c,f).
-While class I binding sites occured predominantly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
+While class I binding sites occurred predominantly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
 Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in wild-type cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
@@
 
+Next, we sought to determine the binding dynamics of Zld and Grh.
+Zld and Grh protein was detectable at 4 or 12 hours following induction, respectively (Extended Data Fig. 3a-b).
+We used CUT&RUN to measure Zld and Grh binding to chromatin at 0, 4, 12, 24 or 48 hours following induction[@Skene2017].
+Both proteins were bound to class I regions even in uninduced cells (0H), suggesting that these sites can be occupied even when protein levels are below the limit of detection of immunoblotting (Extended Data Fig 3c-d).
+Class II and III regions were bound at 4 hours, with slight increases in occupancy at later time points, demonstrating that Zld and Grh bind closed chromatin rapidly following induction of protein expression.
+
 ::: {#fig-1}
@@
 
-**Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
-ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
+**Ectopically expressed pioneer factors open chromatin and activate transcription.
+a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
+ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for wild-type (WT) cells lacking Zld or Grh expression.
 **c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
@@
 i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
-**j,i,** Proportion of class I, II or III regions that are proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
+**j,l,** Proportion of class I, II or III regions that is proximal to a gene that is differentially expressed (DE) upon Zld **(j)** or Grh **(l)** expression.
 Promoter-proximal and promoter-distal binding sites are shown separately.
@@
 :::
-
-Next, we sought to determine the binding dynamics of Zld and Grh.
-Zld and Grh protein was detectable at 4 or 12 hours following induction, respectively (Extended Data Fig. 2a-b).
-We used CUT&RUN to measure Zld and Grh binding to chromatin at 0, 4, 12, 24 or 48 hours following induction[@Skene2017].
-Both proteins were bound to class I regions even in uninduced cells (0H), suggesting that these sites can be occupied even when protein levels are below the limit of detection of immunoblotting (Extended Data Fig 2c-d).
-Class II and III regions were bound at 4 hours, with slight increases in occupancy at later time points, demonstrating that Zld and Grh bind closed chromatin rapidly following induction of protein expression.
 
@@
 To identify whether Zld- and Grh-mediated chromatin accessibility led to transcriptional activation, we performed RNA-seq.
-Differential expression analysis using DESeq2 to compare induced cells to wild-type cells identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig.
-3b,e)[@love2014].
-The genes activated by Grh were enriched for gene ontology (GO) terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig. 3c)[@Nevil2017].
+Differential expression analysis comparing induced cells to wild-type cells identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. 3b,e).
 Genes activated by Zld expression were enriched for embryonic morphogenesis, consistent with the known function of Zld[@Nien2011], as well as imaginal disc development, in which Zld-target genes have not been well characterized (Extended Data Fig.
-3f).
+3c).
+The genes activated by Grh were enriched for gene ontology (GO) terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig. 3f)[@Nevil2017].
 
@@
 As we had done for Zld and Grh, we determined Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
-Using these data, we identified the same three classes of binding sites that we described for Zld and Grh (Fig 2a,b).
+Using these data, we identified the same three classes of binding sites that we described for Zld and Grh (Fig 2a,b; Extended Data Fig. 1g).
 Class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
-Twi bound extensively to closed chromatin, with `r perc_twi_closed`% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
-At a small subset of Twi-binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c; Extended Data Fig. 4b).
+Twi bound extensively to closed chromatin, with `r perc_twi_closed`% of binding sites lacking detectable chromatin accessibility (Fig. 2b).
+At a small subset of Twi-binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-d; Extended Data Fig. 4b).
 RNA-seq revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4c).
@@
 This contrasted with class II and III peaks, which were largely specific to each individual factor.
-Control IP experiments in wild-type S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig. 1e-f).
+Control IP experiments in wild-type S2 cells support that these peaks are not due to nonspecific immunoprecipitation (Extended Data Fig. 1e-g).
 We hypothesized that the overlapping class I binding sites may occur at active *cis*-regulatory regions in wild-type cells.
-To test this, we analyzed published ChIP-seq data from S2 cells for markers of active transcription: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV[@Henriques2018; @Philip2015; @Tettey2019].
+To test this, we analyzed published ChIP-seq data from S2 cells for markers of active chromatin: H3K27ac, CBP, H3K4me1, H3K4me3, and H2AV[@Henriques2018; @Philip2015; @Tettey2019].
 Overlapping class I regions showed high levels of these chromatin marks (Fig. 3b), suggesting that exogenously expressed factors may bind nonspecifically to active regions of the genome.
@@
 Zld, Grh, and Twi ChIP-seq data were generated in this study.
-Remaining ChIP-seq datasets were previously published (see Supplementary table XX).
+Remaining ChIP-seq datasets were previously published (see Supplementary table 2)[@Henriques2018; @Philip2015; @Tettey2019].
 **c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
 Logos plot of the canonical motif are shown alongside the heatmap.
-Percentages are shown for class I, II or III binding sites, or all WT ATAC-seq peaks as a control for the background motif frequency.
+Percentages are shown for class I, II or III binding sites, or all wild-type ATAC-seq peaks as a control for the background motif frequency within regulatory elements.
 **f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
@@
 When compared to all accessible regions, class I regions had low levels of enrichment for the canonical motifs, supporting our prior analysis suggesting that binding to these regions is nonspecific.
-By contrast, class II and III sites were highly enriched for the canonical motifs (Fig. 3c).
+By contrast, class II and III sites were highly enriched for the canonical motifs (Fig. 3c-e).
 Class III sites had the highest percentage of sites containing a canonical motif.
@@
 For all three factors, class III sites tended to have stronger motifs than class II sites (Fig. 3g,i,k).
-These data suggest that a more and stronger motifs at class III sites as compared to class II sites facilitates the ability of Zld, Grh, and Twi to open chromatin.
+These data suggest that more and stronger motifs at class III as compared to class II sites facilitates the ability of Zld, Grh, and Twi to open chromatin.
 
@@
 6d-f).
-For class II and III peaks, some motifs were positioned close to the nuclesome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
-Additionally, the MNase signal in many regions did not show strongly positioned nucleosomes, likely because nucleosomes at silent regions often are not strongly positioned and show variable distributions in different cell types [@Stergachis2020; @Abdulhay2020].
+For class II and III peaks, some motifs were positioned close to the nucleosome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
+Additionally, the MNase signal in many regions did not show strongly positioned nucleosomes, likely because nucleosomes at silent regions often are not strongly positioned [@Stergachis2020; @Abdulhay2020].
 While some PFs may recognize their motif when specifically positioned on a nucleosome, Zld, Grh and Twi do not have a strong preference for motif positioning when exogenously expressed.
@@
 To better understand how cell-type specific variables affect PF binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
-We analyzed Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a], Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs (Nevil et al.
-2017), and Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
-We identified many binding sites that were bound *in vivo*, but were not bound upon ectopic expression in S2 cells (Extended Data Fig. 7a,b).
+We analyzed Zld ChIP-seq data from the early embryo (nuclear cycle 14 embryos)[@Harrison2011] and from larval neural stem cells [@Larson2021a], Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017] and Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
+We identified many binding sites that were bound *in vivo*, but were not bound upon ectopic expression in S2 cells (Extended Data Fig. 7a-c).
 Chromatin enriched for repressive histone modifications H3K9me3 and H3K27me3 is resistant to binding of some PFs [@Mayran2018a; @Soufi2012].
@@
 **Many endogenous binding sites are resistant to ectopic pioneer-factor binding. a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
-ChIP-seq data for repressive histone modifications H3K27me3 and H3K9me3 is also shown.
+ChIP-seq data for repressive histone modifications H3K27me3 [@Jain2020] and H3K9me3[@Eastwood2021] in wild-type S2 cells are also shown.
 **d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
@@
 
-To test the functional relevance of repressive chromatin to PF binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3.
-Despite causing dramatic reductions in H3K27me3 levels, tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig.
-4d-f; Extended Data Fig.
-8a).
+To test the functional relevance of repressive chromatin to PF binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3 (Extended Data Fig 8).
+Despite causing dramatic reductions in H3K27me3 levels, tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig. 4d-f).
 The majority of class IV binding sites remained unbound in tazemetostat-treated cells, demonstrating that loss of PRC2 activity is insufficient to promote widespread binding of any of the three factors studied.
@@
 We performed ChIP-seq on cells induced to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig.
-9; see methods).
-Analysis of our previously defined class I-VI binding sites showed that binding and was concentration dependent (Fig. 5).
+9a-c; see methods).
+Analysis of our previously defined class I-VI sites showed that binding and was concentration dependent (Fig. 5).
 At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i).
 Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i).
-To identify the relationship between PF binding and chromatin accessibility, we performed ATAC-seq on cells induced with the same range in expression.
+To identify the relationship between PF binding and chromatin accessibility, we performed ATAC-seq on cells with the same range in expression.
 As we identified for PF binding, changes in chromatin accessibility were concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig.
@@
 Many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels, suggesting that increased chromatin occupancy, as assayed by ChIP-seq, could result in chromatin accessibility.
-For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites, which were not bound at lower concentrations (Fig.
-5, Extended Data Fig.
-9d-f).
+For Grh and Zld, but not Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites, which were not bound at lower concentrations (Extended Data Fig. 9d-f).
+
+::: {#fig-5}
+![](figures/fig5.pdf)
+
+**Pioneer-factor binding and opening of closed chromatin is concentration dependent.
+a,e,i,** Boxplots of ChIP-seq signal at class I-III regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
+**b,f,j,** Boxplots of ATAC-seq signal at class I-III regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
+**c,g,k,** Distribution of class II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
+Y-axis indicates the total number of peaks called at each concentration.
+Colors indicate the proportion of those peaks that are defined as class II, or III.
+**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
+:::
 
@@
 For Zld and Twi, supraphysiological concentrations led to an increase in class III regions relative to class II regions (Fig. 5c,k).
-For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with most of these sites falling in III (Fig. 5g).
+For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with most of these sites falling in class III (Fig. 5g).
 To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs.
@@
 Together, these data demonstrate that both binding and chromatin opening by PFs are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are resistant to binding at physiological levels.
-
-::: {#fig-5}
-![](figures/fig5.pdf)
-
-**Binding and opening of closed chromatin is concentration dependent.
-a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
-Concetrations of CuSO~4~ used to induce protein expression are indicated above each plot.
-**b,f,j,** Metaplots of ATAC-seq signal at class I-VI regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
-**c,g,k,** Distribution of class I, II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
-Y-axis indicates the total number of peaks called at each concentration.
-Colors indicate the proportion of those peaks that are defined as class I, II, or III.
-**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
-:::
 
@@
 a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
-Shown for Zld **(a)** and Grh **(b)**.
-**c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**.
+Adjacent class I regions are shown where the DBD alone can bind.
+Tracks shown for Zld **(a)** and Grh **(b)**.
+Class III regions are highlighted with gray shading.
+**c,e,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(e)**.
 **d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
@@
 A high level of chromatin occupancy is required for PFs to initiate chromatin opening.
-Chromatin occupancy can be regulated by protein-intrinsic features such as motif affinity and protein domains, and by cell-intrinsic properties such as histone modifications, PF concentration, and cell-type specific cofactors.
+Chromatin occupancy can be regulated by protein-intrinsic features such as motif affinity and protein domains, and by cell-intrinsic properties such as histone modifications, PF concentration, and cell-type specific cofactors (outlined in blue).
 :::

### 2023-03-20 manuscript/manuscript_sections/_04_Discussion.qmd
update MS

@@
 However, these factors are largely excluded from repressed chromatin enriched for H3K9me3 or H3K27me3.
-This binding to naïve chromatin is similar to what has previously been reported for other pioneer factors.
+This binding to naïve chromatin is similar to what has previously been reported for other pioneer factors [@Soufi2012; @Donaghey2018].
 Regions with high levels of H3K9me3 are resistant to binding by the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM).
@@
 Tissue-specific PF occupancy is also likely regulated by the collection of cofactors expressed in a given tissue.
-This may partially explain the presence of class VI sites that lack repressive histone modifications in culture and are bound *in vivo* but not when expressed in S2 cells.
+This may partially explain the presence of class VI sites that lack repressive histone modifications in S2 cells and are bound *in vivo* but not when expressed exogenously.
 For example, in mammalian cell culture FOXA2 occupancy at a subset of tissue-specific binding sites was increased when GATA4 was present [@Donaghey2018].
@@
 Similarly, regions outside the DBDs of human PFs FOXA1 and SOX2 are required for robust binding of closed chromatin [@lerner2022].
-Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and have demonstrated the role that intrinsically disordered regions (IDRs) play in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
+Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and has demonstrated the role that intrinsically disordered regions (IDRs) play in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
 For Zld, this disordered domain is required for transcriptional activation and promotes nucleosome binding *in vitro* [@Hamm2015; @McDaniel2019a].
@@
 Indeed, misexpression of the PFs Grh and DUX4 lead to diseases such as epithelial cancer or facioscapulohumeral muscular dystrophy (FSHD), respectively [@Campbell2018; @Reese2019].
-Together our data identify protein intrinsic and extrinsic features that govern PF binding and activity, providing insights into how these factors define novel cis-regulatory elements and drive gene-regulatory networks to modulate cell fate.
+Together our data identify protein intrinsic and extrinsic features that govern PF binding and activity, providing insights into how these factors define novel *cis*-regulatory elements and drive gene-regulatory networks to modulate cell fate.
 A deeper understanding of how PF activity is regulated will have important implications for determining how diseases are caused by their misexpression and our ability to use these powerful factors for cellular reprogramming.

### 2023-03-20 manuscript/manuscript_sections/_05_Methods.qmd
update MS

@@
 
-Immunostaining was performed as described previously[@angshuman2007].
+Immunostaining was performed as described previously [@angshuman2007].
 Briefly, 1×10^6^ cells in 200 µL PBS were applied to a coverslip pre-coated with 10% poly-L-lysine.
@@
 After washing with PBS, cells were incubated with a 1:500 dilution of goat anti-rabbit IgG DyLight 488 conjugated secondary antibody (Thermo Fisher Scientific #35552) for 1 hour at RT.
-Imaging was performed at 60X on a Nikon Ti2 epi-fluorescent microscope.
+Imaging was performed on a Nikon Ti2 epi-fluorescent microscope with a 60x objective lens.
 
@@
 Peak calling was performed using MACS2 version 2.2 with parameters `-g dm --call-summits`.
-Data from ChIP experiments performed with the same antibody in WT cells were used as a control for peak calling.
+Data from ChIP experiments performed with the same antibody in wild-type cells were used as a control for peak calling.
 Only peaks that were detected in all replicates were considered in downstream analysis.
@@
 
-Previously published datasets analyzed in this study can be found in supplemental table 1.
+Previously published datasets analyzed in this study can be found in supplemental table 2.
 All previously published data were analyzed in parallel with data from this study using analysis parameters described above.

### 2023-03-20 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update MS

@@
 i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
-**j,i,** Proportion of class I, II or III regions that is proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
+**j,l,** Proportion of class I, II or III regions that is proximal to a gene that is differentially expressed (DE) upon Zld **(j)** or Grh **(l)** expression.
 Promoter-proximal and promoter-distal binding sites are shown separately.
@@
 Zld, Grh, and Twi ChIP-seq data were generated in this study.
-Remaining ChIP-seq datasets were previously published (see Supplementary table 1)[@Henriques2018;@Philip2015;@Tettey2019].
+Remaining ChIP-seq datasets were previously published (see Supplementary table 1)[@Henriques2018; @Philip2015; @Tettey2019].
 **c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
 Logos plot of the canonical motif are shown alongside the heatmap.
-Percentages are shown for class I, II or III binding sites, or all wild-type ATAC-seq peaks as a control for the background motif frequency.
+Percentages are shown for class I, II or III binding sites, or all wild-type ATAC-seq peaks as a control for the background motif frequency within regulatory elements.
 **f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
@@
 a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
-ChIP-seq data for repressive histone modifications H3K27me3 [@Jain2020] and H3K9me3[@Eastwood2021] in WT S2 cells is also shown.
+ChIP-seq data for repressive histone modifications H3K27me3 [@Jain2020] and H3K9me3[@Eastwood2021] in wild-type S2 cells are also shown.
 **d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
@@
 **Figure 5: Pioneer-factor binding and opening of closed chromatin is concentration dependent.
-a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
-Concetrations of CuSO~4~ used to induce protein expression are indicated above each plot.
-**b,f,j,** Metaplots of ATAC-seq signal at class I-VI regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
-**c,g,k,** Distribution of class I, II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
+a,e,i,** Boxplots of ChIP-seq signal at class I-III regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
+**b,f,j,** Boxplots of ATAC-seq signal at class I-III regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
+**c,g,k,** Distribution of class II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
 Y-axis indicates the total number of peaks called at each concentration.
-Colors indicate the proportion of those peaks that are defined as class I, II, or III.
+Colors indicate the proportion of those peaks that are defined as class II, or III.
 **d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
@@
 Tracks shown for Zld **(a)** and Grh **(b)**.
-**c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**.
+Class III regions are highlighted with gray shading.
+**c,e,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(e)**.
 **d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
@@
 **b,** mRNA levels of Zld and Grh in S2 cells.
-Top histogram shows the distribution of mRNA levels for all Drosophila genes.
+Top histogram shows the distribution of mRNA levels for all *Drosophila* genes.
 Vertical dashed line indicates a log~2~ RPKM value of 0 as a threshold for considering a gene to be expressed.
@@
 Two independently generated cell lines are shown and compared to 2-3H embryos.
+60,000 cells were loaded in each well, which is equivalent to the approximately 60,000 nuclei present in 10 2-3 hour embryos.
 Black arrowheads indicate Zld **(c)** or Grh **(d)**.
 Gray arrowheads indicate background bands used to assess loading.
-**e-f,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in WT cells.
+**e-g,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in wild-type (WT) cells.
 
-**Extended Data Figure 2: Zld and Grh bind to chromatin rapidly after induction of protein expression.
+**Extended Data Figure 2: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.
+a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
+**b,e,** RNA-seq volcano plots showing gene expression changes in cells expressing Zld **(b)** or Grh **(e)** when compared to wild-type cells treated with the same concentration of CuSO~4~.
+**c,f,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(c)** or Grh **(f)**.
+
+**Extended Data Figure 3: Zld and Grh bind to chromatin rapidly after induction of protein expression.
 a-b,** Immunoblots showing time course of Zld **(a)** or Grh **(b)** protein expression following induction of stable cell lines.
@@
 
-**Extended Data Figure 3: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.
-a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
-**b,e,** RNA-seq volcano plots showing gene expression changes in cells expressing Zld **(b)** or Grh **(e)** when compared to WT cells treated with the same concentration of CuSO~4~.
-**c,f,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(c)** or Grh **(f)**.
-
 **Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
-Protein levels in stable cell lines are compared to 3-4 hour old embryos.
-**b,** Volcano plots showing changes in ATAC-seq signal in cells expressing Twi when compared to WT cells treated with the same concentration of CuSO~4~.
-**c,** RNA-seq volcano plots showing gene expression changes in cells expressing Twi when compared to WT cells treated with the same concentration of CuSO~4~.
+Protein levels in stable cell lines are compared to 3-4 hour (H) old embryos.
+**b,** Volcano plots showing changes in ATAC-seq signal in cells expressing Twi when compared to wild-type cells treated with the same concentration of CuSO~4~.
+**c,** RNA-seq volcano plots showing gene expression changes in cells expressing Twi when compared to wild-type cells treated with the same concentration of CuSO~4~.
 **d,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Twi.
@@
 **Extended Data Figure 6: Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes.
-a-c,** Metaplots showing average MNase signal from WT cells centered on motifs within class I, II, or III regions for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
+a-c,** Metaplots showing average MNase signal from wild-type cells centered on motifs within class I, II, or III regions for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
 **d-f,** Heatmaps showing MNase signal centered on motifs within class II and III regions for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
@@
 
-**Extended Data Figure 8: Specific detection of H3K27me3 in tazemtostat-treated cells. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
+**Extended Data Figure 8: Loss of H3K27me3 in tazemetostat-treated cells. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
 Tubulin levels are shown as a loading control.
@@
 **c-d,** Immunofluorescent microscopy images of stable cell lines expressing full-length protein or DBD only for Zld **(c)** or (Grh).
-Stable cell lines are compared to WT cells.
+Stable cell lines are compared to wild-type (WT) cells.
 
-**Supplementary table 1:** List of previously published genomics datasets that were analyzed in this work.
+**Supplementary table 1: Classification of Zld, Grh and Twi ChIP-seq peaks.** Table contains information about ChIP-seq peaks for Zld, Grh and Twi, including the peak location, information about peak calling by MACS2, classification of peaks as class I, II or III, information about differential accessibility as determined by DESeq2, assignment of peaks to the nearest gene, information about differential expression of nearby genes, and motif content with each peak,
+
+**Supplementary table 2: Previously published ChIP-seq datasets that were analyzed in this work.** For previously published ChIP-seq datasets, this table provides the target of immunoprecipitation, the cell type and/or developmental stage, the Gene Expression Omnibus accession number (GSE), and the PubMed ID for the study in which the dataset was originally published..

### 2023-03-20 manuscript/supplemental_figures.qmd
update MS

@@
   title-delim: "**.**"
-
 ---
@@
 Two independently generated cell lines are shown and compared to 2-3H embryos.
+60,000 cells were loaded in each well, which is equivalent to the approximately 60,000 nuclei present in 10 2-3 hour embryos.
+Black arrowheads indicate Zld **(c)** or Grh **(d)**.
+Gray arrowheads indicate background bands used to assess loading.
+**e-g,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in wild-type (WT) cells.
 :::
@@
 
-**Zld and Grh bind to chromatin rapidly after induction of protein expression.
-a-b,** Immunoblots showing time course of Zld **(a)** or Grh **(b)** protein expression following induction of stable cell lines.
-**c-d,** Metaplots showing average z-score normalized CUT&RUN signal at class I, II or III sites at different time points following induction.
+**Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.
+a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
+**b,e,** RNA-seq volcano plots showing gene expression changes in cells expressing Zld **(b)** or Grh **(e)** when compared to wild-type cells treated with the same concentration of CuSO~4~.
+**c,f,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(c)** or Grh **(f)**.
 :::
@@
 
-**Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.
-a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
-**b,e,** RNA-seq volcano plots showing gene expression changes in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
-**c,f,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(c)** or Grh **(f)**.
+**Zld and Grh bind to chromatin rapidly after induction of protein expression.
+a-b,** Immunoblots showing time course of Zld **(a)** or Grh **(b)** protein expression following induction of stable cell lines.
+Black arrowheads indicate Zld **(a)** or Grh **(b)**.
+Gray arrowheads indicate background bands used to assess loading.
+**c-d,** Metaplots showing average z-score normalized CUT&RUN signal at class I, II or III sites at different time points following induction.
 :::
@@
 **Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
-Protein levels in stable cell lines are compared to 3-4 hour old embryos.
-**b,** Volcano plots showing changes in ATAC-seq signal in cells expressing Twi when compared to WT cells treated with the same concentration of CuSO~4~.
-**c,** RNA-seq volcano plots showing gene expression changes in cells expressing Twi when compared to WT cells treated with the same concentration of CuSO~4~.
+Protein levels in stable cell lines are compared to 3-4 hour (H) old embryos.
+**b,** Volcano plots showing changes in ATAC-seq signal in cells expressing Twi when compared to wild-type cells treated with the same concentration of CuSO~4~.
+**c,** RNA-seq volcano plots showing gene expression changes in cells expressing Twi when compared to wild-type cells treated with the same concentration of CuSO~4~.
 **d,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Twi.
@@
 
-**Chromatin features associated with Zld, Grh and Twi binding sites. a,c,e,** Heatmaps showing the levels of different chromatin marks in class I, II or III regions for Zld (a), Grh (c) or Twi (e).
+**Chromatin features associated with Zld, Grh and Twi binding sites.
+a,c,e,** Heatmaps showing the levels of different chromatin marks in class I, II or III regions for Zld **(a)**, Grh **(c)** or Twi **(e)**.
 The color represents the average z-score normalized read depth across a 1 KB region surrounding the center of class I, II or III ChIP-seq peaks.
-**b,d,f,** Example genome browser tracks for class III regions with high levels of H3K27me3 for Zld (b), Grh (d), or Twi (f).
+**b,d,f,** Example genome browser tracks for class III regions with high levels of H3K27me3 for Zld **(b)**, Grh **(d)**, or Twi **(f)**.
 :::
@@
 **Zld, Grh and Twi do not bind preferentially to motifs in a particular position on nucleosomes.
-a-c,** Metaplots showing average MNase signal at motifs within class I, II, or III regions for Zld (a), Grh (b), or Twi (c).
-**d-f,** Heatmaps showing MNase signal at motifs within class II and III regions for Zld (d), Grh (e), or Twi (f).
+a-c,** Metaplots showing average MNase signal from wild-type cells centered on motifs within class I, II, or III regions for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
+**d-f,** Heatmaps showing MNase signal centered on motifs within class II and III regions for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
 Rows are ordered based on hierarchical clustering to highlight the various patterns of MNase signal around motifs.
@@
 
-**Zld, Grh and Twi display cell-type specific binding. a-c,** Venn diagrams showing overlap between ChIP-seq peaks identified in different tissues for Zld (a), Grh (b) or Twi (c).
-**d-l,** Genome browser tracks showing examples of class IV, V, and VI regions for Zld (d-f), Grh (g-i) or Twi (j-l).
+**Zld, Grh and Twi display cell-type specific binding. a-c,** Venn diagrams showing overlap between ChIP-seq peaks identified in different tissues for Zld **(a)**, Grh **(b)** or Twi **(c)**.
+**d-l,** Genome browser tracks showing examples of class IV, V, and VI regions for Zld **(d-f)**, Grh **(g-i)** or Twi **(j-l)**.
 For each example, the top tracks show H3K27me3 and H3K9me3 signal over a larger region.
@@
 
-**PRC2 inhibition is insufficient to promote Zld, Grh, and Twi binding to repressed targets. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
+**Loss of H3K27me3 in tazemetostat-treated cells. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
 Tubulin levels are shown as a loading control.
+**b,** Heatmap showing specificity of anti-H327me3 antibody in CUT&RUN reactions.
+A panel of barcoded spike-in nucleosomes bearing different modifications was added to each CUT&RUN reaction (see methods).
+For each sample, the heatmap displays the percentage of barcode reads for each sample and histone modification relative to the total number of barcode reads for all modifications.
 :::
@@
 **Expression of Zld, Grh, or Twi at supraphysiological levels results in chromatin opening at a small number of novel binding sites.
-a-c,** Immunoblots showing Zld (a), Grh (b) or Twi (c) protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
-**d-f,** Bar plots showing the percentage of previously defined class I-VI binding sites that are bound by Zld (d), Grh (e), or Twi (f) when expressed at varying concentrations.
-**g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld (g), Grh (h), or Twi (i) are expressed at varying concentrations.
+a-c,** Immunoblots showing Zld **(a)**, Grh **(b)** or Twi **(c)** protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
+**d-f,** Bar plots showing the percentage of previously defined class I-VI binding sites that are bound by Zld **(d)**, Grh **(e)**, or Twi **(f)** when expressed at varying concentrations.
+**g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld **(g)**, Grh **(h)**, or Twi **(i)** are expressed at varying concentrations.
 :::
@@
 **Expression of Zld and Grh DNA-binding domains at protein levels comparable to the full-length proteins.
-a-b,** Immunoblots showing titration of protein levels for Zld (a) or Grh (b) DNA-binding domains to match expression of the full-length proteins.
+a-b,** Immunoblots showing titration of protein levels for Zld **(a)** or Grh **(b)** DNA-binding domains to match expression of the full-length proteins.
 DNA-binding domain protein levels are shown at a range of CuSO~4~ concentrations and compared to an equivalent number of cells expressing full-length Zld or Grh at approximately physiological levels.
-**c-d,** Immunofluorescent microscopy images of stable cell lines expressing full-length protein or DBD only for Zld (c) or (Grh).
-Stable cell lines are compared to WT cells.
-**e,** Metaplots showing the average ChIP-seq signal for full-length Zld or the DNA-binding domain alone at the subset of previously identified class I, II or III binding sites that contain a canonical Zelda CAGGTAG motif.
+**c-d,** Immunofluorescent microscopy images of stable cell lines expressing full-length protein or DBD only for Zld **(c)** or (Grh).
+Stable cell lines are compared to wild-type (WT) cells.
 :::

### 2023-03-20 workflow/scripts/extended_data_fig2.R

@@ -4,56 +4,40 @@ library(tidyverse)
 library(org.Dm.eg.db)
 library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
 library(RColorBrewer)
-library(EBImage)
+library(clusterProfiler)
+library(enrichplot)
+library(rtracklayer)
 
-source("workflow/scripts/plot_heatmap.R")
 
 # define input files ===========================================================
-# specify input files explicitly for testing script interactively
-# S2_Zld_CR_0H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_0H_small.bw"
-# S2_Zld_CR_4H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_4H_small.bw"
-# S2_Zld_CR_12H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_12H_small.bw"
-# S2_Zld_CR_24H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_24H_small.bw"
-# S2_Zld_CR_48H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_48H_small.bw"
+# define input files explicitly for testing
+# zld_RNAseq_results_fn <- "RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results_annotated.tsv"
+# grh_RNAseq_results_fn  <- "RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results_annotated.tsv"
 # 
-# S2_Grh_CR_0H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_0H_small.bw"
-# S2_Grh_CR_4H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_4H_small.bw"
-# S2_Grh_CR_12H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_12H_small.bw"
[…]

### 2023-03-20 workflow/scripts/extended_data_fig3.R

@@ -4,40 +4,56 @@ library(tidyverse)
 library(org.Dm.eg.db)
 library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
 library(RColorBrewer)
-library(clusterProfiler)
-library(enrichplot)
-library(rtracklayer)
+library(EBImage)
 
+source("workflow/scripts/plot_heatmap.R")
 
 # define input files ===========================================================
-# define input files explicitly for testing
-# zld_RNAseq_results_fn <- "RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results_annotated.tsv"
-# grh_RNAseq_results_fn  <- "RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results_annotated.tsv"
+# specify input files explicitly for testing script interactively
+# S2_Zld_CR_0H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_0H_small.bw"
+# S2_Zld_CR_4H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_4H_small.bw"
+# S2_Zld_CR_12H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_12H_small.bw"
+# S2_Zld_CR_24H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_24H_small.bw"
+# S2_Zld_CR_48H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_48H_small.bw"
 # 
-# zld_ATAC_results_fn <- "ATACseq/results/DEseq2_results_filtered/S2-Zld_ATACseq_S2-Zld-FL-vs-S2-WT_results.tsv"
-# grh_ATAC_results_fn <- "ATACseq/results/DEseq2_results_filtered/S2-Grh_ATACseq_S2-Grh-FL-vs-S2-WT_results.tsv"
+# S2_Grh_CR_0H_bw <- "TF_CUTandRUN/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_0H_small.bw"
[…]

### 2023-03-20 workflow/scripts/extended_data_fig6.R

@@ -113,7 +113,15 @@ plot_colors <- c(
   `III` = "#2171B5"
 )
 
-a_plot <- plot_average(bw, regions = regions, row_split = regions$class, line_width = 0.5, upstream = hm_upstream, downstream = hm_downstream) +
+a_plot <- plot_average(
+  bw, 
+  regions = regions, 
+  row_split = regions$class, 
+  line_width = 0.5, 
+  upstream = hm_upstream, 
+  downstream = hm_downstream, 
+  center_label = "motif"
+  ) +
   # scale_color_brewer(palette = "Blues") +
   scale_color_manual(values = plot_colors) +
   theme(text = element_text(size = 5),
@@ -126,7 +134,6 @@ a_plot <- plot_average(bw, regions = regions, row_split = regions$class, line_wi
         legend.position = "bottom",
         legend.margin=margin(-5,-5,-5,-5),
         legend.box.margin=margin(0,0,0,0)
-        
   )
 
 a_plot <- a_plot +
@@ -184,8 +191,15 @@ plot_colors <- c(
   `III` = "#D94701"
 )
 
-b_plot <- plot_average(bw, regions = regions, row_split = regions$class, line_width = 0.5, upstream = hm_upstream, downstream = hm_downstream) +
-  # scale_color_brewer(palette = "Blues") +
+b_plot <- plot_average(
+  bw, 
+  regions = regions, 
+  row_split = regions$class, 
+  line_width = 0.5, 
+  upstream = hm_upstream, 
+  downstream = hm_downstream, 
+  center_label = "motif"
+) +
   scale_color_manual(values = plot_colors) +
   theme(text = element_text(size = 5),
         line = element_line(size = 0.1),
@@ -249,8 +263,15 @@ plot_colors <- c(
   III = "#2B8CBE"
 )
 
-c_plot <- plot_average(bw, regions = regions, row_split = regions$class, line_width = 0.5, upstream = hm_upstream, downstream = hm_downstream) +
-  # scale_color_brewer(palette = "Blues") +
+c_plot <- plot_average(
+  bw, 
+  regions = regions, 
+  row_split = regions$class, 
+  line_width = 0.5, 
+  upstream = hm_upstream, 
+  downstream = hm_downstream, 
+  center_label = "motif"
+) +
   scale_color_manual(values = plot_colors) +
   theme(text = element_text(size = 5),
         line = element_line(size = 0.1),

### 2023-03-20 workflow/scripts/extended_data_fig7.R

@@ -400,13 +400,13 @@ annoZoomLines(
 
 
 plotText(
-  label = "Zld embryo ChIP", params = small_text_params, fontface = "bold",
+  label = "Zld embryo", params = small_text_params, fontface = "bold",
   x = (ref_x + x_offset_browser_label), y = (ref_y + 1.75), just = c("right","center"), default.units = "cm"
 )
 
 
 plotText(
-  label = "Zld S2 ChIP", params = small_text_params, fontface = "bold",
+  label = "Zld S2", params = small_text_params, fontface = "bold",
   x = (ref_x + x_offset_browser_label), y = (ref_y + 2.25), just = c("right","center"), default.units = "cm"
 )
 
@@ -557,13 +557,13 @@ annoZoomLines(
 )
 
 plotText(
-  label = "Zld embryo ChIP", params = small_text_params, fontface = "bold",
+  label = "Zld embryo", params = small_text_params, fontface = "bold",
   x = (ref_x + x_offset_browser_label), y = (ref_y + 1.75), just = c("right","center"), default.units = "cm"
 )
 
 
 plotText(
-  label = "Zld S2 ChIP", params = small_text_params, fontface = "bold",
+  label = "Zld S2", params = small_text_params, fontface = "bold",
   x = (ref_x + x_offset_browser_label), y = (ref_y + 2.25), just = c("right","center"), default.units = "cm"
 )
 
@@ -713,13 +713,13 @@ annoZoomLines(
 )
 
 plotText(
-  label = "Zld embryo ChIP", params = small_text_params, fontface = "bold",
+  label = "Zld embryo", params = small_text_params, fontface = "bold",
   x = (ref_x + x_offset_browser_label), y = (ref_y + 1.75), just = c("right","center"), default.units = "cm"
 )
 
 
 plotText(
-  label = "Zld S2 ChIP", params = small_text_params, fontface = "bold",
+  label = "Zld S2", params = small_text_params, fontface = "bold",
   x = (ref_x + x_offset_browser_label), y = (ref_y + 2.25), just = c("right","center"), default.units = "cm"
 )
 
@@ -870,13 +870,13 @@ annoZoomLines(
 )
 
 plotText(
-  label = "Grh embryo ChIP", params = small_text_params, fontface = "bold",
+  label = "Grh embryo", params = small_text_params, fontface = "bold",
[…]

### 2023-03-20 workflow/scripts/plot_heatmap.R

@@ -425,7 +425,8 @@ plot_average <- function(
     row_split = NULL,
     colors = NULL,
     summary_fun = mean,
-    line_width = 1
+    line_width = 1,
+    center_label = "center"
 ) {
   
   require(rtracklayer)
@@ -532,7 +533,7 @@ plot_average <- function(
     scale_x_continuous(
       
       breaks = c(start,end/2,end), 
-      labels = c(paste0("-", upstream / 1000, "KB"),"center", paste0("+", upstream / 1000, "KB")),
+      labels = c(paste0("-", upstream / 1000, "KB"),center_label, paste0("+", upstream / 1000, "KB")),
       expand = c(0,0,0,0)
       ) 
   

### 2023-03-21 manuscript/manuscript_sections/_05_Methods.qmd
add GEO accession to methods

@@
 
-Sequencing data have been deposited in the Gene Expresion Omnibus under accession GSEXXXX.
+Sequencing data have been deposited in the Gene Expresion Omnibus under accession GSE227884
 All analysis code is available on GitHub: https://github.com/tjgibson/S2_pioneers_manuscript.

### 2023-03-24 manuscript/manuscript_sections/_02_Introduction.qmd
incorporate MS changes suggested by MF and AT

@@
 When ectopically expressed, Zld and Grh bind to closed chromatin, establish chromatin accessibility, and promote transcriptional activation at a subset of binding sites.
-We show that an additional TF, Twist (Twi), can function as a pioneer factor.
-By comparing the features of these three factors and relating exogenous activity to *in vivo* functionality, we demonstrate that pioneering depends on a combination of local chromatin structure, sequence context, and PF concentration.
+By comparing the features of these factors and relating exogenous activity to *in vivo* functionality, we demonstrate that pioneering depends on a combination of local chromatin structure, sequence context, and PF concentration.
 We propose that pioneering activity is not a binary feature intrinsic to a protein, but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type specific features.

### 2023-03-24 manuscript/manuscript_sections/_03_Results.qmd
incorporate MS changes suggested by MF and AT

@@
 To investigate the mechanisms by which pioneer factors bind and open chromatin, we generated S2 cell lines to enable the individual expression of Zld or Grh (Extended Data Fig. 1a).
-These cell lines allowed the inducible and tunable expression of each factor, as neither pioneer factor is normally expressed in wild-type S2 cells (Extended Data Fig. 1b,c,d).
+These cell lines allowed the inducible and tunable expression of each factor, as neither pioneer factor is normally expressed in wild-type S2 cells (Extended Data Fig. 1b-d).
 We expressed each factor at approximately physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
@@
 
-::: {#fig-1}
-![](figures/fig1.pdf)
+<!-- ::: {#fig-1} -->
 
-**Ectopically expressed pioneer factors open chromatin and activate transcription.
+<!-- ![](figures/fig1.pdf) -->
+
+<!-- **Ectopically expressed pioneer factors open chromatin and activate transcription.
 a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
-ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for wild-type (WT) cells lacking Zld or Grh expression.
-**c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
-**d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
-**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h).
-i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
-**j,l,** Proportion of class I, II or III regions that is proximal to a gene that is differentially expressed (DE) upon Zld **(j)** or Grh **(l)** expression.
-Promoter-proximal and promoter-distal binding sites are shown separately.
-The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene.
-:::
+-->
+
+<!-- ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for wild-type (WT) cells lacking Zld or Grh expression. -->
+
+<!-- **c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**. -->
+
+<!-- **d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
+-->
+
+<!-- **e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h).
+i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions (shaded area) where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
+-->
+
+<!-- **j,l,** Proportion of class I, II or III regions that is proximal to a gene that is differentially expressed (DE) upon Zld **(j)** or Grh **(l)** expression. -->
+
+<!-- Promoter-proximal and promoter-distal binding sites are shown separately. -->
+
+<!-- The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene. -->
+
+<!-- ::: -->
 
@@
 Differential expression analysis comparing induced cells to wild-type cells identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. 3b,e).
-Genes activated by Zld expression were enriched for embryonic morphogenesis, consistent with the known function of Zld[@Nien2011], as well as imaginal disc development, in which Zld-target genes have not been well characterized (Extended Data Fig.
+Genes activated by Zld were enriched for the gene ontology (GO) term embryonic morphogenesis, consistent with the known function of Zld[@Nien2011], as well as imaginal disc development, in which Zld-target genes have not been well characterized (Extended Data Fig.
 3c).
-The genes activated by Grh were enriched for gene ontology (GO) terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig. 3f)[@Nevil2017].
+The genes activated by Grh were enriched for GO terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig. 3f)[@Nevil2017].
 
@@
 
-Zld and Grh had previously been shown to have features of PFs.
+Zld and Grh had previously been shown to have features of PFs [@Schulz2015; @Sun2015; @McDaniel2019a; @Jacobs2018; @Nevil2020].
 To determine whether these factors had unique properties, we tested an additional transcription factor with distinctive features, Twist (Twi).
@@
 
-::: {#fig-2}
-![](figures/fig2.pdf)
+<!-- ::: {#fig-2} -->
 
-**Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.
+<!-- ![](figures/fig2.pdf) -->
+
+<!-- **Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.
 a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites.
-**c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions.
-**d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.
-:::
+-->
+
+<!-- **c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions. -->
+
+<!-- **d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions. -->
+
+<!-- ::: -->
 
@@
 
-::: {#fig-3}
-![](figures/fig3.pdf)
+<!-- ::: {#fig-3} -->
 
-**Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
-**b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
-Zld, Grh, and Twi ChIP-seq data were generated in this study.
-Remaining ChIP-seq datasets were previously published (see Supplementary table 2)[@Henriques2018; @Philip2015; @Tettey2019].
-**c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
-Logos plot of the canonical motif are shown alongside the heatmap.
-Percentages are shown for class I, II or III binding sites, or all wild-type ATAC-seq peaks as a control for the background motif frequency within regulatory elements.
-**f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
-**g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**.
-Motif scores were determined by comparing each identified motif instance to the canonical motif.
-:::
+<!-- ![](figures/fig3.pdf) -->
+
+<!-- **Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi. -->
+
+<!-- **b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi. -->
+
+<!-- Zld, Grh, and Twi ChIP-seq data were generated in this study. -->
+
+<!-- Remaining ChIP-seq datasets were previously published (see Supplementary table 2)[@Henriques2018; @Philip2015; @Tettey2019]. -->
+
+<!-- **c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**. -->
+
+<!-- Logos plot of the canonical motif are shown alongside the heatmap. -->
+
+<!-- Percentages are shown for class I, II or III binding sites, or all wild-type ATAC-seq peaks as a control for the background motif frequency within regulatory elements. -->
+
+<!-- **f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**. -->
+
+<!-- **g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**. -->
+
+<!-- Motif scores were determined by comparing each identified motif instance to the canonical motif. -->
+
+<!-- ::: -->
 
@@
 6a-c).
-Hierarchical clustering of MNase-seq data revealed that there was no preference for binding or chromatin opening at sites that had motifs in a particular orientation on nucleosomes (Extended Data Fig.
-6d-f).
+To test if there were different patterns of nucleosome occupancy at individual sites that were obscured by global analysis, we performed hierarchical clustering on the MNase-seq data.
+This analysis revealed that there was no preference for binding or chromatin opening at sites that had motifs in a particular orientation on nucleosomes (Extended Data Fig. 6d-f).
 For class II and III peaks, some motifs were positioned close to the nucleosome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
@@
 We analyzed Zld ChIP-seq data from the early embryo (nuclear cycle 14 embryos)[@Harrison2011] and from larval neural stem cells [@Larson2021a], Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017] and Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
-We identified many binding sites that were bound *in vivo*, but were not bound upon ectopic expression in S2 cells (Extended Data Fig. 7a-c).
+We identified many binding sites that were bound *in vivo* but were not bound upon ectopic expression in S2 cells (Extended Data Fig. 7a-c).
 Chromatin enriched for repressive histone modifications H3K9me3 and H3K27me3 is resistant to binding of some PFs [@Mayran2018a; @Soufi2012].
@@
 
-::: {#fig-4}
-![](figures/fig4.pdf)
+<!-- ::: {#fig-4} -->
 
-**Many endogenous binding sites are resistant to ectopic pioneer-factor binding. a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
-ChIP-seq data for repressive histone modifications H3K27me3 [@Jain2020] and H3K9me3[@Eastwood2021] in wild-type S2 cells are also shown.
-**d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
-Zld, Grh, and Twi ChIP-seq data are z-score normalized.
-CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).
-:::
+<!-- ![](figures/fig4.pdf) -->
+
+<!-- **Many endogenous binding sites are resistant to ectopic pioneer-factor binding.
+a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
+-->
+
+<!-- ChIP-seq data for repressive histone modifications H3K27me3 [@Jain2020] and H3K9me3[@Eastwood2021] in wild-type S2 cells are also shown. -->
+
+<!-- **d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**. -->
+
+<!-- Zld, Grh, and Twi ChIP-seq data are z-score normalized. -->
+
+<!-- CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods). -->
+
+<!-- ::: -->
 
@@
 9a-c; see methods).
-Analysis of our previously defined class I-VI sites showed that binding and was concentration dependent (Fig. 5).
+Analysis of our previously defined class I-VI sites showed that binding was concentration dependent (Fig. 5).
 At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i).
@@
 
-::: {#fig-5}
-![](figures/fig5.pdf)
+<!-- ::: {#fig-5} -->
 
-**Pioneer-factor binding and opening of closed chromatin is concentration dependent.
+<!-- ![](figures/fig5.pdf) -->
+
+<!-- **Pioneer-factor binding and opening of closed chromatin is concentration dependent.
 a,e,i,** Boxplots of ChIP-seq signal at class I-III regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
-**b,f,j,** Boxplots of ATAC-seq signal at class I-III regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
-**c,g,k,** Distribution of class II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
-Y-axis indicates the total number of peaks called at each concentration.
-Colors indicate the proportion of those peaks that are defined as class II, or III.
-**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
-:::
+-->
+
+<!-- **b,f,j,** Boxplots of ATAC-seq signal at class I-III regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels. -->
+
+<!-- **c,g,k,** Distribution of class II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**. -->
+
+<!-- Y-axis indicates the total number of peaks called at each concentration. -->
+
+<!-- Colors indicate the proportion of those peaks that are defined as class II, or III. -->
+
+<!-- **d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**. -->
+
+<!-- ::: -->
 
 To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration.
-For Zld and Twi, supraphysiological concentrations led to an increase in class III regions relative to class II regions (Fig. 5c,k).
-For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with most of these sites falling in class III (Fig. 5g).
+For all three factors, supraphysiological concentrations led to an increase in binding to closed chromatin (Fig. 5c,g,k).
+For Twi, this increase was mostly within class II regions.
+For Grh and Zld, supraphysiological concentrations caused a dramatic increase in class III regions relative to class II (Fig. 5g).
 To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs.
 supraphysiological concentrations (Fig.
-5d,h,i).
+5d,h,l).
 For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration.
@@
 
-::: {#fig-6}
-![](figures/fig6.pdf)
+<!-- ::: {#fig-6} -->
 
-**Regions outside the DNA-binding domain are required for pioneer-factor function.
+<!-- ![](figures/fig6.pdf) -->
+
+<!-- **Regions outside the DNA-binding domain are required for pioneer-factor function.
 a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
-Adjacent class I regions are shown where the DBD alone can bind.
-Tracks shown for Zld **(a)** and Grh **(b)**.
-Class III regions are highlighted with gray shading.
-**c,e,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(e)**.
-**d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
-**g,** Model for the regulation of PFs.
-A high level of chromatin occupancy is required for PFs to initiate chromatin opening.
-Chromatin occupancy can be regulated by protein-intrinsic features such as motif affinity and protein domains, and by cell-intrinsic properties such as histone modifications, PF concentration, and cell-type specific cofactors (outlined in blue).
-:::
+-->
+
+<!-- Adjacent class I regions are shown where the DBD alone can bind. -->
+
+<!-- Tracks shown for Zld **(a)** and Grh **(b)**. -->
+
+<!-- Class III regions are highlighted with gray shading. -->
+
+<!-- **c,e,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(e)**. -->
+
+<!-- **d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**. -->
+
+<!-- **g,** Model for the regulation of PFs. -->
+
+<!-- A high level of chromatin occupancy is required for PFs to initiate chromatin opening. -->
+
+<!-- Chromatin occupancy can be regulated by protein-intrinsic features such as motif affinity and protein domains, and by cell-intrinsic properties such as histone modifications, PF concentration, and cell-type specific cofactors (outlined in blue).
+-->
+
+<!-- ::: -->

### 2023-03-24 manuscript/manuscript_sections/_04_Discussion.qmd
incorporate MS changes suggested by MF and AT

@@
 Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and has demonstrated the role that intrinsically disordered regions (IDRs) play in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
-For Zld, this disordered domain is required for transcriptional activation and promotes nucleosome binding *in vitro* [@Hamm2015; @McDaniel2019a].
-IDRs may contribute to TF binding through protein-protein interactions with other transcription factors, nonspecific interactions with DNA, or by driving condensate formation [@Brodsky2021; @Staller2022].
+For Zld, this disordered domain is required for transcriptional activation and promotes optimal nucleosome binding *in vitro* [@Hamm2015; @McDaniel2019a].
+IDRs may contribute to TF binding through protein-protein interactions with other transcription factors, nonspecific interactions with DNA or histones, or by driving condensate formation [@Brodsky2021; @Staller2022].
 Indeed, Zld is visualized in hubs within the nucleus, and these promote binding of additional transcription factors [@Mir2018; @Yamada2019].
@@
 The failure of the DBD alone to promote chromatin accessibility implicates a separation between the ability to bind nucleosomes and the capacity to promote chromatin accessibility.
-Local restructuring of nucleosome-DNA interactions [@Michael2020; @Dodonova2020] may be insufficient for establishing accessible chromatin and, instead, may stabilize PF binding to allow recruitment of cofactors, such as chromatin remodelers [@Judd2021; @King2017].
+Local restructuring of nucleosome-DNA interactions may be insufficient for establishing accessible chromatin [@Michael2020; @Dodonova2020] and, instead, may stabilize PF binding to allow recruitment of cofactors, such as chromatin remodelers [@Judd2021; @King2017].
 Furthermore, our identification of hundreds of class II regions, which are bound but not opened by exogenous PF expression, further separates chromatin binding and PF activity.

### 2023-03-24 manuscript/manuscript_sections/_06_Figure_Legends.qmd
incorporate MS changes suggested by MF and AT

@@
 **Figure 1: Ectopically expressed pioneer factors open chromatin and activate transcription.
-a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
+a,b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
 ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for wild-type (WT) cells lacking Zld or Grh expression.
@@
 **d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
-**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h).
-i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
+**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h)**.
+**i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions (shaded area) where changes in chromatin accessibility (shaded region) are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
 **j,l,** Proportion of class I, II or III regions that is proximal to a gene that is differentially expressed (DE) upon Zld **(j)** or Grh **(l)** expression.
 Promoter-proximal and promoter-distal binding sites are shown separately.
-The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene.
+The grey dotted line indicates the percentage of all binding sites that are proximal to a differentially expressed gene.
 
@@
 Zld, Grh, and Twi ChIP-seq data were generated in this study.
-Remaining ChIP-seq datasets were previously published (see Supplementary table 1)[@Henriques2018; @Philip2015; @Tettey2019].
+Remaining ChIP-seq datasets were previously published (see Supplementary table 2)[@Henriques2018; @Philip2015; @Tettey2019].
 **c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
@@
 **Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function.
-a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
+a,b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
 Adjacent class I regions are shown where the DBD alone can bind.
@@
 Protein levels in stable cell lines are compared to 3-4 hour (H) old embryos.
+Black arrowhead indicates Twi or HA-Twi.
+Gray arrowhead indicates background band used to assess loading.
 **b,** Volcano plots showing changes in ATAC-seq signal in cells expressing Twi when compared to wild-type cells treated with the same concentration of CuSO~4~.

### 2023-03-24 workflow/scripts/extended_data_fig7.R

@@ -457,6 +457,8 @@ annoYaxis(
 plotGenes(
   params = zoomRegion, assembly = "dm6",
   x = (ref_x + x_offset_browser), y = (ref_y + 2.5), height = 1, width = 3.75, fontsize = small_text_params$fontsize,
+  fill = c("gray50", "gray50"),
+  fontcolor = c("gray50", "gray50"),
   just = c("left", "top"),
   default.units = "cm"
 )
@@ -614,6 +616,8 @@ annoYaxis(
 plotGenes(
   params = zoomRegion, assembly = "dm6",
   x = (ref_x + x_offset_browser), y = (ref_y + 2.5), height = 1, width = 3.75, fontsize = small_text_params$fontsize,
+  fill = c("gray50", "gray50"),
+  fontcolor = c("gray50", "gray50"),
   just = c("left", "top"),
   default.units = "cm"
 )
@@ -770,6 +774,8 @@ annoYaxis(
 plotGenes(
   params = zoomRegion, assembly = "dm6",
   x = (ref_x + x_offset_browser), y = (ref_y + 2.5), height = 1, width = 3.75, fontsize = small_text_params$fontsize,
+  fill = c("gray50", "gray50"),
+  fontcolor = c("gray50", "gray50"),
   just = c("left", "top"),
   default.units = "cm"
 )
@@ -927,6 +933,8 @@ annoYaxis(
 plotGenes(
   params = zoomRegion, assembly = "dm6",
   x = (ref_x + x_offset_browser), y = (ref_y + 2.5), height = 1, width = 3.75, fontsize = small_text_params$fontsize,
+  fill = c("gray50", "gray50"),
+  fontcolor = c("gray50", "gray50"),
   just = c("left", "top"),
   default.units = "cm"
 )
@@ -1083,6 +1091,8 @@ annoYaxis(
 plotGenes(
   params = zoomRegion, assembly = "dm6",
   x = (ref_x + x_offset_browser), y = (ref_y + 2.5), height = 1, width = 3.75, fontsize = small_text_params$fontsize,
+  fill = c("gray50", "gray50"),
+  fontcolor = c("gray50", "gray50"),
   just = c("left", "top"),
   default.units = "cm"
 )
@@ -1244,6 +1254,8 @@ annoYaxis(
 plotGenes(
   params = zoomRegion, assembly = "dm6",
[…]

### 2023-03-24 workflow/scripts/fig1.R

@@ -1083,7 +1083,7 @@ x_offset_browser <- 2.0
 
 # panel label
 plotText(
-  label = "I", params = panel_label_params, fontface = "bold",
+  label = "i", params = panel_label_params, fontface = "bold",
   x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
 )
 # 
@@ -1229,6 +1229,7 @@ g1 <- plotGenes(
   params = region, assembly = "dm6",
   x = (ref_x + x_offset_browser), y = (ref_y + 2.5), height = 0.6, width = 2.75, fontsize = small_text_params$fontsize,
   fill = c("gray50", "gray50"),
+  fontcolor = c("gray50", "gray50"),
   just = c("left", "top"),
   default.units = "cm"
 )
@@ -1444,6 +1445,7 @@ g1 <- plotGenes(
   params = region, assembly = "dm6",
   x = (ref_x + x_offset_browser), y = (ref_y + 2.5), height = 0.6, width = 2.75, fontsize = small_text_params$fontsize, 
   fill = c("gray50", "gray50"),
+  fontcolor = c("gray50", "gray50"),
   just = c("left", "top"),
   default.units = "cm"
 )

### 2023-03-24 workflow/scripts/fig4.R

@@ -28,6 +28,19 @@ source("workflow/scripts/utils.R")
 # S2_Twi_ChIP_bw <-  "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Twi_aTwi_IP.bw"
 # embryo_Twi_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/embryo-1-3H_aTwi.bw"
 
+# S2_Zld_ChIP_DMSO_bw <-  "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_DMSO_aZld.bw"
+# S2_Zld_ChIP_taz_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_taz_aZld.bw"
+# S2_Zld_H3K27me3_DMSO_bw <- "histone_CUTandRUN/results/bigwigs/spikeIn_normalized/merged/S2-Zld_DMSO_aH3K27me3_large.bw"
+# S2_Zld_H3K27me3_taz_bw <- "histone_CUTandRUN/results/bigwigs/spikeIn_normalized/merged/S2-Zld_Taz_aH3K27me3_large.bw"
+# S2_Grh_ChIP_DMSO_bw <-  "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_DMSO_aGrh.bw"
+# S2_Grh_ChIP_taz_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_taz_aGrh.bw"
+# S2_Grh_H3K27me3_DMSO_bw <- "histone_CUTandRUN/results/bigwigs/spikeIn_normalized/merged/S2-Grh_DMSO_aH3K27me3_large.bw"
+# S2_Grh_H3K27me3_taz_bw <- "histone_CUTandRUN/results/bigwigs/spikeIn_normalized/merged/S2-Grh_Taz_aH3K27me3_large.bw"
+# S2_Twi_ChIP_DMSO_bw <-  "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-HA-Twi_DMSO_aHA.bw"
+# S2_Twi_ChIP_taz_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-HA-Twi_taz_aHA.bw"
[…]

### 2023-03-24 workflow/scripts/fig5.R

@@ -70,12 +70,6 @@ twi_color <- "#00A08A"
 ref_x <- 0.5
 ref_y <- 0.5
 
-# panel label
-plotText(
-  label = "a", params = panel_label_params, fontface = "bold",
-  x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
-)
-
 
 a_plot <- zld_titration_ChIP_rpkm_fn |> 
   read_tsv() |> 
@@ -104,17 +98,18 @@ plotGG(
   default.units = "cm"
 )
 
+# panel label
+plotText(
+  label = "a", params = panel_label_params, fontface = "bold",
+  x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
+)
 
 # panel B ======================================================================
 # reference points for positioning figure components
 ref_x <- 7.5
 ref_y <- 0.5
 
-# panel label
-plotText(
-  label = "b", params = panel_label_params, fontface = "bold",
-  x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
-)
+
 
 
 b_plot <- zld_titration_ATAC_rpkm_fn |> 
@@ -142,6 +137,12 @@ plotGG(
   x = (ref_x), y = (ref_y),
   width = 6.5, height = 4.5, just = c("left", "top"),
   default.units = "cm"
+)
+
+# panel label
+plotText(
+  label = "b", params = panel_label_params, fontface = "bold",
+  x = ref_x, y = ref_y, just = "bottom", default.units = "cm"
 )
 
 # panel C ======================================================================
@@ -185,11 +186,6 @@ library(eulerr)
 ref_x <- 14.5
 ref_y <- 2.5
 
-# panel label
-plotText(
-  label = "d", params = panel_label_params, fontface = "bold",
-  x = ref_x, y = ref_y, just = "top", default.units = "cm"
-)
 
 # define input files
 input_files <- c(
@@ -236,16 +232,16 @@ plotText(
   x = (ref_x + 2.8), y = (ref_y + 1), just = "top", default.units = "cm"
 )
 
+# panel label
+plotText(
+  label = "d", params = panel_label_params, fontface = "bold",
+  x = ref_x, y = ref_y, just = "top", default.units = "cm"
+)
+
 # panel E ======================================================================
 # reference points for positioning figure components
 ref_x <- 0.5
 ref_y <- 5.5
-
-# panel label
-plotText(
-  label = "e", params = panel_label_params, fontface = "bold",
[…]

### 2023-03-24 workflow/scripts/fig6.R

@@ -201,6 +201,8 @@ annoYaxis(
 plotGenes(
   params = region, assembly = "dm6",
   x = (ref_x + x_offset_browser), y = (ref_y + 2), height = 0.75, width = 6, fontsize = small_text_params$fontsize,
+  fill = c("gray50", "gray50"),
+  fontcolor = c("gray50", "gray50"),
   just = c("left", "top"),
   default.units = "cm"
 )
@@ -327,6 +329,8 @@ annoYaxis(
 plotGenes(
   params = region, assembly = "dm6",
   x = (ref_x + x_offset_browser), y = (ref_y + 2), height = 0.75, width = 6, fontsize = small_text_params$fontsize,
+  fill = c("gray50", "gray50"),
+  fontcolor = c("gray50", "gray50"),
   just = c("left", "top"),
   default.units = "cm"
 )

### 2023-04-04 manuscript/manuscript.qmd
update figs
update MS
edit manuscript

--- 2023-01-12 manuscript/manuscript.qmd
@@
 ---
-title: "Defining properties regulating pioneer-factor binding and function"
-format: html
+title: "Protein-intrinsic properties and context-dependent effects regulate pioneer-factor binding and function"
+author:
+  - Tyler J. Gibson, Melissa M. Harrison*
+  # - Melissa M. Harrison
+affiliations:
+  - name: Department of Biomolecular Chemistry, University of Wisconsin-Madison
+    city: Madison
+    state: WI
+format:
+  docx:
+    reference-doc: custom-reference-doc.docx
 editor: visual
 bibliography: references.bib
-#csl: nature.csl
+csl: nsmb.csl
 ---
+
+Department of Biomolecular Chemistry, University of Wisconsin School of Medicine and Public Health, Madison, WI
+
+\*Correspondence: mharrison3\@wisc.edu
+
+{{< pagebreak >}}
 
@@
 {{< include manuscript_sections/_01_Abstract.qmd >}}
+
+{{< pagebreak >}}
 

### 2023-04-04 manuscript/manuscript_sections/_01_Abstract.qmd
edit manuscript

@@
-While chromatin presents a barrier to the binding of many transcription factors, pioneer factors access nucleosomal targets and promote chromatin opening.
+---
+editor: 
+  markdown: 
+    wrap: sentence
+---
+
+Chromatin is a barrier to the binding of many transcription factors.
+By contrast, pioneer factors access nucleosomal targets and promote chromatin opening.
 Despite binding to target motifs in closed chromatin, many pioneer factors display cell-type specific binding and activity.
@@
 We studied three *Drosophila* transcription factors with distinct DNA-binding domains and biological functions: Zelda, Grainy head, and Twist.
-We demonstrated that the level of chromatin occupancy is a key determinant of activity.
+We demonstrated that the level of chromatin occupancy is a key determinant of pioneering activity.
 Multiple factors regulate occupancy, including motif content, local chromatin, and protein concentration.

### 2023-04-04 manuscript/manuscript_sections/_02_Introduction.qmd
edit manuscript

@@
 During animal development, the totipotent embryo divides and differentiates to give rise to the diverse cell and tissue types of the adult organism.
-The establishment of cell-type specific gene expression programs is driven by transcription factors (TFs) that bind sequence-specifically to *cis*-regulatory elements.
+The establishment of cell-type specific gene expression programs is driven by transcription factors (TFs) that bind sequence specifically to *cis*-regulatory elements.
 The packaging of DNA into chromatin obstructs TFs from accessing their target motifs [@Luger1997].
 Thus, many TFs only bind the small subset of genomic motifs that fall within regions of open chromatin [@Li2011].
-A specialized subset of transcription factors, termed pioneer factors (PFs), are defined by their ability to bind target motifs on nucleosomes and promote chromatin opening, allowing additional TFs to bind and activate transcription [@Iwafuchi-Doi2016; @Zaret2020].
+A specialized subset of transcription factors, termed pioneer factors (PFs), are defined by their ability to bind target motifs on nucleosomes and promote chromatin opening, allowing additional TFs to bind these newly accessible motifs and activate transcription [@Iwafuchi-Doi2016; @Zaret2020].
 Through this activity, PFs drive cell-type specific gene expression programs that determine cell fate.
-Due to these properties, PFs are essential players in development, induced reprogramming, and disease [@Larson2021].
+PFs are essential players in development, induced reprogramming, and disease [@Larson2021].
 
@@
 Chromatin structure may influence PF binding as cell-type specific occupancy is often associated with distinct chromatin features [@Mayran2018a; @Lupien2008; @Cernilogar2019].
-Many PFs bind primarily to naïve chromatin that is inaccessible, inactive, and lacks most histone modifications [@Soufi2012; @Donaghey2018].
+Many PFs bind primarily to naïve chromatin that is inactive and lacks most histone modifications [@Soufi2012; @Donaghey2018].
 Together, these studies indicate that PF occupancy is regulated by both genomic context and cell-type specific features.
@@
 
-Although PFs are defined by their ability to promote chromatin opening, many PFs are only required for chromatin accessibility at a subset of their binding sites.
-Thus, PF activity is regulated at a step other than chromatin binding.
+PF activity is regulated at a step other than chromatin binding.
+PFs are defined by their ability to promote chromatin opening, yet many PFs are only required for chromatin accessibility at a subset of their binding sites.
 For example, upon depletion of the PFs Zelda, GAGA-factor, Oct4, or Sox2, chromatin accessibility is lost only at a subset of the binding sites for the respective factor [@Schulz2015; @Gaskill2021; @maresca2022].
-This suggests that PF binding does not necessarily translate to chromatin opening.
 Pax7, a PF important for specification of pituitary melanotropes, binds to thousands of "pioneer primed" regions where binding does not result in chromatin opening [@Mayran2018a].
@@
 Nonetheless, PF binding can also perturb nucleosome structure.
-Oct4 binding to nucleosomes can result in local remodeling of histone-DNA interactions, especially at the nucleosome entry-exit site [@Michael2020].
+Oct4 binding to nucleosomes results in local remodeling of histone-DNA interactions, especially at the nucleosome entry-exit site [@Michael2020].
 Together, these observations highlight that the relationship between PF binding and activity is not well understood.
@@
 In contrast to the tissue-specific chromatin occupancy of Zld, Grh-binding sites are largely shared between tissues [@Nevil2017].
-Grh is a master regulator of epithelial cell fate that is conserved among metazoans and functions as a pioneer factor in both *Drosophila* larva and mammalian cell culture [@Wang2012; @Jacobs2018; @Nevil2020].
+Grh is a master regulator of epithelial cell fate that is conserved among metazoans and functions as a pioneer factor in both *Drosophila* larva and mammalian cell culture [@Wang2012; @Jacobs2018;@Chen2018; @Nevil2020].
 However, in the early *Drosophila* embryo Grh is dispensible for chromatin accessibility, and only displays pioneering activity in late embryos or larval tissues.
@@
 The essential roles of Zld and Grh in diverse biological processes and the context-dependent binding and activity of these factors makes them excellent models to explore how PF binding is regulated and how PF occupancy relates to chromatin opening.
-Because many previous studies have relied on loss-of-function experiments, it has been difficult to disentangle causal relationships between chromatin structure and PF activity.
-We therefore expressed Zld and Grh in Drosophila Schneider 2 (S2) cells, where neither factor is endogenously expressed.
-When ectopically expressed, Zld and Grh bind to closed chromatin, establish chromatin accessibility, and promote transcriptional activation at a subset of binding sites.
-By comparing the features of these factors and relating exogenous activity to *in vivo* functionality, we demonstrate that pioneering depends on a combination of local chromatin structure, sequence context, and PF concentration.
+Because both Zld and Grh are expressed immediately following fertilization, previous studies have largely relied on loss-of-function experiments.
+Thus, it has been difficult to disentangle causal relationships between chromatin structure and PF activity.
+We therefore leveraged the well-characterized Drosophila Schneider 2 (S2) cell-culture system in which neither Zld nor Grh is endogenously expressed as a platform to identify features that promote or prevent PF binding and activity.
+When ectopically expressed, Zld and Grh bound to closed chromatin, established chromatin accessibility, and promoted transcriptional activation at a subset of binding sites.
+By comparing the features of these factors and relating exogenous activity to *in vivo* functionality, we demonstrated that pioneering depended on a combination of local chromatin structure, sequence context, and PF concentration.
 We propose that pioneering activity is not a binary feature intrinsic to a protein, but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type specific features.

### 2023-04-04 manuscript/manuscript_sections/_03_Results.qmd
edit manuscript

@@
 
-To investigate the mechanisms by which pioneer factors bind and open chromatin, we generated S2 cell lines to enable the individual expression of Zld or Grh (Extended Data Fig. 1a).
+To investigate the mechanisms by which pioneer factors bind and open chromatin, we generated stable S2 cell lines capable of individually expressing either Zld or Grh (Extended Data Fig. 1a).
 These cell lines allowed the inducible and tunable expression of each factor, as neither pioneer factor is normally expressed in wild-type S2 cells (Extended Data Fig. 1b-d).
 We expressed each factor at approximately physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
-Approximately physiological protein levels were achieved with 1000 mM or 100 mM CuSO4 for Zld or Grh, respectively (Extended Data Fig.
-1c,d).We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
+We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
 Additional ChIP experiments were performed in wild-type cells using anti-Zld, or anti-Grh to control for antibody specificity.
-By using these datasets as controls for peak calling, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh (Extended Data Fig.
-1e-f).
+Despite undetectable levels of protein by immunoblot, ChIP signal for both proteins was identified in wild-type cells (Extended Data Fig 1e-f).
+To ensure that all peaks used for subsequent analysis reflected those that were gained upon induction, we used these wild-type ChIP datasets as controls for peak calling.
 
@@
 
-Having established that Zld and Grh bind thousands of sites in S2 cells, we performed assay for transposase-accessible chromatin (ATAC)-seq on wild-type cells or cells expressing Zld or Grh to determine if these factors bind and open closed chromatin in S2 cells.
-Differential accessibility analysis using DESeq2 to compare Zld or Grh-expressing cells to wild-type cells detected `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
+Having established that Zld and Grh bind thousands of sites in S2 cells, we determined chromatin accessibility in wild-type cells or those expressing Zld or Grh using assay for transposase-accessible chromatin (ATAC)-seq.
+Comparisons between Zld or Grh-expressing cells and wild-type cells identified `r zld_diff` differentially accessible regions for Zld (`r zld_up` increased, `r zld_down` decreased) and `r grh_diff` for Grh (`r grh_up` increased, `r grh_down` decreased) (Extended Data Fig.
 2a,d)[@love2014].
-By integrating our ChIP-seq and ATAC-seq data, we identified three classes of PF-bound sites (Fig. 1a,b; Supplementary Table 1).
+We integrated our ChIP-seq and ATAC-seq data to determine the connection between binding and accessibility.
+We identified three classes of PF-bound sites (Fig. 1a,b; Supplementary Table 1).
 Class I sites are accessible in wild-type cells and remain accessible when bound by the ectopically expressed factor.
 Class II sites are inaccessible in wild-type cells.
-However, binding of Zld or Grh does not lead to chromatin opening.
+They are bound by Zld or Grh but this binding does not lead to chromatin opening.
 Class III sites are inaccessible in wild-type cells, but become accessible upon PF binding.
@@
 For each factor, only a subset of binding sites were within closed chromatin, and at only a subset of these did the chromatin become accessible (Fig. 1c,f).
-While class I binding sites occurred predominantly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
+Class I binding sites occurred predominantly at promoters, while the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g).
 Genome-wide analysis of signal intensities at all class II and III sites confirmed that these sites lack ATAC signal in wild-type cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig.
 1e,h).
-Together, these data show that when expressed exogenously at physiolgical levels, Zld and Grh can bind and open closed chromatin at a subset of binding sites.
+Together, these data show that when expressed exogenously at physiolgical levels, Zld and Grh bind and open closed chromatin at a subset of binding sites.
 
@@
 To identify whether Zld- and Grh-mediated chromatin accessibility led to transcriptional activation, we performed RNA-seq.
-Differential expression analysis comparing induced cells to wild-type cells identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. 3b,e).
+Differential expression analysis comparing induced cells to wild-type cells identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. 2b,e).
 Genes activated by Zld were enriched for the gene ontology (GO) term embryonic morphogenesis, consistent with the known function of Zld[@Nien2011], as well as imaginal disc development, in which Zld-target genes have not been well characterized (Extended Data Fig.
-3c).
-The genes activated by Grh were enriched for GO terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig. 3f)[@Nevil2017].
+2g).
+The genes activated by Grh were enriched for GO terms related to cuticle and epithelial development, consistent with the endogenous function of Grh (Extended Data Fig. 2h)[@Nevil2017].
 
 We compared our RNA-seq and ATAC-seq data, allowing us to investigate the relationship between chromatin opening and transcriptional activation.
-Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l).
+Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l, Extended Data Fig. 2c,f).
 When we individually analyzed the three classes of binding sites for changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
 This was particularly evident at class III promoters.
-Thus, increases in chromatin accessibility were correlated with transcriptional activation, although binding and chromatin opening did not always result in gene activation.
+Thus, PF-dependent increases in chromatin accessibility were correlated with transcriptional activation, although binding and chromatin opening did not always result in gene activation.
 
@@
 Zld and Grh had previously been shown to have features of PFs [@Schulz2015; @Sun2015; @McDaniel2019a; @Jacobs2018; @Nevil2020].
-To determine whether these factors had unique properties, we tested an additional transcription factor with distinctive features, Twist (Twi).
-Twi is a master regulator of mesodermal cell fate and, importantly, was not expressed in our wild-type S2 cells.
-Although like Zld and Grh, Twi is essential for regulating *Drosophila* embryonic development, Twi has properties that are distinct from these PFs.
+To determine how other factors would behave in our system, we tested an additional transcription factor, Twist (Twi).
+Twi is a master regulator of mesodermal cell fate [@Simpson1983] and, importantly, was not expressed in our wild-type S2 cells.
+Like Zld and Grh, Twi is an important regulator of *Drosophila* embryonic development.
+However, Twi has properties that are distinct from these PFs.
 Unlike Grh, many Twi-binding sites are developmentally dynamic and are not maintained through embryonic development [@Wilczynski2010a].
-Additionally, Zld pioneering activity is required for Twi binding in the early embryo, as mutations to Zld-binding sites disrupted Twi binding to the *cactus* enhancer [@Yáñez-Cuna2012].
+Additionally, Twi binding in the early embryo requires Zld pioneering activity, as mutations to Zld-binding sites disrupted Twi binding to the *cactus* enhancer [@Yáñez-Cuna2012].
 Deep learning models trained on ChIP-seq and ATAC-seq data for Zld and Twi suggest that Zld motifs contribute to chromatin accessibility and Twi binding, while Twi motifs are not predictive of chromatin accessibility [@brennan2022].
-To test the properties of Twi in our system, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. 4a).
+To test the properties of Twi in our system, we generated stable cell lines that inducibly express Twi at approximately physiological levels (Extended Data Fig. 4a).
 As we had done for Zld and Grh, we determined Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
-Using these data, we identified the same three classes of binding sites that we described for Zld and Grh (Fig 2a,b; Extended Data Fig. 1g).
+These data revealed the same three classes of binding sites as described for Zld and Grh (Fig 2a,b; Extended Data Fig. 1g).
 Class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
 Twi bound extensively to closed chromatin, with `r perc_twi_closed`% of binding sites lacking detectable chromatin accessibility (Fig. 2b).
-At a small subset of Twi-binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-d; Extended Data Fig. 4b).
-RNA-seq revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4c).
+A relatively small subset of these sites were class III, where binding to closed chromatin resulted in chromatin opening (Fig 2a-d; Extended Data Fig. 4b).
+RNA-seq revealed that Twi activated transcription of hundreds of genes (Extended Data Fig. 4c).
 GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal genes, consistent with the endogenous function of Twi (Extended Data Fig. 4d) [@Wilczynski2010a].
-Despite Twi possessing features distinct from Zld and Grh *in vivo*, ectopically expressed Twi binds closed chromatin and promotes accessibility.
+Despite Twi possessing features distinct from Zld and Grh *in vivo*, ectopically expressed Twi bound closed chromatin and promoted accessibility.
 
@@
 5a,c,e) [@Henriques2018; @Philip2015; @Tettey2019; @Straub2013; @Zouaz2017; @Climent-Cantó2020; @Liu2017a; @Enderle2011; @Jain2020; @Eastwood2021; @Alekseyenko2014; @Liang2014; @Ong2013].
-This supports the hypothesis that PFs preferentially bind naïve chromatin that is inaccessible but devoid of most histone modifications [@Soufi2012].
+This supports the hypothesis that PFs preferentially bind naïve chromatin that is inaccessible and devoid of most histone modifications [@Soufi2012].
 The ability of TF binding to promote chromatin accessibility at class III sites and not class II sites was not explained by a pre-existing lack of histones; the levels of H3 and H1 were higher at class III than at class II sites (Extended Data Fig.
-5).
+5a,c,e).
 While most class III sites were regions of naïve chromatin, a subset were within regions marked with the repressive histone modification H3K27me3 (Extended Data Fig. 5b,d,f).
-Together this analysis suggests that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin, and that the inability of these factors to promote chromatin accessibility at class II regions is unlikely to be due to differences in preexisting features of chromatin.
+In fact, H3K27me3 levels were higher in class III compared to class I or II regions (Extended Data Fig. 5a,c,e).
+Together these analyses suggest that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin, and that the inability of these factors to promote chromatin accessibility at class II regions is unlikely to be due to differences in preexisting features of chromatin.
 
@@
 To further determine the features that drive the different outcomes on chromatin accessibility at PF-bound regions, we analyzed enrichment of the canonical Zld, Grh or Twi motifs for each of the three bound classes.
-We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig. 3c).
+We calculated the percentage of class I, II or III peaks that contained an instance of the canonical motif for the respective factor (Fig. 3c-e).
 When compared to all accessible regions, class I regions had low levels of enrichment for the canonical motifs, supporting our prior analysis suggesting that binding to these regions is nonspecific.
@@
 
-## Ectopic Zld, Grh and Twi fail to access many endogenous binding sites
+## Cell-type specific variables affect Zld, Grh and Twi binding
 
 Prior studies have extensively characterized the *in vivo* binding sites of Zld, Grh, and Twi during normal *Drosophila* development.
-To better understand how cell-type specific variables affect PF binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
+To better understand how cell-type specific variables affect PF binding, we compared ChIP-seq data for our PFs expressed in culture to published datasets for endogenous Zld, Grh and Twi binding.
 We analyzed Zld ChIP-seq data from the early embryo (nuclear cycle 14 embryos)[@Harrison2011] and from larval neural stem cells [@Larson2021a], Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017] and Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
 We identified many binding sites that were bound *in vivo* but were not bound upon ectopic expression in S2 cells (Extended Data Fig. 7a-c).
-Chromatin enriched for repressive histone modifications H3K9me3 and H3K27me3 is resistant to binding of some PFs [@Mayran2018a; @Soufi2012].
-To test if these repressive marks were present in S2 cells at the resistant sites (bound *in vivo* but not upon ectopic expression), we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
+Chromatin enriched for repressive histone modifications H3K9me3 and H3K27me3 is resistant to binding by some PFs [@Mayran2018a; @Soufi2012].
+To test if these repressive marks were present in S2 cells at the resistant sites (bound *in vivo* but not upon ectopic expression), we compared our PF-binding data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
 We divided the sites bound *in vivo* but not in our S2 cell system into three classes: class IV sites have high levels of H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a-c; Extended Data Fig.
@@
 For Twi, this increase was mostly within class II regions.
-For Grh and Zld, supraphysiological concentrations caused a dramatic increase in class III regions relative to class II (Fig. 5g).
+For Grh and Zld, supraphysiological concentrations caused a dramatic increase in class III regions relative to class II (Fig. 5c,g).
 To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs.
@@
 For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration.
-Together, these data demonstrate that both binding and chromatin opening by PFs are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are resistant to binding at physiological levels.
+Together, these data demonstrate that both binding and chromatin opening by PFs are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some endogenous binding sites for these factors during normal development.
 

### 2023-04-04 manuscript/manuscript_sections/_04_Discussion.qmd
edit manuscript

@@
 
-By exogenously expressing Zld, Grh, and Twi in S2 cells where they are not normally expressed, leveraging prior data identifying binding sites for these factors *in vivo*, and analyzing the wealth of data on chromatin structure in S2 cells, we identified features that govern PF activity.
-Studying the two well-defined PFs Zld and Grh, along with Twi, which possess features distinct from these two factors, allowed us to determine general properties regulating PF activity.
-Furthermore, our analysis of both binding and accessibility allowed us to separate PF occupancy from activity.
-We find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin structure, protein concentration, DNA motif content, and protein domains outside the DBD (Fig.
+To identify the mechanisms that regulate tissue-specific PF engagement with chromatin, we exogenously expressed Zld, Grh, and Twi in S2 cells where they are not normally expressed.
+Leveraging prior data identifying binding sites for these factors *in vivo* and analyzing the wealth of data on chromatin structure in S2 cells, we identified features that govern PF activity.
+We studied the two well-defined PFs Zld and Grh, along with Twi, which possess features distinct from these two factors.
+Studying these three distinct TFs, which engage the genome through structurally distinct DBDs, allowed us to determine shared properties that regulate their pioneering activity.
+Furthermore, our genome-wide analysis ofbinding, chromatin accessibility, and gene expression allowed us to separate PF occupancy and activity.
+Collectively, we find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin structure, protein concentration, DNA motif content, and protein domains outside the DBD (Fig.
 6g).
@@
 Unlike Zld and Grh, increased expression of Twi did not result in binding to regions bound in other tissues.
-It is probable that additional cofactors, such as Zld itself, may be required for Twi occupancy at these regions.
-Prior studies demonstrated an important role for Zld in promoting Twi binding in the early embryo [@Yáñez-Cuna2012; @brennan2022].
+It is possible that additional cofactors, may be required for Twi occupancy at these regions.
 Along with prior studies, our data suggest that PF occupancy is regulated by tissue-intrinsic features, including levels of PF expression, the complement of cofactors expressed and chromatin structure.
@@
 Similarly, regions outside the DBDs of human PFs FOXA1 and SOX2 are required for robust binding of closed chromatin [@lerner2022].
-Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and has demonstrated the role that intrinsically disordered regions (IDRs) play in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
+Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and has demonstrated a role for intrinsically disordered regions (IDRs) in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
 For Zld, this disordered domain is required for transcriptional activation and promotes optimal nucleosome binding *in vitro* [@Hamm2015; @McDaniel2019a].
-IDRs may contribute to TF binding through protein-protein interactions with other transcription factors, nonspecific interactions with DNA or histones, or by driving condensate formation [@Brodsky2021; @Staller2022].
+IDRs may contribute to TF binding through protein-protein interactions with other TFs, nonspecific interactions with DNA or histones, or by driving condensate formation [@Brodsky2021; @Staller2022].
 Indeed, Zld is visualized in hubs within the nucleus, and these promote binding of additional transcription factors [@Mir2018; @Yamada2019].
@@
 The failure of the DBD alone to promote chromatin accessibility implicates a separation between the ability to bind nucleosomes and the capacity to promote chromatin accessibility.
+While the DBD of Zld is sufficient to bind nucleosomes in vitro, regions outside the DBD promote this interaction [@McDaniel2019a].
+Thus, nucleosome binding in vitro does not necessarily translate into in vivo occupancy of closed chromatin.
 Local restructuring of nucleosome-DNA interactions may be insufficient for establishing accessible chromatin [@Michael2020; @Dodonova2020] and, instead, may stabilize PF binding to allow recruitment of cofactors, such as chromatin remodelers [@Judd2021; @King2017].
-Furthermore, our identification of hundreds of class II regions, which are bound but not opened by exogenous PF expression, further separates chromatin binding and PF activity.
-Increased motif content at the sites in which PFs can promote accessibility suggests a model whereby PF chromatin occupancy is correlated with PF activity.
-Supporting this model, we identified a strong concentration dependence of chromatin opening by Zld and Grh.
+Furthermore, our identification of hundreds of class II regions, which are bound but not opened by exogenous PF expression, distinguishes chromatin binding and PF activity.
+We propose that PF-mediated chromatin accessibility requires high chromatin occupancy.
+This is supported by our data demonstrating a higher motif content at sites in which PF binding promotes accessibility as compared to sites which remain closed despite PF occupancy.
+Furthermore, we identified a strong concentration dependence of Zld and Grh for chromatin opening.
 This model would suggest that PF expression levels must be tightly controlled during development.
-Indeed, misexpression of the PFs Grh and DUX4 lead to diseases such as epithelial cancer or facioscapulohumeral muscular dystrophy (FSHD), respectively [@Campbell2018; @Reese2019].
+Indeed, misexpression of the PFs Grh and DUX4 leads to diseases such as epithelial cancer or facioscapulohumeral muscular dystrophy (FSHD), respectively [@Campbell2018; @Reese2019].
 Together our data identify protein intrinsic and extrinsic features that govern PF binding and activity, providing insights into how these factors define novel *cis*-regulatory elements and drive gene-regulatory networks to modulate cell fate.

### 2023-04-04 manuscript/manuscript_sections/_05_Methods.qmd
edit manuscript

--- 2023-03-20 manuscript/manuscript_sections/_05_Methods.qmd
@@
 
-Cells were plated at 1×10^6^ cells per mL and DMSO or tazemetostat was added to a final concentration of 10 µM.
+Cells were plated at 1×10^6^ cells per mL and DMSO or tazemetostat (Fisher Scientific) was added to a final concentration of 10 µM.
 Cells were split on days 2 and 5 and tazemetostat was added to maintain a final concentration of 10 µM.
@@
 
-Previously published datasets analyzed in this study can be found in supplemental table 2.
+Previously published datasets analyzed in this study can be found in supplementary table 2.
 All previously published data were analyzed in parallel with data from this study using analysis parameters described above.
@@
 
-Sequencing data have been deposited in the Gene Expresion Omnibus under accession GSEXXXX.
+Sequencing data have been deposited in the Gene Expresion Omnibus under accession GSE227884. Reviewers can access the data using reviewer token whsjesemxzqllyj.
 All analysis code is available on GitHub: https://github.com/tjgibson/S2_pioneers_manuscript.
+
+## Acknowledgements
+We thank Alex Theis, Meghan Freund, and Andrew Mehle for helpful discussions and advice.
+We thank Julia Zeitlinger for generously sharing the Twist antibody.
+We acknowledge the University of Wisconsin-Madison Biotechnology Center and the NUSeq Core Facility for sequencing.
+TJG was supported by the National Institutes of Health (NIH) National Research Service Award T32 GM007215.
+Experiments were supported by a R35 GM136298 from NIH to MMH.
+MMH was also supported by a Vallee Scholar Award.

### 2023-04-04 manuscript/manuscript_sections/_06_Figure_Legends.qmd
edit manuscript

@@
 Motif scores were determined by comparing each identified motif instance to the canonical motif.
+For all boxplots, line shows the median, boxes extend from the 25th to the 75th percentile, and whiskers show 1.5 × the interquartile range.
+Outlier points beyond the range of the whiskers are shown individually.
 
@@
 a,** Schematic of generation of stable cell lines and induction of protein expression.
-**b,** mRNA levels of Zld and Grh in S2 cells.
+**b,** mRNA levels of *zld* and *grh* in S2 cells.
 Top histogram shows the distribution of mRNA levels for all *Drosophila* genes.
@@
 **c-d,** Immunoblots showing titration of Zld **(c)** or Grh **(d)** protein levels in stable cell lines.
-Two independently generated cell lines are shown and compared to 2-3H embryos.
-60,000 cells were loaded in each well, which is equivalent to the approximately 60,000 nuclei present in 10 2-3 hour embryos.
+Two independently generated cell lines are shown and compared to 2-3 hours (H) embryos.
+60,000 cells were loaded in each well, which is equivalent to the approximately 60,000 nuclei present in 10 2-3 hours embryos.
 Black arrowheads indicate Zld **(c)** or Grh **(d)**.
@@
 **Extended Data Figure 2: Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.
-a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
+a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to wild-type cells treated with the same concentration of CuSO~4~.
 **b,e,** RNA-seq volcano plots showing gene expression changes in cells expressing Zld **(b)** or Grh **(e)** when compared to wild-type cells treated with the same concentration of CuSO~4~.
-**c,f,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(c)** or Grh **(f)**.
+**c,f,** Violin plots showing the correlation between changes in chromatin accessibility and gene expression upon expression of Zld **(c)** or Grh **(f)**.
+On the x-axis, all ATAC-seq peaks are grouped based on increased, decreased, or non-significant (ns) changes to chromatin accessibility in Zld- or Grh-expresing cells compared to wild-type cells.
+Groups were compared using a Wilcoxon rank sum test and Bonferroni-corrected p-values are shown.
+**g,h,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(g)** or Grh **(h)**.
 
@@
 DNA-binding domain protein levels are shown at a range of CuSO~4~ concentrations and compared to an equivalent number of cells expressing full-length Zld or Grh at approximately physiological levels.
+The concentration of CuSO~4~ used to induce DBD expression for ChIP, ATAC and immunofluorescence experiments is indicated in red.
 **c-d,** Immunofluorescent microscopy images of stable cell lines expressing full-length protein or DBD only for Zld **(c)** or (Grh).

### 2023-04-04 manuscript/supplemental_figures.qmd
incorporate MS changes suggested by MF and AT
edit manuscript

@@
 **Stable cell lines allow inducible expression of transcription factors at physiological concentrations. a,** Schematic of generation of stable cell lines and induction of protein expression.
-**b,** mRNA levels of Zld and Grh in S2 cells.
+**b,** mRNA levels of *zld* and *grh* in S2 cells.
 Top histogram shows the distribution of mRNA levels for all *Drosophila* genes.
@@
 **Ectopic expression of pioneer factors in S2 cells leads to widespread changes to chromatin accessibility and gene expression.
-a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to WT cells treated with the same concentration of CuSO~4~.
+a,d,** Volcano plots showing changes in ATAC-seq signal in cells expressing Zld **(a)** or Grh **(d)** when compared to wild-type cells treated with the same concentration of CuSO~4~.
 **b,e,** RNA-seq volcano plots showing gene expression changes in cells expressing Zld **(b)** or Grh **(e)** when compared to wild-type cells treated with the same concentration of CuSO~4~.
-**c,f,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(c)** or Grh **(f)**.
+**c,f,** Violin plots showing the correlation between changes in chromatin accessibility and gene expression upon expression of Zld **(c)** or Grh **(f)**.
+On the x-axis, all ATAC-seq peaks are grouped based on increased, decreased, or non-significant (ns) changes to chromatin accessibility in Zld- or Grh-expresing cells compared to wild-type cells.
+Groups were compared using a Wilcoxon rank sum test and Bonferroni-corrected p-values are shown.
+**g,h,** Bar plots showing enrichment of gene ontology terms in genes significantly upregulated upon expression of Zld **(g)** or Grh **(h)**.
 :::

### 2023-10-20 manuscript/manuscript_sections/_02_Introduction.qmd
update analysis and MS based on reviewer comments

@@
 PFs are defined by their ability to promote chromatin opening, yet many PFs are only required for chromatin accessibility at a subset of their binding sites.
-For example, upon depletion of the PFs Zelda, GAGA-factor, Oct4, or Sox2, chromatin accessibility is lost only at a subset of the binding sites for the respective factor [@Schulz2015; @Gaskill2021; @maresca2022].
+For example, upon depletion of the PFs Zelda, GAGA-factor, Oct4, or Sox2, chromatin accessibility is lost only at a subset of the binding sites for the respective factor [@Schulz2015; @Gaskill2021; @Maresca2023].
 Pax7, a PF important for specification of pituitary melanotropes, binds to thousands of "pioneer primed" regions where binding does not result in chromatin opening [@Mayran2018a].
@@
 In contrast to the tissue-specific chromatin occupancy of Zld, Grh-binding sites are largely shared between tissues [@Nevil2017].
-Grh is a master regulator of epithelial cell fate that is conserved among metazoans and functions as a pioneer factor in both *Drosophila* larva and mammalian cell culture [@Wang2012; @Jacobs2018;@Chen2018; @Nevil2020].
+Grh is a master regulator of epithelial cell fate that is conserved among metazoans and functions as a pioneer factor in both *Drosophila* larva and mammalian cell culture [@Wang2012; @Jacobs2018; @Chen2018; @Nevil2020].
 However, in the early *Drosophila* embryo Grh is dispensible for chromatin accessibility, and only displays pioneering activity in late embryos or larval tissues.

### 2023-10-20 manuscript/manuscript_sections/_03_Results.qmd
update analysis and MS based on reviewer comments

@@
 Additional ChIP experiments were performed in wild-type cells using anti-Zld, or anti-Grh to control for antibody specificity.
-Despite undetectable levels of protein by immunoblot, ChIP signal for both proteins was identified in wild-type cells (Extended Data Fig 1e-f).
-To ensure that all peaks used for subsequent analysis reflected those that were gained upon induction, we used these wild-type ChIP datasets as controls for peak calling.
+Despite undetectable levels of protein by immunoblot, ChIP signal for both proteins was identified in wild-type, unedited cells (Extended Data Fig 1e,f).
+To ensure that all peaks used for subsequent analysis reflected those that were gained upon induction, we used these wild-type ChIP datasets (S2-WT) as controls for peak calling (see methods).
 
@@
 Next, we sought to determine the binding dynamics of Zld and Grh.
-Zld and Grh protein was detectable at 4 or 12 hours following induction, respectively (Extended Data Fig. 3a-b).
+Zld and Grh protein was detectable at 4 or 12 hours following induction, respectively (Extended Data Fig. 3a,b).
 We used CUT&RUN to measure Zld and Grh binding to chromatin at 0, 4, 12, 24 or 48 hours following induction[@Skene2017].
@@
 
-## Chromatin opening correlates with transcriptional activation
+## Chromatin opening correlates with increased transcript levels
 
@@
 
-We compared our RNA-seq and ATAC-seq data, allowing us to investigate the relationship between chromatin opening and transcriptional activation.
-Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l, Extended Data Fig. 2c,f).
+We compared our RNA-seq and ATAC-seq data, allowing us to investigate the relationship between chromatin opening and expression levels.
+Overall, chromatin opening was correlated with transcript levels (Fig. 1i-l, Extended Data Fig. 2c,f).
 When we individually analyzed the three classes of binding sites for changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l).
 This was particularly evident at class III promoters.
-Thus, PF-dependent increases in chromatin accessibility were correlated with transcriptional activation, although binding and chromatin opening did not always result in gene activation.
+Thus, PF-dependent increases in chromatin accessibility were correlated with increased transcript levels.
 
@@
 Additionally, Twi binding in the early embryo requires Zld pioneering activity, as mutations to Zld-binding sites disrupted Twi binding to the *cactus* enhancer [@Yáñez-Cuna2012].
-Deep learning models trained on ChIP-seq and ATAC-seq data for Zld and Twi suggest that Zld motifs contribute to chromatin accessibility and Twi binding, while Twi motifs are not predictive of chromatin accessibility [@brennan2022].
+Deep learning models trained on ChIP-seq and ATAC-seq data for Zld and Twi suggest that Zld motifs contribute to chromatin accessibility and Twi binding, while Twi motifs are not predictive of chromatin accessibility [@Brennan2023].
 To test the properties of Twi in our system, we generated stable cell lines that inducibly express Twi at approximately physiological levels (Extended Data Fig. 4a).
@@
 Overlapping class I regions showed high levels of these chromatin marks (Fig. 3b), suggesting that exogenously expressed factors may bind nonspecifically to active regions of the genome.
+As would be predicted if active chromatin promoted non-specific binding, those regions co-bound by all three factors were more enriched than regions bound by fewer factors (Fig. 3b).
+To focus our analysis on specific pioneer-factor binding, we excluded the set of cobound class I regions from downstream analysis.
 
@@
 In fact, H3K27me3 levels were higher in class III compared to class I or II regions (Extended Data Fig. 5a,c,e).
-Together these analyses suggest that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin, and that the inability of these factors to promote chromatin accessibility at class II regions is unlikely to be due to differences in preexisting features of chromatin.
+Together these analyses suggest that Zld, Grh and Twi pioneer at regions within naïve or repressed chromatin, and that the inability of these factors to promote chromatin accessibility at class II regions is not due to a preexisting permissive chromatin state.
+Instead, our data are consistent with the model that nucleosomes may promote PF function, as has been reported for other PFs [@miao2022; @veil2019; @ballaré2013].
 
@@
 
-To test the functional relevance of repressive chromatin to PF binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3 (Extended Data Fig 8).
+To test the functional relevance of repressive chromatin to PF binding, we used the chemical inhibitor tazemetostat to disrupt activity of the PRC2 complex responsible for depositing H3K27me3 (Extended Data Fig 8 a,b).
 Despite causing dramatic reductions in H3K27me3 levels, tazemetostat treatment did not result in global changes to Zld, Grh, or Twi binding when compared to DMSO controls (Fig. 4d-f).
 The majority of class IV binding sites remained unbound in tazemetostat-treated cells, demonstrating that loss of PRC2 activity is insufficient to promote widespread binding of any of the three factors studied.
+We therefore further examined the chromatin state of the class IV regions in S2 cells, but did not identify enrichment of other chromatin marks that might explain the continued lack of binding even upon tazemetostat treatment (Extended Data Fig.
+8c-e).
+
+Given the lack of a clear role for chromatin features in driving the tissue-specific PF occupancy, we investigated whether cofactors might contribute by analyzing motif enrichment at class I-VI regions (Extended Data Fig.
+8f-h).
+For all three PF, class I regions were enriched for motifs bound by factors that are robustly expressed in S2 cells.
+By contrast, regions bound in other tissues (class IV, V, VI) were enriched for factors lowly expressed in S2 cells and more highly expressed in other tissues.
+Class IV sites for Twist are enriched for the Zld-binding motif, and Zld has been demonstrated to promote Twist binding in the embryo[@Yáñez-Cuna2012].
+Together this analysis identifies cofactors that may facilitate PF binding in specific tissues and suggests that highly expressed factors in S2 cells promote occupancy of class I regions.
 
@@
 Therefore, we tested if differences in the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.
-We performed ChIP-seq on cells induced to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig.
+We performed ChIP-seq on cells induced to express Zld, Grh and Twi at multiple protein concentrations within a range including physiological levels: uninduced, low, medium, or high (Extended Data Fig.
 9a-c; see methods).
 Analysis of our previously defined class I-VI sites showed that binding was concentration dependent (Fig. 5).
-At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i).
-Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i).
+At class II and III binding sites, low levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i).
+Expression of high protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i).
 To identify the relationship between PF binding and chromatin accessibility, we performed ATAC-seq on cells with the same range in expression.
@@
 Many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels, suggesting that increased chromatin occupancy, as assayed by ChIP-seq, could result in chromatin accessibility.
-For Grh and Zld, but not Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites, which were not bound at lower concentrations (Extended Data Fig. 9d-f).
+For Grh and Zld, but not Twi, high concentrations allowed some binding and opening to class IV-VI sites, which were not bound at lower concentrations (Extended Data Fig. 9d-f).
 
@@
 To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration.
-For all three factors, supraphysiological concentrations led to an increase in binding to closed chromatin (Fig. 5c,g,k).
+For all three factors, high concentrations led to an increase in binding to closed chromatin (Fig. 5c,g,k).
 For Twi, this increase was mostly within class II regions.
-For Grh and Zld, supraphysiological concentrations caused a dramatic increase in class III regions relative to class II (Fig. 5c,g).
-To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs.
-supraphysiological concentrations (Fig.
+For Grh and Zld, high concentrations caused a dramatic increase in class III regions relative to class II (Fig. 5c,g).
+To assess the extent to which increased concentrations led to binding to new regions that were not bound when induced at medium levels, we overlapped the binding sites detected at medium vs.
+high levels (Fig.
 5d,h,l).
 For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration.
-Together, these data demonstrate that both binding and chromatin opening by PFs are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some endogenous binding sites for these factors during normal development.
 
-## Regions outside the DNA-binding domain are required for binding and opening of closed chromatin
+To explore whether concentration similarly effects PF binding in vivo, we built on our prior studies of Zld in neural stem cells[@Larson2021a].
+We induced expression of HA-tagged Zld in the neural stem cells, allowing us to specifically query binding of this exogenous, overexpressed Zld using CUT&RUN.
+We identified 4557 novel Zld-binding sites upon over-expression (Extended Data Fig. 9j).
+Analysis of ATAC-seq on brains enriched for neural stem cells demonstrated that these newly occupied sites were enriched for the Zld-binding (CAGGTA) motif in regions of closed chromatin (Extended Data Fig.
+9j,k).
+Thus, similar to increased expression in S2 cells, exogenous expression of Zld in neural stem cells results in increased binding to regions of closed chromatin enriched for the PF-binding motif.
+Prior studies of Grh demonstrated that exogenous expression in neurons results in increased chromatin accessibility at regions enriched for the Grh motif[@Jacobs2018].
+Together, these data both in culture and in larval brains demonstrate that binding and chromatin opening by PFs are concentration dependent, and that increasing protein concentration leads to novel binding sites.
+
+## The DNA-binding domain is not sufficient for binding and opening closed chromatin
 
@@
 Binding of the Grh DBD to class II-III regions was strongly reduced compared to the full-length protein, although some closed regions retained low levels of binding (Fig. 6b,e).
-For both Zld and Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
+For both Zld and Grh, when expressed at approximately physiological levels the DBD alone was insufficient to open chromatin (Fig 6f,i).
+Because levels of Zld and Grh correlated with binding to closed chromatin, we tested whether higher levels of DBD expression could overcome the lack of binding at lower concentrations.
+We induced increased levels Zld DBD or Grh DBD and assayed binding and accessibility.
+In contrast to the full-length proteins, increased expression of the DBD alone did not result in increased binding to closed chromatin as compared to lower concentrations nor did it result in significant chromatin opening at bound regions (Fig 6 c-j).
 Together, these experiments show that regions outside the DBD are essential for binding and opening closed chromatin.

### 2023-10-20 manuscript/manuscript_sections/_04_Discussion.qmd
update analysis and MS based on reviewer comments

@@
 To identify the mechanisms that regulate tissue-specific PF engagement with chromatin, we exogenously expressed Zld, Grh, and Twi in S2 cells where they are not normally expressed.
-Leveraging prior data identifying binding sites for these factors *in vivo* and analyzing the wealth of data on chromatin structure in S2 cells, we identified features that govern PF activity.
+Leveraging prior data identifying binding sites for these factors *in vivo* and analyzing the wealth of data on chromatin structure and gene expression in S2 cells, we identified features that govern PF activity.
 We studied the two well-defined PFs Zld and Grh, along with Twi, which possess features distinct from these two factors.
 Studying these three distinct TFs, which engage the genome through structurally distinct DBDs, allowed us to determine shared properties that regulate their pioneering activity.
-Furthermore, our genome-wide analysis ofbinding, chromatin accessibility, and gene expression allowed us to separate PF occupancy and activity.
-Collectively, we find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin structure, protein concentration, DNA motif content, and protein domains outside the DBD (Fig.
+Furthermore, our genome-wide analysis of binding, accessibility, and gene expression allowed us to separate PF occupancy and activity.
+Collectively, we find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin structure, cofactor expression, protein concentration, DNA motif content, and protein domains outside the DBD (Fig.
 6g).
 
-Even when expressed at low levels Zld, Grh and Twi bind promiscuously to active, accessible regions with degenerate motifs or no detectable motifs at all.
+Even when expressed at low levels Zld, Grh and Twi bind promiscuously to active, accessible regions with degenerate motifs or no detectable motifs for the TF.
 When expressed at approximately physiological levels these factors all possess the capacity to bind to canonical motifs in closed chromatin.
@@
 While H3K27me3 levels were decreased in our treatment, PF binding was largely unchanged.
-Thus, unlike H3K9me3 in the case of OSKM, H3K27me3 is not a barrier to binding by the three factors assayed and suggest other features are restricting binding to these regions.
+Thus, unlike H3K9me3 in the case of OSKM, H3K27me3 is not a barrier to binding by the three factors assayed and our analysis suggests other features, including the expression of additional TFs, are restricting binding to these regions.
 
@@
 These results suggest that protein concentration and affinity for the target motif drives tissue-specific PF binding rather than the presence or absence of Polycomb-mediated silencing.
+This is consistent with recent work demonstrating the Zld-mediated chromatin opening in early embryos depends on motif affinity[@Brennan2023].
 Tissue-specific PF occupancy is also likely regulated by the collection of cofactors expressed in a given tissue.
@@
 Unlike Zld and Grh, increased expression of Twi did not result in binding to regions bound in other tissues.
-It is possible that additional cofactors, may be required for Twi occupancy at these regions.
+It is possible that additional cofactors may be particularly required for Twi occupancy at these regions, and it is clear that in the embryo Twi depends on Zld for binding.
+However, our data also show that upon exogenous expression Twi possess some features of a PF.
+This is similar to recent data demonstrating that TWIST1 is required for chromatin accessibility in human neural crest cells, and together support a model in which Twi possesses some features of a PF when expressed in certain tissues [@kim2023].
 Along with prior studies, our data suggest that PF occupancy is regulated by tissue-intrinsic features, including levels of PF expression, the complement of cofactors expressed and chromatin structure.
@@
 We show that regions outside the DBD of Zld and Grh are important for binding-site selection.
-Similarly, regions outside the DBDs of human PFs FOXA1 and SOX2 are required for robust binding of closed chromatin [@lerner2022].
+Similarly, regions outside the DBDs of human PFs FOXA1 and SOX2 are required for robust binding of closed chromatin [@Lerner2023], and domains both N-terminal and C-terminal to the zinc-finger DBD of GAGA factor are required for stable genomic occupancy [@tang2022].
 Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and has demonstrated a role for intrinsically disordered regions (IDRs) in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
@@
 This model would suggest that PF expression levels must be tightly controlled during development.
-Indeed, misexpression of the PFs Grh and DUX4 leads to diseases such as epithelial cancer or facioscapulohumeral muscular dystrophy (FSHD), respectively [@Campbell2018; @Reese2019].
+Indeed, misexpression of the PFs GRHL2 and DUX4 leads to diseases such as epithelial cancer or facioscapulohumeral muscular dystrophy (FSHD), respectively [@Campbell2018; @Reese2019].
 Together our data identify protein intrinsic and extrinsic features that govern PF binding and activity, providing insights into how these factors define novel *cis*-regulatory elements and drive gene-regulatory networks to modulate cell fate.

### 2023-10-20 manuscript/manuscript_sections/_05_Methods.qmd
update analysis and MS based on reviewer comments

@@
 Transcription factor expression was induced by adding CuSO~4~ to the cell culture media.
-Expression of protein at physiological levels was achieved with 1000, 100, or 40 µM CuSO~4~ for Zld, Grh or Twi, respectively (Extended Data Fig. 1c-d; Extended Data Fig. 4a).
-For experiments with sub- or supraphysiological protein levels (Fig.
+Expression of protein at approximately physiological levels was achieved with 1000, 100, or 40 µM CuSO4 for Zld, Grh or Twi, respectively (Extended Data Fig. 1c-d; Extended Data Fig. 4a).
+For experiments with multiple protein levels (Fig.
 5; Extended Data Fig.
-9a-c), the following CuSO~4~ concentrations were used: 500 µM (Zld subphysiological), 1000 µM (Zld physiological) or 1500 µM (Zld supraphysiological); 25 µM (Grh subphysiological), 100 µM (Grh physiological) or 400 µM (Grh supraphysiological); 10 µM (Twi subphysiological), 40 µM (Twi physiological) or 160 µM (Twi supraphysiological).
+9a-c), the following CuSO~4~ concentrations were used: 500 µM (Zld low), 1000 µM (Zld medium) or 1500 µM (Zld high); 25 µM (Grh low), 100 µM (Grh medium) or 400 µM (Grh high); 10 µM (Twi low), 40 µM (Twi medium) or 160 µM (Twi high); 100 µM (Zld DNA-binding domain low); 400 µM (Zld DNA-binding domain high); 20 µM (Grh DNA-binding domain low); 100 µM (Grh DNA-binding domain high).
+The Zld or Grh DNA-binding domains were induced with 100 µM, or 20 µM CuSO~4~, respectively for immunofluorescence experiments (Extended Data Fig. 10).
 For all induction experiments, cells were plated at 1×10^6^ cells per mL.
@@
 Data from ChIP experiments performed with the same antibody in wild-type cells were used as a control for peak calling.
+Thus, only peaks with statistically significantly enrichment (MACS2 q-value \< 0.5) in induced cells relative to wild-type cells are called as peaks.
 Only peaks that were detected in all replicates were considered in downstream analysis.
@@
 
+## Drosophila strains and collection of brains for CUT&RUN
+
+Drosophila strains and genetic crosses for the isolation of type II neuroblasts (neural stem cells) and for the ectopic expression of Zld in neural stem cells were described previously [@Larson2021a].
+Brains from *brat^11/DF^* mutant larvae were used to perform CUT&RUN for endogenous Zld in neural stem cells.
+For ectopic expression of Zld, brains were collected from *brat^11/DF,Tub-Gal80ts;UAS-HA-Zld^* larvae, in which *Worniu-Gal4* drives overexpression of HA-Zld in the neural stem cells.
+Larvae were grown at 18°C until reaching the L3 stage, then shifted to 33°C for 24 hours to induce Zld overexpression.
+50 brains were dissected into Schneider's medium.
+Brains were centrifuged at 600 × *g* for 3 minutes and transferred to CUT&RUN wash buffer supplemented with 0.1% BSA.
+Brains were dounced 10 times in a 1 mL dounce, centrifuged at 600 × *g* for 3 minutes and resuspended in 100 µL CUT&RUN wash buffer.
+
 ## CUT&RUN
@@
 CUT&RUN was performed using EpiCypher reagents according to the manufacturer protocol.
-2×10^5^ cells were used for each CUT&RUN reaction.
+2×10^5^ cells, or material from 50 brains prepared as described above, were used for each CUT&RUN reaction.
 Overnight incubation with antibodies was performed in antibody buffer containing 0.05% digitonin.
@@
 
+## Motif analysis
+
+Motif analysis was performed using the MEME suite[@Bailey2015], the memes R package[@Nystrom2021], and the universalmotif R package[@universalmotif].
+To facilitate analysis of canonical Zld, Grh, or Twi motifs, position-weight matrices were derived from ChIP-seq data for each factor in the *Drosophila* embryos.
+FIMO[@Grant2011] was used to find instances of the canonical motifs within ChIP-seq peaks for S2 cells.
+Motifs with an adjusted p-value \< 1×10^-3^ were considered for analysis of motif number and score.
+AME[@mcleay2010] was used to test for enrichment of known motifs from the FlyFactorSurvey database[@zhu2010].
+Previously published RNA-seq data was used to pre-filter motifs to include only those for which the associate factor is expressed (RPKM \> 1) in at least one of the tissues being analyzed (embryo, neural stem cells, and wing discs).
+Motifs enrichment was visualized using the normalized motif rank implemented by the memes package.
+
 ## Analysis of previously published data
 
-Previously published datasets analyzed in this study can be found in supplementary table 2.
+Previously published datasets analyzed in this study can be found in Supplementary Table 2.
 All previously published data were analyzed in parallel with data from this study using analysis parameters described above.
@@
 
-Sequencing data have been deposited in the Gene Expresion Omnibus under accession GSE227884. Reviewers can access the data using reviewer token whsjesemxzqllyj.
+Sequencing data have been deposited in the Gene Expresion Omnibus under accession GSE227884.
+Reviewers can access the data using reviewer token whsjesemxzqllyj.
 All analysis code is available on GitHub: https://github.com/tjgibson/S2_pioneers_manuscript.
@@
 ## Acknowledgements
+
 We thank Alex Theis, Meghan Freund, and Andrew Mehle for helpful discussions and advice.
 We thank Julia Zeitlinger for generously sharing the Twist antibody.
+We particularly thank Elizabeth Larson for sharing the neural stem cell data for HA-Zld.
 We acknowledge the University of Wisconsin-Madison Biotechnology Center and the NUSeq Core Facility for sequencing.
 TJG was supported by the National Institutes of Health (NIH) National Research Service Award T32 GM007215.
-Experiments were supported by a R35 GM136298 from NIH to MMH.
+Experiments were supported by a R35 GM136298 and R01 NS111647 from NIH to MMH.
 MMH was also supported by a Vallee Scholar Award.
+MMH is a Romnes Faculty Fellow and Vilas Faculty Mid-Career Investigator.

### 2023-10-20 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update analysis and MS based on reviewer comments

@@
 The grey dotted line indicates the percentage of all binding sites that are proximal to a differentially expressed gene.
+Expression was induced for 48 hours.
 
@@
 **d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.
+Expression was induced for 48 hours.
 
@@
 **b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
+Cobound indicates regions bound by all three factors.
+Not cobound indicates regions bound by any combination of two factors.
 Zld, Grh, and Twi ChIP-seq data were generated in this study.
@@
 CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).
+Expression was induced for 48 hours.
 
 **Figure 5: Pioneer-factor binding and opening of closed chromatin is concentration dependent.
-a,e,i,** Boxplots of ChIP-seq signal at class I-III regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
+a,e,i,** Boxplots of ChIP-seq signal at class I-III regions previously identified for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
 **b,f,j,** Boxplots of ATAC-seq signal at class I-III regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
-**c,g,k,** Distribution of class II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
-Y-axis indicates the total number of peaks called at each concentration.
-Colors indicate the proportion of those peaks that are defined as class II, or III.
-**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
+**c,g,k,** Distribution of redefined class II and III peaks when redefined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
+Y-axis indicates the total number of peaks called at each level of induction.
+Colors indicate the proportion of those peaks that are defined as class II, or III for each level of induction.
+**d,h,l,** Venn diagrams showing overlap of peaks called at medium or high expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
+Expression was induced for 48 hours.
 
-**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function.
-a,b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
+**Figure 6: The DNA-binding domain is not sufficient for pioneer-factor function.
+a,b,** Genome browser tracks showing example of class III regions where 48 hours of expression of the DBD alone is insufficient for robust binding and chromatin opening.
 Adjacent class I regions are shown where the DBD alone can bind.
@@
 Class III regions are highlighted with gray shading.
-**c,e,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(e)**.
-**d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
-**g,** Model for the regulation of PFs.
+**c,d,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(d)**.
+CuSO~4~ levels used for induction are indicated for each DBD.
+(1000 µM CuSO~4~ was used for full-length Zld.
+100 µM CuSO~4~ was used for full-length Grh.) **e-j,** Volcano plots for ATAC-seq data upon expression of full-length Zld **(e)**, Zld DBD at low concentration **(f)**, Zld DBD at high concentration **(g),** full-length Grh **(h)**, Grh DBD at low concentration **(i)**, or Grh DBD at high concentration **(j)**.
+Regions bound by full-length Zld are indicated in blue.
+Regions bound by full-length -Grh are indicated in orange.
+**k,** Model for the regulation of PFs.
 A high level of chromatin occupancy is required for PFs to initiate chromatin opening.
@@
 **e-g,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in wild-type (WT) cells.
+ATAC-seq signal in wild-type cells is shown for reference.
 
@@
 
-**Extended Data Figure 8: Loss of H3K27me3 in tazemetostat-treated cells. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
+**Extended Data Figure 8: Analysis of chromatin and motif content at class I-VI regions. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
 Tubulin levels are shown as a loading control.
@@
 For each sample, the heatmap displays the percentage of barcode reads for each sample and histone modification relative to the total number of barcode reads for all modifications.
+**c-e,** Heatmaps showing the levels of different chromatin marks in class I-VI regions for Zld **(c)**, Grh **(d)** or Twi **(e)**.
+**f-h,** Enrichment of known motifs in class I-VI sites for Zld **(f)**, Grh **(g)** or Twi **(h)**.
+Left heatmap shows normalized motif rank within each class, with 1 being more enriched and 0 being less enriched.
+Right heatmap shows the normalized expression (log~2~ RPKM) in RNA-seq datasets from each of the tissues that was analyzed.
 
-**Extended Data Figure 9: Expression of Zld, Grh, or Twi at supraphysiological levels results in chromatin opening at a small number of novel binding sites.
+**Extended Data Figure 9: Expression of Zld, Grh, or Twi at high protein levels results in chromatin opening at a small number of novel binding sites.
 a-c,** Immunoblots showing Zld **(a)**, Grh **(b)** or Twi **(c)** protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
@@
 **g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld **(g)**, Grh **(h)**, or Twi **(i)** are expressed at varying concentrations.
+**j,** Heatmap showing z-score normalized, anti-Zld CUT&RUN data generated from either wild-type neural stem cells (Type II neuroblasts) or neural stem cells over-expressing Zld.
+Heatmaps are divided to show those peaks detected in wild-type vs. novel peaks that were only observed with over-expression of Zld.
+ATAC-seq data from wild-type neural stem cells is also shown.
+**i,** Heatmap of adjusted p-values for the top *de novo* motifs enriched in either the endogenous or novel (Zld overexpression) Zld binding sites.
 
@@
 
-**Supplementary table 2: Previously published ChIP-seq datasets that were analyzed in this work.** For previously published ChIP-seq datasets, this table provides the target of immunoprecipitation, the cell type and/or developmental stage, the Gene Expression Omnibus accession number (GSE), and the PubMed ID for the study in which the dataset was originally published..
+**Supplementary table 2: Previously published genomics datasets that were analyzed in this work.** This table provides the target of immunoprecipitation (for ChIP-seq datasets), the cell type and/or developmental stage, the Gene Expression Omnibus accession number (GSE), and the PubMed ID for the study in which the dataset was originally published.

### 2023-10-20 manuscript/supplemental_figures.qmd
update analysis and MS based on reviewer comments

@@
 **e-g,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in wild-type (WT) cells.
+ATAC-seq signal in wild-type cells is shown for reference.
 :::
@@
 
-**Loss of H3K27me3 in tazemetostat-treated cells. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
+**Analysis of chromatin and motif content at class I-VI regions. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
 Tubulin levels are shown as a loading control.
@@
 For each sample, the heatmap displays the percentage of barcode reads for each sample and histone modification relative to the total number of barcode reads for all modifications.
+**c-e,** Heatmaps showing the levels of different chromatin marks in class I-VI regions for Zld **(c)**, Grh **(d)** or Twi **(e)**.
+**f-h,** Enrichment of known motifs in class I-VI sites for Zld **(f)**, Grh **(g)** or Twi **(h)**.
+Left heatmap shows normalized motif rank within each class, with 1 being more enriched and 0 being less enriched.
+Right heatmap shows the normalized expression (log~2~ RPKM) in RNA-seq datasets from each of the tissues that was analyzed.
 :::
@@
 
-**Expression of Zld, Grh, or Twi at supraphysiological levels results in chromatin opening at a small number of novel binding sites.
+**Expression of Zld, Grh, or Twi at high protein levels results in chromatin opening at a small number of novel binding sites.
 a-c,** Immunoblots showing Zld **(a)**, Grh **(b)** or Twi **(c)** protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
@@
 **g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld **(g)**, Grh **(h)**, or Twi **(i)** are expressed at varying concentrations.
+**j,** Heatmap showing z-score normalized, anti-Zld CUT&RUN data generated from either wild-type neural stem cells (Type II neuroblasts) or neural stem cells over-expressing Zld.
+Heatmaps are divided to show those peaks detected in wild-type vs.
+novel peaks that were only observed with over-expression of Zld.
+ATAC-seq data from wild-type neural stem cells is also shown.
+**i,** Heatmap of adjusted p-values for the top *de novo* motifs enriched in either the endogenous or novel (Zld overexpression) Zld binding sites.
 :::

### 2023-10-20 workflow/scripts/AME_motif_enrichment.R

@@ -1,0 +1,324 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(universalmotif))
+suppressPackageStartupMessages(library(Biostrings))
+library(pheatmap)
+library(GenomicRanges)
+library(memes)
+
+# define input files ===========================================================
+zld_tissue_classes_fn <- "results/ChIP_tissue_classes/zld_tissue_classes.tsv"
+grh_tissue_classes_fn <- "results/ChIP_tissue_classes/grh_tissue_classes.tsv"
+twi_tissue_classes_fn <- "results/ChIP_tissue_classes/twi_tissue_classes.tsv"
+
+embryo_nc14_RNAseq_fn <- "RNAseq/results/count_tables/embryo_RNAseq_RPKM.tsv"
+NB_RNAseq_fn <- "RNAseq/results/count_tables/neuroblast_RNA-seq_RPKM.tsv"
+
+`embryo_15-16H_RNAseq_fn` <- "RNAseq/results/count_tables/embryo-15-16H_RNAseq_RPKM.tsv"
+wing_disc_RNAseq_fn <- "RNAseq/results/count_tables/wing-disc_RNAseq_RPKM.tsv"
+
+flybase_synonym_table_fn <- "resources/fb_synonyms.tsv.gz"
+
+
+# import and clean motif database ==============================================
+# import Drosophila motif database
+meme_db_path <- system.file("extdata/flyFactorSurvey_cleaned.meme", package = "memes", mustWork = TRUE)
+# meme_db_path <- "/Users/tylergibson/meme/motif_databases/FLY/combined_drosophila_database.meme"
+
+meme_db <- read_meme(meme_db_path) |> 
+  to_df() 
+
+# import gene synonym table
+fb_synonym_table <- read_tsv(flybase_synonym_table_fn, comment = "#", 
+                             col_names = c("primary_FBid",  
+                                           "organism_abbreviation",
+                                           "current_symbol",
+                                           "current_fullname",
+                                           "fullname_synonym",
+                                           "symbol_synonym")) |> 
+  filter(startsWith(primary_FBid, "FBgn")) |> 
+  filter(organism_abbreviation == "Dmel") |> 
+  separate_longer_delim(symbol_synonym, delim = "|") |> 
[…]

### 2023-10-20 workflow/scripts/brain_zld_OE_peak_classes.R

@@ -1,0 +1,68 @@
+# setup ========================================================================
+library(tidyverse)
+library(GenomicRanges)
+library(rtracklayer)
+library(here)
+suppressPackageStartupMessages(library(BSgenome.Dmelanogaster.UCSC.dm6))
+suppressPackageStartupMessages(library(Biostrings))
+
+source(here("../useful_functions/peak_processing.R"))
+
+# define input files ===========================================================
+# define input files explicitly for testing script
+brat_aZld_peaks_fn <-  "TF_CUTandRUN/results/peaks/filtered/brain_brat_aZld.narrowPeak"
+OE_aZld_peaks_fn <-  "TF_CUTandRUN/results/peaks/filtered/brain_UAS-HA-Zld_aZld.narrowPeak"
+
+# get input files from snakemake
+snakemake@input[[1]]
+snakemake@input[[2]]
+
+# define peak classes ==========================================================
+NB_peaks <- c(
+  brat_aZld = brat_aZld_peaks_fn,
+  OE_aZld = OE_aZld_peaks_fn) |> 
+  map(rtracklayer::import) |>
+  GRangesList() |>
+  peak_overlap_table() |> 
+  mutate(
+    class = case_when(
+      brat_aZld ~ "endogenous",
+      !brat_aZld ~ "novel"
+    )
+  )
+
+# export novel binding sites ===================================================
+NB_peaks |> 
+  filter(class == "endogenous") |> 
+  makeGRangesFromDataFrame() |> 
+  rtracklayer::export(snakemake@output[[1]])
+
+NB_peaks |> 
+  filter(class == "novel") |> 
+  makeGRangesFromDataFrame() |> 
+  rtracklayer::export(snakemake@output[[2]])
+
+# export sequences for novel binding sites =====================================
+classes <- c("endogenous", "novel")
+
+for (i in seq(classes)) {
+  peaks <- NB_peaks %>% 
+    filter(class == classes[i]) %>% 
+    makeGRangesFromDataFrame(keep.extra.columns = TRUE)
+  
+  seqlevels(peaks) <- seqlevels(BSgenome.Dmelanogaster.UCSC.dm6)
+  seqinfo(peaks) <- seqinfo(BSgenome.Dmelanogaster.UCSC.dm6)
+  peaks <- trim(peaks)
+  
+  peak_seqs <- getSeq(BSgenome.Dmelanogaster.UCSC.dm6, peaks)
[…]

### 2023-10-20 workflow/scripts/classes_motif_heatmap_grh.R

@@ -1,0 +1,219 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(universalmotif))
+suppressPackageStartupMessages(library(Biostrings))
+library(pheatmap)
+library(GenomicRanges)
+library(memes)
+
+# define input files ===========================================================
+meme_files <- c(
+  # "results/ChIP_tissue_classes/motifs/grh_class_I/combined.meme",
+  # "results/ChIP_tissue_classes/motifs/grh_class_II/combined.meme",
+  # "results/ChIP_tissue_classes/motifs/grh_class_III/combined.meme",
+  "results/ChIP_tissue_classes/motifs/grh_class_IV/combined.meme",
+  "results/ChIP_tissue_classes/motifs/grh_class_V/combined.meme",
+  "results/ChIP_tissue_classes/motifs/grh_class_VI/combined.meme"
+)
+
+class_names <- c(
+  # "class_I",
+  # "class_II",
+  # "class_III",
+  "class_IV",
+  "class_V",
+  "class_VI"
+)
+
+tissue_classes <- "results/ChIP_tissue_classes/grh_tissue_classes.tsv" |> 
+  read_tsv() |> 
+  makeGRangesFromDataFrame(keep.extra.columns = TRUE)
+
+# import motfs for all classes =================================================
+all_motifs <- meme_files |> 
+  map(read_meme) 
+names(all_motifs) <- class_names
+
+
+# combine all motifs into data frame ===========================================
+all_motifs <- all_motifs |> 
+  map(convert_type, "PWM", pseudocount = 0.1) |> 
+  map(to_df) |> 
+  bind_rows(.id = "class") |> 
+  mutate(name = paste0(class, "_", name))
+
+# compute motif similarity =====================================================
+motif_similarity <- all_motifs %>%
+  # Convert to universalmotif format 
+  to_list() %>%
+  # Compute the pearson correlation for each motif with all other motifs
+  universalmotif::compare_motifs(method = "PCC")
+
+# group motifs based on motif similarity =======================================
+threshold <- 0.9
+similarity_binary <- motif_similarity > threshold
+d <- dist(similarity_binary, method = "binary")
+hc <- hclust(d)
+motif_groups <- cutree(hc, h = threshold)
+motif_groups |> unique() |> length()
+
[…]

### 2023-10-20 workflow/scripts/classes_motif_heatmap_twi.R

@@ -1,0 +1,219 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(universalmotif))
+suppressPackageStartupMessages(library(Biostrings))
+library(pheatmap)
+library(GenomicRanges)
+library(memes)
+
+# define input files ===========================================================
+meme_files <- c(
+  # "results/ChIP_tissue_classes/motifs/twi_class_I/combined.meme",
+  # "results/ChIP_tissue_classes/motifs/twi_class_II/combined.meme",
+  # "results/ChIP_tissue_classes/motifs/twi_class_III/combined.meme",
+  "results/ChIP_tissue_classes/motifs/twi_class_IV/combined.meme",
+  "results/ChIP_tissue_classes/motifs/twi_class_V/combined.meme",
+  "results/ChIP_tissue_classes/motifs/twi_class_VI/combined.meme"
+)
+
+class_names <- c(
+  # "class_I",
+  # "class_II",
+  # "class_III",
+  "class_IV",
+  "class_V",
+  "class_VI"
+)
+
+tissue_classes <- "results/ChIP_tissue_classes/twi_tissue_classes.tsv" |> 
+  read_tsv() |> 
+  makeGRangesFromDataFrame(keep.extra.columns = TRUE)
+
+# import motfs for all classes =================================================
+all_motifs <- meme_files |> 
+  map(read_meme) 
+names(all_motifs) <- class_names
+
+
+# combine all motifs into data frame ===========================================
+all_motifs <- all_motifs |> 
+  map(convert_type, "PWM", pseudocount = 0.1) |> 
+  map(to_df) |> 
+  bind_rows(.id = "class") |> 
+  mutate(name = paste0(class, "_", name))
+
+# compute motif similarity =====================================================
+motif_similarity <- all_motifs %>%
+  # Convert to universalmotif format 
+  to_list() %>%
+  # Compute the pearson correlation for each motif with all other motifs
+  universalmotif::compare_motifs(method = "PCC")
+
+# group motifs based on motif similarity =======================================
+threshold <- 0.9
+similarity_binary <- motif_similarity > threshold
+d <- dist(similarity_binary, method = "binary")
+hc <- hclust(d)
+motif_groups <- cutree(hc, h = threshold)
+motif_groups |> unique() |> length()
+
[…]

### 2023-10-20 workflow/scripts/classes_motif_heatmap_zld.R

@@ -1,0 +1,170 @@
+# setup ========================================================================
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(universalmotif))
+suppressPackageStartupMessages(library(Biostrings))
+library(pheatmap)
+library(GenomicRanges)
+library(memes)
+
+# define input files ===========================================================
+meme_files <- c(
+  "results/ChIP_tissue_classes/motifs/zld_class_I/combined.meme",
+  "results/ChIP_tissue_classes/motifs/zld_class_II/combined.meme",
+  "results/ChIP_tissue_classes/motifs/zld_class_III/combined.meme",
+  "results/ChIP_tissue_classes/motifs/zld_class_IV/combined.meme",
+  "results/ChIP_tissue_classes/motifs/zld_class_V/combined.meme",
+  "results/ChIP_tissue_classes/motifs/zld_class_VI/combined.meme"
+)
+
+class_names <- c(
+  "class_I",
+  "class_II",
+  "class_III",
+  "class_IV",
+  "class_V",
+  "class_VI"
+)
+
+tissue_classes <- "results/ChIP_tissue_classes/zld_tissue_classes.tsv" |> 
+  read_tsv() |> 
+  makeGRangesFromDataFrame(keep.extra.columns = TRUE)
+
+# import motfs for all classes =================================================
+all_motifs <- meme_files |> 
+  map(read_meme) 
+names(all_motifs) <- class_names
+
+
+# combine all motifs into data frame ===========================================
+all_motifs <- all_motifs |> 
+  map(convert_type, "PWM", pseudocount = 0.1) |> 
+  map(to_df) |> 
+  bind_rows(.id = "class") |> 
+  mutate(name = paste0(class, "_", name))
+
+# compute motif similarity =====================================================
+motif_similarity <- all_motifs %>%
+  # Convert to universalmotif format 
+  to_list() %>%
+  # Compute the pearson correlation for each motif with all other motifs
+  universalmotif::compare_motifs(method = "PCC")
+
+# group motifs based on motif similarity =======================================
+threshold <- 0.9
+similarity_binary <- motif_similarity > threshold
+d <- dist(similarity_binary, method = "binary")
+hc <- hclust(d)
+motif_groups <- cutree(hc, h = threshold)
+motif_groups |> unique() |> length()
+
+all_motifs$group <- motif_groups
+
[…]

### 2023-10-20 workflow/scripts/extended_data_fig1.R

@@ -20,6 +20,9 @@ source("workflow/scripts/plot_heatmap.R")
 # S2_Zld_aZld_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_IP.bw"
 # S2_WT_aZld_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-WT_aZld_IP.bw"
 # S2_Zld_IgG_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aIgG_IP.bw"
+# Zld_WT_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/S2-WT_1000uM_small.bw"
+# Grh_WT_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/FL_ATAC_S2-WT_100uM_small.bw"
+# Twi_WT_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/Twi_ATAC_S2-WT_40uM_small.bw"
 # 
 # zld_ChIP_classes_fn <- "results/ChIP_peak_classes/zld_ChIP_classes.tsv"
 # 
@@ -44,6 +47,9 @@ grh_blot_image <- snakemake@input[["grh_blot_image"]]
 S2_Zld_aZld_ChIP_bw <- snakemake@input[["S2_Zld_aZld_ChIP_bw"]]
 S2_WT_aZld_ChIP_bw <- snakemake@input[["S2_WT_aZld_ChIP_bw"]]
 S2_Zld_IgG_ChIP_bw <- snakemake@input[["S2_Zld_IgG_ChIP_bw"]]
+Zld_WT_ATAC_bw <- snakemake@input[["Zld_WT_ATAC_bw"]]
+Grh_WT_ATAC_bw <- snakemake@input[["Grh_WT_ATAC_bw"]]
+Twi_WT_ATAC_bw <- snakemake@input[["Twi_WT_ATAC_bw"]]
 
 zld_ChIP_classes_fn <- snakemake@input[["zld_ChIP_classes_fn"]]
 
@@ -61,13 +67,12 @@ twi_ChIP_classes_fn <- snakemake@input[["twi_ChIP_classes_fn"]]
 
 
 # # create blank layout for plot ===============================================
-# # create blank layout for plot ===============================================
 fig_width <-  18
 fig_height <- 18.5
 
 # open pdf
[…]

### 2023-10-20 workflow/scripts/extended_data_fig10.R

@@ -105,7 +105,7 @@ grh_ChIP_classes <- grh_FL_ChIP_classes_fn |>
 # create blank layout for plot =================================================
 # define figure dimensions in cm
 fig_width <-  18
-fig_height <- 21
+fig_height <- 17
 
 # open pdf
 pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)

### 2023-10-20 workflow/scripts/extended_data_fig7.R

@@ -66,7 +66,7 @@ fig_height <- 18
 
 # open pdf
 pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
-# pdf("manuscript/figures/extended_data_fig7.pdf", useDingbats = FALSE)
+pdf("manuscript/figures/extended_data_fig7.pdf", useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
 
 # generate plotGardener page
 pageCreate(width = fig_width, height = fig_height, default.units = "cm", showGuides = FALSE)
@@ -91,7 +91,7 @@ H3K27me3_color <- "gray40"
   
   # reference points for positioning figure components
 x_offset_class_label <- 0.25
-x_offset_browser_label <- 1
+x_offset_browser_label <- 0.9
 x_offset_browser <- 1.5
 
 # set genome browser height
@@ -306,7 +306,7 @@ plotText(
 # panel D ==================================================================
 # panel label
 ref_x <- 0.5
-ref_y <- 4.5
+ref_y <- 5
 
 plotText(
   label = "d", params = panel_label_params, fontface = "bold",
@@ -321,7 +321,7 @@ plotText(
 
 # add broad H3K27me3 region
 plotText(
-  label = paste("H3K27me3"),
+  label = paste0("S2-WT","\n","anti-H3K27me3"),
   params = small_text_params,
   fontface = "bold",
   x = (ref_x + x_offset_browser_label),
@@ -331,11 +331,11 @@ plotText(
 )
 
 plotText(
-  label = paste("H3K9me3"),
+  label = paste0("S2-WT","\n","anti-H3K9me3"),
   params = small_text_params,
   fontface = "bold",
   x = (ref_x + x_offset_browser_label),
-  y = (ref_y + 0.75),
+  y = (ref_y + 1),
   just = c("right", "center"),
   default.units = "cm"
 )
@@ -371,7 +371,7 @@ annoYaxis(
 
 s2 <- plotSignal(
   data = `H3K9me3_ChIP`, params = region,
-  x = (ref_x + x_offset_browser), y = (ref_y + 0.75), width = 3.75, height = gb_height,
+  x = (ref_x + x_offset_browser), y = (ref_y + 1), width = 3.75, height = gb_height,
   just = c("left", "center"), default.units = "cm",
[…]

### 2023-10-20 workflow/scripts/extended_data_fig8.R

@@ -1,10 +1,10 @@
 # setup ========================================================================
 library(plotgardener)
 library(tidyverse)
-library(org.Dm.eg.db)
-library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
 library(RColorBrewer)
 library(EBImage)
+library(rtracklayer)
+library(memes)
 
 source("workflow/scripts/plot_heatmap.R")
 source("workflow/scripts/utils.R")
@@ -15,20 +15,194 @@ source("workflow/scripts/utils.R")
 # 
 # taz_blot_aH3K27me3_image <- "data/immunoblot_raw_images/2021-06-29_taz/anti-H3K27me3_3.tif"
 # taz_blot_aTub_image <- "data/immunoblot_raw_images/2021-06-29_taz/anti-tubulin_2.tif"
+# 
+# bw <- c(
+#   H3K27ac = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE85191_aH3K27ac.bw",
+#   Nej = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE64464_aNej.bw",
+#   H3K4me1 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE85191_aH3K4me1.bw",
+#   H3K4me3 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE85191_aH3K4me3.bw",
+#   H4K16ac = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE37865_aH4K16ac.bw",
+#   H2AV = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_H2Av_IP.bw",
+# 
+#   Rpb3 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_Rpb3_IP.bw",
+#   `PolII-pS2` = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE101557_S2_PolII_phosphoSer2_IP.bw",
+#   H3K36me3 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_H3K36me3_IP.bw",
+#   SSRP1 = "published_ChIPseq/results/bigwigs/zscore_normalized/merged/GSE129236_SSRP1_IP.bw",
[…]

### 2023-10-20 workflow/scripts/extended_data_fig9.R

@@ -6,13 +6,16 @@ library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
 library(RColorBrewer)
 library(EBImage)
 library(rtracklayer)
+library(memes)
+library(universalmotif)
 
 source("workflow/scripts/plot_heatmap.R")
 source("workflow/scripts/utils.R")
 
 # define input files ===========================================================
 # define input files explicitly for interactively testing script
-# zld_titration_blot_aZld <- "data/immunoblot_raw_images/2022-7-12_titration/Zld-titration_aZld.tif"
+# zld_titration_blot_aZld <- "data/immunoblot_raw_images/2023-03-10_Zld_titration/Zld_exposure_2.tif"
+# zld_titration_blot_aTub <- "data/immunoblot_raw_images/2023-03-10_Zld_titration/tubulin.tif"
 # 
 # grh_titration_blot_aGrh <- "data/immunoblot_raw_images/2022-7-12_titration/Grh-titration_aGrh_short.tif"
 # grh_titration_blot_aTub <- "data/immunoblot_raw_images/2022-7-12_titration/Grh-titration_aTub.tif"
@@ -20,23 +23,33 @@ source("workflow/scripts/utils.R")
 # twi_titration_blot_aTwi <- "data/immunoblot_raw_images/2022-7-12_titration/Twi-titration_aTwi.tif"
 # twi_titration_blot_aTub <- "data/immunoblot_raw_images/2022-7-12_titration/Twi-titration_aTub.tif"
 # 
-# zld_tissue_classes <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv") |> 
-#   mutate(class = replace(class, class == "repressed_H3K27me3", "class IV")) |> 
-#   mutate(class = replace(class, class == "repressed_H3K9me3", "class V")) |> 
+# zld_tissue_classes <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv") |>
+#   mutate(class = replace(class, class == "repressed_H3K27me3", "class IV")) |>
+#   mutate(class = replace(class, class == "repressed_H3K9me3", "class V")) |>
 #   mutate(class = replace(class, class == "repressed_other", "class VI"))
 # 
-# grh_tissue_classes <- read_tsv("results/ChIP_tissue_classes/grh_tissue_classes.tsv") |> 
[…]

### 2023-10-20 workflow/scripts/fig3.R

@@ -170,7 +170,12 @@ plotText(
 )
 
 # read in overlapping class I sites
-ns_sites <- rtracklayer::import(ns_sites_bed_fn)
+regions <-  nonspecific_sites |> 
+  rowwise() |> 
+  mutate(n_samples = sum(c(zld_class_I, grh_class_I, twi_class_I))) |>
+  ungroup() |> 
+  mutate(ns_class = ifelse(n_samples == 3, "overlapping_class_I", "non-overlapping_class_I")) |> 
+  makeGRangesFromDataFrame(keep.extra.columns = TRUE)
 
 # generate heatmap
 bw <- c(
@@ -187,7 +192,6 @@ bw <- c(
 hm_upstream <- 500
 hm_downstream <- 500
 
-regions <- ns_sites 
 
 hm <- plot_heatmap_minimal(
   bw, regions, 
@@ -196,6 +200,7 @@ hm <- plot_heatmap_minimal(
   order_by_samples = 1:3, 
   individual_scales = TRUE,
   return_heatmap_list = TRUE,
+  row_split = regions$ns_class,
   use_raster = TRUE, raster_by_magick = TRUE, raster_magick_filter = "Triangle",
   border = "black",
   border_gp = gpar(lwd = 0.5)
@@ -212,14 +217,19 @@ plotGG(
 )
 
 # add axes to heatmaps
-seekViewport(name = "matrix_1_heatmap_body_1_1")
+seekViewport(name = "matrix_1_heatmap_body_2_1")
 grid.xaxis(at = c(0, 1), label = c(paste0("-", hm_upstream / 1000, "KB"), paste0("+",hm_downstream / 1000, "KB")), gp = gpar(lwd = 0.5, fontsize = small_text_params$fontsize))
 seekViewport(name = "page")
 
 # add heatmap labels
 plotText(
-  label = paste(length(regions), "overlapping", "\n", "class I sites"), params = small_text_params, fontface = "bold",
-  x = (ref_x), y = (ref_y + 1.5), just = c("center"), default.units = "cm", rot = 90
+  label = "not cobound", params = small_text_params, fontface = "bold",
+  x = (ref_x), y = (ref_y + 1), just = c("center"), default.units = "cm", rot = 90
+)
+
+plotText(
+  label = "cobound", params = small_text_params, fontface = "bold",
[…]

### 2023-10-20 workflow/scripts/fig4.R

@@ -27,7 +27,7 @@ source("workflow/scripts/utils.R")
 # 
 # S2_Twi_ChIP_bw <-  "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Twi_aTwi_IP.bw"
 # embryo_Twi_ChIP_bw <-  "published_ChIPseq/results/bigwigs/zscore_normalized/merged/embryo-1-3H_aTwi.bw"
-
+# 
 # S2_Zld_ChIP_DMSO_bw <-  "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_DMSO_aZld.bw"
 # S2_Zld_ChIP_taz_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_taz_aZld.bw"
 # S2_Zld_H3K27me3_DMSO_bw <- "histone_CUTandRUN/results/bigwigs/spikeIn_normalized/merged/S2-Zld_DMSO_aH3K27me3_large.bw"
@@ -76,7 +76,7 @@ S2_Twi_H3K27me3_taz_bw <-  snakemake@input[["S2_Twi_H3K27me3_taz_bw"]]
 ## create blank layout for plot =================================================
 # define figure dimensions in cm
 fig_width <-  18
-fig_height <- 12.5
+fig_height <- 13.5
 
 # open pdf
 pdf(snakemake@output[[1]], useDingbats = FALSE, width = fig_width / 2.54,height = fig_height / 2.54)
@@ -155,11 +155,11 @@ hm <- plot_heatmap_minimal(
   return_heatmap_list = TRUE
 )
 
-b_hm <- grid.grabExpr(draw(hm, show_heatmap_legend = FALSE, padding = unit(c(0, 0, 0, 0), "mm")))
+a_hm <- grid.grabExpr(draw(hm, show_heatmap_legend = FALSE, padding = unit(c(0, 0, 0, 0), "mm")))
 
 # place heatmap on page
 plotGG(
-  plot = b_hm,
+  plot = a_hm,
   x = (ref_x + 0.25), y = (ref_y + 0.25),
   width = 5, height = 5, just = c("left", "top"),
   default.units = "cm"
@@ -207,27 +207,27 @@ plotText(
 )
 
 plotText(
-  label = paste0("Zld", "\n", "S2 cells"), params = small_text_params, fontface = "bold",
+  label = paste0("S2-Zld", "\n", "anti-Zld"), params = small_text_params, fontface = "bold",
[…]

### 2023-10-20 workflow/scripts/fig5.R

@@ -11,18 +11,18 @@ source("workflow/scripts/plot_heatmap.R")
 source("workflow/scripts/utils.R")
 
 # define input files ===========================================================
-# zld_tissue_occupancy_fn <- "results/ChIP_tissue_classes/zld_tissue_classes.tsv"
-# grh_tissue_occupancy_fn <- "results/ChIP_tissue_classes/grh_tissue_classes.tsv"
-# twi_tissue_occupancy_fn <- "results/ChIP_tissue_classes/twi_tissue_classes.tsv"
-# zld_titration_classes_fn <- "results/ChIP_titration_classes/zld_titration_classes.tsv"
-# grh_titration_classes_fn <- "results/ChIP_titration_classes/grh_titration_classes.tsv"
-# twi_titration_classes_fn <- "results/ChIP_titration_classes/twi_titration_classes.tsv"
-# zld_titration_ChIP_rpkm_fn <- "results/ChIP_tissue_classes/zld_classes_titration_ChIP_rpkm.tsv"
-# zld_titration_ATAC_rpkm_fn <- "results/ChIP_tissue_classes/zld_classes_titration_ATAC_rpkm.tsv"
-# grh_titration_ChIP_rpkm_fn <- "results/ChIP_tissue_classes/grh_classes_titration_ChIP_rpkm.tsv"
-# grh_titration_ATAC_rpkm_fn <- "results/ChIP_tissue_classes/grh_classes_titration_ATAC_rpkm.tsv"
-# twi_titration_ChIP_rpkm_fn <- "results/ChIP_tissue_classes/twi_classes_titration_ChIP_rpkm.tsv"
-# twi_titration_ATAC_rpkm_fn <- "results/ChIP_tissue_classes/twi_classes_titration_ATAC_rpkm.tsv"
+zld_tissue_occupancy_fn <- "results/ChIP_tissue_classes/zld_tissue_classes.tsv"
+grh_tissue_occupancy_fn <- "results/ChIP_tissue_classes/grh_tissue_classes.tsv"
+twi_tissue_occupancy_fn <- "results/ChIP_tissue_classes/twi_tissue_classes.tsv"
+zld_titration_classes_fn <- "results/ChIP_titration_classes/zld_titration_classes.tsv"
+grh_titration_classes_fn <- "results/ChIP_titration_classes/grh_titration_classes.tsv"
[…]

### 2023-10-20 workflow/scripts/fig6.R

@@ -3,7 +3,6 @@ suppressPackageStartupMessages(library(plotgardener))
 suppressPackageStartupMessages(library(tidyverse))
 suppressPackageStartupMessages(library(org.Dm.eg.db))
 suppressPackageStartupMessages(library(TxDb.Dmelanogaster.UCSC.dm6.ensGene))
-library(grImport)
 suppressPackageStartupMessages(library(grid))
 suppressPackageStartupMessages(library(RColorBrewer))
 
@@ -13,12 +12,15 @@ source("workflow/scripts/utils.R")
 # define input files ===========================================================
 # define input files explicitly for interactive testing
 # Zld_FL_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld_aZld_IP.bw"
-# Zld_DBD_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-DBD_aZld_IP.bw"
+# Zld_DBD_ChIP_100uM_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-DBD-100uM_aZld_IP.bw"
+# Zld_DBD_ChIP_400uM_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Zld-DBD-400uM_aZld_IP.bw"
 # Zld_WT_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/S2-WT_1000uM_small.bw"
 # Zld_Zld_ATAC_bw <- "ATACseq/results/bigwigs/zscore_normalized/merged/S2-Zld_1000uM_small.bw"
 # 
 # Grh_FL_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh_aGrh_IP.bw"
-# Grh_DBD_ChIP_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-DBD_aGrh_IP.bw"
+# Grh_DBD_ChIP_20uM_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-DBD-20uM_aGrh_IP.bw"
+# Grh_DBD_ChIP_80uM_bw <- "ChIPseq/results/bigwigs/zscore_normalized/merged/S2-Grh-DBD-80uM_aGrh_IP.bw"
+# 
[…]

### 2023-10-20 workflow/scripts/generate_fig5_RPKM_tables.R

@@ -7,11 +7,12 @@ source("workflow/scripts/utils.R")
 # Zld ChIP =====================================================================
 # import unit names
 units <- read_tsv("config/ChIP_units.tsv")
-units <- units[c(37:84),]
 
 units <- units |> 
   filter(call_peaks) |> 
-  filter(grepl("Zld", sample_name))
+  filter(grepl("Zld", sample_name)) |> 
+  filter(grepl("uM", sample_name)) |> 
+  filter(!grepl("DBD", sample_name))
 
 # import regions of interest
 zld_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv")
@@ -71,10 +72,11 @@ rpkm_counts |>
 # Zld ATAC =====================================================================
 # import unit names
 units <- read_tsv("config/ATACseq_units.tsv")
-units <- units[25:48,]
 
 units <- units |> 
-  filter(grepl("Zld", sample_name))
+  filter(grepl("titration", sample_name)) |> 
+  filter(grepl("Zld", sample_name)) |> 
+  filter(!grepl("DBD", sample_name))
 
 # import regions of interest
 zld_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/zld_tissue_classes.tsv")
@@ -135,11 +137,12 @@ rpkm_counts |>
 # Grh ==========================================================================
 # import unit names
 units <- read_tsv("config/ChIP_units.tsv")
-units <- units[37:84,]
 
 units <- units |> 
   filter(call_peaks) |> 
-  filter(grepl("Grh", sample_name))
+  filter(grepl("Grh", sample_name)) |> 
+  filter(grepl("uM", sample_name)) |> 
+  filter(!grepl("DBD", sample_name))
 
 # import regions of interest
 grh_tissue_occupancy <- read_tsv("results/ChIP_tissue_classes/grh_tissue_classes.tsv")
@@ -199,10 +202,11 @@ rpkm_counts |>
 # Grh ATAC =====================================================================
 # import unit names
 units <- read_tsv("config/ATACseq_units.tsv")
-units <- units[25:48,]
 
 units <- units |> 
-  filter(grepl("Grh", sample_name))
+  filter(grepl("titration", sample_name)) |> 
+  filter(grepl("Grh", sample_name)) |> 
+  filter(!grepl("DBD", sample_name))
 
 # import regions of interest
[…]

### 2023-10-20 workflow/scripts/quantify_blots.R

@@ -1,0 +1,158 @@
+# setup ========================================================================
+library(tidyverse)
+
+# ==============================================================================
+# for ZLD
+# ==============================================================================
+
+pdf("data/immunoblot_raw_images/2018-10-17_Zld_induction/quantification_plot.pdf", useDingbats = FALSE)
+# defin input files ============================================================
+raw_band_intensities <- "data/immunoblot_raw_images/2018-10-17_Zld_induction/quantification.tsv"
+
+# analysis of band intensities =================================================
+# get normalized intensity
+gel_quantification <- raw_band_intensities |> 
+  read_tsv(col_names = c("lane", "band", "Rf", "raw_volume", "cal_volume", "MW"), skip = 1) |> 
+  select(lane, band, raw_volume) |> 
+  pivot_wider(names_from = band, names_prefix = "band_", values_from = raw_volume) |> 
+  add_column(sample =
+               c(
+                 "Zld_lineA_0", 
+                 "Zld_lineA_750", 
+                 "Zld_lineA_1000", 
+                 "Zld_lineA_1250",
+                 "Zld_lineA_1500",
+                 
+                 "Zld_lineB_0", 
+                 "Zld_lineB_750", 
+                 "Zld_lineB_1000", 
+                 "Zld_lineB_1250",
+                 "Zld_lineB_1500",
+                 
+                 "5 2-3H embryos A",
+                 "5 2-3H embryos B",
+                 "10 2-3H embryos A",
+                 "10 2-3H embryos B"
+               )
+  ) |>
+  mutate(normalized_intensity = band_1 / band_2) 
+
+
+# get estimated WT level
+wt_level <- gel_quantification[11:14, "normalized_intensity"] |> 
+  pull() |> 
+  mean()
+
+# plot band intensities
+gel_quantification |> 
+  mutate(normalized_intensity = normalized_intensity / wt_level) |> 
+  ggplot(aes(x = fct_inorder(sample), y = normalized_intensity)) + geom_bar(stat = "identity") +
+  geom_hline(yintercept = wt_level, lty = 2, color = "grey") +
+  theme_classic(base_size = 24) +
+  theme(axis.text.x = element_text(angle = 90, hjust=1,), axis.title.x = element_blank()) 
+
+dev.off()
+# ==============================================================================
+# for Grh
[…]

### 2023-10-20 workflow/scripts/split_tissue_classes.R

@@ -1,0 +1,46 @@
+# setup ------------------------------------------------------------------------
+suppressPackageStartupMessages(library(rtracklayer))
+suppressPackageStartupMessages(library(GenomicRanges))
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(BSgenome.Dmelanogaster.UCSC.dm6))
+suppressPackageStartupMessages(library(Biostrings))
+
+# get input file names ---------------------------------------------------------
+tissue_classes <- snakemake@input[[1]] %>% 
+  read_tsv()
+
+# export each class to bed file ------------------------------------------------
+classes <- 
+  c(
+    "class I",           
+    "class II",       
+    "class III",
+    "repressed_H3K27me3",
+    "repressed_H3K9me3",
+    "repressed_other"
+    )
+
+for (i in seq(classes)) {
+tissue_classes %>% 
+  filter(class ==  classes[i]) %>% 
+  makeGRangesFromDataFrame() %>% 
+  rtracklayer::export(snakemake@output[[i]])
+}
+
+
+
+# get underlying sequences and write to fasta file -----------------------------
+for (i in seq(classes)) {
+  peaks <- tissue_classes %>% 
+    filter(class == classes[i]) %>% 
+    makeGRangesFromDataFrame(keep.extra.columns = TRUE)
+  
+  seqlevels(peaks) <- seqlevels(BSgenome.Dmelanogaster.UCSC.dm6)
+  seqinfo(peaks) <- seqinfo(BSgenome.Dmelanogaster.UCSC.dm6)
+  peaks <- trim(peaks)
+  
+  peak_seqs <- getSeq(BSgenome.Dmelanogaster.UCSC.dm6, peaks)
+  names(peak_seqs) <-  paste0("peak_", 1:length(peak_seqs))
+  
+  writeXStringSet(peak_seqs, snakemake@output[[i + 6]])
+}

### 2023-10-20 workflow/scripts/supplemental_tables.R

@@ -24,4 +24,21 @@ published_ChIP_data <- read_tsv("config/published_ChIP_units_all.tsv") |>
   distinct(sample_group, .keep_all = TRUE) |> 
   select(sample_group, GSE_accession)
 
-openxlsx::write.xlsx(published_ChIP_data, file = "manuscript/supplemental_tables/supplemental_table_2.xlsx") 
+published_RNAseq_data <- read_tsv("config/RNAseq_units.tsv") |> 
+  filter(!is.na(GSE_accession)) |> 
+  distinct(sample_group, .keep_all = TRUE) |> 
+  select(sample_group, GSE_accession)
+
+published_ATACseq_data <- read_tsv("config/ATACseq_units.tsv") |> 
+  filter(!is.na(GSE_accession)) |> 
+  distinct(sample_group, .keep_all = TRUE) |> 
+  select(sample_group, GSE_accession)
+
+
+supp_table_2 <- list(
+  published_ChIPseq_data = published_ChIP_data,
+  published_RNAseq_data = published_RNAseq_data,
+  published_ATACseq_data = published_ATACseq_data
+)
+
+openxlsx::write.xlsx(supp_table_2, file = "manuscript/supplemental_tables/supplemental_table_2.xlsx") 
